### 2021-04-27 manuscriptPGPE.tex
Updates from Overleaf

%% ****** Start of file aiptemplate.tex ****** %
%%
%%   This file is part of the files in the distribution of AIP substyles for REVTeX4.
%%   Version 4.1 of 9 October 2009.
%%
%
% This is a template for producing documents for use with 
% the REVTEX 4.1 document class and the AIP substyles.
% 
% Copy this file to another name and then work on that file.
% That way, you always have this original template file to use.

%\documentclass[aip,graphicx]{revtex4-1}
%\documentclass[aip,reprint]{revtex4-1}

%\usepackage{graphicx}

%\draft % marks overfull lines with a black rule on the right
%\documentclass[pre,aps,floatfix,authordate1-4,twocolumn]{revtex4-1}
%\documentclass[pre,aps,floatfix,authordate1-4]{revtex4-1}

\documentclass[aps,prl,superscriptaddress,twocolumn]{revtex4}

%\documentclass[aps,prl,preprint,groupedaddress]{revtex4}

\usepackage{rotating} 
\usepackage{times}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{epstopdf}
\usepackage[obeyFinal]{easy-todo}
\usepackage{csquotes}
\usepackage{xr-hyper}
\usepackage{hyperref} % to make TOC titles clickable
\hypersetup{
    colorlinks=true, %set true if you want colored links
    linktoc=all,     %set to all if you want both sections and subsections linked
    linkcolor=black,  %choose some color if you want links to stand out
    citecolor=black,
    filecolor=black,
    urlcolor=black
}

\makeatletter
\newcommand*{\addFileDependency}[1]{% argument=file name and extension
  \typeout{(#1)}
  \@addtofilelist{#1}
  \IfFileExists{#1}{}{\typeout{No file #1.}}
}
\makeatother

\newcommand*{\myexternaldocument}[1]{%
    \externaldocument{#1}%
    \addFileDependency{#1.tex}%
    \addFileDependency{#1.aux}%
}

\myexternaldocument{manuscriptPGPEsuppl}

\begin{document}

% Use the \preprint command to place your local institutional report number 
% on the title page in preprint mode.
% Multiple \preprint commands are allowed.
%\preprint{}

\title{Conformational plasticity of phospholipid headgroups in simulations and experiments} %Title of paper

% repeat the \author .. \affiliation  etc. as needed
% \email, \thanks, \homepage, \altaffiliation all apply to the current author.
% Explanatory text should go in the []'s, 
% actual e-mail address or url should go in the {}'s for \email and \homepage.
% Please use the appropriate macro for the type of information

% \affiliation command applies to all authors since the last \affiliation command. 
% The \affiliation command should follow the other information.

\author{Am{\'e}lie Bacle}
\affiliation{Laboratoire Coop{\'e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit{\'e} de Poitiers, 1 rue Georges Bonnet, 86000 Poitiers, France }

\author{Pavel Buslaev}
\affiliation{Nanoscience Center and Department of Chemistry, University of Jyväskylä, P.O. Box 35, 40014 Jyväskylä, Finland}
\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, 141701 Dolgoprudny, Russia}

\author{Rebeca Garc{\'i}a Fandi{\~n}o}
\affiliation{Center for Research in Biological Chemistry and Molecular Materials (CiQUS), Universidade de Santiago de Compostela, E-15782 Santiago de Compostela, Spain}
\affiliation{CIQUP, Centro de Investigação em Qu{\'i}mica, Departamento de Qu{\'i}mica e Bioqu{\'i}mica, Faculdade de Ci{\^e}ncias, Universidade do Porto, Porto, Portugal}

\author{Fernando Favela-Rosales}
\affiliation{Departamento de Ciencias B\'{a}sicas, Tecnol\'{o}gico Nacional de M\'{e}xico - ITS Zacatecas Occidente, M\'{e}xico}

\author{Tiago M. Ferreira}
\affiliation{NMR group - Institute for Physics, Martin Luther University Halle-Wittenberg, 06120 Halle (Saale), Germany}

\author{Patrick F.J. Fuchs}
\affiliation{Sorbonne Universit{\'e}, Ecole Normale Sup{\'e}rieure, PSL University, CNRS, Laboratoire des Biomol{\'e}cules (LBM), 75005 Paris, France}
\affiliation{Universit{\'e} de Paris, UFR Sciences du Vivant, 75013, Paris, France}

\author{Ivan Gushchin}
\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, 141701 Dolgoprudny, Russia}

\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}

\author{Anne M. Kiirikki}
\affiliation{Institute of Biotechnology, University of Helsinki}

\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
\affiliation{Global and Planetary Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}

\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\affiliation{Groningen Biomolecular Sciences and Biotechnology Institute 
and The Zernike Institute for Advanced Materials, 
University of Groningen, 9747 AG Groningen, The Netherlands}

\author{Paula Mil{\'a}n Rodr{\'i}guez}
\affiliation{Sorbonne Universit{\'e}, Ecole Normale Sup{\'e}rieure, PSL University, CNRS, Laboratoire des Biomol{\'e}cules (LBM), 75005 Paris, France}

\author{Markus S. Miettinen}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
% \affiliation[Max Planck Institute of Colloids and Interfaces]{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}

\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
\affiliation{Institute of Biotechnology, University of Helsinki}

\author{Chris G. Papadopoulos}
\affiliation{Universit{\'e} Paris-Saclay, CEA, CNRS, Institute for Integrative Biology of the Cell (I2BC), 91198 Gif-sur-Yvette, France}

\author{Antonio Pe{\'o}n}
\affiliation{Spain}

\author{Thomas J. Piggot}
\affiliation{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, United Kingdom}

\author{{\'A}ngel Pi{\~n}eiro}
\affiliation{Departamento de F{\'i}sica Aplicada, Facultade de F{\'i}sica, Universidade de Santiago de Compostela, E-15782 Santiago de Compostela, Spain}

\author{Salla I. Virtanen}
\affiliation{Institute of Biotechnology, University of Helsinki}

% Collaboration name, if desired (requires use of superscriptaddress option in \documentclass). 
% \noaffiliation is required (may also be used with the \author command).
%\collaboration{}
%\noaffiliation

\date{\today}

\begin{abstract}
% insert abstract here
  Chemistry of lipid headgroups, the water facing components of cell membranes that regulate cell functions via lipid--protein interactions, varies between organisms and organelles. Because membranes are in liquid state under physiological conditions,
  individual lipids are not fixed to single conformations but they sample an ensemble of conformations with the Boltzmann weighted probabilities.
  However, these ensembles and their dependence on lipid types have not yet been experimentally determined. 
  Therefore, it has not been clear if lipid molecules exchange between few rigid conformations,
  or if they can freely fluctuate in all possible conformations.
  Here, we combine solid state NMR experiments and molecular dynamics simulations from the NMRlipids open collaboration to resolve the conformational ensembles of the headgroups of key lipid types in their liquid lamellar phase under various biologically relevant conditions. Interpretation of NMR experiments using the plethora of simulation data collected in the NMRlipids project suggests that all lipid headgroups sample a wide range of conformations in neutral and charged cellular membranes.
  Differences in the headgroup chemistry between different lipid types manifest
  in probability distributions of conformations, but all lipid types can access almost any of the possible conformations.
  Together with the analysis of protein-bound lipids from the protein data bank (PDB), this suggests that lipids can bind to proteins in a wide range of conformations independently of their headgroup chemistry. Therefore, the selective adsorption of proteins to membranes is likely regulated by specific protein--lipid interactions rather than conformational restrictions of the bilayer lipids.
  Our results pave the way to comprehensive understanding of
  lipid mediated signaling and lipid--protein interactions in biomedical applications.

\end{abstract}

%\pacs{}% insert suggested PACS numbers in braces on next line

\maketitle

% Body of paper goes here. Use proper sectioning commands. 
% References should be done using the \cite, \ref, and \label commands

%\label{}
\section{Introduction}

Chemical compositions of hydrophilic lipid headgroups vary between different
organelles and organisms, and different lipid types
regulate protein functions in many different ways \cite{lee03,vanmeer08}.
Lipids can directly bind to proteins or indirectly affect protein
functions by altering membrane properties such as charge or elasticity \cite{lee03,lemmon08}.
Specific interactions with certain lipid headgroups
are known to be essential for the function of several proteins \cite{lee11,lemmon08},
but it is not clear if the specificity is driven by the
differences in accessible conformations between lipid types or
by specific lipid--protein interactions.

Stuctures of protein-bound lipids are available in the protein data bank (PDB) \cite{berman00}
and crystal structures of lipids have been determined \cite{buldt81,pascher92},
but their relation to the conformational ensembles in bulk membranes in the liquid lamellar phase remains unclear \cite{marsh13b}.
Most accurate experimental information on conformational ensembles of lipids
in liquid lamellar phase are typically derived from NMR experiments, particularly from 
C--H bond order parameters measured using $^2$H NMR~\cite{seelig77c,davis83,Semchyschyn04}.
According to these experiments, the glycerol backbone conformations are largely similar irrespectively of the headgroup \cite{gally81} and
the headgroup conformations are similar in phosphatidylcholine (PC), phosphatidylethanolamine (PE) and phosphatidylglycerol (PG) lipids,
whereas the headgroup of phosphatidylserine (PS) lipids is more rigid \cite{wohlgemuth80,buldt81}. 
However, order parameter signs are not accessible in $^2$H NMR experiments \cite{ollila16}
and universal models to map order parameters to structural ensembles are not available \cite{pezeshkian18,akutsu20}.
Therefore, it is not clear if the conformational ensemble of lipid headgroups in liquid lamellar bilayer
is composed of exchange between few restricted conformations, or if lipid headgroups can fluctuate freely across a wide range of conformations.

Here, we use natural abundance $^{13}$C NMR experiments %(enabling the determination of order parameter signs)
and MD simulations from the NMRlipids open collaboration
to resolve the differences in the conformational ensembles between PC, PE, PG and PS lipid headgroups.
We elucidate also the effect of charges and protein binding to lipid headgroup conformations.
Zwitterionic PC and PE are the most common lipids in eukaryotes and bacteria, respectively \cite{vanmeer08,sohlenkamp16}.
PE is also the second most abundant glycerophospholipid in eukaryotic cells
and has been related to various diseases \cite{vance15,calzada16,patel17}.
PS and PG are the most common negatively charged lipids in eukaryotes and bacteria, respectively,
and their presence affect membrane protein functionality and signaling \cite{lemmon08,leventis10,sohlenkamp16,hariharan18}.
All the studied lipids specifically bind to various proteins \cite{yeagle14}.

Similarity of glycerol backbone and headgroup order parameters in model membranes and bacteria \cite{gally81,scherer87,seelig90}
suggest that our results can be used to 
understand the biological role of lipid headgroups in lipid--protein interactions.
These interactions are crucial, for example, in lipid mediated signaling \cite{lemmon08} and
design of phoshoplipid-specific antibodies \cite{vigant15}.

\section{Methods}
\subsection{Experimental C--H bond order parameters}
The headgroup and glycerol backbone C--H bond order parameters of 1-palmitoyl-2-oleoyl-sn-glycero-3-phosphoethanolamine (POPE) and 1-palmitoyl-2-oleoyl-sn-glycero-3-phospho-(1'-rac-glycerol) (POPG), purchased from Avanti polar lipids, were measured using natural abundance $^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}. The magnitudes of order parameters were determined from the chemical-shift resolved dipolar splittings using a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04}, and the signs from S-DROSS experiments~\cite{gross97} combined with SIMPSON simulations \cite{bak00}.
%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
The NMR experiments were identical as in our previous work~\cite{antila19}. The POPE experiments were recorded at 310~K and POPG experiments at 298~K, where the bilayers are in the liquid disordered phase \cite{marsh13}.

Glycerol backbone peaks from both lipids, and the $\alpha$-carbon peak from POPE in the INEPT spectra 
were assigned based on previously measured POPC spectra~\cite{ferreira13}.
The $\beta$-carbon peak from POPE was assigned based on $^{13}$C chemical shift table for amines available
at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm}.
\todo{How were $\alpha$ and  $\gamma$-carbon peaks assigned in POPG?}.
The $\beta$-carbon peak from POPG overlapped with the g$_2$ peak from glycerol backbone
because their chemical environments are similar.
\todo{Details to be checked by Tiago}.

\subsection{Molecular dynamics simulations}

Molecular dynamics simulation data were collected and analyzed using
the methods from the NMRlipids Open Collaboration project (\url{nmrlipids.blogspot.fi}) \cite{botan15,catte16,ollila16,antila19}.
%\url{github.com/NMRlipids/NMRlipidsIVotherHGs}. 
Simulation details, accessibility information and quality evaluation of almost 100 combinations of lipid headgroup and force field simulated for this work are in the supplementary information.

Best models for the interpretation of lipid headgroup conformational ensembles from the experimental data were selected
using quality evaluation measures defined in the NMRlipids project.
Conformational ensembles of headgroup and glycerol backbone in PE and PG simulations were evaluated using the C--H bond order parameters \cite{botan15}. Interactions between different headgroups were evaluated by monitoring the changes in headgroup order parameters upon mixing the lipids \cite{antila19}. The ion binding affinities and response of lipids to bound charge were evaluated by monitoring the changes in lipid headgroup order parameters \cite{catte16,antila19}.

The differences in headgroup conformation ensembles were analyzed calculating the distributions of the last six heavy atom dihedral angles, i.e., dihedrals O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$, P-O$_\alpha$-C$_\alpha$-C$_\beta$, O$_{g3}$-P-O$_\alpha$-C$_\alpha$, g$_3$-O$_{g3}$-P-O$_\alpha$, g$_2$-g$_3$-O$_{g3}$-P, and g$_1$-g$_2$-g$_3$-O$_{g3}$ labeled in Fig. \ref{HGorderParameters} C. Relative energy costs for turning dihedral angles with respect to the most probable value (lowest energy) were estimated from the inverse Boltzmann formula $\Delta E(\theta) = -kT \left[\ln\left[p(\theta)\right]-\ln\left[p(\theta_0)\right] \right]$, where $p(\theta)$ is the dihedral angle distribution and $\theta_0$ is the most probable angle from MD simulation.

%\clearpage
\begin{figure*}[]
  \centering
%  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
  %  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCPSPEPG.eps}
   \includegraphics[width=18.0cm]{./Figs/figure_PG_PE.pdf}
   \caption{\label{HGorderParameters}
     Chemical structure, refocused-INEPT spectrum, 2D R-PDLF spectra, and S-DROSS dipolar slices (from top to bottom) for \textbf{A)} POPE  and \textbf{B)} POPG MLVs measured by $^1$H-$^{13}$C solid-state NMR. The S-DROSS dipolar slices show both experimental data (open circles) and the result of NMR numerical simulations (blue solid lines).
      Full NMR spectra are shown in Figs. \ref{POPEspectra} and \ref{POPGspectra}.
     \textbf{C)} Chemical structure of POPC and POPS.
    \textbf{D)} Headgroup and glycerol backbone C-H bond order parameters for different phospholipids measured in the lamellar liquid disordered phase.
    The $S_{\rm{CH}}$ magnitudes and signs determined by $^1$H-$^{13}$C NMR spectroscopy (filled squares) for POPE (310~K) and POPG (298~K)
    measured in this work are shown together with the data previously reported for POPS (298~K) \cite{antila19} and POPC (300~K) \cite{ferreira13,ferreira16}. The headgroup and glycerol backbone $S_{\rm{CH}}$ magnitudes measured previously with $^2$H NMR (empty symbols) are also plotted as real values using the signs determined in this work for 
    POPG with 10nM PIPES (298~K) \cite{borle85},
    DPPG with 10mM PIPES and 100mM NaCl (314~K) \cite{wohlgemuth80}, 
    DPPE (341~K) \cite{seelig76},
    E.coli PE and E.coli PG (310~K) \cite{gally81}
    .
   }
   %\todo{This is a sketch, Tiago Ferreira will make a new figure.} \\
%  \todo{D) could be clarified as Fig. 2 in the NMRlipids IVps paper.}
\end{figure*}

\subsection{Analysis of protein-bound lipid conformations}
Lipid structures from the Protein Data Bank (PDB, \cite{berman2000})
were searched using PDBe REST API (\url{www.ebi.ac.uk/pdbe/pdbe-rest-api})
using the ligand names listed in the supplementary information.
Heavy atom dihedral distributions from the lipid structures were calculated
using the MDAnalysis Python library \cite{agrawal11,gowers16} and
Jupyter notebook available from {\it scripts/PDBanalysis.ipynb} folder in Ref~\citenum{NMRlipidsIVbgit}.
Only the structures determined using X-ray crystallography or Cryo-EM with a resolution higher than 3.2~\text{\AA} were used in the analysis. Some structures of multimeric proteins contained multiple lipids in the same conformation (analyzed dihedral angles within 3 degrees of each other). In that case, this conformation was counted once for the subsequent analyses
to avoid its overweighting.

To demonstrate the existence of similar structures of different lipids bound to different proteins, we searched for pairs having the last five headgroup heavy atom dihedral angles (excluding g$_1$-g$_2$-g$_3$-O$_{g3}$) within
30 degrees of each other among the structures with resolution higher than 2.5~\text{\AA}. The two most representative examples
were handpicked from the results.

\section{Results and Discussion}

\begin{figure*}[bt]
  \centering
   \includegraphics[width=18.0cm]{./Figs/figure2.eps}
   \caption{\label{structures}
     Results from the best simulation model (CHARMM36) simulations demonstrating the differences in conformational ensembles between different lipids. 
     \textbf{A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurence of different conformations.
     \textbf{B)} Headgroup and glycerol backbone region order parameters of different different lipids.
     \textbf{C)} Relative free energies for individual heavy atom dihedral angles estimated from the inverse Boltzmann formula.
     Angles corresponding energies above 7~$k\mathrm{B}T$ are not shown because they are not observed in simulations.
     \textbf{D)} Distributions of P--N vector angle with respect to membrane normal.
  }
\end{figure*}

\subsection{Differences between lipid headgroups from $^{13}$C NMR experiments}

To experimentally characterize the differences in headgroup conformational ensembles of lipids that are not
bound to proteins, we measured the C--H bond order parameters
and their signs of POPG and POPE in the liquid lamellar phase, as we did previously for POPC and POPS \cite{ferreira13,ferreira16,antila19}.
Determination of headgroup and glycerol backbone order parameters and their signs
was straightforward from the data in Figs.~\ref{HGorderParameters}, \ref{POPEspectra} and \ref{POPGspectra}
for all the C--H bonds, except for the $\beta$ and g$_2$ carbons in POPG.
These carbons have overlapping peaks in the INEPT spectra due to their similar chemical environments,
and only the magnitude of the larger order parameter could be determined from the R-PDLF spectra (Fig.~\ref{HGorderParameters}B).
Based on previous $^2$H NMR measurements \cite{wohlgemuth80,gally81,borle85},
we assigned the larger order parameter to the g$_2$ carbon
and used the literature value for the $\beta$-carbon in SIMPSON simulations to determine the signs.
The decrease in the beginning of the S-DROSS curve suggests that the sign of larger g$_2$ order parameter
is negative and later increase suggests that sign of smaller $\beta$ order parameter is positive (Fig.~\ref{HGorderParameters}B).
This interpretation is confirmed by SIMPSON calculations in Fig.~\ref{POPGsimpson}.

Experimental order parameters of POPC, POPE, POPG and POPS glycerol backbones and headgroups from this and previous studies are collected in Fig.~\ref{HGorderParameters}D, where the signs from $^{13}$C NMR experiments are used also for the $^2$H NMR data from the literature. The overall agreement of order parameters determined by different research teams and different techniques for the same lipid headgroup is excellent here and in previous studies \cite{botan15,ollila16,antila19}. Therefore, the differences between lipid types are dictated by the headgroup chemistry rather than inaccuracies in experiments, differences in the acyl chains, or in experimental conditions.

The most distinct order parameters are observed for PS headgroups, for which the $\alpha$-carbon order parameter exhibits significant forking (different values for different hydrogens bound to the same carbon) and the $\beta$-carbon has a more negative value than other studied lipid types. On the other hand, the $\beta$-carbon order parameter of PG headgroup has a positive sign, in contrast to all the other lipid types. This has not been previously observed  in $^2$H NMR experiments detecting only the absolute values~\cite{wohlgemuth80,gally81,borle85}. The glycerol backbone order parameters are similar for all the lipid types, although they move slightly toward positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG. Only minor differences between PC and PE headgroups are observed.

\subsection{Conformational ensembles of different lipid headgroups from MD simulations}

\begin{figure*}[bt]
  \centering
  \includegraphics[width=18.0cm]{./Figs/HGorderparametersPCvsSURFchangeDIHEDRALSlog.eps}
  \caption{\label{changesWITHsurf}
    \textbf{A)} Modulation of PC headgroup order parameters and P--N vector angle upon addition of cationic surfactant (dihexadecyldimethylammonium)
    from CHARMM36 simulations compared with the experimental data \cite{scherer89}.
    \textbf{B)} Relative energies for individual dihedral angles estimated from inverse Boltzmann distributions of heavy atom dihedral angles
    with different amounts of cationic surfactant from CHARMM36 simulations.
  }
\end{figure*}

MD simulations could be used to interpret the lipid conformational ensembles from experiments due to their direct connection to the measured C-H bond order parameters. To find the best models for this purpose, we have evaluated all the available MD simulation force fields of PC~\cite{botan15}, PS~\cite{antila19}, PE (Fig.~\ref{HGorderParametersPE}), and PG (Fig.~\ref{HGorderParametersPOPG}) lipids against NMR data in the NMRlipids project. The comparison of hundreds of simulation trajectories simulated with approximately 20 different force fields reveal that none of the current parameters can correctly capture the lipid conformational ensemble and reproduce the experimental C-H bond order parameters of all segments. However, the best parameters (CHARMM36) correctly capture the essential differences between PC, PE, PG and PS headgroup order parameters (Fig.~\ref{structures} B).

To understand the structural origin of distinct order parameters for PG and PS lipids,
we first calculated the heavy atom dihedral angle distributions from the best performing simulations that reproduced the differences between headgroups~(Fig. \ref{DIHdists}).
Then, we used the inverse Boltzmann formula to estimate the
energy costs for different dihedral angle orientations.
The results in Fig.~\ref{structures}C suggest that all lipid headgroups are very flexible and
energy costs for rotating individual dihedrals to almost any angle is low (below 7~$k_\mathrm{B}T$).
Only $cis$ states of P-O$_\alpha$-C$_\alpha$-C$_\beta$ and g$_2$-g$_3$-O$_{g_3}$-P have larger relative
energies and are not observed for any lipids during the simulations.

Major differences between headgroups are observed for the last two dihedrals in the headgroup end,
O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ and P-O$_\alpha$-C$_\alpha$-C$_\beta$,
which prefer \textit{gauche$^-$} conformations for PG and \textit{gauche$^+$} for PS,
whereas PC and PE exhibit symmetric distributions.
Also, the energy barriers for O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ dihedral
rotations between \textit{gauche} and \textit{trans} states are larger for
PS and PG lipids than for PC and PE lipids. 
Rest of the dihedrals are similar between different lipids,
with the exception of PS lipids for which slightly larger energy for \textit{eclipsed anti} conformation
in g$_3$-O$_{{\mathrm g}3}$-P-O$_\alpha$ dihedral was observed.
Therefore we suggest that the main differences between lipid headgroups leading to distinct order parameters occur in the headgroup end, while phosphate region remains similar in different lipids with the exception of PS. The increased barriers for dihedral rotations may explain the more rigid headgroup structures in PS~\cite{browning80,buldt81}. Furthermore, the angle between headgroup dipole and membrane normal decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig.~\ref{structures}D). However, the differences between PC and PE in P-O$_\alpha$-C$_\alpha$-C$_\beta$ dihedral
and P--N vector dipole may be artificial as the $\beta$-carbon order parameter in PC is too negative even in the best available force field, thereby not being equal to the order parameter in PE as observed in experiments \cite{botan15}.

In conclusion, all studied lipid headgroups sample very broad conformational ensembles in the liquid lamellar phase and the sampled dihedral angles are within similar ranges for all headgroup types. % (Fig. \ref{structures}).
Since the rotation of dihedral angles to almost any position bears relatively low free energy cost, the lipid headgroups are able to adopt a wide range of conformations when interacting with proteins, ions, or other biomolecules.
%Our results support the models proposing free rotation around phosphate group for PC lipids, which decouples the dynamics and conformations between acyl chains and headgroups \cite{klauda08c,antila21b}, and suggests that these models could be applied for all lipids containing the phosphorus group.
%The wide range of observed conformations suggest that
Therefore, the structures in lipid crystals \cite{buldt81,pascher92} are not representative of the liquid state, and models aiming to explain NMR data using only a few conformations \cite{seelig77c,davis83,Semchyschyn04,akutsu20} are not sufficient to capture the large conformational space of lipids in the liquid lamellar state.
However, MD simulations with accurate force fields can be expected to capture the correct conformational ensembles of lipids.

\subsection{Lipid conformational ensembles in charged membranes}

Charged entities, such as lipids, proteins, surfactants, drugs, and ions incorporated in membranes can reorient the headgroup dipole in PC lipids and affect the order parameters of the lipid headgroups \cite{seelig87}. However, the detailed understanding of the structural and energetic responses of lipids to membrane-bound electric charge is still lacking \cite{Semchyschyn04}.

To resolve lipid headgroup conformational ensembles in cellular membranes bearing positive charge,
we calculated the heavy atom dihedral angle distributions from 
the subset of simulations %(CHARMM36)
that correctly captured the experimentally measured decrease in PC headgroup order parameters upon addition of cationic surfactants into a bilayer (figure \ref{changesWITHsurf} A).
The dihedral angle distributions and relative energies
in figures \ref{HGorderparametersPCvsSURFdihedrals} and \ref{changesWITHsurf} B) reveal that the
addition of positive charge into a membrane 
decreases the abundance of \textit{trans} states in g$_2$-g$_3$-O$_{g_3}$-P and g$_3$-O$_{g_3}$-P-O$_\alpha$
dihedrals.
The choline region remains essentially unchanged and only minor changes are observed in other dihedrals even at a surfactant molar fraction of 0.47.

In addition, binding of ions to the membrane may affect the lipid headgroup conformational ensembles under physiological conditions.
The bound Ca$^{2+}$ ion to PC headgroup leads to similar decrease in trans state probability for g$_3$-O$_{g3}$-P-O$_\alpha$ dihedral
in the most realistic MD simulation models
as observed for cationic surfactants
(lipid17ecc and CHARMM36 in Figs.~\ref{DIHSwithCAlipid17eccPOPC}, \ref{DIHSwithCAcharmm36POPC} and Fig.~\ref{changesWITHCaClPG}).
The dihedral distributions of the PG headgroup are more sensitive to the bound ions in the most realistic simulations,
but upward tilting of the headgroup dipole upon the addition of CaCl$_2$ is weaker than in PC
(Lipid17 and Slipids in Figs.~\ref{changesWITHCaClPG},~\ref{DIHSwithCAslipidsPOPG} and \ref{DIHSwithCAlipid17POPG}).
However, the changes in PG lipid dihedrals upon the addition of CaCl$_2$ differ between the best models (Figs.~\ref{DIHSwithCAslipidsPOPG} and~\ref{DIHSwithCAlipid17POPG}), and none of the simulations captures the Ca$^{2+}$ ion binding affinity and conformational ensemble of PG lipids simultaneously. Moreover, experimental data to evaluate the response of $\alpha$-carbon order parameters to the added CaCl$_2$ in PG is not available.
Additionally, the headgroup conformational ensembles in mixtures of PC and charged (PG or PS) or zwitterionic (PE)
lipids could not be resolved with the currently available force fields and experimental data
(Figs.~\ref{HGorderparametersPCvsPEPG} and~\ref{HGorderparametersPGvsPCchange}, and Ref.~\cite{antila19,melcr20}).

In conclusion, the structural response of lipid headgroups to 
membrane-bound charges arise
from mild changes in dihedral angle distributions, rather than restriction of lipids to
fixed conformations. Therefore, lipid headgroups sample a large space of different conformations
also in charged membranes, thereby retaining the capability to adopt multiple conformations when
interacting with proteins or other molecules.

\subsection{Protein-bound lipid conformations}

\begin{figure*}[]
  \centering
  \includegraphics[width=19.0cm]{./Figs/figure4.eps}
  \caption{\label{dihedralsFROMpdb}
    A) Heavy atom dihedral angle distributions calculated from lipid structures in PDB.
    B) The structure of PE headgroup bound to cytochrome {\it bc}$_1$ complex (PDB ID: 1PP9 \cite{huang05})
    with identical conformation as the PC headgroup bound to yeast Sec14 (PDB ID: 3B7Q \cite{schaaf08}),
    and the structure of PG headgroup bound to bovine cytochrome {\it c} oxidase (PDB ID: 3AG3 \cite{muramoto10})
    with identical conformation as PC to FraC, a pore-forming toxin (PDB ID: 4TSQ \cite{tanaka15}).
  }
\end{figure*}

Interpretation of experimental order parameters using MD simulations
in previous sections suggest that PC, PE, PG and PS lipid headgroups are very flexible,
allowing them to bind proteins in various different conformations.
To test this prediction, we analyzed the conformations of all protein-bound lipids 
from structures deposited in the PDB \cite{berman00}.
We found 311 PC, 394 PE, 154 PG, and 35 PS lipid conformations that were bound to different types of proteins and were determined as a part of protein structure using crystallography or cryo-EM. The full list of structures with bound lipids is available from {\it Data/PDBdistributions/pdbtable.pdf} in Ref.~\cite{NMRlipidsIVbgit}.

The heavy atom dihedral angle distributions calculated from these conformations 
(Fig.~\ref{dihedralsFROMpdb} A) reveal that the protein-bound lipids indeed exhibit wide
range of conformations independently on the headgroup type.
As for bulk lipid bilayers, only $cis$ conformations
of P-O$_\alpha$-C$_\alpha$-C$_\beta$ and g$_2$-g$_3$-O$_{g_3}$-P dihedrals are almost
completely absent in all lipids, and significant differences between
different headgroups were not observed.
Structures deviating from lipid crystals 
have been previously proposed to indicate inaccuracies in lipid structures in PDB \cite{marsh13b,pezeshkian18}. 
However, we observe large deviations from lipid crystals structures also in conformational ensembles
that reproduce the NMR data in the liquid lamellar phase, indicating that such deviations
are realistic also in protein-bound states.

Our results suggest that flexible lipid headgroups can optimize the intermolecular interactions with proteins by binding
in a wide range of conformations.
Therefore, the specific binding of lipids to proteins is not driven by the structural differences between headgroups.
This is demonstrated in Fig.~\ref{dihedralsFROMpdb} B with two examples where different lipids bound to
different proteins have almost identical headgroup conformations:
PE in cytochrome {\it bc}$_1$ complex is similar to PC bound to yeast Sec14,
and PG bound to bovine cytochrome {\it c} oxidase is similar to PC bound to pore-forming toxin (FraC).
On the other hand, a single lipid headgroup type is capable of accommodating various binding positions and as such would 
be able to specifically bind to many different kinds of binding sites in different proteins.

%However, it is important to note that lipids are often not the main target in
%structures deposited in PDB and therefore their conformations may be less reliable than those of proteins.

\section{Conclusions}

C--H bond order parameters from NMR experiments %in this and previous studies \cite{wohlgemuth80,buldt81,seelig87}
suggest that lipid headgroup conformational ensembles depend on lipid type (PC, PE, PG or PS) 
and accumulated membrane charge (cationic lipids, surfactants, ions or drugs). 
Our interpretation of these data, using MD simulations collected
within the NMRlipids open collaboration,
revealed that the differences in order parameters can be explained
by relatively small changes in dihedral angle probability distributions.
All studied headgroups (PC, PE, PG and PS) are flexible
and access a similar wide range of conformations with only very few restictions in dihedral orientations
also when charges are bound to membranes.
The observed differences in order parameters originate from
reweighting conformational probabilities rather than changes in accessible structures.

The flexibility and wide conformational space of headgroups suggest that protein--bound lipids can
adapt to various binding sites to optimize the intermolecular lipid--protein interactions.
We tested this prediction by analyzing the conformations of lipids that are tightly bound to proteins in the PDB.
Indeed, also protein--bound lipids exhibit wide range of conformations without significant 
differences between different lipid types. Therefore, the specificify of lipid binding to proteins is not
regulated by accessible structures of lipids, and a single lipid type can adapt to various binding sites in proteins.

Our results pave the way toward the understanding of lipid-mediated cell signaling and how lipids regulate membrane protein function in general.
We suggest that the key to understand selective binding of certain lipid types to proteins
are intermolecular lipid--protein interactions, rather than conformational restrictions of the lipids.
On the other hand, broad conformational ensembles in bulk bilayers suggest that lipid crystal structures
play a minor role while the entropic cost of lipid binding may be significant.
Finally, we demonstrated the power of open access MD simulation data from the NMRlipids open collaboration
to complement the data in the PDB for elucidating how complex systems made up of disordered biomolecules behave.

% If you have acknowledgments, this puts in the proper section head.
\begin{acknowledgments}
AP is grateful to the Centro de
Supercomputación de Galicia (CESGA) for use of the Finis
Terrae computer
%
MJ thanks CSC -- IT Center for Science for computational resources and the Emil Aaltonen foundation for financial support.
    %     Put your acknowledgments here.

P.B. was supportedby the Academy of Finland (Grant 311031)

F.F.-R. acknowledges Tecnol\'{o}gico Nacional de M\'{e}xico Proyecto IT16C431, Direcci\'{o}n General de Asuntos del Personal Acad\'{e}mico (DGAPA) Programa de Apoyo a Proyectos de Investigaci\'{o}n e Innovaci\'{o}n Tecnol\'{o}gica (PAPIIT) IG100920, CONACyT Ciencia de Frontera 74884 for financial support and Miztli-Direcci\'{o}n de C\'{o}mputo y de Tecnolog\'{i}as de Informaci\'{o}n y Comunicaci\'{o}n (DGTIC) - Universidad Nacional Aut\'{o}noma de M{\'e}xico (UNAM) (Project LANCAD-UNAM-DGTIC-057) facilities for computing-time allocation.

I.G. was supported by the Ministry of Science and Higher Education of the Russian Federation (agreement \#075-00337-20-03, project FSMG-2020-0003).

J.J.M. gratefully acknowledges financial support from the Carlsberg 
Foundation in the form of a postdoctoral fellowship while at 
the University of Chicago (grants CF15-0552, CF16-0639, and 
CF17-0783) and the research framework provided by the 
Research Computing Center at the University of Chicago.

O.H.S.O, A.M.K, and S.I.V acknowledge CSC -- IT Center for Science for computational resources and Academy of Finland (315596 and 319902) for financial support.

T.J.P acknowledges use of the Iridis high-performance computing resources at the University of Southampton.

\end{acknowledgments}

% Create the reference section using BibTe
\bibliography{refs.bib}

%\newpage
%\section{APPENDIX: The NMR results reported by Tiago Ferreira}

\listoftodos

\end{document}
%
% ****** End of file aiptemplate.tex ******

### 2021-04-27 manuscriptPGPEsuppl.tex
Updates from Overleaf

\documentclass[journal=jpcbfk]{achemso}

\usepackage[version=3]{mhchem}
\usepackage[T1]{fontenc}
\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
%\newcommand{\todo}[1]{\textcolor{red}{#1}}
\usepackage[obeyFinal]{easy-todo}
\usepackage{rotating}
\usepackage{upgreek}				
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{xr-hyper}
\usepackage{soul} % for highlights with \hl{} 
\usepackage{hyperref} % to make TOC titles clickable
\hypersetup{
    colorlinks=true, %set true if you want colored links
    linktoc=all,     %set to all if you want both sections and subsections linked
    linkcolor=black,  %choose some color if you want links to stand out
    citecolor=black,
    filecolor=black,
    urlcolor=black
}

\makeatletter
\newcommand*{\addFileDependency}[1]{% argument=file name and extension
  \typeout{(#1)}
  \@addtofilelist{#1}
  \IfFileExists{#1}{}{\typeout{No file #1.}}
}
\makeatother

\newcommand*{\myexternaldocument}[1]{%
    \externaldocument{#1}%
    \addFileDependency{#1.tex}%
    \addFileDependency{#1.aux}%
}

\myexternaldocument{manuscriptPGPE}

\author{\scriptsize Am{\'e}lie Bacle}
\affiliation{Laboratoire Coop\'{e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit\'{e} de Poitiers, 1 rue Georges Bonnet, 86000 Poitiers, France }

\author{\scriptsize Pavel Buslaev}
\affiliation{\tiny Nanoscience Center and Department of Chemistry, University of Jyväskylä, P.O. Box 35, 40014 Jyväskylä, Finland}
\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, 141701 Dolgoprudny, Russia}

\author{Rebeca Garc{\'i}a Fandi{\~n}o}
\affiliation{Center for Research in Biological Chemistry and Molecular Materials (CiQUS), Universidade de Santiago de Compostela, E-15782 Santiago de Compostela, Spain}
\affiliation{CIQUP, Centro de Investigação em Química, Departamento de Química e Bioquímica, Faculdade de Ciências, Universidade do Porto, Porto, Portugal}

\author{Fernando Favela-Rosales}
\affiliation[Tecnol\'{o}gico Nacional de M\'{e}xico]{Departamento de Ciencias B\'{a}sicas, Tecnol\'{o}gico Nacional de M\'{e}xico, ITS Zacatecas Occidente, M\'{e}xico}

\author{Tiago Ferreira}
\affiliation{Halle, Germany}

\author{Patrick F.J. Fuchs}
\affiliation{Sorbonne Universit\'{e}, Ecole Normale Sup\'{e}rieure, PSL University, CNRS, Laboratoire des Biomol\'{e}cules (LBM), 75005 Paris, France}
\alsoaffiliation{Universit\'{e} de Paris, UFR Sciences du Vivant, 75013, Paris, France}

\author{Ivan Gushchin}
\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, 141701 Dolgoprudny, Russia}

\author{Matti Javanainen}
\affiliation[Czech Academy of Sciences]{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}

\author{Anne M. Kiirikki}
\affiliation{Institute of Biotechnology, University of Helsinki}

\author{Jesper J. Madsen}
\affiliation[University of Chicago]{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
\alsoaffiliation[University of South Florida]{Global and Planetary Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}

\author{Josef Melcr}
\affiliation[Czech Academy of Sciences]{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\alsoaffiliation{Groningen Biomolecular Sciences and Biotechnology Institute 
and The Zernike Institute for Advanced Materials, 
University of Groningen, 9747 AG Groningen, The Netherlands}

\author{Paula Milán Rodríguez}
\affiliation{Sorbonne Université, Ecole Normale Supérieure, PSL University, CNRS, Laboratoire des Biomolécules (LBM), 75005 Paris, France}

\author{Markus S. Miettinen}
% \affiliation[Max Planck Institute of Colloids and Interfaces]{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}

\author{O. H. Samuli Ollila}
\email{samuli.ollila@helsinki.fi}
\affiliation{Institute of Biotechnology, University of Helsinki}

\author{Chris G. Papadopoulos}
\affiliation{Université Paris-Saclay, CEA, CNRS, Institute for Integrative Biology of the Cell (I2BC), 91198, Gif-sur-Yvette, France}

\author{Antonio Pe{\'o}n}
\affiliation[]{Spain}

\author{Thomas J. Piggot}
\affiliation[University of Southampton]{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, United Kingdom}

\author{{\'A}ngel Pi{\~n}eiro}
\affiliation{Departamento de F{\'i}sica Aplicada, Facultade de F{\'i}sica, Universidade de Santiago de Compostela, E-15782 Santiago de Compostela, Spain}

\author{Salla I. Virtanen}
\affiliation{Institute of Biotechnology, University of Helsinki}

%\author{O. H. Samuli Ollila}
%\email{samuli.ollila@helsinki.fi}
%%\homepage[]{Your web page}
%\affiliation{Institute of Organic Chemistry and Biochemistry,
%Academy of Sciences of the Czech Republic, 
%Prague 6, Czech Republic}
%\affiliation{Institute of Biotechnology, University of Helsinki}

\SectionNumbersOn

\renewcommand{\thetable}{S\arabic{table}}%
\renewcommand{\thefigure}{S\arabic{figure}}%
\renewcommand{\thesection}{S\arabic{section}}%
\renewcommand{\thepage}{S\arabic{page}}%

\title{ Supporting Information:\\Conformational plasticity of phospholipid headgroups in simulations and experiments}

\begin{document}
\tableofcontents{}

\clearpage
\section{R-PDLF and SDROSS experiments}

\begin{figure}[]
%  \centering
  \includegraphics[width=0.9\textwidth]{./Figs/POPE_SI.pdf}
  \caption{\label{POPEspectra}
    Solid-state NMR spectroscopy for determining the C--H bond order parameters, $S_{\rm{CH}}$, in the POPE headgroup and glycerol backbone. (A) Chemical structure of POPE showing the labels used to identify different carbons.
    (B) Refocused INEPT $^{13}$C spectrum of POPE MLVs. 
    (C) Headgroup and glycerol backbone R-PDLF dipolar slices. The arrows indicate the splittings used to determine the C--H bond order parameters by using  $|S_{\rm{CH}}|=\Delta\nu/(0.315 \times d_{\rm{CH}})$. The rigid coupling $d_{\rm{CH}}$ used was 22 kHz.
    (D) Experimental S-DROSS curves giving signs of the order parameters measured.  Grey lines are a set of slices taken from a region of the spectrum without peaks, i.e. grey lines are noise profiles, to highlight the signal-to-noise ratio of the measured modulations.}
\end{figure}

\begin{figure}[]
%  \centering
  \includegraphics[width=\textwidth]{./Figs/POPG_SI.pdf}
  \caption{\label{POPGspectra}
    Solid-state NMR spectroscopy for determining the C--H bond order parameters, $S_{\rm{CH}}$, in the POPG headgroup and glycerol backbone. (A) Chemical structure of POPG showing the labels used to identify different carbons.
    (B) Refocused INEPT $^{13}$C spectrum of POPG MLVs.
    (C) Headgroup and glycerol backbone R-PDLF dipolar slices. The arrows indicate the splittings used to determine the C--H bond order parameters by using  $|S_{\rm{CH}}|=\Delta\nu/(0.315 \times d_{\rm{CH}})$. The rigid coupling $d_{\rm{CH}}$ used was 22 kHz.
    (D) Experimental S-DROSS curves giving signs of the order parameters measured.  Note that the S-DROSS modulation for $g_3$ is very close to the noise level. In this case, we have confirmed that the order parameter sign of $g_3$ was negative by performing a measurement with less points in the indirect dimension, more scans and using a higher MAS frequency of 8 kHz (not shown). 
  }
\end{figure}

\clearpage
\section{Lipid ligand names in PDB used in the analysis of conformations of protein-bound lipids}
{\bf PC:} PLC, PX4, 6PL, LIO, HGX, PC7, PC8, P1O, 6O8, XP5, EGY, PLD, SBM, HXG, and PCW \\
{\bf PE:} 8PE, PTY, 3PE, PEH, PEF, 6OE, 6O9, 9PE, PEV, 46E, SBJ, L9Q, PEK, EPH, ZPE, 9TL, 9Y0, 6OU, LOP, and PEE \\
{\bf PG:} PGT, PGK, LHG, 44G, PGV, OZ2, D3D, PGW, DR9, P6L, PG8, H3T, and GOT \\
{\bf PS:} PSF, PS6, Q3G, P5S, D39, PS2, 17F, and 8SP.

\clearpage
\section{Evaluation of simulations against NMR experiments}
\subsection{Conformational ensembles of headgroup and glycerol backbone in PE and PG lipids}

The quality of PE and PG headgroup conformational ensembles in different simulations are evaluated against NMR experiments in figures \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG} using C-H bond order parameters as in our previous studies for PC and PS lipids \cite{botan15,antila19}. Conclusions are the same for all lipids: None of the force fields correctly captures the lipid headgroup conformational ensembles, but CHARMM36 gives results closest to experiments. Most importantly for this work, the CHARMM36 captures the distinct headgroup order parameters for PG and PS lipids observed in NMR experiments (Figs. \ref{HGorderParameters} and \ref{structures} in the main text).

It should be noted that the PG headgroup is biologically abundant R enantiomer in all simulations, while our $^{13}$C NMR experiments has a racemic mixture. Nevertheless, previous $^{2}$H NMR experiments comparing results between different enantiomers concluded that the structural differences between these are minor \cite{wohlgemuth80}.

%The poor performance of headgroup order parameters in Berger model can be probably explained by
%ring like structures seen in Fig. 6 in Ref. \citenum{mukhopadhyay04}, which is a typical feature
%for Berger based lipid force fields containing explicit hydrogen atoms in the head group \cite{zhao08,henin09,dahlberg10}.
%The poor performance of glycerol backbone of Slipids simulations is systemically observed also
%for other lipids in previous studies \cite{botan15,antila19}.
%\todo{Should we comment more the relative quality of different force fields and/or make the subjective force field ranking figures?
%https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/8}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{./Figs/HGorderparametersPE.pdf}
  \caption{\label{HGorderParametersPE}
    C--H bond order parameters, $S_{\rm CH}$, of the PE headgroup ($\beta$ and $\alpha$) and
    glycerol backbone (g$_3$, g$_2$, g$_1$) carbons from NMR experiments (horizontal lines;
    POPE and signs this work, DPPE from Ref.~\citenum{seelig76},
    {\it Escherichia coli} PE from Ref.~\citenum{gally81})
    and MD simulations with different force fields (symbols).
    The light blue areas span 0.04 units around the average of the extremal experimental values,
    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
    The vertical bars shown for most simulation values are not error bars,
    but demonstrate that for these systems we had at least two data sets;
    the ends of the bars mark the extreme values from the sets,
    the symbol marks the measurement-time-weighted average.
  }
\end{figure}

\begin{figure}[!h]
  \centering
  \includegraphics[width=9.0cm]{./Figs/HGorderparametersPG.pdf}
  \caption{\label{HGorderParametersPOPG}
    The headgroup and glycerol backbone order parameters of PG lipids
    from experiments (POPG and signs from this work and from Ref.~\citenum{borle85}, %contains 10mM of PIPES,
    DPPG with 100mM NaCl from Ref.~\citenum{wohlgemuth80},% contains 10mM PIPES and,
    and E.Coli PG results from Ref.~\citenum{gally81})
    and simulations with different force fields.
  }
\end{figure}

\clearpage

\subsection{PC headgroup in mixtures with PE or PG lipids}
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{./Figs/HGorderparametersPCvsPEPG.eps}
  \caption{\label{HGorderparametersPCvsPEPG}
    Modulation of POPC headgroup order parameters with increasing amount of POPE (left) and POPG (right) in bilayer
    from experiments at 298 K \cite{scherer87,macdonald87} and simulations with different force fields
    (temperatures listed in tables \ref{systemsMIX} and \ref{systemsMIX2} are between 298-310 K).
    Signs are determined as discussed in Refs. \citenum{botan15,ollila16}.
  }
\end{figure*}

Headgroup order parameters of PC lipids are unchanged upon addition
of zwitterionic lipids or cholesterol in experiments, but increase
upon addition of negatively charged PG or PS lipids because
headgroup dipole tilts more parallel to the membrane plane after incorporation
of negative charges into the membrane~\cite{seelig87, scherer87,antila19}.
The response of PC headgroup order parameters to the addition of PE or PG lipids
from different simulations is compared with experiments in figure~\ref{HGorderparametersPCvsPEPG}.
None of the simulations reproduce neither the experimentally observed increase in PC headgroup order parameters
with increasing amount of PG nor the related tilting of the headgroup more parallel with the membrane.
Similar observations in our previous work for PS lipids were explained by the overestimated counterion
binding affinity that neutralizes the effect of added negative charge \cite{antila19}.
All simulations except Berger-OPLS predict tilting of P-N headgroup outwards from the membrane and
decrease of PC headgroup order parameters upon addition of PE lipids.
These results are not in line with experiments where the PC headgroup order parameters are not affected by zwitterionic lipids \cite{scherer87}.
The good performance of Berger-OPLS simulations is surprising here because
headgroup conformational ensemble is not very close to experiments in this model and
the response of headgroup order parameters
to cholesterol was significantly overestimated by the Berger/H{\"o}ltje force field in our previous work \cite{botan15}.

In conclusion, more accurate force fields are needed to correctly simulate the interactions between different headgroups.

\clearpage
\subsection{PG headgroup in mixtures with PC lipids}
Changes in other than PC lipid headgroup with changing membrane composition are less
extensively characterized in the literature.
The $\beta$-carbon order parameter in PG headgroup increases
mildly \cite{macdonald87} or is unchanged~\cite{borle85} upon increasing amount
of PC lipids (Fig. \ref{HGorderparametersPGvsPCchange}), but experimental data from
$\alpha$-carbon is not available. Also the tested force fields predict very small changes
for the $\beta$-carbon order parameter, while the P-N vector tilt and its response to the
increased amount of PC varies significantly between force fields in figure \ref{HGorderparametersPGvsPCchange}. 
Therefore, more experimental data and more accurate force fields are still required
to resolve the PG conformational ensembles in mixtures with other lipids.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{./Figs/HGorderparametersPGvsPCchange.eps}
  \caption{\label{HGorderparametersPGvsPCchange}
    Modulation of PG lipid headgroup order parameters with the increasing amount of PC in lipid bilayer
    from experiments at 298 K \cite{borle85,macdonald87} and simulations with different force fields at 310 K.
  }
\end{figure}

\clearpage
\subsection{Calcium binding to POPC:POPG mixtures}

The changes of headgroup order parameters in POPC:POPG mixtures upon addition of CaCl$_2$ between different simulations
and experiments \cite{borle85,macdonald87} are compared in figures \ref{changesWITHCaClPG} (molar ratio 1:1) and \ref{changesWITHCaClPG1PC4} (molar ratio 4:1).
The results are in line with our previous studies:
most force fields overestimate the calcium binding \cite{catte16,antila19}, but CHARMM36 with the NBfix correction underestimates the binding affinity \cite{antila19}, and the implicit inclusion of electronic polarizability using the electronic continuum correction (ECC) improves the results \cite{melcr18,melcr20}.

\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{./Figs/CHANGESwithCaClPG1PC1.eps}
  \caption{\label{changesWITHCaClPG}
    Modulation of headgroup order parameters of POPC ({\it left}) and POPG ({\it right}) in POPC:POPG (1:1)
    mixture upon addition of CaCl$_2$ in 298 K temperature from experiments \cite{borle85,macdonald87} and simulations.
    The $\beta$-carbon order parameter of POPC (dashed line on top left) is not directly measured but
    calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
    The changes with respect to the systems without CaCl$_2$ are shown for other data than
    for the $\alpha$-carbon of POPG for which experimental order parameter is not available.
    Calsium density distributions are shown in figure \ref{CAdensPG}.
    %(right) The headgroup order parameters of PG from PC:PG mixtures as a function CaCl$_2$
    %concentration from experiments \cite{borle85} and CHARMM36 simulations.
  }
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{./Figs/CAdensPG1PC1.eps}
  \caption{\label{CAdensPG}
    Calcium ion density profiles along membrane normal from simulations of POPC:POPG (1:1) mixtures with different force fields.
    The changes in the order parameters upon addition of CaCl$_2$ are compared with experiments in figure \ref{changesWITHCaClPG}.
  }
\end{figure}

\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{./Figs/CHANGESwithCaClPG1PC4.eps}
  \caption{\label{changesWITHCaClPG1PC4}
    Modulation of headgroup order parameters of POPC ({\it left}) and POPG ({\it right}) in POPC:POPG (4:1)
    mixture upon addition of CaCl$_2$ in 298 K temperature from experiments \cite{macdonald87} and simulations.
    The changes with respect to the systems without CaCl$_2$ are shown for other data than
    for the $\alpha$-carbon of POPG for which experimental order parameter is not available.
  }
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{./Figs/CAdensPG1PC4.eps}
  \caption{\label{CAdensPG1PC4}
    Calcium ion density profiles along membrane normal from simulations of POPC:POPG (4:1) mixtures with different force fields.
  }
\end{figure}

\clearpage
\section{Dihedral angle distributions}

\subsection{Dihedral angles of PC, PE, PG and PS headgroups}

%The dihedral angle distributions are shown in Fig. \ref{DIHdists}
%and relative energies in Fig. \ref{structures} in the main text.

\begin{figure}[]
  \centering
  \includegraphics[width=7.0cm]{./Figs/DIHEDRALS.eps}
  \caption{\label{DIHdists}
    Heavy atom dihedral angle distributions from CHARMM36 simulations that correctly capture the order parameter differences between lipid headgroups.
  }
\end{figure}

\clearpage
\subsection{Changes in headgroup conformations upon addition of charged surfactants or CaCl$_2$}

\begin{figure}[]
  \centering
  \includegraphics[width=12.0cm]{./Figs/HGorderparametersPCvsSURFdihedrals.eps}
  \caption{\label{HGorderparametersPCvsSURFdihedrals}
    Changes in PC headgroup conformational ensembles upon increasing the amount of positive charge in bilayer,
    characterized by the heavy atom dihedral distributions, from CHARMM36 simulations.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=12.0cm]{./Figs/DIHEDRALSlipid17eccWITHCaClPOPC.eps}
  \caption{\label{DIHSwithCAlipid17eccPOPC}
    Changes in POPC lipid17ecc dihedrals with increasing amount of CaCl$_2$.
  }
\end{figure}

\begin{figure}[]
  \centering
   \includegraphics[width=12.0cm]{./Figs/DIHEDRALScharmm36WITHCaClPOPC.eps}
  \caption{\label{DIHSwithCAcharmm36POPC}
    Changes in POPC CHARMM36 dihedrals with increasing amount of CaCl$_2$.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=16.0cm]{./Figs/DIHEDRALSslipidsWITHCaClPOPG.eps}
  \caption{\label{DIHSwithCAslipidsPOPG}
    Changes in POPG Slipids dihedrals with increasing amount of CaCl$_2$.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=16.0cm]{./Figs/DIHEDRALSlipid17WITHCaClPOPG.eps}
  \caption{\label{DIHSwithCAlipid17POPG}
    Changes in POPG lipid17 dihedrals with increasing amount of CaCl$_2$.
  }
\end{figure}

\clearpage
\section{Simulated systems}

The simulated systems of pure PE and PG bilayers without additional ions
are listed in Tables~\ref{systemsPE} and \ref{systemsPG},
and lipid mixtures with additional ions in Tables~\ref{systemsMIX} and \ref{systemsMIX2}.
The simulations were analyzed using preliminary versions of the NMRlipids databank
(\url{www.nmrlipids.fi}, \url{github.com/NMRlipids/MATCH} and \url{https://github.com/NMRLipids/NMRlipidsIVPEandPG/tree/master/Data/Simulations})
and unique naming convention for lipid atoms (\url{http://nmrlipids.blogspot.com/2015/03/mapping-scheme-for-lipid-atom-names-for.html}),
which enable automatic analysis of simulations with different force fields with varying atom naming conventions.
The automatic analyses were implemented using MDAnalysis \cite{agrawal11,gowers16} and MDTraj \cite{mcgibbon15}
python libraries, and tools in the GROMACS sofware package \cite{gromacsMANUAL}.
All codes are available from the project's GitHub repository \cite{NMRlipidsIVbgit}.

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\left\langle 3\cos^2\theta -1 \right\rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid in the system. The average and
the standard error of the mean were then calculated over different lipids.
Code for all atom simulations is available in Ref.~\citenum{MATCHgit} (\texttt{scripts/calcOrderParameters.py}).
For united atom simulations, we first constructed trajectories including hydrogens with ideal geometry using either \texttt{buildH} program~\cite{buildH} or (\texttt{scratch/opAAUA\_prod.py}) in  Ref.~\citenum{MATCHgit}, and the order parameters were then calculated from these trajectories. This approach has been tested against trajectories with explicit hydrogens and the deviations in order parameters are small \cite{buildH,piggot17}.\\
%\todo{BuildH program is now cited with a direct link to the GitHub repo. I think that a release to Zenodo would be nice in the final publication.}\\

\begin{table*}[htb]
%
%\begin{sidewaystable*}[!p]
  \centering
  \caption{List of MD simulations with PE lipids.
    % The salt concentrations calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
    % these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
    % The lipid force fields named as in our previous work~\cite{botan15}.
  }\label{systemsPE}
  \begin{minipage}[t]{\textwidth}
  \resizebox{\columnwidth}{!}{
%    \begin{tabular}{l c c r r r r r r c c}
    \begin{tabular}{l c r r r r r c c}
      %\hline
      % some footnotes are not visible in typeset-MS (pdf)
      lipid 	& force field for lipids  %& NaCl (M) 
      		& \footnote{Number of lipid molecules}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   %& \footnote{Number of additional cations}N$_{\rm c}$   
		& \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
      \hline
      POPE & CHARMM36 \cite{klauda10}           %&0       
      		& 144	& 5760  %&0    
		& 310  & 500          & 400          & \citenum{charmm36POPEfiles} \\
      POPE & CHARMM36 \cite{klauda10}           %& 0      
      		& 500       & 25000 %& 0   
		&  310  & 500 & 100 & \citenum{POPEcharmm} \\
%   POPE	& CHARMM36 \cite{klauda10}           %& 0.11
%		& 500       & 25000 & 50  &  310  & 500 & 100 & \cite{POPEcharmm150mMNaCl} \\
      POPE	& CHARMM36-UA \cite{lee14}         %&0       
      		& 336	& 15254 %&0    
		& 310  & 2$\times$200 & 2$\times$100 & \citenum{charmm36uaPOPEfiles}  \\
      \hline
      DPPE	& Slipids \cite{jambeck12b}    %&0    
      		& 288 	& 9386  %&0    
		& 336  & 200 & 100 & \citenum{slipidsDPPEfiles}  \\
      POPE	& Slipids \cite{jambeck12b} %&0    
      		& 336	& ?     %&0    
		& 310  & 2$\times$200 &  2$\times$100 & \citenum{slipidsPOPEfiles}  \\
      POPE	& Slipids \cite{jambeck12b}            %& 0
		& 500 & 25000 %& 0   
		&  310  & 500 & 100 & \citenum{POPEslipids} \\
%      POPE  & Slipids / {\AA}qvist \cite{jambeck12b,aqvist90}  %& 0.11
%		& 500 & 25000 & 50  &  310  & 500 & 100 & \cite{POPEslipids150mMNaCl} \\
      \hline
      DPPE	& GROMOS-CKP    \cite{piggot11}      %&0
          	& 128	& 3655  %&0    
		& 342  & 2$\times$500 & 2$\times$400 & \citenum{gromosCKPdppe} \\
%      POPE  & GROMOS-CKP    \cite{??}      %&0
%    		& 128	& 3552  &0    & 313  & 2$\times$500 & 2$\times$400 & \cite{gromosCKPpope} \\
      POPE	& GROMOS-CKP    \cite{piggot11}      %&0
          	& 500	& 25000 %&0    
		& 310  & 500 & 100 & \citenum{gromosCKPpopeT310} \\
%      POPE  & GROMOS-CKP    \cite{??}      &0.11 & 500	& 25000 &50   & 310  & 500 & 100 & \cite{gromosCKPpopeT310150mMNaCl} \\
%      DOPE  & GROMOS-CKP    \cite{??}      &0    & 128	& 4789  &0    & 271  & 2$\times$500 & 2$\times$400 & \cite{gromosCKPdope} \\
      \hline
      POPE	& GROMOS 43A1-S3 \cite{chiu09}     %&0
          	& 128	& 3552     %&0    
		& 313  & 2$\times$200 & 2$\times$100 & \citenum{gromos43a1s3POPEfiles}  \\
      \hline
      POPE	& OPLS-UA/HG-H \cite{Ulmschneider09}   %&0    
      		& 128	& 3328     %&0    
		& 303  & 2$\times$200 & 2$\times$100 & \citenum{OPLSuaWvdWPOPEfiles} \\
      POPE	& OPLS-UA \cite{Ulmschneider09}            %&0
      		& 128	& 3328     %&0    
		& 303  & 2$\times$200 & 2$\times$100 & \citenum{OPLSuaPOPEfiles} \\
      \hline
      POPE	& OPLS-MacRog \cite{rog16}     %&0    
      		& 144	& 5760     %&0    
		& 310  & 500 & 350 & \citenum{MacRogPOPEfiles} \\
%   POPE	& OPLS-MacRog \cite{rog16}     &0    & 128	& 5120     &0    & 300  & 500 & 300 & \cite{MacRogPOPEfilesT300K} \\
      \hline
      POPE	& Berger-POPE-2004 \cite{devries04}       %&0    
      		& 128	& 3552  %&0    
		& 303  & 2$\times$200 & 2$\times$100 & \citenum{bergerPOPEfiles}  \\
      POPE	& Berger-POPE-2018 \cite{devries04}      %&0    
      		& 128	& 3552  %&0    
		& 303  & 2$\times$200 & 2$\times$100 & \citenum{berger2POPEfiles}  \\
%   DOPE  & Berger-POPE-2004 \cite{??}       &0    & 128	& 4789  &0    & 271  & 2$\times$200 & 2$\times$100 & \cite{bergerDOPEfiles}  \\
%   DOPE  & Berger-POPE-2018 \cite{??}      &0    & 128	& 4789  &0    & 271  & 2$\times$300 & 2$\times$100 & \cite{berger2DOPEfiles} \\ 
      \hline
      POPE	& Lipid17 \cite{gould18} %& 0      
      		& 500 & 25000 %& 0  
		&  310  & 500 & 100 & \citenum{POPElipid17} \\
%      POPE             & Lipid17 \cite{gould18} & 0.11   & 500 & 25000 & 50  &  310  & 500 & 100 & \cite{POPElipid17150mMNaCl} \\
    \end{tabular}
    }
  \end{minipage}
  %\todo{Which ion model is used in \cite{POPEcharmm150mMNaCl}?} \\
\end{table*}
%  \end{sidewaystable*}

      \begin{table*}[htb]
  %\begin{sidewaystable*}[!p]
  \centering
  \caption{List of MD simulations with PG lipids.
    % The salt concentrations calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
    % these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
    % The lipid force fields named as in our previous work~\cite{botan15}.
  }\label{systemsPG}
  \begin{minipage}[t]{\textwidth}
  \resizebox{\columnwidth}{!}{
    \begin{tabular}{l c c r r r r r r c c}
      %\hline
      % some footnotes are not visible in typeset-MS (pdf)
      lipid/counter-ions & force field for lipids / ions &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
      \hline
      POPG/K$^+$  & CHARMM36 \cite{venable13}        & 118& 4110    & 298  & 100 & 100 & \citenum{CHARMM36popg}  \\
%      POPG             & CHARMM36 \cite{venable13}        & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGcharmm150mMNaCl} \\
      POPG             & CHARMM36 \cite{venable13}                & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGcharmm} \\
      \hline
      POPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}            & 288 	& 10664   & 298  & 250 & 100 & \citenum{slipidsPOPGfiles} \\
      DPPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}           & 288 	& 11232   & 314  & 200 & 100 & \citenum{slipidsDPPGfiles} \\
      DPPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}           & 288 	& 11232   & 298  & 400 & 100 & \citenum{slipidsDPPGfilesT298K} \\
      POPG         & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}      & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGslipids} \\
%      POPG         & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    & 0.11    & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGslipids150mMNaCl} \\
      \hline
      POPG             & LIPID17 / Dang \cite{gould18,smith94,dang06}     & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGlipid17} \\
%      POPG             & LIPID17 \cite{gould18}         & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGlipid17150mMNaCl} \\
      \hline
      POPG             & GROMOS-CKP \cite{piggot11}         & 500 & 25000 &  310  & 500 & 100 & \citenum{POPGgromosCKP} \\
%      POPG             & GROMOS-CKP \cite{??}         & 0.11           & 500 & 25000 & 49 &  310  & 500 & 100 & \cite{POPGgromosCKP150mMNaCl} \\
    \end{tabular}
    }
  \end{minipage}
\end{table*}
% \end{sidewaystable*}

\begin{table*}[htb]
%  \begin{sidewaystable*}[!p]
  \centering
  \caption{List of MD simulations with PE and PG lipids mixed with PC.
    % The salt concentrations calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
    % these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
    % The lipid force fields named as in our previous work~\cite{botan15}.
  }\label{systemsMIX}
  \begin{minipage}[t]{\textwidth}
    \resizebox{\columnwidth}{!}{
    \begin{tabular}{l c c r r r r r r c c}
      %\hline
      % some footnotes are not visible in typeset-MS (pdf)
      lipid/counter-ions & force field for lipids / ions & CaCl$_2$\,(M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
      \hline
%      POPC                   & CHARMM36 \cite{??}        & 0.11      & 0  & 500     & 25000 & 48  &  310  & 500 & 100 & \cite{POPCcharmm150mMNaCl301K}  \\
%      POPC:POPG (7:3)        & CHARMM36 \cite{??}        & 0.11      & 0  & 350     & ?     & ?   &  310  & 500 & 100 & \cite{POPC7POPG1charmm36NaCl}  \\
      POPC                   & CHARMM36 \cite{klauda10}            & 0  & 500     & 25000 & 0   &  310  & 500 & 100 & \cite{POPCcharmmT310K}  \\
      POPC:POPG (7:3)        & CHARMM36 \cite{klauda10,venable13}    & 0  & 350     & 25000 & 0   &  310  & 500 & 100 & \cite{POPC7POPG1charmm36}  \\
      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}     & 0  & 150:150 & 31500 & 0   &  298  & 500 & 400 & \cite{CHARMM36POPCPOPG5050} \\ 
      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}     & 0.1 & 150:150 & 31329 & 57 &  298  & 400 & 300 & \cite{CHARMM36POPCPOPG5050100mMCaCl} \\
      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}     & 1.08 & 150:150 & 29766 & 578  &  298  & 500 & 400 & \cite{CHARMM36POPCPOPG50501000mMCaCl} \\
      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 0  & 350:88 & 26280 & 0  &  298  & 500 & 400 & \cite{CHARMM36POPCPOPG4010} \\
      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 0.1 & 350:88 & 26280 & 47  &  298  & 500 & 400 & \cite{CHARMM36POPCPOPG4010100mMCaCl} \\
      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 1.0 & 350:88 & 24927 & 451  &  298  & 500 & 400 & \cite{CHARMM36POPCPOPG40101000mMCaCl} \\
      \hline
      POPC             & CHARMM36 \cite{klauda10}        &0          & 256 & 8704 & 0  &  300  & 300 & 250 & \cite{POPCcharmm300K} \\
      POPC:POPE (1:1)  & CHARMM36 \cite{klauda10,venable13}         & 0  & 128 & 8704 & 0  &  300  & 300 & 250 & \cite{POPC1POPE1charmm36} \\
      \hline
      POPC             & OPLS-MacRog \cite{rog16}     &0           & 128 & 5120 & 0  &  300  & 500 & 300 & \cite{POPCmacrog300K} \\
      POPC:POPE (1:1)  & OPLS-MacRog \cite{rog16}     &0           & 128 & 5120 & 0  &  300  & 500 & 300 & \cite{POPC1POPE1macrogT300K} \\
      \hline
      POPC             & Slipid \cite{jambeck12b}     &0          & 512 & 23943 & 0  &  298  & 170 & 100 & \cite{POPCslipid298K} \\
      POPC:POPE (1:1)  & Slipid \cite{jambeck12b}     &0          & 128 & 5120  & 0  &  298  & 500 & 300 & \cite{POPC1POPE1slipidT298K} \\
     \hline
      POPC                   & GROMOS-CKP / ?? \cite{piggot12,??}  & 0  & 500     & 25000 & 0   &  310  & 500 & 100 & \cite{POPCgromosCKPT310K}  \\
      POPC:POPG (7:3)        & GROMOS-CKP / ?? \cite{piggot11,piggot12,??}  & 0  & 350:150 & 25000 & 0   &  310  & 500 & 100 & \cite{POPC7POPG3gromosCKPT310K} \\
     \hline
      POPC                   & Slipid \cite{jambeck12b}  & 0      & 500     & 25000 & 0   &  310  & 500 & 100 & \cite{POPCslipid301K}  \\
      POPC:POPG (7:3)        & Slipid / {\AA}qvist \cite{jambeck12b,aqvist90} & 0  & 350:150 & 25000 & 0   &  310  & 500 & 100 & \cite{slipidPOPC70POPG30T310K} \\
      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0  & 128:128 & 12800 & 0   &  298  & 500 & 400 & \cite{slipidPOPC50POPG50T298K} \\
      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0.1  & 128:128 & 12800 & 23  &  298  & 500 & 400 & \cite{slipidPOPC50POPG50T298K} \\
      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0.2  & 128:128 & 12800 & 46  &  298  & 1500 & 500 & \cite{slipidPOPC50POPG50T298K} \\
      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0.5  & 128:128 & 12800 & 115 &  298  & 1500 & 500 & \cite{slipidPOPC50POPG50T298K} \\
      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 1.0  & 128:128 & 12800 & 230 &  298  & 1500 & 500 & \cite{slipidPOPC50POPG50T298K} \\
    \end{tabular}
    }
  \end{minipage}
    \todo{ion model for GROMOS-CKP?} \\ % TJP: Most likely standard GROMOS but not my simulations so I've left the ion reference
\end{table*}
%\end{sidewaystable*}

\begin{table*}
%  \begin{sidewaystable*}[!p]
  \centering
  \caption{List of MD simulations with PE and PG lipids mixed with PC.
    % The salt concentrations calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
    % these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
    % The lipid force fields named as in our previous work~\cite{botan15}.
  }\label{systemsMIX2}
  \begin{minipage}[t]{\textwidth}
    \resizebox{\columnwidth}{!}{
    \begin{tabular}{l c c r r r r r r c c}
      %\hline
      % some footnotes are not visible in typeset-MS (pdf)
      lipid/counter-ions & force field for lipids / ions & CaCl$_2$\,(M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
      \hline
      POPC:POPG (4:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}       & 0  & 350:88 & 26265 & 0  &  298  & 400 & 350 & \cite{Lipid17POPCPOPG8020} \\
      POPC:POPG (4:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}    & 0.1& 350:88 & 26124 & 47 &  298  & 400 & 250 & \cite{Lipid17POPCPOPG8020100mMCaCl} \\
      POPC:POPG (4:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}    & 1.0& 350:88 & 24840 & 475 &  298  & 1200 & 200 & \cite{Lipid17POPCPOPG80201000mMCaCl} \\
      POPC:POPG (1:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}   & 0  & 150:150 & 31572 & 0  &  298  & 320 & 200 & \cite{Lipid17POPCPOPG5050} \\
      POPC:POPG (1:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}   & 0.1& 150:150 & 31401 & 57 &  298  & 718 & 198 & \cite{Lipid17POPCPOPG5050100mMCaCl} \\
      POPC:POPG (1:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}  & 1.0& 150:150 & 29865 & 569 &  298  & 720 & 200 & \cite{Lipid17POPCPOPG50501000mMCaCl} \\
      \hline
      POPC:POPG (4:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     &0     & 350:88 & 26265 & 0  &  298  & 400 & 300 & \cite{Lipid17eccPOPCPOPG8020} \\
      POPC:POPG (4:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 0.1& 350:88 & 26124 & 47 &  298  & 400 & 300 & \cite{Lipid17eccPOPCPOPG8020100mMCaCl} \\
      POPC:POPG (4:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 1.0& 350:88 & 24840 & 475 &  298  & 400 & 300 & \cite{Lipid17eccPOPCPOPG80201000mMCaCl} \\
      POPC:POPG (1:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 0  & 150:150 & 31572 & 0  &  298  & 347.8 & 333 & \cite{Lipid17eccPOPCPOPG5050} \\
      POPC:POPG (1:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 0.1& 150:150 & 29865 & 54 &  298  & 400 & 300 & \cite{Lipid17eccPOPCPOPG5050100mMCaCl} \\
      POPC:POPG (1:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 1.0& 150:150 & 29865 & 569 &  298  & 600 & 400 & \cite{Lipid17eccPOPCPOPG50501000mMCaCl} \\
      \hline
      POPC             & Berger-POPC-?? \cite{??} \todoi{This is probable not plain berger, correct force filed should be described.}  & 0  & 256 & 10240 & 0  &  300  & 300 & 200 & \cite{POPCberger300K} \\
      POPC:POPE (1:1)  & Berger-POPE-04~\cite{devries04}  & 0  & 128 & 11008 & 0  &  300  & 300 & 200 & \cite{POPC1POPE1berger} \\
      %POPC:DOPE (1:1)  & Berger \cite{??}  \todoi{This is probable not plain berger, correct force filed should be described.} &0          & 0  & 128 & 10240 & 0  &  300  & 300 & 200 & \cite{POPC1DOPE1berger} \\
    % \hline
    %  DOPC             & Berger \cite{??}  \todoi{This is probable not plain berger, correct force filed should be described.} &0          & 0  & 256 & 11008 & 0  &  300  & 300 & 200 & \cite{DOPCberger300K} \\
      %DOPC:DOPE (1:1)  & Berger \cite{??}   \todoi{This is probable not plain berger, correct force filed should be described.}  &0          & 0  & 128 & 11008 & 0  &  300  & 300 & 200 & \cite{DOPC1DOPE1berger} \\
    \end{tabular}
    }
  \end{minipage}
  \todo{Citation and description for Berger-POPC-?? model?} \\
\end{table*}
%\end{sidewaystable*}

\clearpage

\subsection{CHARMM36}

\noindent {\it POPE} A lipid bilayer, consisting of a total of 144 POPE molecules, distributed equally between the two leaflets was set using CHARMM-GUI \cite{lee16}. The bilayer was solvated by 5760 water molecules (40 per lipid). The random initial configuration and topologies were generated using the CHARMM-GUI web portal, which provides  GROMACS-compatible simulation input files \cite{lee16}. CHARMM36 lipid parameters \cite{klauda10} were used for POPE, whereas the CHARMM-specific TIP3P water model \cite{jorgensen83} was used for water. The bilayer was simulated for 500~ns using GROMACS 2018.6 at 310~K.

The recommended simulation parameters for CHARMM36 force field in GROMACS were used \cite{lee16}. Namely, buffered Verlet lists were used to keep track of neighbouring atoms \cite{Pall13}. The Lennard-Jones potential was cut off at 1.2~nm with the forces switched to 0 between 1.0~nm and the cutoff value. The smooth PME algorithm \cite{darden93,essman95} was used to account for long-range electrostatics. The temperatures of the lipid and the solvent were separately coupled to a Nos\'{e}-Hoover thermostat \cite{nose84,hoover85} with a time constant of 1~ps and a target temperature of 310~K. The system was coupled to a semi-isotropic (isotropic on the membrane plane) Parrinello--Rahman barostat \cite{parrinello81} with a time constant of 5~ps, a reference pressure of 1~bar, and compressibility of 4.5$\times$10$^{-5}$~1/bar. The bonds with hydrogen atoms were constrained using p-LINCS \cite{hess07,hess97}. The simulation files are available at Ref.~\citenum{charmm36POPEfiles}.

%\noindent {\it POPE with additional NaCl} \todo{Simulation details by A. Peon.}

\noindent {\it POPG} Lipid bilayer containing 118 POPG molecules, 4110 TIP3P water molecules, and 118 potassium ions was build using CHARMM-GUI \cite{lee16}.
The system was simulated 100~ns, coupled to 298 K using Nose-Hoover~\cite{nose84,hoover85} thermostat and 1 bar with
semi-isotropic Parrinello-Rahman~\cite{parrinello81} pressure coupling.
The used default parameters and force field files from CHARMM-GUI were used.
The used files are available from \citenum{CHARMM36popg}.

\todo{Simulation details for larger simulation by A. Peon.}

%\noindent {\it POPG with additional NaCl} \todo{Simulation details by A. Peon.}

\noindent {\it POPC:POPE mixtures} 
A pure POPC system and a 50:50 POPC:POPE mixture were built and equilibrated using CHARMM-GUI~\cite{lee16}.
They contained 256 lipids (for the mixture 64 lipids per leaflet for each species) and 34 water molecules per lipid.
No ions were added.
The production simulations were run for 300 ns with a time step of 2 fs.
The first 50 ns were discarded for the analysis.
The simulations were run with the GROMACS 2016.4~\cite{abraham2015gromacs} version.
The v-rescale thermostat~\cite{bussi07} was used with a temperature of 300 K and a time constant of 1 ps; lipids and water were coupled separately to the heat bath.
Pressure was kept constant at 1 bar using a semi-isotropic Parrinello--Rahman barostat~\cite{parrinello81} with a time constant of 5.0 ps.
A real space cut-off of 1.2 nm was employed for electrostatic interactions while the long-range part was evaluated using the PME method~\cite{darden93,essman95}.
A force-based switching function was used to switch the Lennard-Jones forces to zero over a range of 1--1.2 nm.
All bonds with hydrogen atoms were constrained with the LINCS algorithm~\cite{hess07,hess97}.
Water molecules were kept rigid with the SETTLE algorithm~\cite{miyamoto92}.
The simulation files are available from Ref.~\citenum{POPCcharmm300K} (pure POPC) and~\citenum{POPC1POPE1charmm36} (POPC:POPE mixture).

\noindent {\it POPC:POPG 1:1 and POPC:POPG 4:1 mixtures with additional calcium} 
The initial structures were built with CHARMM-GUI Membrane Builder \cite{lee16}. The TIP3P water model was used to solvate the systems. The simulations were run for 400~ns with timestep 2~fs and the first 100~ns were discarded as equilibration time. The simulations were run with GROMACS version 2020.2~\cite{pall20}. The Nose-Hoover thermostat~\cite{nose84,hoover85} was used with temperature of 298~K and the time constant for temperature coupling was 1.0~ps. The semi-isotropic Parrinello-Rahman barostat~\cite{parrinello81} was used with reference pressure 1.0~bar and with a time constant of 5.0~ps with compressibility of 4.5e-5~bar$^{-1}$. Long range electrostatic interactions were calculated with the PME method. All bonds with hydrogen atoms were constrained with LINCS algorithm.
The simulation files are available from Refs.~\citenum{CHARMM36POPCPOPG5050,CHARMM36POPCPOPG5050100mMCaCl,CHARMM36POPCPOPG50501000mMCaCl,CHARMM36POPCPOPG4010,CHARMM36POPCPOPG4010100mMCaCl,CHARMM36POPCPOPG40101000mMCaCl}.

\noindent {\it POPC and POPC:POPG (7:3) mixture} \todo{Simulation details by A. Peon.}

\noindent {\it POPC and cationic surfactant (dihexadecyldimethylammonium) mixture}
Intial structures were taken from similar previously published \cite{melcr18} simulations with Amber lipid14 force field,
which are available from Ref. \citenum{POPClipid14T313K,POPClipid1410perCATsurfT313K,POPClipid1420perCATsurfT313K,POPClipid1430perCATsurfT313K,POPClipid1442perCATsurfT313K,POPClipid1450perCATsurfT313K}.
Default simulations parameters and force field files from CHARMM-GUI \cite{lee16} were used, except for dihexadecyldimethylammonium
for which the atom types and partial charges of Amber lipid14 parameters from previous work \cite{melcr18} were modified to correspond Charmm36
force field. 
Systems contained 50 POPC molecules, 3983 water molecules, and 12, 30, 44, or 88 dihexadecyldimethylammonium molecules.
Chloride ions were used as counterions for dihexadecyldimethylammonium.
Reference system without cationic surfactants contained 200 POPC and 9000 water molecules.
Systems were simulated 200 ns (the first 20 ns was discarded as an equilibration period)
using Gromacs 5 \cite{abraham2015gromacs} at the temperature of 313 K.
All simulation files are available from Refs. \citenum{charmmPOPC313K,CHARMM36cationicSURF}.

\subsection{CHARMM36-UA}

\noindent {\it POPE} Data is available at \cite{charmm36uaPOPEfiles}. CHARMM36-UA POPE simulations were performed for 200 ns using GROMACS version 5.0.6. A topology for POPE was constructed using the standard CHARMM36 all-atom PE head group combined with the original CHARMM36-UA lipid tail parameters \cite{lee14}. Simulations were performed using a hexagonal periodic box containing 336 POPE lipids. This membrane was constructed using a MARTINI force field \cite{marrink07} coarse-grained self-assembly simulation followed by reverse-mapping to an atomic resolution. Simulations were performed using standard CHARMM lipid simulation settings; further details are available at \cite{charmm36uaPOPEfiles}.

%\subsection{MacRog}

%\subsection{Lipid17}

\subsection{Slipids}
\noindent {\it POPE} Data is available at \cite{slipidsPOPEfiles}. Slipids POPE simulations were performed for 200 ns using GROMACS version 5.0.6. Simulations employed the original Slipids POPE parameters \cite{jambeck12b} and employed the same starting structure as the CHARMM36-UA POPE simulations. Slipids simulations employed standard AMBER force field cut-offs of 1.0 nm, previously validated for use with the Slipids force field \cite{piggot17}. Further simulation details are available at \cite{slipidsPOPEfiles}.

%\noindent {\it POPE with additional NaCl} \todo{Simulation details by A. Peon.
%  I have assumed that ion parameters are default Slipids, i.e., {\AA}qvist, please correct if this is not true.
%}

\noindent {\it DPPE with 288 lipids.} The starting structure for simulation
with 288 DPPE lipids and 9386 water molecules
was constructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}. The TIP3P~\cite{jorgensen83} water model was used to solvate the system.
Simulation was performed for 200~ns, and the last 100~ns were used for the analysis. Simulation was carried out within the NPT ensemble using the GROMACS 5.0.4 package~\cite{abraham2015gromacs}. Timestep of 2~fs was used with the leapfrog integrator. The Nos\'{e}--Hoover thermostat~\cite{nose84,hoover85} was used with reference temperature of 336~K and a relaxation time constant of 0.5~ps; lipids and water were coupled separately to the heat bath. Pressure was kept constant at 1.013~bar using a semi--isotropic Parrinello--Rahman
barostat~\cite{parrinello81} with a time constant of 10.0~ps. Long-range electrostatic interactions were calculated using the PME method~\cite{darden93,essman95}. A real space cut-off of 1.0~nm was employed with grid spacing of 0.12~nm in the reciprocal space. Lennard-Jones potentials were cut off at 1.4~nm, with a dispersion correction applied to both energy and pressure. All covalent bonds in lipids were constrained using the LINCS algorithm~\cite{hess97}, whereas water molecules were constrained using SETTLE~\cite{miyamoto92}. Twin-range cutoffs, 1.0~nm and 1.6~nm, were used for the neighbor lists with the long-range neighbor list updated every 10 steps.

\noindent {\it POPG with 288 lipids.} The starting structure for simulation
with 288 POPG lipids, 10664 water molecules and 288 Na ions
was constructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}. The TIP3P~\cite{jorgensen83} water model was used to solvate the system and Ions are described by the parameters derived by \AA qvist~\cite{aqvist90}.
Simulation was performed for 250~ns, and the last 100~ns were used for the analysis. Same simulation conditions as DPPE with reference temperature of 298~K.

%\noindent {\it POPG with additional NaCl} \todo{Simulation details by A. Peon.
%  I have assumed that ion parameters are default Slipids, i.e., {\AA}qvist, please correct if this is not true.}

\noindent {\it DPPG with 288 lipids.} The starting structure for simulation
with 288 DPPG lipids, 11232 water molecules and 288 Na ions
was constructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}. The TIP3P~\cite{jorgensen83} water model was used to solvate the system and Ions are described by the parameters derived by \AA qvist~\cite{aqvist90}. For the 298~K temperature, simulation was performed for 400~ns, and the last 100~ns were used for the analysis. For the 314~K temperature, simulation was performed for 200~ns, and the last 100~ns were used for the analysis. Same simulation conditions as DPPE for both temperatures.

\noindent{\it POPC:POPE mixture}
A POPC/POPE bilayer with its lipids distributed evenly among the leaflets was generated by from a pure POPC bilayer by removing and renaming atoms in the head group region. The bilayer contained a total of 100 POPC and 100 POPE lipids, and it was solvated by 45 water molecules per lipid (for a total of 9000 water molecules).
The Slipids force field \cite{jambeck12,jambeck12b,jambeck13} was used for lipids, and the TIP3P model \cite{jorgensen83} for water. A 300~ns-long simulation was performed using GROMACS 2019.4 \cite{abraham2015gromacs}. The simulation parameters were equal to those used for the POPC/POPG mixture with additional CaCl$_2$ (see below). 
The simulation data are available at Ref.~\citenum{POPC1POPE1slipidT298K}.

%\noindent {\it POPC:POPG mixture with additional NaCl} \todo{Simulation details by A. Peon.
%  I have assumed that ion parameters are default Slipids, i.e., {\AA}qvist, please correct if this is not true.}

\noindent {\it POPC:POPG mixture with additional CaCl$_2$} 
A lipid bilayer consisting of a total of 256 lipids (128 POPC + 128 POPG) spread equally between the two leaflets was generated using CHARMM-GUI \cite{lee16}. The membrane was solvated by 50 water molecules per lipid (a total of 12800), and 128 Na$^{+}$ counter ions for the POPG charges. The initial random configuration was equilibrated using the CHARMM-GUI protocol and using the CHARMM36 force field \cite{klauda10}. Next, additional ions were added to obtain initial CaCl$_2$ concentrations of 0, 100, 200, 500, or 1000~mM (0/0, 23/46, 46/92, 115/230, 230/460 Ca$^{2+}$/Cl$^{-}$ ions) The systems were energy-minimized and equilibrated using the Slipids force field \cite{jambeck12,jambeck12b,jambeck13} before production simulations at 298~K. The ion parameters by \citeauthor{dang06} were used \cite{dang06}.

The neighbour lists with a cutoff of 1.0~nm were updated every 10 simulation steps. The smooth PME algorithm was used to calculate long-range electrostatics \cite{darden93,essman95}. The Lennard-Jones potential was cut off at 1.0~nm, and the dispersion corrections \cite{shirts07} were applied to energy and pressure. The stochastic velocity rescaling thermostat \cite{bussi07} with a time constant of 0.5~ps and a target temperature of 298~K was applied separately to lipids and the solvent. A constant pressure of 1~bar was maintained by a Parrinello--Rahman barostat \cite{parrinello81} with a time constant of 10~ps and compressibility of 4.5$\times$10$^{-5}$~1/bar. The pressure coupling was performed semi-isotropically with the two simulation box vectors aligned along the membrane plane considered isotropic. All bonds were constrained using the p-LINCS algorithm \cite{hess97,hess07}.

The systems simulated for 1500~ns (CaCl$_2$-containing systems) or 500~ns (systems without CaCl$_2$). The simulations were performed using GROMACS 2019.4 \cite{abraham2015gromacs}, and the simulation files are available at Ref.~\citenum{slipidPOPC50POPG50T298K}.

\subsection{Berger}

Following the earlier convention in the NMRlipids Project~\cite{botan15}, 
for the Berger-based models we use the following naming convention: 
Berger - \{{\it molecule name}\} - \{{\it year when model published first time}\} \{{\it citation}\}.
%The reason is that there are several different molecular topologies which are using the non-bonded parameters originally
%developed by Berger et al.~\cite{berger97}. Thus the common factor in the Berger based models are the non-bonded parameters,
%while the molecule specific parameters might somewhat vary. However, the majority of the molecular level topologies are 
%relying (especially for the glycerol backbone and headgroup) on the parameters originally introduced by Marrink et al.~\cite{marrink98}.

\noindent {\it POPE} Data are available at Ref.~\citenum{bergerPOPEfiles} for Berger-POPE-2004~\cite{devries04} and at Ref.~\citenum{berger2POPEfiles} for Berger-POPE-2018~\cite{berger2POPEfiles}. Simulations of POPE membranes using the Berger force field employed two variants of the Berger force field for PE lipids. The first (Berger-POPE-2004) used the de Vries modifications which includes additional repulsive Lennard-Jones interactions on the ethanolamine hydrogen atoms. The second (Berger-POPE-2018) used this same approach but doubled the repulsive strength employed for the hydrogen atom van der Waals interactions. The latter variant results in more disorder within the membrane and a closer agreement with experimental properties such as the area per lipid of POPE. POPE simulations were performed for 200 ns using GROMACS 5.0.6 and employed a POPE membrane with 64 lipids per leaflet taken from a published GROMOS-CKP POPE membrane \cite{piggot11}. Simulations used a 1.0 nm cut-off for both electrostatic and van der Waals interactions with PME and a dispersion correction doe the energy and pressure employed respectively. Further simulation details can be found at Ref.~\citenum{bergerPOPEfiles} and at Ref.~\citenum{berger2POPEfiles}.

%\noindent {\it DOPE} Data is available at \cite{bergerDOPEfiles,berger2DOPEfiles}. \todo{Simulation details by  T. Piggot.}

\noindent {\it POPC:POPE mixtures} 
Two systems were simulated using the Berger force field~\cite{Berger97}, pure POPC and a mixture 50:50 POPC:POPE. For POPE, we additionnally used the de Vries modification implementing a repulsion of the hydrogen atoms located on the amino group of ethanolamine~\cite{devries04}.
Starting from a PDB file of a pure POPC system with Berger atom names, a 50:50 POPC:POPE mixture was built by mutating randomly methyl groups to hydrogens (POPC -> POPE). Each system contained 256 lipids (for the mixture 64 lipids per leaflet for each species) and about 40 water molecules per lipid. No ions were added. Both systems were minimized and equilibrated. The production simulations were run for 300 ns with a time step of 2 fs. The first 100 ns were discarded for the analysis. The simulations were run with GROMACS 4.5.3~\cite{pronk13} version. The v-rescale thermostat~\cite{bussi07} was used with a temperature of 300 K and a time constant of 0.1 ps; lipids and water were coupled separately to the heat bath. Pressure was kept constant at 1 bar using a semi–isotropic Parrinello–Rahman barostat~\cite{parrinello81} with a time constant of 4.0 ps. A real space cut-off of 1.0 nm was employed for van der Waals and electrostatic interactions. The long-range part of electrostatic interactions was evaluated using the PME method~\cite{darden93,essman95} with a grid spacing of 0.12 nm and an interpolation order of 4. All bonds with hydrogen atoms were constrained with the LINCS algorithm~\cite{hess07,hess97}. Water molecules were kept rigid with the SETTLE algorithm~\cite{miyamoto92}. The simulation files are available from
Ref.~\citenum{POPCberger300K} (pure POPC) and~\citenum{POPC1POPE1berger} (POPC:POPE mixture).

\subsection{GROMOS 43A1-S3}

\noindent {\it POPE} Data is available at \cite{gromos43a1s3POPEfiles}. GROMOS 43A1-S3 simulations were performed for 200 ns using GROMACS 4.0.7 employing the standard GROMOS 43A1-S3 POPE force field \cite{chiu09}. The simulation structure had 64 lipids per leaflet and was constructed from a GROMOS 43A1-S3 POPC membrane \cite{piggot12}. Simulations employed standard GROMOS 43A1-S3 settings. Further details can be found at Ref.\citenum{gromos43a1s3POPEfiles}.

\subsection{OPLS-UA}

\noindent {\it POPE} Data is available at \cite{OPLSuaPOPEfiles}. POPE simulations with the OPLS-UA force field were performed for 200 ns using GROMACS 4.5.7. POPE parameters were constructed by modifying the OPLS-UA POPC of Ulmschneider and Ulmschneider \cite{Ulmschneider09} with standard OPLS lysine parameters. The starting membrane structure, containing 64 POPE lipids per leaflet, was created by modifying an OPLS-UA POPC membrane. Simulations employed a 1.0 nm cut-off with PME and no dispersion correction. Further simulation details can be found at \citenum{OPLSuaPOPEfiles}.

\noindent {\it POPE with vdW interaction in H (OPLS-UA/HG-H)} Data is available at \cite{OPLSuaWvdWPOPEfiles}. In addition to the OPLS simulations mentioned above, further simulations employing the same starting structure and simulation settings were employed using slightly modified parameters. These modified parameters, termed OPLS-UA/HG-H, were designed to increase the area per lipid through employing a small repulsive potential on the ethanolamine hydrogen atoms, as per the approach of de Vries with the Berger POPE parameters \cite{devries04}. Further details of these simulations can be found at Ref.\citenum{OPLSuaWvdWPOPEfiles}.

\subsection{GROMOS-CKP}

\noindent {\it POPE} Data is available at \cite{gromosCKPpopeT310}. \todo{Simulation details by  A. Peon.}

%\noindent {\it DOPE} Data is available at \cite{gromosCKPdope}.

\noindent {\it DPPE} Data is available at \cite{gromosCKPdppe}. GROMOS-CKP DPPE simulations were performed for 500 ns using GROMACS 5.0.6. These simulations used the original GROMOS-CKP parameters \cite{piggot11} but with the charges in the ethanolamine head group taken from a GROMOS 53A6 lysine side-chain rather than a PC lipid head group, as done in the original parameters. The starting structure contained 64 lipids per leaflet and was taken from a previous GROMOS-CKP simulation \cite{piggot11}. Simulations employed standard GROMOS-CKP settings with PME employed for the long-range electrostatic interactions. Further details can be found at Ref.\citenum{gromosCKPdppe}.

\noindent {\it POPG} \todo{Simulation details by A. Peon.}

\noindent {\it POPC:POPG mixture} \todo{Simulation details by A. Peon.}

\subsection{OPLS-MacRog}

\noindent {\it POPE} A bilayer patch with a total of 144 POPE molecules, distributed evenly between two leaflets, was created by reordering atoms in a final structure of a POPE simulation performed using the CHARMM36 force field. The bilayer was hydrated by 40 water molecules per lipid for a total of 5760 water molecules. The OPLS-based MacRog force field was used \cite{Kulig15b} for lipids and the TIP3P model \cite{jorgensen83} for water.

Buffered Verlet lists \cite{Pall13} were used to keep track of neighbours. The Lennard-Jones potential was cut off at 1.0~nm, and dispersion corrections were applied to energy and pressure \cite{shirts07}. Smooth PME algorithm was used to calculate long-range electrostatics \cite{darden93,essman95}. The Nos\'{e}--Hoover thermostat was used to keep the temperatures of the lipids and water at 310~K. These groups were coupled separately, and a time constant of 0.4~ps was used. The Parrinello--Rahman barostat~\cite{parrinello81} was used to keep the pressures in the membrane plane as well as normal to it constant at 1~bar. For the barostat, a time constant of 10~ps was used, and the membrane compressibility was set to 4.5$\times$10$^{-5}$~1/bar. All bonds were constrained using p-LINCS \cite{hess97,hess07}.

The system was simulated for 500~ns using GROMACS 2019.2 \cite{abraham2015gromacs}, and the simulation data are available at Ref.~\citenum{MacRogPOPEfiles}.

\noindent {\it POPC:POPE mixtures} The initial force field parameter files in GROMACS format were taken from Ref.~\citenum{rog16}. 
However, a number of errors were detected in the published files so we fixed them in the following way. 
For POPE, the two aliphatic chains sn-1 and sn-2 were switched (in the file the lipid was in fact OPPE); 
one atom (named C27) was not connected to the previous atom in the aliphatic chain so bonds, angles, dihedrals, 
pairs were included to create the connection; two atoms were called C27 thus one of them was renamed C28. 
For POPC, some impropers were missing leading to non planar systems for double bonds 
(in particular the carbonyls of sn-1 and sn-2 as well as the double bond of the oleoyl chain), 
we thus added back those impropers. The resulting fixed files can be found from Refs.~\citenum{POPCmacrog300K} (popc\_fixed.itp) and~\citenum{POPC1POPE1macrogT300K} (pope\_fixed.itp). 
To check our fix, one simulation of pure POPC was also run and the order parameter compared to previous published results~\cite{botan15}. We found very similar values.
All the details of the procedure described here are described 
in file {\it scratch/report\_results\_comparison.pdf} in Ref.~\citenum{NMRlipidsIVbgit}.

Two systems of pure POPC and 50:50 POPC:POPE mixture were built using CHARMM-GUI~\cite{lee16}. They contained 128 lipids and 5120 TIP3 water molecules. The initial PDB file was modified to match OPLS-MacRog nomenclature and atom order. No ions were added. The production simulations were run for 500 ns with a time step of 2 fs. The first 200 ns were discarded for the analysis. The simulations were run with the GROMACS 2018.5~\cite{abraham2015gromacs} version. The v-rescale thermostat~\cite{bussi07} was used with a temperature of 300 K and a time constant of 0.1 ps; lipids and water were coupled separately to the heat bath. Pressure was kept constant at 1 bar using a semi–isotropic Parrinello–Rahman barostat~\cite{parrinello81} with a time constant of 4.0 ps. Long-range electrostatic interactions were calculated using the PME method~\cite{darden93,essman95}. A real space cut-off of 1.0 nm was employed with a grid spacing of 0.1 nm in the reciprocal space. Lennard-Jones potentials were cut off at 1.0 nm. All covalent bonds in lipids were constrained using the LINCS algorithm~\cite{hess07,hess97}. Water molecules were kept rigid with the SETTLE algorithm~\cite{miyamoto92}.

The data for pure POPC are available from Refs.~\citenum{POPCmacrog300K}, and for the POPC:POPE mixture from Ref.~\citenum{POPC1POPE1macrogT300K}.

\subsection{Lipid17}

\noindent {\it POPE} \todo{Simulation details by A. Peon.}

\noindent {\it POPG} \todo{Simulation details by A. Peon.}

\noindent {\it POPC:POPG 4:1 and POPC:POPG 1:1 mixtures with different CaCl$_2$ concentrations}
Initial structures were build by removing appropriate amount of lipids from POPC:POPG 7:3 mixture available from Ref.~\citenum{POPCPOPG73lipid17}.
Force field parameters from the same reference were used \todo{We still need description from A. Peon how these were obtained},
except that incorrect dihedrals with type 1 were changed to type 9
(for details, see discussion in \url{https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/12}).
Simulations were performed using the Gromacs simulation package~\cite{pall20} with the time step of 2~fs. 
The non-bonded interactions were calculated directly within 1.0~nm cutoff; the Verlet scheme was used\cite{Pall13};
and the long-range electrostatic forces were calculated using particle mesh Ewald~\cite{essman95}. 
The bond lengths of hydrogen atoms were constrained using LINCS~\cite{hess97}.
Temperature was coupled to the velocity rescaling thermostat \cite{bussi07} at 298~K with a coupling constant of 1~ps.
Pressure was coupled to the Parrinello--Rahman barostat \cite{parrinello81} at 1~bar with a coupling constant of 10~ps. 
For simulations with CaCl$_2$, appropriate amount of ions with Dang~\cite{smith94,dang06} parameters were added into the solvent.
The simulation files are available from
Refs.~\citenum{Lipid17POPCPOPG8020,Lipid17POPCPOPG8020100mMCaCl,Lipid17POPCPOPG80201000mMCaCl,Lipid17POPCPOPG5050,Lipid17POPCPOPG5050100mMCaCl,Lipid17POPCPOPG50501000mMCaCl}

\subsection{Lipid17ecc}
\noindent {\it POPC:POPG 4:1 and POPC:POPG 1:1 mixtures with different CaCl$_2$ concentrations}
Implicit inclusion of electronic polarizability by electronic continuum correction (ECC),
implemented by scaling the partial charges in force fields, can be used to improve ion interactions with lipids
and other biomolecules in classical MD simulations~\cite{dijon20}.
For Amber Lipid14/17 force fields, ECC has been previously implemented by scaling the
charges and Lennard-Jones~$\sigma$s of headgroup, glycerol backbone, and carbonyl regions
by constant factors~\cite{melcr18,melcr20}.
Here, we apply similar ECC approach to Amber Lipid17 PG parameters as done previously for PS \cite{melcr20}:
charges and Lennard-Jones $\sigma$s of headgroup, glycerol backbone, and carbonyl regions
of parameters POPG from Ref.~\citenum{POPCPOPG73lipid17} were scaled by factors of $f_q$=0.75 and $f_\sigma$=0.89, respectively
(and the dihedral types were corrected to type 9 as in previous section).
Previously introduced ECC-POPC parameters (scaling factors $f_q$=0.8 and $f_\sigma$=0.89 applied to Lipid14 POPC parameters)
were used for POPC~\cite{melcr18}.
ECC-ion parameters with the scaled charges \cite{pluharova14,kohagen16,martinek18}
from \url{bitbucket.org/hseara/ions/src/master/}, and SPC/E water model~\cite{berendsen87} were used in these simulations. 
Rest of the simulation parameters and initial configurations were taken from Lipid17
simulations~\cite{Lipid17POPCPOPG8020,Lipid17POPCPOPG8020100mMCaCl,Lipid17POPCPOPG80201000mMCaCl,Lipid17POPCPOPG5050,Lipid17POPCPOPG5050100mMCaCl,Lipid17POPCPOPG50501000mMCaCl}.
Simulation files of Lipid17ecc simulations are available from
Refs.~\citenum{Lipid17eccPOPCPOPG8020,Lipid17eccPOPCPOPG8020100mMCaCl,Lipid17eccPOPCPOPG80201000mMCaCl,Lipid17eccPOPCPOPG5050,Lipid17eccPOPCPOPG5050100mMCaCl,Lipid17eccPOPCPOPG50501000mMCaCl}.

\clearpage

\section{Author contributions}
\noindent
{\it Am{\'e}lie Bacle} set up, performed and analysed POPC and POPC:POPE (1:1) simulations with the Berger Force Field.
%% Please write description of your contribution here

\noindent
{\it Pavel Buslaev}
%% Please write description of your contribution here
performed the analysis of dihedrals and isomers of lipids. Analysed lipid structures from Protein Data Bank. Prepared panels for figures 2 and 4. Participated in discussions.

\noindent
{\it Rebeca Garc{\'i}a Fandi{\~n}o}
designed and supervised the molecular dynamics simulations carried out by Antonio Pe{\'o}n, and contributed to some discussions.
%% Please write description of your contribution here

\noindent
{\it Fernando Favela-Rosales} 
set up and performed DPPE, POPG and DPPG simulations with the Slipids Force Field.
%% Please write description of your contribution here

\noindent
{\it Tiago M. Ferreira}
was responsible for the solid-state NMR experiments/figures and took part in writing the manuscript.

\noindent
{\it Patrick F.J. Fuchs} supervised Paula Mil{\'a}n Rodr{\'i}guez, Am{\'e}lie Bacle and Chris Papadopoulos, created the buildH software, contributed to many discussions.
%% Please write description of your contribution here

\noindent
{\it Ivan Gushchin} supervised Pavel Buslaev and contributed to the analysis of lipid structures in the Protein Data Bank.
%% Please write description of your contribution here

\noindent
{\it Matti Javanainen} set up and performed simulations using CHARMM36, Slipids, and MacRog lipid models. He contributed to the organization of the manuscript.
%% Please write description of your contribution here

\noindent
{\it Anne M. Kiirikki} set up and performed POPC:POPG(1:1) and POPC:POPG(4:1) simulations with CHARMM36 and POPC:POPG(4:1) with Lipid17ecc. She also contributed to the NMRlipids databank.
%% Please write description of your contribution here

\noindent
{\it Jesper J. Madsen}
set up, performed, and analysed several of the CHARMM36 simulations.
Provided comments on the manuscript.

\noindent
{\it Josef Melcr}
%% Please write description of your contribution here

\noindent
{\it Paula Mil{\'a}n Rodr{\'i}guez} set up, performed and analysed POPC and POPC:POPE (1:1) simulations with the MacRog Force Field. She also fixed the initial force field (itp) files prior to the simulations.
%% Please write description of your contribution here

\noindent
{\it Markus S. Miettinen}
%% Please write description of your contribution here

\noindent
{\it O. H. Samuli Ollila}
designed the project and managed the work.
Ran and analysed several simulations. Wrote the manuscript.

\noindent
{\it Chris G. Papadopoulos} set up, performed and analysed POPC and POPC:POPE (1:1) simulations with the CHARMM36 Force Field.
%% Please write description of your contribution here

\noindent
{\it Antonio Pe{\'o}n}
%% Please write description of your contribution here

\noindent
{\it Thomas J. Piggot}
Setup and performed many of the PE simulations including those with Berger, CHARMM36-UA, GROMOS 43A1-S3, GROMOS-CKP, and OPLS-UA force fields.

\noindent
{\it {\'A}ngel Pi{\~n}eiro}
Created the opAAUA\_prod.py code for the calculation of order parameters. Contributed to some discussions. 
%% Please write description of your contribution here

\noindent
{\it Salla I. Virtanen}
%% Please write description of your contribution here

\bibliography{refs.bib}

\end{document}

### 2021-04-27 manuscriptPGPEsuppl.tex

@@
 \author{Antonio Pe{\'o}n}
-\affiliation[]{Spain}
+\affiliation{CIQUP, Centro de Investigação em Qu\'{i}mica, Departamento de Qu\'{i}mica e Bioqu\'{i}mica, Faculdade de Ciências, Universidade do Porto, Porto, Portugal}
 
@@
 
-%\noindent {\it POPE with additional NaCl} \todo{Simulation details by A. Peon.}
+%\noindent {\it POPE with additional NaCl} \todo{Simulation details by A.
+Peon.}.
+A pure POPE system was built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935].
+A symmetric lipid distribution was used.
+The bilayer contains 500 lipids and 50 CHARMM TIP3P Water model per lipid.
+The simulated system was minimized, equilibrated and production run.
+Production simulation was 500-ns-long, which the last 100 ns were analysed.
+For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
+For water, the SETTLE method was applied.
+The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps.
+The pressure was kept constant with a semi-isotropic scheme.
+The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1.
+Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
+van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
+Long-range dispersion correction to the pressure and energy was added.
 
@@
 
-%\noindent {\it POPG with additional NaCl} \todo{Simulation details by A. Peon.}
+%\noindent {\it POPG with additional NaCl} \todo{Simulation details by A.
+Peon.} A pure POPG system was built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935].
+A symmetric lipid distribution was used.
+The bilayer contains 500 lipids and 50 CHARMM TIP3P Water model per lipid.
+The simulated system was minimized, equilibrated and production run.
+The production simulation was 500-ns-long, which the last 100 ns were analysed.
+For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
+For water, the SETTLE method was applied.
+The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps.
+The pressure was kept constant with a semi-isotropic scheme.
+The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1.
+Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
+van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
+Long-range dispersion correction to the pressure and energy was added.
 
@@
 
-\noindent {\it POPC and POPC:POPG (7:3) mixture} \todo{Simulation details by A. Peon.}
+\noindent {\it POPC and POPC:POPG (7:3) mixture} \todo{Simulation details by A.
+Peon.} A pure POPC and POPC:POPG (7:3) mixture were built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935].
+Bilayer contains 500 lipids and 50 CHARMM TIP3P Water model per lipid.
+All simulated systems were minimized, equilibrated and production run.
+Production simulations were 500-ns-long, which the last 100 ns were analysed.
+For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
+For water, the SETTLE method was applied.
+All simulations were performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps.
+The pressure was kept constant with a semi-isotropic scheme.
+The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1.
+Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
+van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
+Long-range dispersion correction to the pressure and energy was added.
 
@@
 \noindent {\it POPE} Data is available at \cite{gromosCKPpopeT310}. \todo{Simulation details by  A. Peon.}
+A pure POPE system was built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935].
+A symmetric lipid distribution was used.
+The bilayer contains 500 lipids and 50 SPC Water model per lipid.
+GROMOS-CKP force field was used for lipid and GROMOS54A7 for the rest.
+The simulated system was minimized, equilibrated and production run.
+The production simulation was 500-ns-long, which the last 100 ns were analysed.
+For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
+For water, the SETTLE method was applied.
+The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps.
+The pressure was kept constant with a semi-isotropic scheme.
+The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1.
+Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
+van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
+Long-range dispersion correction to the pressure and energy was added.
 
@@
 \noindent {\it POPG} \todo{Simulation details by A. Peon.}
-
+A pure L-POPG system was built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935].
+A symmetric lipid distribution was used.
+The bilayer contains 500 lipids (http://pubs.acs.org/doi/abs/10.1021/jp207013v) and 50 SPC Water model per lipid with 500 sodium ions.
+GROMOS-CKP force field was used for lipid and GROMOS54A7 for the rest.
+The simulated system was minimized, equilibrated and production run.
+The production simulation was 500-ns-long, which the last 100 ns were analysed.
+For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
+For water, the SETTLE method was applied.
+The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps.
+The pressure was kept constant with a semi-isotropic scheme.
+The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1.
+Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
+van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
+Long-range dispersion correction to the pressure and energy was added.
 \noindent {\it POPC:POPG mixture} \todo{Simulation details by A. Peon.}
@@
 \noindent {\it POPE} \todo{Simulation details by A. Peon.}
+From a pure POPE system built with CHARMM-GUI was used to create the Amber input files.
+Amber prmtop and inpcrd files were obtained with LEaP from AmberTools 16 distribution (http://ambermd.org/).
+Then, in order to create the Gromacs input files, ParmEd (v 3.0.0, Swails J.
+ParmEd [Internet].
+https://github.com/ParmEd/ParmEd; https://dx.doi.org/10.1007/s10822-016-9977-1) was used to convert the Amber prmtop and inpcrd files into Gromacs top and gro files.
+The bilayer contains 500 lipids and 50 TIP3P waters per lipid.
+The simulated system was minimized, equilibrated and production run.
+Production simulation was 500-ns-long, which the last 100 ns were analysed.
+For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
+For water, the SETTLE method was applied.
+The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps.
+The pressure was kept constant with a semi-isotropic scheme.
+The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1.
+Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
+van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
+Long-range dispersion correction to the pressure and energy was added.
 
 \noindent {\it POPG} \todo{Simulation details by A. Peon.}
+From a pure POPG system built with CHARMM-GUI was used to create the Amber input files.
+Amber prmtop and inpcrd files were obtained with LEaP from AmberTools 16 distribution (http://ambermd.org/).
+Then, in order to create the Gromacs input files, ParmEd (v 3.0.0, Swails J.
+ParmEd [Internet].
+https://github.com/ParmEd/ParmEd; https://dx.doi.org/10.1007/s10822-016-9977-1) was used to convert the Amber prmtop and inpcrd files into Gromacs top and gro files.
+The bilayer contains 500 lipids and 50 TIP3P waters per lipid and 500 sodium ions.
+The simulated system was minimized, equilibrated and production run.
+Production simulation was 500-ns-long, which the last 100 ns were analysed.
+For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
+For water, the SETTLE method was applied.
+The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps.
+The pressure was kept constant with a semi-isotropic scheme.
+The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1.
+Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
+van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
+Long-range dispersion correction to the pressure and energy was added.
 
@@
 \noindent
-{\it Antonio Pe{\'o}n}
+{\it Antonio Pe{\'o}n} set up, performed and analysed POPC, POPE, POPG and POPC:POPG (7:3) simulations with the CHARMM36, SLIPIDS, GROMOS-CKP and Lipid17 Force Fields.
 %% Please write description of your contribution here

### 2021-04-27 manuscriptPGPEsuppl.tex
Update pds

@@
 \subsection{Conformational ensembles of headgroup and glycerol backbone in PE and PG lipids}
-
 The quality of PE and PG headgroup conformational ensembles in different simulations are evaluated against NMR experiments in figures \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG} using C-H bond order parameters as in our previous studies for PC and PS lipids \cite{botan15,antila19}.
@@
 \ref{HGorderParameters} and \ref{structures} in the main text).
-
 It should be noted that the PG headgroup is biologically abundant R enantiomer in all simulations, while our $^{13}$C NMR experiments has a racemic mixture.
@@
 
-%\noindent {\it POPE with additional NaCl} \todo{Simulation details by A.
-Peon.}.
-A pure POPE system was built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935].
+%\noindent {\it POPE with additional NaCl} \todo{Simulation details by A. Peon.}.
+%A pure POPE system was built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935].
 A symmetric lipid distribution was used.
@@
 Production simulation was 500-ns-long, which the last 100 ns were analysed.
-For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
+%For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
 For water, the SETTLE method was applied.
@@
 
-%\noindent {\it POPG with additional NaCl} \todo{Simulation details by A.
+\noindent {\it POPG with additional NaCl} \todo{Simulation details by A.
 Peon.} A pure POPG system was built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935].

### 2021-04-27 manuscriptPGPEsuppl.tex
Update details in SI

@@
 
-\noindent {\it POPG} Lipid bilayer containing 118 POPG molecules, 4110 TIP3P water molecules, and 118 potassium ions was build using CHARMM-GUI \cite{lee16}.
+\noindent {\it POPG with 118 lipids} Lipid bilayer containing 118 POPG molecules, 4110 TIP3P water molecules, and 118 potassium ions was build using CHARMM-GUI \cite{lee16}.
 The system was simulated 100~ns, coupled to 298 K using Nose-Hoover~\cite{nose84,hoover85} thermostat and 1 bar with
@@
 
-\todo{Simulation details for larger simulation by A. Peon.}
-
-\noindent {\it POPG with additional NaCl} \todo{Simulation details by A.
-Peon.} A pure POPG system was built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935].
+\noindent {\it POPG with 500 lipids} A pure POPG system was built with CHARMM-GUI Membrane builder \cite{lee16}.
 A symmetric lipid distribution was used.
-The bilayer contains 500 lipids and 50 CHARMM TIP3P Water model per lipid.
-The simulated system was minimized, equilibrated and production run.
-The production simulation was 500-ns-long, which the last 100 ns were analysed.
-For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
+The bilayer contains 500 lipids and 50 CHARMM TIP3P water molecules per lipid.
+The 500-ns-long production simulation was performed after energy minimization and equilibration.
+The last 100 ns was used for analysis.
+For production simulation, a Leap-Frog algorithm was applied for integrating Newton's equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
 For water, the SETTLE method was applied.
-The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps.
+The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat~\cite{bussi07}.
+Solvent and lipids were coupled to separate thermostats with a coupling constant of 0.1 ps.
 The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1.
+The Parrinello-Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 $\times$ 10$^5$ bar$^{-1}$.
 Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
-van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
+Van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
 Long-range dispersion correction to the pressure and energy was added.
@@
 
-\noindent {\it POPC and POPC:POPG (7:3) mixture} \todo{Simulation details by A.
-Peon.} A pure POPC and POPC:POPG (7:3) mixture were built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935].
-Bilayer contains 500 lipids and 50 CHARMM TIP3P Water model per lipid.
-All simulated systems were minimized, equilibrated and production run.
-Production simulations were 500-ns-long, which the last 100 ns were analysed.
-For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
+\noindent {\it POPC and POPC:POPG (7:3) mixture} Pure POPC bilayer and POPC:POPG (7:3) mixture were built
+with CHARMM-GUI Membrane builder \cite{lee16}.
+Both systems contained 500 lipids and 50 CHARMM TIP3P Water model per lipid.
+The 500-ns-long production simulation was performed after energy minimization and equilibration.
+The last 100 ns was used for analysis.
+For production simulation, a Leap-Frog algorithm was applied for integrating Newton's equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
 For water, the SETTLE method was applied.
-All simulations were performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps.
+All simulations were performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat~\cite{bussi07}.
+Solvent and lipids were coupled to separate thermostats with a coupling constant of 0.1 ps.
 The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1.
+The Parrinello-Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility
+of 4.6 $\times$ 10$^5$ bar$^{-1}$.
 Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
@@
 
-\noindent {\it POPE} Data is available at \cite{charmm36uaPOPEfiles}.
-CHARMM36-UA POPE simulations were performed for 200 ns using GROMACS version 5.0.6.
+\noindent {\it POPE} CHARMM36-UA POPE simulations were performed for 200 ns using GROMACS version 5.0.6.
 A topology for POPE was constructed using the standard CHARMM36 all-atom PE head group combined with the original CHARMM36-UA lipid tail parameters \cite{lee14}.
@@
 This membrane was constructed using a MARTINI force field \cite{marrink07} coarse-grained self-assembly simulation followed by reverse-mapping to an atomic resolution.
-Simulations were performed using standard CHARMM lipid simulation settings; further details are available at \cite{charmm36uaPOPEfiles}.
+Simulations were performed using standard CHARMM lipid simulation settings; further details and simulation data are available at Ref.
+\citenum{charmm36uaPOPEfiles}.
 
@@
 \subsection{Slipids}
-\noindent {\it POPE} Data is available at \cite{slipidsPOPEfiles}.
-Slipids POPE simulations were performed for 200 ns using GROMACS version 5.0.6.
-Simulations employed the original Slipids POPE parameters \cite{jambeck12b} and employed the same starting structure as the CHARMM36-UA POPE simulations.
+\noindent {\it POPE} Slipids POPE simulations were performed for 200 ns using GROMACS version 5.0.6.
+Simulations employed the original Slipids POPE parameters \cite{jambeck12b} and the same starting structure as the CHARMM36-UA POPE simulations.
 Slipids simulations employed standard AMBER force field cut-offs of 1.0 nm, previously validated for use with the Slipids force field \cite{piggot17}.
-Further simulation details are available at \cite{slipidsPOPEfiles}.
+Further simulation details and data are available at Ref.
+\citenum{slipidsPOPEfiles}.
 
@@
 
-\noindent {\it POPE} Data are available at Ref.~\citenum{bergerPOPEfiles} for Berger-POPE-2004~\cite{devries04} and at Ref.~\citenum{berger2POPEfiles} for Berger-POPE-2018~\cite{berger2POPEfiles}.
+\noindent {\it POPE} 
 Simulations of POPE membranes using the Berger force field employed two variants of the Berger force field for PE lipids.
-The first (Berger-POPE-2004) used the de Vries modifications which includes additional repulsive Lennard-Jones interactions on the ethanolamine hydrogen atoms.
-The second (Berger-POPE-2018) used this same approach but doubled the repulsive strength employed for the hydrogen atom van der Waals interactions.
+The first (Berger-POPE-2004) used the de Vries modifications which includes additional repulsive Lennard-Jones interactions on the ethanolamine hydrogen atoms~\cite{devries04}.
+The second (Berger-POPE-2018) used this same approach but doubled the repulsive strength employed for the hydrogen atom van der Waals interactions~\cite{berger2POPEfiles}.
 The latter variant results in more disorder within the membrane and a closer agreement with experimental properties such as the area per lipid of POPE.
@@
 Simulations used a 1.0 nm cut-off for both electrostatic and van der Waals interactions with PME and a dispersion correction doe the energy and pressure employed respectively.
-Further simulation details can be found at Ref.~\citenum{bergerPOPEfiles} and at Ref.~\citenum{berger2POPEfiles}.
+Further simulation details can be found at Ref.~\citenum{bergerPOPEfiles} for Berger-POPE-2004 and at Ref.~\citenum{berger2POPEfiles} Berger-POPE-2018.
 
@@
 
-\noindent {\it POPE} Data is available at \cite{gromos43a1s3POPEfiles}.
-GROMOS 43A1-S3 simulations were performed for 200 ns using GROMACS 4.0.7 employing the standard GROMOS 43A1-S3 POPE force field \cite{chiu09}.
+\noindent {\it POPE} GROMOS 43A1-S3 simulations were performed for 200 ns using GROMACS 4.0.7 employing the standard GROMOS 43A1-S3 POPE force field \cite{chiu09}.
 The simulation structure had 64 lipids per leaflet and was constructed from a GROMOS 43A1-S3 POPC membrane \cite{piggot12}.
 Simulations employed standard GROMOS 43A1-S3 settings.
-Further details can be found at Ref.\citenum{gromos43a1s3POPEfiles}.
+Further details and simulation data can be found at Ref.\citenum{gromos43a1s3POPEfiles}.
 
@@
 
-\noindent {\it POPE} Data is available at \cite{OPLSuaPOPEfiles}.
-POPE simulations with the OPLS-UA force field were performed for 200 ns using GROMACS 4.5.7.
+\noindent {\it POPE} POPE simulations with the OPLS-UA force field were performed for 200 ns using GROMACS 4.5.7.
 POPE parameters were constructed by modifying the OPLS-UA POPC of Ulmschneider and Ulmschneider \cite{Ulmschneider09} with standard OPLS lysine parameters.
@@
 Simulations employed a 1.0 nm cut-off with PME and no dispersion correction.
-Further simulation details can be found at \citenum{OPLSuaPOPEfiles}.
-
-\noindent {\it POPE with vdW interaction in H (OPLS-UA/HG-H)} Data is available at \cite{OPLSuaWvdWPOPEfiles}.
-In addition to the OPLS simulations mentioned above, further simulations employing the same starting structure and simulation settings were employed using slightly modified parameters.
+Further simulation details and data can be found at \citenum{OPLSuaPOPEfiles}.
+
+\noindent {\it POPE with vdW interaction in H (OPLS-UA/HG-H)} In addition to the OPLS simulations mentioned above, further simulations employing the same starting structure and simulation settings were employed using slightly modified parameters.
 These modified parameters, termed OPLS-UA/HG-H, were designed to increase the area per lipid through employing a small repulsive potential on the ethanolamine hydrogen atoms, as per the approach of de Vries with the Berger POPE parameters \cite{devries04}.
-Further details of these simulations can be found at Ref.\citenum{OPLSuaWvdWPOPEfiles}.
+Further details and data of these simulations can be found at Ref.\citenum{OPLSuaWvdWPOPEfiles}.
 
@@
 
-\noindent {\it POPE} Data is available at \cite{gromosCKPpopeT310}. \todo{Simulation details by  A. Peon.}
-A pure POPE system was built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935].
-A symmetric lipid distribution was used.
+\noindent {\it POPE} 
+Pure POPE system was built with CHARMM-GUI Membrane builder \cite{lee16}.
+Symmetric lipid distribution was used.
 The bilayer contains 500 lipids and 50 SPC Water model per lipid.
-GROMOS-CKP force field was used for lipid and GROMOS54A7 for the rest.
-The simulated system was minimized, equilibrated and production run.
-The production simulation was 500-ns-long, which the last 100 ns were analysed.
-For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
+GROMOS-CKP force field was used for lipids \cite{piggot11}.
+%and GROMOS54A7 for the rest.
+The 500-ns-long production simulation was performed after energy minimization and equilibration.
+The last 100 ns was used for analysis.
+For production simulation, a Leap-Frog algorithm was applied for integrating Newton's equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
 For water, the SETTLE method was applied.
-The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps.
+The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat \cite{bussi07}.
+Solvent and lipids were coupled to separate thermostats with a coupling constant of 0.1 ps.
 The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1.
+The Parrinello-Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 $\times$ 10$^5$ bar$^{-1}$.
 Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
@@
 Long-range dispersion correction to the pressure and energy was added.
+The simulation data and files are available at Ref. \citenum{gromosCKPpopeT310}.
 
@@
 
-\noindent {\it DPPE} Data is available at \cite{gromosCKPdppe}.
-GROMOS-CKP DPPE simulations were performed for 500 ns using GROMACS 5.0.6.
+\noindent {\it DPPE} GROMOS-CKP DPPE simulations were performed for 500 ns using GROMACS 5.0.6.
 These simulations used the original GROMOS-CKP parameters \cite{piggot11} but with the charges in the ethanolamine head group taken from a GROMOS 53A6 lysine side-chain rather than a PC lipid head group, as done in the original parameters.
@@
 Simulations employed standard GROMOS-CKP settings with PME employed for the long-range electrostatic interactions.
-Further details can be found at Ref.\citenum{gromosCKPdppe}.
-
-\noindent {\it POPG} \todo{Simulation details by A. Peon.}
-A pure L-POPG system was built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935].
+Further details and simulation data can be found at Ref.\citenum{gromosCKPdppe}.
+
+\noindent {\it POPG} 
+A pure L-POPG system was built with CHARMM-GUI Membrane builder \cite{lee16}.
 A symmetric lipid distribution was used.
-The bilayer contains 500 lipids (http://pubs.acs.org/doi/abs/10.1021/jp207013v) and 50 SPC Water model per lipid with 500 sodium ions.
-GROMOS-CKP force field was used for lipid and GROMOS54A7 for the rest.
-The simulated system was minimized, equilibrated and production run.
-The production simulation was 500-ns-long, which the last 100 ns were analysed.
-For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
+The bilayer contains 500 lipids and 50 SPC Water model per lipid with 500 sodium ions.
+GROMOS-CKP force field was used for lipid \cite{piggot11} and GROMOS54A7 \cite{schmid11} for the rest.
+The 500-ns-long production simulation was performed after energy minimization and equilibration.
+The last 100 ns was used for analysis.
+For production simulation, a Leap-Frog algorithm was applied for integrating Newton's equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
 For water, the SETTLE method was applied.
-The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps.
+The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat \cite{bussi07}.
+Solvent and lipids were coupled to separate thermostats with a coupling constant of 0.1 ps.
 The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1.
+The Parrinello-Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 $\times$ 10$^5$ bar$^{-1}$.
 Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
-van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
+Van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
 Long-range dispersion correction to the pressure and energy was added.
+
 \noindent {\it POPC:POPG mixture} \todo{Simulation details by A. Peon.}
@@
 The v-rescale thermostat~\cite{bussi07} was used with a temperature of 300 K and a time constant of 0.1 ps; lipids and water were coupled separately to the heat bath.
-Pressure was kept constant at 1 bar using a semi–isotropic Parrinello–Rahman barostat~\cite{parrinello81} with a time constant of 4.0 ps.
+Pressure was kept constant at 1 bar using a semi-isotropic Parrinello-Rahman barostat~\cite{parrinello81} with a time constant of 4.0 ps.
 Long-range electrostatic interactions were calculated using the PME method~\cite{darden93,essman95}.

### 2021-04-28 manuscriptPGPEsuppl.tex
Update simulation details

@@
 are listed in Tables~\ref{systemsPE} and \ref{systemsPG},
-and lipid mixtures with additional ions in Tables~\ref{systemsMIX} and \ref{systemsMIX2}.
+and lipid mixtures with additional ions in Table~\ref{systemsMIX}.
 The simulations were analyzed using preliminary versions of the NMRlipids databank
@@
       \hline
-      POPG             & GROMOS-CKP \cite{piggot11}         & 500 & 25000 &  310  & 500 & 100 & \citenum{POPGgromosCKP} \\
+      POPG             & GROMOS-CKP / GROMOS54A7 \cite{piggot11,schmid11}         & 500 & 25000 &  310  & 500 & 100 & \citenum{POPGgromosCKP} \\
 %      POPG             & GROMOS-CKP \cite{??}         & 0.11           & 500 & 25000 & 49 &  310  & 500 & 100 & \cite{POPGgromosCKP150mMNaCl} \\
@@
       POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 1.0  & 128:128 & 12800 & 230 &  298  & 1500 & 500 & \cite{slipidPOPC50POPG50T298K} \\
-    \end{tabular}
-    }
-  \end{minipage}
-    \todo{ion model for GROMOS-CKP?} \\ % TJP: Most likely standard GROMOS but not my simulations so I've left the ion reference
-\end{table*}
-%\end{sidewaystable*}
-
-\begin{table*}
-%  \begin{sidewaystable*}[!p]
-  \centering
-  \caption{List of MD simulations with PE and PG lipids mixed with PC.
-    % The salt concentrations calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
-    % these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
-    % The lipid force fields named as in our previous work~\cite{botan15}.
-  }\label{systemsMIX2}
-  \begin{minipage}[t]{\textwidth}
-    \resizebox{\columnwidth}{!}{
-    \begin{tabular}{l c c r r r r r r c c}
-      %\hline
-      % some footnotes are not visible in typeset-MS (pdf)
-      lipid/counter-ions & force field for lipids / ions & CaCl$_2$\,(M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
       \hline
@@
       \hline
-      POPC             & Berger-POPC-??
-\cite{??} \todoi{This is probable not plain berger, correct force filed should be described.}  & 0  & 256 & 10240 & 0  &  300  & 300 & 200 & \cite{POPCberger300K} \\
+      POPC             & Berger-POPC-07 \cite{ollila07a} & 0  & 256 & 10240 & 0  &  300  & 300 & 200 & \cite{POPCberger300K} \\
       POPC:POPE (1:1)  & Berger-POPE-04~\cite{devries04}  & 0  & 128 & 11008 & 0  &  300  & 300 & 200 & \cite{POPC1POPE1berger} \\
@@
   \end{minipage}
-  \todo{Citation and description for Berger-POPC-?? model?} \\
+    \todo{ion model for GROMOS-CKP?} \\ % TJP: Most likely standard GROMOS but not my simulations so I've left the ion reference
 \end{table*}
@@
 The last 100 ns was used for analysis.
-For production simulation, a Leap-Frog algorithm was applied for integrating Newton's equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
+The simulation details are the same as in POPE system.
+%For production simulation, a Leap-Frog algorithm was applied for integrating Newton's equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
 For water, the SETTLE method was applied.
 The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat \cite{bussi07}.
-Solvent and lipids were coupled to separate thermostats with a coupling constant of 0.1 ps.
+%Solvent and lipids were coupled to separate thermostats with a coupling constant of 0.1 ps.
 The pressure was kept constant with a semi-isotropic scheme.
@@
 Long-range dispersion correction to the pressure and energy was added.
+The data and simulation files are available at Ref.~\citenum{POPGgromosCKP}.
 
@@
 
-\noindent {\it POPE} \todo{Simulation details by A. Peon.}
-From a pure POPE system built with CHARMM-GUI was used to create the Amber input files.
-Amber prmtop and inpcrd files were obtained with LEaP from AmberTools 16 distribution (http://ambermd.org/).
+\noindent {\it POPE} 
+A pure POPE system built with CHARMM-GUI was used to create the Amber input files.
+Amber prmtop and inpcrd files were obtained with LEaP from AmberTools 16 distribution (\url{http://ambermd.org/}).
+ParmEd version 3.0.0~\cite{shirts17} was used to convert the Amber prmtop and inpcrd files into Gromacs top and gro files. 
+After conversion, the incorrect dihedrals with type 1 were changed to type 9
+(for details, see discussion in \url{https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/12}).
+
+The bilayer contains 500 lipids and 50 TIP3P waters per lipid.
+The 500-ns-long production simulation was performed after energy minimization and equilibration.
+The last 100 ns was used for analysis.
+For production simulation, a Leap-Frog algorithm was applied for integrating Newton's equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
+For water, the SETTLE method was applied.
+The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat~\cite{bussi07}.
+Solvent and lipids were coupled to separate thermostats with a coupling constant of 0.1 ps.
+The pressure was kept constant with a semi-isotropic scheme.
+The Parrinello-Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 $\times$ 10$^5$ bar$^{-1}$.
+Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
+van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
+Long-range dispersion correction to the pressure and energy was added.
+
+\noindent {\it POPG} 
+A pure POPG system built with CHARMM-GUI was used to create the Amber input files.
+The system was further processed and simulated similarly to the POPE system.
+
+%Amber prmtop and inpcrd files were obtained with LEaP from AmberTools 16 distribution (http://ambermd.org/).
 Then, in order to create the Gromacs input files, ParmEd (v 3.0.0, Swails J.
@@
 https://github.com/ParmEd/ParmEd; https://dx.doi.org/10.1007/s10822-016-9977-1) was used to convert the Amber prmtop and inpcrd files into Gromacs top and gro files.
-The bilayer contains 500 lipids and 50 TIP3P waters per lipid.
+%The bilayer contains 500 lipids and 50 TIP3P waters per lipid and 500 sodium ions.
 The simulated system was minimized, equilibrated and production run.
 Production simulation was 500-ns-long, which the last 100 ns were analysed.
-For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
+%For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
 For water, the SETTLE method was applied.
@@
 
-\noindent {\it POPG} \todo{Simulation details by A. Peon.}
-From a pure POPG system built with CHARMM-GUI was used to create the Amber input files.
-Amber prmtop and inpcrd files were obtained with LEaP from AmberTools 16 distribution (http://ambermd.org/).
-Then, in order to create the Gromacs input files, ParmEd (v 3.0.0, Swails J.
-ParmEd [Internet].
-https://github.com/ParmEd/ParmEd; https://dx.doi.org/10.1007/s10822-016-9977-1) was used to convert the Amber prmtop and inpcrd files into Gromacs top and gro files.
-The bilayer contains 500 lipids and 50 TIP3P waters per lipid and 500 sodium ions.
-The simulated system was minimized, equilibrated and production run.
-Production simulation was 500-ns-long, which the last 100 ns were analysed.
-For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
-For water, the SETTLE method was applied.
-The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps.
-The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1.
-Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
-van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
-Long-range dispersion correction to the pressure and energy was added.
-
 \noindent {\it POPC:POPG 4:1 and POPC:POPG 1:1 mixtures with different CaCl$_2$ concentrations}
-Initial structures were build by removing appropriate amount of lipids from POPC:POPG 7:3 mixture available from Ref.~\citenum{POPCPOPG73lipid17}.
-Force field parameters from the same reference were used \todo{We still need description from A. Peon how these were obtained},
-except that incorrect dihedrals with type 1 were changed to type 9
+Initial structures were build by removing appropriate amount of lipids from POPC:POPG 7:3 mixture available from Ref.~\citenum{POPCPOPG73lipid17},
+which was build similarly to pure POPG and POPE systems.
+Force field parameters from the same reference 
+(Amber prmtop and inpcrd files were obtained with LEaP from AmberTools 16 distribution (\url{http://ambermd.org/})
+and converted to Gromacs top and gro files with ParmEd version 3.0.0~\cite{shirts17})
+were used, except that incorrect dihedrals with type 1 were changed to type 9
 (for details, see discussion in \url{https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/12}).

### 2021-04-28 manuscriptPGPEsuppl.tex
Updates from Overleaf

@@
 \affiliation{Center for Research in Biological Chemistry and Molecular Materials (CiQUS), Universidade de Santiago de Compostela, E-15782 Santiago de Compostela, Spain}
-\affiliation{CIQUP, Centro de Investigação em Química, Departamento de Química e Bioquímica, Faculdade de Ciências, Universidade do Porto, Porto, Portugal}
+\alsoaffiliation{CIQUP, Centro de Investigação em Qu\'{i}mica, Departamento de Qu\'{i}mica e Bioqu\'{i}mica, Faculdade de Ciências, Universidade do Porto, Porto, Portugal}
 
@@
 For united atom simulations, we first constructed trajectories including hydrogens with ideal geometry using either \texttt{buildH} program~\cite{buildH} or (\texttt{scratch/opAAUA\_prod.py}) in  Ref.~\citenum{MATCHgit}, and the order parameters were then calculated from these trajectories.
-This approach has been tested against trajectories with explicit hydrogens and the deviations in order parameters are small \cite{buildH,piggot17}.\\
+This approach has been tested against trajectories with explicit hydrogens and the deviations in order parameters are small \cite{buildHvalidation,piggot17}.\\
 %\todo{BuildH program is now cited with a direct link to the GitHub repo. I think that a release to Zenodo would be nice in the final publication.}\\
@@
       POPE	& Slipids \cite{jambeck12b} %&0    
-      		& 336	& ?     %&0    
+      		& 336	& 15254     %&0    
 		& 310  & 2$\times$200 &  2$\times$100 & \citenum{slipidsPOPEfiles}  \\
@@
 Two systems were simulated using the Berger force field~\cite{Berger97}, pure POPC and a mixture 50:50 POPC:POPE.
+For POPC, parameters with the Bachar~\cite{bachar04} correction for the dihedrals next to double bond were used (Berger-POPC-07).
 For POPE, we additionnally used the de Vries modification implementing a repulsion of the hydrogen atoms located on the amino group of ethanolamine~\cite{devries04}.

### 2021-05-02 manuscriptPGPE.tex

@@
 \author{Antonio Pe{\'o}n}
-\affiliation{Spain}
+\affiliation{CIQUP, Centro de Investigação em Qu{\'i}mica, Departamento de Qu{\'i}mica e Bioqu{\'i}mica, Faculdade de Ci{\^e}ncias, Universidade do Porto, Porto, Portugal}
 
@@
 MD simulations could be used to interpret the lipid conformational ensembles from experiments due to their direct connection to the measured C-H bond order parameters.
-To find the best models for this purpose, we have evaluated all the available MD simulation force fields of PC~\cite{botan15}, PS~\cite{antila19}, PE (Fig.~\ref{HGorderParametersPE}), and PG (Fig.~\ref{HGorderParametersPOPG}) lipids against NMR data in the NMRlipids project.
+To find the best models for this purpose, we took advantage of the power of the NMRlipids project which has gathered a massive number of trajectories using open collaboration.
+We have thereby evaluated all the available MD simulation force fields of PC~\cite{botan15}, PS~\cite{antila19}, PE (Fig.~\ref{HGorderParametersPE}), and PG (Fig.~\ref{HGorderParametersPOPG}) lipids against NMR data in the NMRlipids project.
 The comparison of hundreds of simulation trajectories simulated with approximately 20 different force fields reveal that none of the current parameters can correctly capture the lipid conformational ensemble and reproduce the experimental C-H bond order parameters of all segments.
@@
 from structures deposited in the PDB \cite{berman00}.
-We found 311 PC, 394 PE, 154 PG, and 35 PS lipid conformations that were bound to different types of proteins and were determined as a part of protein structure using crystallography or cryo-EM.
+We found 311 PC, 394 PE, 154 PG, and 35 PS lipid conformations that were bound to different types of proteins 
+(including both integral and peripheral membrane proteins) and were determined as a part of protein structure using crystallography or cryo-EM.
 The full list of structures with bound lipids is available from {\it Data/PDBdistributions/pdbtable.pdf} in Ref.~\cite{NMRlipidsIVbgit}.

### 2021-05-02 manuscriptPGPEsuppl.tex
PG OP plot in SI.

--- 2021-04-27 manuscriptPGPEsuppl.tex
@@
 \author{\scriptsize Am{\'e}lie Bacle}
-\affiliation{Laboratoire Coop\'{e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit\'{e} de Poitiers, 1 rue Georges Bonnet, 86000 Poitiers, France }
+\affiliation{\tiny Laboratoire Coop\'{e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit\'{e} de Poitiers, 1 rue Georges Bonnet, 86000 Poitiers, France}
 
 \author{\scriptsize Pavel Buslaev}
-\affiliation{\tiny Nanoscience Center and Department of Chemistry, University of Jyväskylä, P.O. Box 35, 40014 Jyväskylä, Finland}
+\affiliation{Nanoscience Center and Department of Chemistry, University of Jyväskylä, P.O. Box 35, 40014 Jyväskylä, Finland}
 \affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, 141701 Dolgoprudny, Russia}
@@
 \author{Antonio Pe{\'o}n}
-\affiliation{CIQUP, Centro de Investigação em Qu\'{i}mica, Departamento de Qu\'{i}mica e Bioqu\'{i}mica, Faculdade de Ciências, Universidade do Porto, Porto, Portugal}
+\affiliation[]{Spain}
 
@@
 \subsection{Conformational ensembles of headgroup and glycerol backbone in PE and PG lipids}
+
 The quality of PE and PG headgroup conformational ensembles in different simulations are evaluated against NMR experiments in figures \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG} using C-H bond order parameters as in our previous studies for PC and PS lipids \cite{botan15,antila19}.
@@
 \ref{HGorderParameters} and \ref{structures} in the main text).
+
 It should be noted that the PG headgroup is biologically abundant R enantiomer in all simulations, while our $^{13}$C NMR experiments has a racemic mixture.
@@
 %		& 500       & 25000 & 50  &  310  & 500 & 100 & \cite{POPEcharmm150mMNaCl} \\
-      POPE	& CHARMM36-UA \cite{lee14}         %&0       
+      POPE	& CHARMM36-UA \cite{??}         %&0       
       		& 336	& 15254 %&0    
@@
       \hline
-      DPPE	& GROMOS-CKP    \cite{piggot11}      %&0
+      DPPE	& GROMOS-CKP    \cite{??}      %&0
           	& 128	& 3655  %&0    
@@
 %    		& 128	& 3552  &0    & 313  & 2$\times$500 & 2$\times$400 & \cite{gromosCKPpope} \\
-      POPE	& GROMOS-CKP    \cite{piggot11}      %&0
+      POPE	& GROMOS-CKP    \cite{??}      %&0
           	& 500	& 25000 %&0    
@@
       \hline
-      POPE	& GROMOS 43A1-S3 \cite{chiu09}     %&0
+      POPE	& GROMOS 43A1-S3 \cite{??}     %&0
           	& 128	& 3552     %&0    
@@
       \hline
-      POPE	& OPLS-UA/HG-H \cite{Ulmschneider09}   %&0    
+      POPE	& OPLS-UA/HG-H \cite{??}   %&0    
       		& 128	& 3328     %&0    
 		& 303  & 2$\times$200 & 2$\times$100 & \citenum{OPLSuaWvdWPOPEfiles} \\
-      POPE	& OPLS-UA \cite{Ulmschneider09}            %&0
+      POPE	& OPLS-UA \cite{??}            %&0
       		& 128	& 3328     %&0    
@@
       \hline
-      POPE	& Berger-POPE-2004 \cite{devries04}       %&0    
+      POPE	& Berger-POPE-2004 \cite{??}       %&0    
       		& 128	& 3552  %&0    
 		& 303  & 2$\times$200 & 2$\times$100 & \citenum{bergerPOPEfiles}  \\
-      POPE	& Berger-POPE-2018 \cite{devries04}      %&0    
+      POPE	& Berger-POPE-2018 \cite{??}      %&0    
       		& 128	& 3552  %&0    
@@
   \end{minipage}
+  \todo{Citation for CHARMM36ua?} \\
   %\todo{Which ion model is used in \cite{POPEcharmm150mMNaCl}?} \\
+  \todo{Citation for GROMOS-CKP?} \\
+  \todo{Citation for GROMOS 43A1-S3?} \\
+  \todo{Citation for OPLS-UA models?} \\
+  \todo{Citations for Berger-* simulations?} \\
 \end{table*}
@@
       % some footnotes are not visible in typeset-MS (pdf)
-      lipid/counter-ions & force field for lipids / ions &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+      lipid/counter-ions & force field for lipids / ions &  \footnote{Number of lipid molecules}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
       \hline
@@
 %      POPG             & CHARMM36 \cite{venable13}        & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGcharmm150mMNaCl} \\
-      POPG             & CHARMM36 \cite{venable13}                & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGcharmm} \\
+%      POPG/?             & CHARMM36 \cite{venable13}                & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGcharmm} \\
       \hline
@@
       DPPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}           & 288 	& 11232   & 314  & 200 & 100 & \citenum{slipidsDPPGfiles} \\
-      DPPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}           & 288 	& 11232   & 298  & 400 & 100 & \citenum{slipidsDPPGfilesT298K} \\
-      POPG         & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}      & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGslipids} \\
+%      DPPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}           & 288 	& 11232   & 298  & 400 & 100 & \citenum{slipidsDPPGfilesT298K} \\
+%      POPG/?         & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}      & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGslipids} \\
 %      POPG         & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    & 0.11    & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGslipids150mMNaCl} \\
       \hline
-      POPG             & LIPID17 / Dang \cite{gould18,smith94,dang06}     & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGlipid17} \\
+      POPG/?             & Lipid17 / Dang \cite{gould18,smith94,dang06}     & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGlipid17} \\
 %      POPG             & LIPID17 \cite{gould18}         & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGlipid17150mMNaCl} \\
       \hline
-      POPG             & GROMOS-CKP \cite{piggot11}         & 500 & 25000 &  310  & 500 & 100 & \citenum{POPGgromosCKP} \\
+      POPG/?             & GROMOS-CKP \cite{??}         & 500 & 25000 &  310  & 500 & 100 & \citenum{POPGgromosCKP} \\
 %      POPG             & GROMOS-CKP \cite{??}         & 0.11           & 500 & 25000 & 49 &  310  & 500 & 100 & \cite{POPGgromosCKP150mMNaCl} \\
@@
   \end{minipage}
+  \todo{Citation for GROMOS-CKP?}
 \end{table*}
@@
      \hline
-      POPC                   & GROMOS-CKP / ?? \cite{piggot12,??}  & 0  & 500     & 25000 & 0   &  310  & 500 & 100 & \cite{POPCgromosCKPT310K}  \\
-      POPC:POPG (7:3)        & GROMOS-CKP / ?? \cite{piggot11,piggot12,??}  & 0  & 350:150 & 25000 & 0   &  310  & 500 & 100 & \cite{POPC7POPG3gromosCKPT310K} \\
+      POPC                   & GROMOS-CKP / ?? \cite{??,??}  & 0  & 500     & 25000 & 0   &  310  & 500 & 100 & \cite{POPCgromosCKPT310K}  \\
+      POPC:POPG (7:3)        & GROMOS-CKP / ?? \cite{??,??}  & 0  & 350:150 & 25000 & 0   &  310  & 500 & 100 & \cite{POPC7POPG3gromosCKPT310K} \\
      \hline
@@
   \end{minipage}
-    \todo{ion model for GROMOS-CKP?} \\ % TJP: Most likely standard GROMOS but not my simulations so I've left the ion reference
+    \todo{Citation and ion model for GROMOS-CKP?} \\
 \end{table*}
@@
 
-%\noindent {\it POPE with additional NaCl} \todo{Simulation details by A. Peon.}.
-%A pure POPE system was built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935].
-A symmetric lipid distribution was used.
-The bilayer contains 500 lipids and 50 CHARMM TIP3P Water model per lipid.
-The simulated system was minimized, equilibrated and production run.
-Production simulation was 500-ns-long, which the last 100 ns were analysed.
-%For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
-For water, the SETTLE method was applied.
-The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps.
-The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1.
-Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
-van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
-Long-range dispersion correction to the pressure and energy was added.
-
-\noindent {\it POPG with 118 lipids} Lipid bilayer containing 118 POPG molecules, 4110 TIP3P water molecules, and 118 potassium ions was build using CHARMM-GUI \cite{lee16}.
+%\noindent {\it POPE with additional NaCl} \todo{Simulation details by A. Peon.}
+
+\noindent {\it POPG} Lipid bilayer containing 118 POPG molecules, 4110 TIP3P water molecules, and 118 potassium ions was build using CHARMM-GUI \cite{lee16}.
 The system was simulated 100~ns, coupled to 298 K using Nose-Hoover~\cite{nose84,hoover85} thermostat and 1 bar with
@@
 
-\noindent {\it POPG with 500 lipids} A pure POPG system was built with CHARMM-GUI Membrane builder \cite{lee16}.
-A symmetric lipid distribution was used.
-The bilayer contains 500 lipids and 50 CHARMM TIP3P water molecules per lipid.
-The 500-ns-long production simulation was performed after energy minimization and equilibration.
-The last 100 ns was used for analysis.
-For production simulation, a Leap-Frog algorithm was applied for integrating Newton's equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
-For water, the SETTLE method was applied.
-The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat~\cite{bussi07}.
-Solvent and lipids were coupled to separate thermostats with a coupling constant of 0.1 ps.
-The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello-Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 $\times$ 10$^5$ bar$^{-1}$.
-Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
-Van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
-Long-range dispersion correction to the pressure and energy was added.
+\todo{Simulation details for larger simulation by A. Peon.}
+
+%\noindent {\it POPG with additional NaCl} \todo{Simulation details by A. Peon.}
 
@@
 
-\noindent {\it POPC and POPC:POPG (7:3) mixture} Pure POPC bilayer and POPC:POPG (7:3) mixture were built
-with CHARMM-GUI Membrane builder \cite{lee16}.
-Both systems contained 500 lipids and 50 CHARMM TIP3P Water model per lipid.
-The 500-ns-long production simulation was performed after energy minimization and equilibration.
-The last 100 ns was used for analysis.
-For production simulation, a Leap-Frog algorithm was applied for integrating Newton's equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
-For water, the SETTLE method was applied.
-All simulations were performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat~\cite{bussi07}.
-Solvent and lipids were coupled to separate thermostats with a coupling constant of 0.1 ps.
-The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello-Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility
-of 4.6 $\times$ 10$^5$ bar$^{-1}$.
-Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
-van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
-Long-range dispersion correction to the pressure and energy was added.
+\noindent {\it POPC and POPC:POPG (7:3) mixture} \todo{Simulation details by A. Peon.}
 
@@
 
-\subsection{CHARMM36-UA}
-
-\noindent {\it POPE} CHARMM36-UA POPE simulations were performed for 200 ns using GROMACS version 5.0.6.
-A topology for POPE was constructed using the standard CHARMM36 all-atom PE head group combined with the original CHARMM36-UA lipid tail parameters \cite{lee14}.
-Simulations were performed using a hexagonal periodic box containing 336 POPE lipids.
-This membrane was constructed using a MARTINI force field \cite{marrink07} coarse-grained self-assembly simulation followed by reverse-mapping to an atomic resolution.
-Simulations were performed using standard CHARMM lipid simulation settings; further details and simulation data are available at Ref.
-\citenum{charmm36uaPOPEfiles}.
+\subsection{CHARMM36ua}
+
+\noindent {\it POPE} Data is available at \cite{charmm36uaPOPEfiles}. \todo{Simulation details by  T. Piggot.}
 
@@
 \subsection{Slipids}
-\noindent {\it POPE} Slipids POPE simulations were performed for 200 ns using GROMACS version 5.0.6.
-Simulations employed the original Slipids POPE parameters \cite{jambeck12b} and the same starting structure as the CHARMM36-UA POPE simulations.
-Slipids simulations employed standard AMBER force field cut-offs of 1.0 nm, previously validated for use with the Slipids force field \cite{piggot17}.
-Further simulation details and data are available at Ref.
-\citenum{slipidsPOPEfiles}.
+\noindent {\it POPE} Data is available at \cite{slipidsPOPEfiles}. \todo{Simulation details by  T. Piggot.}
 
@@
 
-\noindent {\it POPE} 
-Simulations of POPE membranes using the Berger force field employed two variants of the Berger force field for PE lipids.
-The first (Berger-POPE-2004) used the de Vries modifications which includes additional repulsive Lennard-Jones interactions on the ethanolamine hydrogen atoms~\cite{devries04}.
-The second (Berger-POPE-2018) used this same approach but doubled the repulsive strength employed for the hydrogen atom van der Waals interactions~\cite{berger2POPEfiles}.
-The latter variant results in more disorder within the membrane and a closer agreement with experimental properties such as the area per lipid of POPE.
-POPE simulations were performed for 200 ns using GROMACS 5.0.6 and employed a POPE membrane with 64 lipids per leaflet taken from a published GROMOS-CKP POPE membrane \cite{piggot11}.
-Simulations used a 1.0 nm cut-off for both electrostatic and van der Waals interactions with PME and a dispersion correction doe the energy and pressure employed respectively.
-Further simulation details can be found at Ref.~\citenum{bergerPOPEfiles} for Berger-POPE-2004 and at Ref.~\citenum{berger2POPEfiles} Berger-POPE-2018.
+\noindent {\it POPE} Data are available at Ref.~\citenum{bergerPOPEfiles} for Berger-POPE-2004~\cite{??} and at Ref.~\citenum{berger2POPEfiles} for Berger-POPE-2018~\cite{??}.
+\todo{Simulation details by  T.
+Piggot.}
 
@@
 
-\noindent {\it POPE} GROMOS 43A1-S3 simulations were performed for 200 ns using GROMACS 4.0.7 employing the standard GROMOS 43A1-S3 POPE force field \cite{chiu09}.
-The simulation structure had 64 lipids per leaflet and was constructed from a GROMOS 43A1-S3 POPC membrane \cite{piggot12}.
-Simulations employed standard GROMOS 43A1-S3 settings.
-Further details and simulation data can be found at Ref.\citenum{gromos43a1s3POPEfiles}.
+\noindent {\it POPE} Data is available at \cite{gromos43a1s3POPEfiles}. \todo{Simulation details by  T. Piggot.}
 
@@
 
-\noindent {\it POPE} POPE simulations with the OPLS-UA force field were performed for 200 ns using GROMACS 4.5.7.
-POPE parameters were constructed by modifying the OPLS-UA POPC of Ulmschneider and Ulmschneider \cite{Ulmschneider09} with standard OPLS lysine parameters.
-The starting membrane structure, containing 64 POPE lipids per leaflet, was created by modifying an OPLS-UA POPC membrane.
-Simulations employed a 1.0 nm cut-off with PME and no dispersion correction.
-Further simulation details and data can be found at \citenum{OPLSuaPOPEfiles}.
-
-\noindent {\it POPE with vdW interaction in H (OPLS-UA/HG-H)} In addition to the OPLS simulations mentioned above, further simulations employing the same starting structure and simulation settings were employed using slightly modified parameters.
-These modified parameters, termed OPLS-UA/HG-H, were designed to increase the area per lipid through employing a small repulsive potential on the ethanolamine hydrogen atoms, as per the approach of de Vries with the Berger POPE parameters \cite{devries04}.
-Further details and data of these simulations can be found at Ref.\citenum{OPLSuaWvdWPOPEfiles}.
-
-\subsection{GROMOS-CKP}
-
-\noindent {\it POPE} 
-Pure POPE system was built with CHARMM-GUI Membrane builder \cite{lee16}.
-Symmetric lipid distribution was used.
-The bilayer contains 500 lipids and 50 SPC Water model per lipid.
-GROMOS-CKP force field was used for lipids \cite{piggot11}.
-%and GROMOS54A7 for the rest.
-The 500-ns-long production simulation was performed after energy minimization and equilibration.
-The last 100 ns was used for analysis.
-For production simulation, a Leap-Frog algorithm was applied for integrating Newton's equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
-For water, the SETTLE method was applied.
-The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat \cite{bussi07}.
-Solvent and lipids were coupled to separate thermostats with a coupling constant of 0.1 ps.
-The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello-Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 $\times$ 10$^5$ bar$^{-1}$.
-Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
-van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
-Long-range dispersion correction to the pressure and energy was added.
-The simulation data and files are available at Ref. \citenum{gromosCKPpopeT310}.
-
-%\noindent {\it DOPE} Data is available at \cite{gromosCKPdope}.
-
-\noindent {\it DPPE} GROMOS-CKP DPPE simulations were performed for 500 ns using GROMACS 5.0.6.
-These simulations used the original GROMOS-CKP parameters \cite{piggot11} but with the charges in the ethanolamine head group taken from a GROMOS 53A6 lysine side-chain rather than a PC lipid head group, as done in the original parameters.
-The starting structure contained 64 lipids per leaflet and was taken from a previous GROMOS-CKP simulation \cite{piggot11}.
-Simulations employed standard GROMOS-CKP settings with PME employed for the long-range electrostatic interactions.
-Further details and simulation data can be found at Ref.\citenum{gromosCKPdppe}.
-
-\noindent {\it POPG} 
-A pure L-POPG system was built with CHARMM-GUI Membrane builder \cite{lee16}.
-A symmetric lipid distribution was used.
-The bilayer contains 500 lipids and 50 SPC Water model per lipid with 500 sodium ions.
-GROMOS-CKP force field was used for lipid \cite{piggot11} and GROMOS54A7 \cite{schmid11} for the rest.
-The 500-ns-long production simulation was performed after energy minimization and equilibration.
-The last 100 ns was used for analysis.
-For production simulation, a Leap-Frog algorithm was applied for integrating Newton's equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
-For water, the SETTLE method was applied.
-The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat \cite{bussi07}.
-Solvent and lipids were coupled to separate thermostats with a coupling constant of 0.1 ps.
-The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello-Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 $\times$ 10$^5$ bar$^{-1}$.
-Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
-Van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
-Long-range dispersion correction to the pressure and energy was added.
+\noindent {\it POPE} Data is available at \cite{OPLSuaPOPEfiles}. \todo{Simulation details by  T. Piggot.}
+%From Zenodo: The PE parameters were constructed by modifying the OPLS-UA POPC of Ulmschneider and Ulmschneider with the standard OPLS lysine parameters.
+
+\noindent {\it POPE with vdW interaction in H (OPLS-UA/HG-H)} Data is available at \cite{OPLSuaWvdWPOPEfiles}. \todo{Simulation details by  T. Piggot.}
+%From Zenodo: The PE parameters were constructed by modifying the OPLS-UA POPC of Ulmschneider and Ulmschneider with the standard OPLS lysine parameters and the addition of LJ parameters on the hydrogen atoms in the head group (designed to increase the area per lipid).
+
+\subsection{GROMOS-CKP and GROMOS-CKPM}
+
+\noindent {\it POPE} Data is available at \cite{gromosCKPpopeT310}. \todo{Simulation details by  A. Peon.}
+
+%\noindent {\it DOPE} Data is available at \cite{gromosCKPdope}. \todo{Simulation details by  T. Piggot.}
+
+\noindent {\it DPPE} Data is available at \cite{gromosCKPdppe}. \todo{Simulation details by  T. Piggot.}
+
+\noindent {\it POPG} \todo{Simulation details by A. Peon.}
 
@@
 The Nos\'{e}--Hoover thermostat was used to keep the temperatures of the lipids and water at 310~K.
-These groups were coupled separately, and a time constant of 0.4~ps was used.
+These groups were coupled separately, and a time constat of 0.4~ps was used.
 The Parrinello--Rahman barostat~\cite{parrinello81} was used to keep the pressures in the membrane plane as well as normal to it constant at 1~bar.
@@
 The v-rescale thermostat~\cite{bussi07} was used with a temperature of 300 K and a time constant of 0.1 ps; lipids and water were coupled separately to the heat bath.
-Pressure was kept constant at 1 bar using a semi-isotropic Parrinello-Rahman barostat~\cite{parrinello81} with a time constant of 4.0 ps.
+Pressure was kept constant at 1 bar using a semi–isotropic Parrinello–Rahman barostat~\cite{parrinello81} with a time constant of 4.0 ps.
 Long-range electrostatic interactions were calculated using the PME method~\cite{darden93,essman95}.
@@
 \noindent {\it POPE} \todo{Simulation details by A. Peon.}
-From a pure POPE system built with CHARMM-GUI was used to create the Amber input files.
-Amber prmtop and inpcrd files were obtained with LEaP from AmberTools 16 distribution (http://ambermd.org/).
-Then, in order to create the Gromacs input files, ParmEd (v 3.0.0, Swails J.
-ParmEd [Internet].
-https://github.com/ParmEd/ParmEd; https://dx.doi.org/10.1007/s10822-016-9977-1) was used to convert the Amber prmtop and inpcrd files into Gromacs top and gro files.
-The bilayer contains 500 lipids and 50 TIP3P waters per lipid.
-The simulated system was minimized, equilibrated and production run.
-Production simulation was 500-ns-long, which the last 100 ns were analysed.
-For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
-For water, the SETTLE method was applied.
-The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps.
-The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1.
-Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
-van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
-Long-range dispersion correction to the pressure and energy was added.
 
 \noindent {\it POPG} \todo{Simulation details by A. Peon.}
-From a pure POPG system built with CHARMM-GUI was used to create the Amber input files.
-Amber prmtop and inpcrd files were obtained with LEaP from AmberTools 16 distribution (http://ambermd.org/).
-Then, in order to create the Gromacs input files, ParmEd (v 3.0.0, Swails J.
-ParmEd [Internet].
-https://github.com/ParmEd/ParmEd; https://dx.doi.org/10.1007/s10822-016-9977-1) was used to convert the Amber prmtop and inpcrd files into Gromacs top and gro files.
-The bilayer contains 500 lipids and 50 TIP3P waters per lipid and 500 sodium ions.
-The simulated system was minimized, equilibrated and production run.
-Production simulation was 500-ns-long, which the last 100 ns were analysed.
-For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
-For water, the SETTLE method was applied.
-The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps.
-The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1.
-Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
-van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
-Long-range dispersion correction to the pressure and energy was added.
 
@@
 \noindent
-{\it Anne M. Kiirikki} set up and performed POPC:POPG(1:1) and POPC:POPG(4:1) simulations with CHARMM36 and POPC:POPG(4:1) with Lipid17ecc. She also contributed to the NMRlipids databank.
+{\it Anne M. Kiirikki}
 %% Please write description of your contribution here
@@
 \noindent
-{\it Antonio Pe{\'o}n} set up, performed and analysed POPC, POPE, POPG and POPC:POPG (7:3) simulations with the CHARMM36, SLIPIDS, GROMOS-CKP and Lipid17 Force Fields.
+{\it Antonio Pe{\'o}n}
 %% Please write description of your contribution here
@@
 {\it Thomas J. Piggot}
-Setup and performed many of the PE simulations including those with Berger, CHARMM36-UA, GROMOS 43A1-S3, GROMOS-CKP, and OPLS-UA force fields.
+%% Please write description of your contribution here
 

### 2021-05-02 manuscriptPGPEsuppl.tex
PG OP plot.

--- 2021-04-28 manuscriptPGPEsuppl.tex
@@
 
-\author{\scriptsize Am{\'e}lie Bacle}
-\affiliation{Laboratoire Coop\'{e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit\'{e} de Poitiers, 1 rue Georges Bonnet, 86000 Poitiers, France }
-
-\author{\scriptsize Pavel Buslaev}
-\affiliation{\tiny Nanoscience Center and Department of Chemistry, University of Jyväskylä, P.O. Box 35, 40014 Jyväskylä, Finland}
+\author{\small Am{\'e}lie Bacle}
+\affiliation{\tiny Laboratoire Coop\'{e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit\'{e} de Poitiers, 1 rue Georges Bonnet, 86000 Poitiers, France }
+
+\author{Pavel Buslaev}
+\affiliation{Nanoscience Center and Department of Chemistry, University of Jyväskylä, P.O. Box 35, 40014 Jyväskylä, Finland}
 \affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, 141701 Dolgoprudny, Russia}

### 2021-05-02 manuscriptPGPEsuppl.tex
Edits to match the final PG OP plot.

@@
   \caption{\label{HGorderParametersPOPG}
-    The headgroup and glycerol backbone order parameters of PG lipids
-    from experiments (POPG and signs from this work and from Ref.~\citenum{borle85}, %contains 10mM of PIPES,
-    DPPG with 100mM NaCl from Ref.~\citenum{wohlgemuth80},% contains 10mM PIPES and,
-    and E.Coli PG results from Ref.~\citenum{gally81})
-    and simulations with different force fields.
+    C--H bond order parameters, $S_{\rm CH}$, of the PG headgroup ($\beta$ and $\alpha$) and
+    glycerol backbone (g$_3$, g$_2$, g$_1$) carbons
+    from MD simulations with different force fields (symbols) and
+    NMR experiments (horizontal lines;
+    $^{13}$C NMR POPG and signs from this work,
+    $^2$H NMR POPG from Ref.~\citenum{borle85}, % contains 10 mM of PIPES,
+    DPPG from Ref.~\citenum{wohlgemuth80}, % and 10 mM PIPES
+    and {\it Escherichia coli} PG from Ref.~\citenum{gally81}).
+    ///
+    Experimental data for DPPG had been measured with 0.1\,M NaCl~\cite{wohlgemuth80}, while all of the other data are with
+    (Na$^+$, except for CHARMM36 K$^+$) counterions only.
+    The DPPG is [R],
+    2H POPG beta racemic, alpha 3,3' averaged, which should equal 3,1'~\cite{wohlgemuth80}
+    13C POPG racemic.
+    $^2$ NMR DPPG and POPG in 10\,mM PIPES buffer.
+    ///
+    The light blue areas span 0.04 units around the average of the extremal experimental values,
+    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
   }
@@
 %      POPG             & CHARMM36 \cite{venable13}        & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGcharmm150mMNaCl} \\
-      POPG             & CHARMM36 \cite{venable13}                & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGcharmm} \\
+%      POPG             & CHARMM36 \cite{venable13}                & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGcharmm} \\
       \hline
@@
       DPPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}           & 288 	& 11232   & 314  & 200 & 100 & \citenum{slipidsDPPGfiles} \\
-      DPPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}           & 288 	& 11232   & 298  & 400 & 100 & \citenum{slipidsDPPGfilesT298K} \\
-      POPG         & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}      & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGslipids} \\
+%      DPPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}           & 288 	& 11232   & 298  & 400 & 100 & \citenum{slipidsDPPGfilesT298K} \\
+%      POPG         & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}      & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGslipids} \\
 %      POPG         & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    & 0.11    & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGslipids150mMNaCl} \\
       \hline
-      POPG             & LIPID17 / Dang \cite{gould18,smith94,dang06}     & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGlipid17} \\
-%      POPG             & LIPID17 \cite{gould18}         & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGlipid17150mMNaCl} \\
-      \hline
-      POPG             & GROMOS-CKP / GROMOS54A7 \cite{piggot11,schmid11}         & 500 & 25000 &  310  & 500 & 100 & \citenum{POPGgromosCKP} \\
+      POPG/Na$^+$         & Lipid17 / Dang \cite{gould18,smith94,dang06}     & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGlipid17} \\
+%      POPG             & Lipid17 \cite{gould18}         & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGlipid17150mMNaCl} \\
+      \hline
+      POPG/Na$^+$             & GROMOS-CKP / GROMOS 54A7 \cite{piggot11,schmid11}         & 500 & 25000 &  310  & 500 & 100 & \citenum{POPGgromosCKP} \\
 %      POPG             & GROMOS-CKP \cite{??}         & 0.11           & 500 & 25000 & 49 &  310  & 500 & 100 & \cite{POPGgromosCKP150mMNaCl} \\

### 2021-05-03 manuscriptPGPEsuppl.tex
Updates from Overleaf

--- 2021-05-02 manuscriptPGPEsuppl.tex
@@
 
-\author{\small Am{\'e}lie Bacle}
-\affiliation{\tiny Laboratoire Coop\'{e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit\'{e} de Poitiers, 1 rue Georges Bonnet, 86000 Poitiers, France }
-
-\author{Pavel Buslaev}
-\affiliation{Nanoscience Center and Department of Chemistry, University of Jyväskylä, P.O. Box 35, 40014 Jyväskylä, Finland}
+\author{\scriptsize Am{\'e}lie Bacle}
+\affiliation{Laboratoire Coop\'{e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit\'{e} de Poitiers, 1 rue Georges Bonnet, 86000 Poitiers, France }
+
+\author{\scriptsize Pavel Buslaev}
+\affiliation{\tiny Nanoscience Center and Department of Chemistry, University of Jyväskylä, P.O. Box 35, 40014 Jyväskylä, Finland}
 \affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, 141701 Dolgoprudny, Russia}
@@
       \hline
-%      POPC                   & CHARMM36 \cite{??}        & 0.11      & 0  & 500     & 25000 & 48  &  310  & 500 & 100 & \cite{POPCcharmm150mMNaCl301K}  \\
+%      POPC                   & CHARMM36 \citenum{??}        & 0.11      & 0  & 500     & 25000 & 48  &  310  & 500 & 100 & \cite{POPCcharmm150mMNaCl301K}  \\
 %      POPC:POPG (7:3)        & CHARMM36 \cite{??}        & 0.11      & 0  & 350     & ?     & ?   &  310  & 500 & 100 & \cite{POPC7POPG1charmm36NaCl}  \\
-      POPC                   & CHARMM36 \cite{klauda10}            & 0  & 500     & 25000 & 0   &  310  & 500 & 100 & \cite{POPCcharmmT310K}  \\
-      POPC:POPG (7:3)        & CHARMM36 \cite{klauda10,venable13}    & 0  & 350     & 25000 & 0   &  310  & 500 & 100 & \cite{POPC7POPG1charmm36}  \\
-      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}     & 0  & 150:150 & 31500 & 0   &  298  & 500 & 400 & \cite{CHARMM36POPCPOPG5050} \\ 
-      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}     & 0.1 & 150:150 & 31329 & 57 &  298  & 400 & 300 & \cite{CHARMM36POPCPOPG5050100mMCaCl} \\
-      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}     & 1.08 & 150:150 & 29766 & 578  &  298  & 500 & 400 & \cite{CHARMM36POPCPOPG50501000mMCaCl} \\
-      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 0  & 350:88 & 26280 & 0  &  298  & 500 & 400 & \cite{CHARMM36POPCPOPG4010} \\
-      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 0.1 & 350:88 & 26280 & 47  &  298  & 500 & 400 & \cite{CHARMM36POPCPOPG4010100mMCaCl} \\
-      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 1.0 & 350:88 & 24927 & 451  &  298  & 500 & 400 & \cite{CHARMM36POPCPOPG40101000mMCaCl} \\
-      \hline
-      POPC             & CHARMM36 \cite{klauda10}        &0          & 256 & 8704 & 0  &  300  & 300 & 250 & \cite{POPCcharmm300K} \\
-      POPC:POPE (1:1)  & CHARMM36 \cite{klauda10,venable13}         & 0  & 128 & 8704 & 0  &  300  & 300 & 250 & \cite{POPC1POPE1charmm36} \\
-      \hline
-      POPC             & OPLS-MacRog \cite{rog16}     &0           & 128 & 5120 & 0  &  300  & 500 & 300 & \cite{POPCmacrog300K} \\
-      POPC:POPE (1:1)  & OPLS-MacRog \cite{rog16}     &0           & 128 & 5120 & 0  &  300  & 500 & 300 & \cite{POPC1POPE1macrogT300K} \\
-      \hline
-      POPC             & Slipid \cite{jambeck12b}     &0          & 512 & 23943 & 0  &  298  & 170 & 100 & \cite{POPCslipid298K} \\
-      POPC:POPE (1:1)  & Slipid \cite{jambeck12b}     &0          & 128 & 5120  & 0  &  298  & 500 & 300 & \cite{POPC1POPE1slipidT298K} \\
-     \hline
-      POPC                   & GROMOS-CKP / ?? \cite{piggot12,??}  & 0  & 500     & 25000 & 0   &  310  & 500 & 100 & \cite{POPCgromosCKPT310K}  \\
-      POPC:POPG (7:3)        & GROMOS-CKP / ?? \cite{piggot11,piggot12,??}  & 0  & 350:150 & 25000 & 0   &  310  & 500 & 100 & \cite{POPC7POPG3gromosCKPT310K} \\
-     \hline
-      POPC                   & Slipid \cite{jambeck12b}  & 0      & 500     & 25000 & 0   &  310  & 500 & 100 & \cite{POPCslipid301K}  \\
-      POPC:POPG (7:3)        & Slipid / {\AA}qvist \cite{jambeck12b,aqvist90} & 0  & 350:150 & 25000 & 0   &  310  & 500 & 100 & \cite{slipidPOPC70POPG30T310K} \\
-      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0  & 128:128 & 12800 & 0   &  298  & 500 & 400 & \cite{slipidPOPC50POPG50T298K} \\
-      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0.1  & 128:128 & 12800 & 23  &  298  & 500 & 400 & \cite{slipidPOPC50POPG50T298K} \\
-      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0.2  & 128:128 & 12800 & 46  &  298  & 1500 & 500 & \cite{slipidPOPC50POPG50T298K} \\
-      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0.5  & 128:128 & 12800 & 115 &  298  & 1500 & 500 & \cite{slipidPOPC50POPG50T298K} \\
-      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 1.0  & 128:128 & 12800 & 230 &  298  & 1500 & 500 & \cite{slipidPOPC50POPG50T298K} \\
-      \hline
-      POPC:POPG (4:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}       & 0  & 350:88 & 26265 & 0  &  298  & 400 & 350 & \cite{Lipid17POPCPOPG8020} \\
-      POPC:POPG (4:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}    & 0.1& 350:88 & 26124 & 47 &  298  & 400 & 250 & \cite{Lipid17POPCPOPG8020100mMCaCl} \\
-      POPC:POPG (4:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}    & 1.0& 350:88 & 24840 & 475 &  298  & 1200 & 200 & \cite{Lipid17POPCPOPG80201000mMCaCl} \\
-      POPC:POPG (1:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}   & 0  & 150:150 & 31572 & 0  &  298  & 320 & 200 & \cite{Lipid17POPCPOPG5050} \\
-      POPC:POPG (1:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}   & 0.1& 150:150 & 31401 & 57 &  298  & 718 & 198 & \cite{Lipid17POPCPOPG5050100mMCaCl} \\
-      POPC:POPG (1:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}  & 1.0& 150:150 & 29865 & 569 &  298  & 720 & 200 & \cite{Lipid17POPCPOPG50501000mMCaCl} \\
-      \hline
-      POPC:POPG (4:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     &0     & 350:88 & 26265 & 0  &  298  & 400 & 300 & \cite{Lipid17eccPOPCPOPG8020} \\
-      POPC:POPG (4:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 0.1& 350:88 & 26124 & 47 &  298  & 400 & 300 & \cite{Lipid17eccPOPCPOPG8020100mMCaCl} \\
-      POPC:POPG (4:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 1.0& 350:88 & 24840 & 475 &  298  & 400 & 300 & \cite{Lipid17eccPOPCPOPG80201000mMCaCl} \\
-      POPC:POPG (1:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 0  & 150:150 & 31572 & 0  &  298  & 347.8 & 333 & \cite{Lipid17eccPOPCPOPG5050} \\
-      POPC:POPG (1:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 0.1& 150:150 & 29865 & 54 &  298  & 400 & 300 & \cite{Lipid17eccPOPCPOPG5050100mMCaCl} \\
-      POPC:POPG (1:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 1.0& 150:150 & 29865 & 569 &  298  & 600 & 400 & \cite{Lipid17eccPOPCPOPG50501000mMCaCl} \\
-      \hline
-      POPC             & Berger-POPC-07 \cite{ollila07a} & 0  & 256 & 10240 & 0  &  300  & 300 & 200 & \cite{POPCberger300K} \\
-      POPC:POPE (1:1)  & Berger-POPE-04~\cite{devries04}  & 0  & 128 & 11008 & 0  &  300  & 300 & 200 & \cite{POPC1POPE1berger} \\
+      POPC                   & CHARMM36 \cite{klauda10}            & 0  & 500     & 25000 & 0   &  310  & 500 & 100 & \citenum{POPCcharmmT310K}  \\
+      POPC:POPG (7:3)        & CHARMM36 \cite{klauda10,venable13}    & 0  & 350     & 25000 & 0   &  310  & 500 & 100 & \citenum{POPC7POPG1charmm36}  \\
+      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}     & 0  & 150:150 & 31500 & 0   &  298  & 500 & 400 & \citenum{CHARMM36POPCPOPG5050} \\ 
+      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}     & 0.1 & 150:150 & 31329 & 57 &  298  & 400 & 300 & \citenum{CHARMM36POPCPOPG5050100mMCaCl} \\
+      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}     & 1.08 & 150:150 & 29766 & 578  &  298  & 500 & 400 & \citenum{CHARMM36POPCPOPG50501000mMCaCl} \\
+      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 0  & 350:88 & 26280 & 0  &  298  & 500 & 400 & \citenum{CHARMM36POPCPOPG4010} \\
+      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 0.1 & 350:88 & 26280 & 47  &  298  & 500 & 400 & \citenum{CHARMM36POPCPOPG4010100mMCaCl} \\
+      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 1.0 & 350:88 & 24927 & 451  &  298  & 500 & 400 & \citenum{CHARMM36POPCPOPG40101000mMCaCl} \\
+      \hline
+      POPC             & CHARMM36 \cite{klauda10}        &0          & 256 & 8704 & 0  &  300  & 300 & 250 & \citenum{POPCcharmm300K} \\
+      POPC:POPE (1:1)  & CHARMM36 \cite{klauda10,venable13}         & 0  & 128 & 8704 & 0  &  300  & 300 & 250 & \citenum{POPC1POPE1charmm36} \\
+      \hline
+      POPC             & OPLS-MacRog \cite{rog16}     &0           & 128 & 5120 & 0  &  300  & 500 & 300 & \citenum{POPCmacrog300K} \\
+      POPC:POPE (1:1)  & OPLS-MacRog \cite{rog16}     &0           & 128 & 5120 & 0  &  300  & 500 & 300 & \citenum{POPC1POPE1macrogT300K} \\
+      \hline
+      POPC             & Slipid \cite{jambeck12b}     &0          & 512 & 23943 & 0  &  298  & 170 & 100 & \citenum{POPCslipid298K} \\
+      POPC:POPE (1:1)  & Slipid \cite{jambeck12b}     &0          & 128 & 5120  & 0  &  298  & 500 & 300 & \citenum{POPC1POPE1slipidT298K} \\
+     \hline
+      POPC                   & GROMOS-CKP \cite{piggot12} & 0  & 500     & 25000 & 0   &  310  & 500 & 100 & \citenum{POPCgromosCKPT310K}  \\
+      POPC:POPG (7:3)        & GROMOS-CKP / GROMOS54A7 \cite{piggot11,piggot12,schmid11} & 0  & 350:150 & 25000 & 0   &  310  & 500 & 100 & \citenum{POPC7POPG3gromosCKPT310K} \\
+     \hline
+      POPC                   & Slipid \cite{jambeck12b}  & 0      & 500     & 25000 & 0   &  310  & 500 & 100 & \citenum{POPCslipid301K}  \\
+      POPC:POPG (7:3)        & Slipid / {\AA}qvist \cite{jambeck12b,aqvist90} & 0  & 350:150 & 25000 & 0   &  310  & 500 & 100 & \citenum{slipidPOPC70POPG30T310K} \\
+      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0  & 128:128 & 12800 & 0   &  298  & 500 & 400 & \citenum{slipidPOPC50POPG50T298K} \\
+      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0.1  & 128:128 & 12800 & 23  &  298  & 500 & 400 & \citenum{slipidPOPC50POPG50T298K} \\
+      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0.2  & 128:128 & 12800 & 46  &  298  & 1500 & 500 & \citenum{slipidPOPC50POPG50T298K} \\
+      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0.5  & 128:128 & 12800 & 115 &  298  & 1500 & 500 & \citenum{slipidPOPC50POPG50T298K} \\
+      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 1.0  & 128:128 & 12800 & 230 &  298  & 1500 & 500 & \citenum{slipidPOPC50POPG50T298K} \\
+      \hline
+      POPC:POPG (4:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}       & 0  & 350:88 & 26265 & 0  &  298  & 400 & 350 & \citenum{Lipid17POPCPOPG8020} \\
+      POPC:POPG (4:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}    & 0.1& 350:88 & 26124 & 47 &  298  & 400 & 250 & \citenum{Lipid17POPCPOPG8020100mMCaCl} \\
+      POPC:POPG (4:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}    & 1.0& 350:88 & 24840 & 475 &  298  & 1200 & 200 & \citenum{Lipid17POPCPOPG80201000mMCaCl} \\
+      POPC:POPG (1:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}   & 0  & 150:150 & 31572 & 0  &  298  & 320 & 200 & \citenum{Lipid17POPCPOPG5050} \\
+      POPC:POPG (1:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}   & 0.1& 150:150 & 31401 & 57 &  298  & 718 & 198 & \citenum{Lipid17POPCPOPG5050100mMCaCl} \\
+      POPC:POPG (1:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}  & 1.0& 150:150 & 29865 & 569 &  298  & 720 & 200 & \citenum{Lipid17POPCPOPG50501000mMCaCl} \\
+      \hline
+      POPC:POPG (4:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     &0     & 350:88 & 26265 & 0  &  298  & 400 & 300 & \citenum{Lipid17eccPOPCPOPG8020} \\
+      POPC:POPG (4:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 0.1& 350:88 & 26124 & 47 &  298  & 400 & 300 & \citenum{Lipid17eccPOPCPOPG8020100mMCaCl} \\
+      POPC:POPG (4:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 1.0& 350:88 & 24840 & 475 &  298  & 400 & 300 & \citenum{Lipid17eccPOPCPOPG80201000mMCaCl} \\
+      POPC:POPG (1:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 0  & 150:150 & 31572 & 0  &  298  & 347.8 & 333 & \citenum{Lipid17eccPOPCPOPG5050} \\
+      POPC:POPG (1:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 0.1& 150:150 & 29865 & 54 &  298  & 400 & 300 & \citenum{Lipid17eccPOPCPOPG5050100mMCaCl} \\
+      POPC:POPG (1:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 1.0& 150:150 & 29865 & 569 &  298  & 600 & 400 & \citenum{Lipid17eccPOPCPOPG50501000mMCaCl} \\
+      \hline
+      POPC             & Berger-POPC-07 \cite{ollila07a} & 0  & 256 & 10240 & 0  &  300  & 300 & 200 & \citenum{POPCberger300K} \\
+      POPC:POPE (1:1)  & Berger-POPE-04~\cite{devries04}  & 0  & 128 & 11008 & 0  &  300  & 300 & 200 & \citenum{POPC1POPE1berger} \\
       %POPC:DOPE (1:1)  & Berger \cite{??}  \todoi{This is probable not plain berger, correct force filed should be described.} &0          & 0  & 128 & 10240 & 0  &  300  & 300 & 200 & \cite{POPC1DOPE1berger} \\
@@
   \end{minipage}
-    \todo{ion model for GROMOS-CKP?} \\ % TJP: Most likely standard GROMOS but not my simulations so I've left the ion reference
 \end{table*}
@@
 
-\noindent {\it POPC:POPG mixture} \todo{Simulation details by A. Peon.}
+\noindent {\it POPC:POPG mixture} 
+Symmetric membranes of pure POPC and POPC:POPG 7:3 mixtures containing 500 lipids and 50 SPC water molecules per lipid were composed with CHARMM-GUI \cite{lee16}.
+The POPC:POPG 7:3 mixture contained also 150 sodium counter ions to balance the charge of PG lipids.
+GROMOS-CKP force field~\cite{piggot11,piggot12} was used for lipids and GROMOS54A7 \cite{schmid11} for the rest.
+Other simulation details are as in POPE and POPG systems.
 
@@
 {\it Josef Melcr}
-%% Please write description of your contribution here
+created the analysis script \texttt{scripts/calcOrderParameters.py}, 
+prepared the Lipid17ecc topology, 
+and contributed to discussions and manuscript preparation.
 

### 2021-05-03 CoverLetter.tex

\documentclass[11pt]{letter}
\usepackage{graphicx,baskervald,microtype}
\usepackage{hyperref,amsmath,xcolor}
\usepackage{pgfplots,marginnote}
\usepackage[top=0.5in,bottom=0.5in,left=1in,right=1in]{geometry}
%\newgeometry{margin=2.5cm}
\newcommand{\omamargin}[1]{\marginnote{\textbf{#1}}[7pt]}
\begin{document}
\reversemarginpar
\pagestyle{empty}
\noindent O. H. Samuli Ollila \\
\noindent Institute of Biotechnology \\
\noindent University of Helsinki, Finland \\
\noindent samuli.ollila@helsinki.fi, Tel. +358503746963 \\

Dear Editor,

Please find enclosed our manuscript entitled {\it Conformational plasticity of phospholipid headgroups in simulations and experiments}
by A. Bacle et al. which we would like to submit for consideration of publication in the \textit{ACS Central Science}.

In the manuscript, we demonstrate that lipids in cellular membranes exhibit more significant plasticity than previously anticipated.
Our results update the current understanding on how lipids selectively bind to proteins.

Lipids with different headgroups, the water facing components of cell membranes, are known to regulate cell functions
by selective binding to different proteins. It is, however, not known if the selectivity is driven by suitable
lipid conformations that fit to binding pockets in proteins (lock and key model), or by some other interactions.
One of the main obstacles in understanding the selective binding has been the lack of
experimental methods to resolve the conformational ensembles of lipids when they are not bound to proteins.

In 2013 we set out to solve the conformational ensembles of lipid headgroups by establishing the NMRlipids open collaboration project
which combines data from molecular dynamics (MD) simulations and NMR experiments.
Using the extensive data collected within this project, we have now resolved the differences in conformational ensembles
between most common lipids in eukaryotes and bacteria. 
Our results reveal significant structural plasticity in lipid headgroup and only minor differences in conformational ensembles
between lipids with different headgroup chemistry.
Bioinformatical analysis of protein-bound lipid structures from the protein data bank (PDB) revealed the plasticity in
lipid headgroups also when bound to proteins.

Novel conclusions from our work are that lipid headgroups can adapt to multiple binding sites in proteins,
and the selectivity of lipid-protein binding is driven by intermolecular interactions rather than by restrictions in lipid conformations.
Furthermore, we demonstrated the power of conceptually novel open collaboration approach
to complement the data in the PDB for elucidating how complex systems made up of disordered biomolecules behave.

Sincerely yours,

On behalf of the authors,

O. H. Samuli Ollila

\end{document}

### 2021-05-03 manuscriptPGPE.tex

@@
 \author{Josef Melcr}
-\affiliation{Institute of Organic Chemistry and Biochemistry of the 
-Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
 \affiliation{Groningen Biomolecular Sciences and Biotechnology Institute 
@@
 % insert abstract here
-  Chemistry of lipid headgroups, the water facing components of cell membranes that regulate cell functions via lipid--protein interactions, varies between organisms and organelles.
-Because membranes are in liquid state under physiological conditions,
-  individual lipids are not fixed to single conformations but they sample an ensemble of conformations with the Boltzmann weighted probabilities.
-  However, these ensembles and their dependence on lipid types have not yet been experimentally determined. 
-  Therefore, it has not been clear if lipid molecules exchange between few rigid conformations,
-  or if they can freely fluctuate in all possible conformations.
-  Here, we combine solid state NMR experiments and molecular dynamics simulations from the NMRlipids open collaboration to resolve the conformational ensembles of the headgroups of key lipid types in their liquid lamellar phase under various biologically relevant conditions.
-Interpretation of NMR experiments using the plethora of simulation data collected in the NMRlipids project suggests that all lipid headgroups sample a wide range of conformations in neutral and charged cellular membranes.
+  The chemistry of lipid headgroups, the water facing components of cell membranes that regulate cell functions via lipid--protein interactions, varies between organisms and organelles. 
+  In biological membranes at physiological conditions,
+  lipid molecules sample an ensemble of conformations.
+  However, the individual sets of conformations in the ensemble and their dependence on lipid types have not yet been experimentally determined. 
+  It then remains unclear how the interactions with membrane proteins
+  are related to the structural configurations of the lipids. 
+  Here, we combine solid-state NMR experiments and molecular dynamics simulations from the NMRlipids open collaboration to resolve the conformational ensembles of the headgroups of several key lipid types in the liquid phase under various biologically relevant conditions.
+Interpretation of the NMR observables using the plethora of simulation data collected in the NMRlipids project suggests that all lipid headgroups sample a wide range of conformations in both neutral and charged cellular membranes.
+  Only the overall orientations of the headgroups 
+  measured by P-N/C$_\gamma$ vector 
+  is affected by membrane charge
+  and is related to the C$_\beta$ headgroup order parameter. 
   Differences in the headgroup chemistry between different lipid types manifest
-  in probability distributions of conformations, but all lipid types can access almost any of the possible conformations.
-  Together with the analysis of protein-bound lipids from the protein data bank (PDB), this suggests that lipids can bind to proteins in a wide range of conformations independently of their headgroup chemistry.
-Therefore, the selective adsorption of proteins to membranes is likely regulated by specific protein--lipid interactions rather than conformational restrictions of the bilayer lipids.
+  in the conformation probability distributions, 
+  nevertheless all lipid types can access almost any of the possible conformations.
+  This is in line with our analysis of protein-bound lipids from the protein data bank (PDB), 
+  which suggests that all lipid types can bind to proteins in a wide range of conformations independently of their headgroup chemistry.
+Therefore, the selective adsorption of peripheral proteins to membranes and binding of lipids to transmembrane proteins are likely regulated by specific protein--lipid interactions rather than conformational restrictions of the bilayer lipids
+  justifying the use of simplified coarse-grained models,
+  which account for the intermolecular interactions 
+  as well as for the entropic effects 
+  in their effective empirical potentials.
   Our results pave the way to comprehensive understanding of
@@
 are known to be essential for the function of several proteins \cite{lee11,lemmon08},
-but it is not clear if the specificity is driven by the
+but it is not clear if the specificity arises from the
 differences in accessible conformations between lipid types or
-by specific lipid--protein interactions.
+only from specific non-covalent lipid--protein interactions.
 
-Stuctures of protein-bound lipids are available in the protein data bank (PDB) \cite{berman00}
+Structures of protein-bound lipids are available in the protein data bank (PDB) \cite{berman00}
 and crystal structures of lipids have been determined \cite{buldt81,pascher92},
@@
 We elucidate also the effect of charges and protein binding to lipid headgroup conformations.
-Zwitterionic PC and PE are the most common lipids in eukaryotes and bacteria, respectively \cite{vanmeer08,sohlenkamp16}.
+Zwitterionic PC and PE lipids are the most common lipids in eukaryotes and bacteria, respectively \cite{vanmeer08,sohlenkamp16}.
 PE is also the second most abundant glycerophospholipid in eukaryotic cells
@@
 
-Similarity of glycerol backbone and headgroup order parameters in model membranes and bacteria \cite{gally81,scherer87,seelig90}
+The similarity of glycerol backbone and headgroup order parameters in model membranes and bacteria \cite{gally81,scherer87,seelig90}
 suggest that our results can be used to 
@@
 The headgroup and glycerol backbone C--H bond order parameters of 1-palmitoyl-2-oleoyl-sn-glycero-3-phosphoethanolamine (POPE) and 1-palmitoyl-2-oleoyl-sn-glycero-3-phospho-(1'-rac-glycerol) (POPG), purchased from Avanti polar lipids, were measured using natural abundance $^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
-The magnitudes of order parameters were determined from the chemical-shift resolved dipolar splittings using a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04}, and the signs from S-DROSS experiments~\cite{gross97} combined with SIMPSON simulations \cite{bak00}.
+The samples were prepared by simply mixing the lipids with water until a homogeneous preparation of multi-lamellar vesicles (MLVs) was attained.
+The magnitudes of order parameters were determined from chemical-shift resolved dipolar splittings using a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04}, and the order parameter signs from S-DROSS experiments~\cite{gross97} combined with SIMPSON simulations \cite{bak00}.
 %The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
 %Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
-The NMR experiments were identical as in our previous work~\cite{antila19}.
+The set-up of the solid-state NMR experiments were identical as in our previous work~\cite{antila19}.
 The POPE experiments were recorded at 310~K and POPG experiments at 298~K, where the bilayers are in the liquid disordered phase \cite{marsh13}.
@@
 at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm}.
-\todo{How were $\alpha$ and  $\gamma$-carbon peaks assigned in POPG?}.
+The asignment of the $\alpha$, $\beta$, and $\gamma$ peaks of POPG was based on the measured R-PDLF splittings in comparison to the known order parameters from previous $^2$ NMR experiments~\cite{borle85,wohlgemuth80}.
 The $\beta$-carbon peak from POPG overlapped with the g$_2$ peak from glycerol backbone
 because their chemical environments are similar.
-\todo{Details to be checked by Tiago}.
 
@@
 %\url{github.com/NMRlipids/NMRlipidsIVotherHGs}. 
-Simulation details, accessibility information and quality evaluation of almost 100 combinations of lipid headgroup and force field simulated for this work are in the supplementary information.
+Simulation details, accessibility information and quality evaluation of approximately 70 combinations of lipid headgroup and force field simulated for this work are in the supplementary information.
 
@@
 using the ligand names listed in the supplementary information.
-Heavy atom dihedral distributions from the lipid structures were calculated
+Dihedral distributions from the lipid structures were calculated
 using the MDAnalysis Python library \cite{agrawal11,gowers16} and
@@
 Only the structures determined using X-ray crystallography or Cryo-EM with a resolution higher than 3.2~\text{\AA} were used in the analysis. 
-Some structures of multimeric proteins contained multiple lipids in the same conformation (analyzed dihedral angles within 3 degrees of each other).
-In that case, this conformation was counted once for the subsequent analyses
-to avoid its overweighting.
+Copies of the lipids from symmetric structures of multimeric proteins 
+that exhibited identical configurations 
+(dihedral angles matching within 3 degrees)
+were counted only once
+to aviod putting an artificially large weight on the conformation.
 
@@
 were handpicked from the results.
+
+The pairs of different protein-attached/interacting lipids 
+with similar configuration
+were searched by comparing the following torsion angles
+A LIST OF THE COMPARED ANGLES.
+Pairs of lipids were selected as having similar structure
+if their torsion angles agreed within 30$^/circ$. 
+Only the crystal structures structures with a resolution higher than 2.5~\text{\AA} were used.
 
@@
      \textbf{B)} Headgroup and glycerol backbone region order parameters of different different lipids.
-     \textbf{C)} Relative free energies for individual heavy atom dihedral angles estimated from the inverse Boltzmann formula.
+     \textbf{C)} Relative energies for individual heavy atom dihedral angles estimated from the inverse Boltzmann formula.
      Angles corresponding energies above 7~$k\mathrm{B}T$ are not shown because they are not observed in simulations.
-     \textbf{D)} Distributions of P--N vector angle with respect to membrane normal.
+     \textbf{D)} Distributions of P--N/C$_\gamma$ vector angle with respect to membrane normal without entropic correction.
   }
@@
 This has not been previously observed  in $^2$H NMR experiments detecting only the absolute values~\cite{wohlgemuth80,gally81,borle85}.
+This means that compared to the other phospholipids 
+the orientation of the C$_\beta$–H bond in the PG headgroup
+has a small preference for being parallel to the membrane normal.
+At the same time,
+increasing the $\beta$-carbon order parameters in the phospholipids
+yields corresponding systematic shifts 
+of the P-N/C$_\gamma$ vector angle distributions 
+towards configurations perpendicular to the membrane normal 
+(Fig \ref{structures}~B and D).
+
 The glycerol backbone order parameters are similar for all the lipid types, although they move slightly toward positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG.
@@
 However, the differences between PC and PE in P-O$_\alpha$-C$_\alpha$-C$_\beta$ dihedral
-and P--N vector dipole may be artificial as the $\beta$-carbon order parameter in PC is too negative even in the best available force field, thereby not being equal to the order parameter in PE as observed in experiments \cite{botan15}.
+and P--N/C$_\gamma$ vector dipole may be artificial as the $\beta$-carbon order parameter in PC is too negative even in the best available force field, thereby not being equal to the order parameter in PE as observed in experiments \cite{botan15}.
 
@@
 \ref{structures}).
-Since the rotation of dihedral angles to almost any position bears relatively low free energy cost, the lipid headgroups are able to adopt a wide range of conformations when interacting with proteins, ions, or other biomolecules.
+Since the rotation of dihedral angles to almost any position bears relatively low energy cost, the lipid headgroups are able to adopt a wide range of conformations when interacting with proteins, ions, or other biomolecules.
+MD simulations can represent such conformational ensembles 
+with accurate enough force fields.
 %Our results support the models proposing free rotation around phosphate group for PC lipids, which decouples the dynamics and conformations between acyl chains and headgroups \cite{klauda08c,antila21b}, and suggests that these models could be applied for all lipids containing the phosphorus group.
@@
 
-\subsection{Lipid conformational ensembles in charged membranes}
+\subsection{Lipid conformational ensembles in charged membranes are not restrained}
 
-Charged entities, such as lipids, proteins, surfactants, drugs, and ions incorporated in membranes can reorient the headgroup dipole in PC lipids and affect the order parameters of the lipid headgroups \cite{seelig87}.
+Charged entities, such as lipids, proteins, surfactants, drugs, and ions incorporated in membranes can change the orientation of the headgroup dipole in PC lipids and affect the order parameters of the lipid headgroups \cite{seelig87, botan15, melcr18}.
 However, the detailed understanding of the structural and energetic responses of lipids to membrane-bound electric charge is still lacking \cite{Semchyschyn04}.
@@
 dihedrals.
-The choline region remains essentially unchanged and only minor changes are observed in other dihedrals even at a surfactant molar fraction of 0.47.
+The choline region remains essentially unchanged and only minor changes are observed in the remaining headgroup dihedrals even at a surfactant molar fraction of 0.47 (Fig S\ref{HGorderparametersPCvsSURFdihedrals} in SI).
 
@@
 lipids could not be resolved with the currently available force fields and experimental data
-(Figs.~\ref{HGorderparametersPCvsPEPG} and~\ref{HGorderparametersPGvsPCchange}, and Ref.~\cite{antila19,melcr20}).
+(Figs.~\ref{HGorderparametersPCvsPEPG} and~\ref{HGorderparametersPGvsPCchange}, and Ref.~\cite{melcrova16,antila19,melcr20}).
 
@@
 play a minor role while the entropic cost of lipid binding may be significant.
+This provides further justification for using simplified coarse-grained models,
+which account for the intermolecular interactions 
+as well as for the entropic effects 
+in their effective empirical potentials. \cite{LipidFingerprints2018}
 Finally, we demonstrated the power of open access MD simulation data from the NMRlipids open collaboration
@@
 \begin{acknowledgments}
+    %     Put your acknowledgments here.
 AP is grateful to the Centro de
@@
 MJ thanks CSC -- IT Center for Science for computational resources and the Emil Aaltonen foundation for financial support.
-    %     Put your acknowledgments here.
 
@@
 
-O.H.S.O, A.M.K, and S.I.V acknowledge CSC -- IT Center for Science for computational resources and Academy of Finland (315596 and 319902) for financial support.
+O.H.S.O, A.M.K, and S.I.V acknowledge CSC -- IT Center for Science for computational resources and Academy of Finland (grants 315596 and 319902) for financial support.
 
 T.J.P acknowledges use of the Iridis high-performance computing resources at the University of Southampton.
+
+J.M. thanks the Center for Information Technology of the University of Groningen for their support and for providing access to the Peregrine high performance computing cluster.
 

### 2021-05-03 manuscriptPGPE.tex
Updates from Overleaf

@@
   The chemistry of lipid headgroups, the water facing components of cell membranes that regulate cell functions via lipid--protein interactions, varies between organisms and organelles.
-  In biological membranes at physiological conditions,
-  lipid molecules sample an ensemble of conformations.
-  However, the individual sets of conformations in the ensemble and their dependence on lipid types have not yet been experimentally determined. 
-  It then remains unclear how the interactions with membrane proteins
-  are related to the structural configurations of the lipids. 
-  Here, we combine solid-state NMR experiments and molecular dynamics simulations from the NMRlipids open collaboration to resolve the conformational ensembles of the headgroups of several key lipid types in the liquid phase under various biologically relevant conditions.
+In physiologically relevant liquid state,
+  lipid molecules are not fixed to a single structure but they sample an ensemble of conformations with the Boltzmann weighted probabilities.
+  However, the distributions of conformations in these ensembles and their dependence on lipid types have not yet been experimentally determined. 
+  Therefore, it has not been clear if lipid molecules exchange between few rigid conformations,
+  or if they can freely fluctuate in all possible conformations.
+  Here, we combine solid-state NMR experiments and molecular dynamics simulations from the NMRlipids open collaboration to resolve the conformational ensembles of headgroups of several key lipid types in various biologically relevant conditions.
 Interpretation of the NMR observables using the plethora of simulation data collected in the NMRlipids project suggests that all lipid headgroups sample a wide range of conformations in both neutral and charged cellular membranes.
-  Only the overall orientations of the headgroups 
-  measured by P-N/C$_\gamma$ vector 
-  is affected by membrane charge
-  and is related to the C$_\beta$ headgroup order parameter. 
   Differences in the headgroup chemistry between different lipid types manifest
-  in the conformation probability distributions, 
-  nevertheless all lipid types can access almost any of the possible conformations.
-  This is in line with our analysis of protein-bound lipids from the protein data bank (PDB), 
-  which suggests that all lipid types can bind to proteins in a wide range of conformations independently of their headgroup chemistry.
-Therefore, the selective adsorption of peripheral proteins to membranes and binding of lipids to transmembrane proteins are likely regulated by specific protein--lipid interactions rather than conformational restrictions of the bilayer lipids
-  justifying the use of simplified coarse-grained models,
-  which account for the intermolecular interactions 
-  as well as for the entropic effects 
-  in their effective empirical potentials.
+  in probability distributions of conformations, but all lipid types can access almost any of the possible conformations.
+  Together with the analysis of 894 protein-bound lipid structures from the protein data bank (PDB), this suggests that lipids can bind to proteins in wide range of conformations which are not limited by the headgroup chemistry.
+Therefore, the selective adsorption of peripheral proteins to membranes and binding of lipids to transmembrane proteins are likely regulated by specific protein--lipid interactions rather than conformational restrictions of the lipids.
   Our results pave the way to comprehensive understanding of
@@
 differences in accessible conformations between lipid types or
-only from specific non-covalent lipid--protein interactions.
+from specific non-covalent lipid--protein interactions.
 
@@
 We elucidate also the effect of charges and protein binding to lipid headgroup conformations.
-Zwitterionic PC and PE lipids are the most common lipids in eukaryotes and bacteria, respectively \cite{vanmeer08,sohlenkamp16}.
+Zwitterionic PC and PE are the most common lipids in eukaryotes and bacteria, respectively \cite{vanmeer08,sohlenkamp16}.
 PE is also the second most abundant glycerophospholipid in eukaryotic cells
@@
 
-The similarity of glycerol backbone and headgroup order parameters in model membranes and bacteria \cite{gally81,scherer87,seelig90}
+Similarity of glycerol backbone and headgroup order parameters in model membranes and bacteria \cite{gally81,scherer87,seelig90}
 suggest that our results can be used to 
@@
 
-The differences in headgroup conformation ensembles were analyzed calculating the distributions of the last six heavy atom dihedral angles, i.e., dihedrals O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$, P-O$_\alpha$-C$_\alpha$-C$_\beta$, O$_{g3}$-P-O$_\alpha$-C$_\alpha$, g$_3$-O$_{g3}$-P-O$_\alpha$, g$_2$-g$_3$-O$_{g3}$-P, and g$_1$-g$_2$-g$_3$-O$_{g3}$ labeled in Fig.
+The differences in headgroup conformation ensembles were analyzed calculating the distributions of six dihedral angles, O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$, P-O$_\alpha$-C$_\alpha$-C$_\beta$, O$_{g3}$-P-O$_\alpha$-C$_\alpha$, g$_3$-O$_{g3}$-P-O$_\alpha$, g$_2$-g$_3$-O$_{g3}$-P, and g$_1$-g$_2$-g$_3$-O$_{g3}$, labeled in Fig.
 \ref{HGorderParameters} C.
@@
 Only the structures determined using X-ray crystallography or Cryo-EM with a resolution higher than 3.2~\text{\AA} were used in the analysis.
-Copies of the lipids from symmetric structures of multimeric proteins 
-that exhibited identical configurations 
-(dihedral angles matching within 3 degrees)
-were counted only once
-to aviod putting an artificially large weight on the conformation.
+Some structures of multimeric proteins contained multiple lipids in the same conformation.
+The conformations were considered identical if all dihedral angles were the same within 3 degrees.
+In these cases, this conformation was counted only once for the subsequent analyses
+to avoid its overweighting.
 
-To demonstrate the existence of similar structures of different lipids bound to different proteins, we searched for pairs having the last five headgroup heavy atom dihedral angles (excluding g$_1$-g$_2$-g$_3$-O$_{g3}$) within
+To demonstrate the existence of similar structures of different lipids bound to different proteins, we searched for pairs having the last five heavy atom dihedrals angles in the headgroup end (excluding g$_1$-g$_2$-g$_3$-O$_{g3}$ from the list in previous section) within
 30 degrees of each other among the structures with resolution higher than 2.5~\text{\AA}. The two most representative examples
 were handpicked from the results.
-
-The pairs of different protein-attached/interacting lipids 
-with similar configuration
-were searched by comparing the following torsion angles
-A LIST OF THE COMPARED ANGLES.
-Pairs of lipids were selected as having similar structure
-if their torsion angles agreed within 30$^/circ$. 
-Only the crystal structures structures with a resolution higher than 2.5~\text{\AA} were used.
 
@@
      Angles corresponding energies above 7~$k\mathrm{B}T$ are not shown because they are not observed in simulations.
-     \textbf{D)} Distributions of P--N/C$_\gamma$ vector angle with respect to membrane normal without entropic correction.
+     \textbf{D)} Distributions of P--N/C$_\gamma$ vector angle with respect to membrane normal.
   }
@@
 This has not been previously observed  in $^2$H NMR experiments detecting only the absolute values~\cite{wohlgemuth80,gally81,borle85}.
-This means that compared to the other phospholipids 
-the orientation of the C$_\beta$–H bond in the PG headgroup
-has a small preference for being parallel to the membrane normal.
-At the same time,
-increasing the $\beta$-carbon order parameters in the phospholipids
-yields corresponding systematic shifts 
-of the P-N/C$_\gamma$ vector angle distributions 
-towards configurations perpendicular to the membrane normal 
-(Fig \ref{structures}~B and D).
-
 The glycerol backbone order parameters are similar for all the lipid types, although they move slightly toward positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG.
@@
 However, the differences between PC and PE in P-O$_\alpha$-C$_\alpha$-C$_\beta$ dihedral
-and P--N/C$_\gamma$ vector dipole may be artificial as the $\beta$-carbon order parameter in PC is too negative even in the best available force field, thereby not being equal to the order parameter in PE as observed in experiments \cite{botan15}.
+and P--N vector dipole may be artificial as the $\beta$-carbon order parameter in PC is too negative even in the best available force field, thereby not being equal to the order parameter in PE as observed in experiments \cite{botan15}.
 
 In conclusion, all studied lipid headgroups sample very broad conformational ensembles in the liquid lamellar phase and the sampled dihedral angles are within similar ranges for all headgroup types.
-% (Fig.
-\ref{structures}).
 Since the rotation of dihedral angles to almost any position bears relatively low energy cost, the lipid headgroups are able to adopt a wide range of conformations when interacting with proteins, ions, or other biomolecules.
-MD simulations can represent such conformational ensembles 
-with accurate enough force fields.
-%Our results support the models proposing free rotation around phosphate group for PC lipids, which decouples the dynamics and conformations between acyl chains and headgroups \cite{klauda08c,antila21b}, and suggests that these models could be applied for all lipids containing the phosphorus group.
-%The wide range of observed conformations suggest that
 Therefore, the structures in lipid crystals \cite{buldt81,pascher92} are not representative of the liquid state, and models aiming to explain NMR data using only a few conformations \cite{seelig77c,davis83,Semchyschyn04,akutsu20} are not sufficient to capture the large conformational space of lipids in the liquid lamellar state.
@@
 
-\subsection{Lipid conformational ensembles in charged membranes are not restrained}
+\subsection{Lipid conformational ensembles in charged membranes}
 
-Charged entities, such as lipids, proteins, surfactants, drugs, and ions incorporated in membranes can change the orientation of the headgroup dipole in PC lipids and affect the order parameters of the lipid headgroups \cite{seelig87, botan15, melcr18}.
+Charged entities, such as lipids, proteins, surfactants, drugs, and ions incorporated in membranes can change the orientation of the headgroup dipole in PC lipids and affect the order parameters of the lipid headgroups~\cite{seelig87}.
 However, the detailed understanding of the structural and energetic responses of lipids to membrane-bound electric charge is still lacking \cite{Semchyschyn04}.
@@
 dihedrals.
-The choline region remains essentially unchanged and only minor changes are observed in the remaining headgroup dihedrals even at a surfactant molar fraction of 0.47 (Fig S\ref{HGorderparametersPCvsSURFdihedrals} in SI).
+The choline region remains essentially unchanged and only minor changes are observed in the other dihedrals even at a surfactant molar fraction of 0.47.
 
@@
 play a minor role while the entropic cost of lipid binding may be significant.
-This provides further justification for using simplified coarse-grained models,
-which account for the intermolecular interactions 
-as well as for the entropic effects 
-in their effective empirical potentials. \cite{LipidFingerprints2018}
 Finally, we demonstrated the power of open access MD simulation data from the NMRlipids open collaboration
@@
 
-\listoftodos
+%\listoftodos
 

### 2021-05-05 manuscriptPGPE.tex
Updates from Overleaf

@@
 \usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{epstopdf}
@@
 % insert abstract here
-  The chemistry of lipid headgroups, the water facing components of cell membranes that regulate cell functions via lipid--protein interactions, varies between organisms and organelles.
-In physiologically relevant liquid state,
-  lipid molecules are not fixed to a single structure but they sample an ensemble of conformations with the Boltzmann weighted probabilities.
-  However, the distributions of conformations in these ensembles and their dependence on lipid types have not yet been experimentally determined. 
-  Therefore, it has not been clear if lipid molecules exchange between few rigid conformations,
-  or if they can freely fluctuate in all possible conformations.
-  Here, we combine solid-state NMR experiments and molecular dynamics simulations from the NMRlipids open collaboration to resolve the conformational ensembles of headgroups of several key lipid types in various biologically relevant conditions.
-Interpretation of the NMR observables using the plethora of simulation data collected in the NMRlipids project suggests that all lipid headgroups sample a wide range of conformations in both neutral and charged cellular membranes.
-  Differences in the headgroup chemistry between different lipid types manifest
-  in probability distributions of conformations, but all lipid types can access almost any of the possible conformations.
-  Together with the analysis of 894 protein-bound lipid structures from the protein data bank (PDB), this suggests that lipids can bind to proteins in wide range of conformations which are not limited by the headgroup chemistry.
-Therefore, the selective adsorption of peripheral proteins to membranes and binding of lipids to transmembrane proteins are likely regulated by specific protein--lipid interactions rather than conformational restrictions of the lipids.
-  Our results pave the way to comprehensive understanding of
-  lipid mediated signaling and lipid--protein interactions in biomedical applications.
+  The chemistry of lipid headgroups, the water-facing interface of cellular membranes, regulates cell functions via lipid--protein interactions. In the physiologically relevant liquid state,
+  lipid molecules are not fixed to a single structure but sample an ensemble of conformations with Boltzmann-weighted probabilities.
+  However, the distributions of conformations in such ensembles and their lipid-type dependence have not yet been experimentally determined: 
+  It is not clear if lipid molecules exchange between a few rigid structures,
+  or fluctuate freely across a practically continuous spectrum of  conformations.
+  Here, we combine solid-state NMR experiments and molecular dynamics simulations from the NMRlipids Project to resolve the conformational ensembles of headgroups of four key lipid types in various biologically relevant conditions.
+We find all lipid headgroup types to sample a wide range of conformations in both neutral and charged cellular membranes.
+  Differences in the headgroup chemistry do manifest
+  in probability distributions of conformations, but all four lipid types can access almost the whole conformation range.
+  Together with the analysis of 894 protein-bound lipid structures from the Protein Data Bank (PDB), this suggests that lipids can bind to proteins in a wide range of conformations, which are not limited by the headgroup chemistry.
+Therefore, the selective adsorption of peripheral proteins to membranes and the binding of lipids to transmembrane proteins are likely to be regulated by specific protein--lipid interactions, not by any inherent conformational restrictions of the lipids.
+  Our results pave the way to comprehensive understanding of lipid--protein interactions in biomedical applications, in particular in
+  lipid-mediated signaling.
 
@@
 regulate protein functions in many different ways \cite{lee03,vanmeer08}.
-Lipids can directly bind to proteins or indirectly affect protein
+Lipids can directly bind to proteins, or indirectly affect protein
 functions by altering membrane properties such as charge or elasticity \cite{lee03,lemmon08}.
@@
 differences in accessible conformations between lipid types or
-from specific non-covalent lipid--protein interactions.
+from specific intermolecular lipid--protein interactions.
 
-Structures of protein-bound lipids are available in the protein data bank (PDB) \cite{berman00}
+Structures of protein-bound lipids are available in the Protein Data Bank (PDB) \cite{berman00},
 and crystal structures of lipids have been determined \cite{buldt81,pascher92},
 but their relation to the conformational ensembles in bulk membranes in the liquid lamellar phase remains unclear \cite{marsh13b}.
-Most accurate experimental information on conformational ensembles of lipids
-in liquid lamellar phase are typically derived from NMR experiments, particularly from 
-C--H bond order parameters measured using $^2$H NMR~\cite{seelig77c,davis83,Semchyschyn04}.
-According to these experiments, the glycerol backbone conformations are largely similar irrespectively of the headgroup \cite{gally81} and
-the headgroup conformations are similar in phosphatidylcholine (PC), phosphatidylethanolamine (PE) and phosphatidylglycerol (PG) lipids,
-whereas the headgroup of phosphatidylserine (PS) lipids is more rigid \cite{wohlgemuth80,buldt81}. 
-However, order parameter signs are not accessible in $^2$H NMR experiments \cite{ollila16}
+The most accurate experimental information on conformational ensembles of lipids
+in this biologically relevant phase are typically derived from NMR experiments, particularly from the
+C--H bond order parameters $S_\mathrm{CH}$ measured using $^2$H\,NMR~\cite{seelig77c,davis83,Semchyschyn04}.
+According to these experiments, the glycerol backbone conformations are largely similar irrespective of the headgroup \cite{gally81}, and
+the headgroup conformations are similar in the phosphatidylcholine (PC), phosphatidylethanolamine (PE), and phosphatidylglycerol (PG) lipids,
+whereas the phosphatidylserine (PS) headgroup is more rigid \cite{wohlgemuth80,buldt81}. 
+Notably, however, the signs of $S_\mathrm{CH}$ are not accessible from $^2$H\,NMR experiments \cite{ollila16},
 and universal models to map order parameters to structural ensembles are not available \cite{pezeshkian18,akutsu20}.
-Therefore, it is not clear if the conformational ensemble of lipid headgroups in liquid lamellar bilayer
-is composed of exchange between few restricted conformations, or if lipid headgroups can fluctuate freely across a wide range of conformations.
+Therefore, it is not clear if the lipid headgroups in a liquid lamellar bilayer exchange between a few restricted conformations, or can fluctuate freely across a wide conformational range.
 
-Here, we use natural abundance $^{13}$C NMR experiments %(enabling the determination of order parameter signs)
-and MD simulations from the NMRlipids open collaboration
-to resolve the differences in the conformational ensembles between PC, PE, PG and PS lipid headgroups.
-We elucidate also the effect of charges and protein binding to lipid headgroup conformations.
-Zwitterionic PC and PE are the most common lipids in eukaryotes and bacteria, respectively \cite{vanmeer08,sohlenkamp16}.
+Here, we use natural abundance $^1$H-$^{13}$C solid-state NMR experiments %(enabling the determination of order parameter signs)
+and MD simulations from the NMRlipids Project
+to resolve differences in the conformational ensembles of PC, PE, PG, and PS lipid headgroups.
+Zwitterionic PC are the most common lipids in eukaryotes and PE in bacteria \cite{vanmeer08,sohlenkamp16};
 PE is also the second most abundant glycerophospholipid in eukaryotic cells
 and has been related to various diseases \cite{vance15,calzada16,patel17}.
-PS and PG are the most common negatively charged lipids in eukaryotes and bacteria, respectively,
-and their presence affect membrane protein functionality and signaling \cite{lemmon08,leventis10,sohlenkamp16,hariharan18}.
-All the studied lipids specifically bind to various proteins \cite{yeagle14}.
-
+PS are the most common negatively charged lipids in eukaryotes and PG in bacteria,
+and both are known to affect membrane protein functionality and signaling \cite{lemmon08,leventis10,sohlenkamp16,hariharan18}.
+All the four studied lipid types specifically bind to various proteins \cite{yeagle14};
+we therefore elucidate also the effect of protein binding to headgroup conformations.
 Similarity of glycerol backbone and headgroup order parameters in model membranes and bacteria \cite{gally81,scherer87,seelig90}
@@
 understand the biological role of lipid headgroups in lipid--protein interactions.
-These interactions are crucial, for example, in lipid mediated signaling \cite{lemmon08} and
-design of phoshoplipid-specific antibodies \cite{vigant15}.
+These interactions are crucial, for example, in lipid-mediated signaling \cite{lemmon08} and for the
+design of phoshopolipid-specific antibodies \cite{vigant15}.
 
@@
 \subsection{Experimental C--H bond order parameters}
-The headgroup and glycerol backbone C--H bond order parameters of 1-palmitoyl-2-oleoyl-sn-glycero-3-phosphoethanolamine (POPE) and 1-palmitoyl-2-oleoyl-sn-glycero-3-phospho-(1'-rac-glycerol) (POPG), purchased from Avanti polar lipids, were measured using natural abundance $^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
-The samples were prepared by simply mixing the lipids with water until a homogeneous preparation of multi-lamellar vesicles (MLVs) was attained.
-The magnitudes of order parameters were determined from chemical-shift resolved dipolar splittings using a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04}, and the order parameter signs from S-DROSS experiments~\cite{gross97} combined with SIMPSON simulations \cite{bak00}.
+The headgroup and glycerol backbone C--H bond order parameters of 1-palmitoyl-2-oleoyl-sn-glycero-3-phosphoethanolamine (POPE) and 1-palmitoyl-2-oleoyl-sn-glycero-3-phospho-(1'-rac-glycerol) (POPG), purchased from Avanti polar lipids, were measured using natural abundance $^1$H-$^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
+The samples were prepared by simply mixing the lipids with water until a homogeneous preparation of multi-lamellar vesicles (MLVs) was attained and left to equilibrate for approximately 24 hours before measurements.
+The magnitudes of the order parameters were determined using a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04}, and the order parameter signs from S-DROSS experiments~\cite{gross97} combined with SIMPSON simulations \cite{bak00}.
 %The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
@@
 at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm}.
-The asignment of the $\alpha$, $\beta$, and $\gamma$ peaks of POPG was based on the measured R-PDLF splittings in comparison to the known order parameters from previous $^2$ NMR experiments~\cite{borle85,wohlgemuth80}.
+The asignment of the $\alpha$, $\beta$, and $\gamma$ peaks of POPG was based on the measured R-PDLF dipolar couplings in comparison to the known order parameters from previous $^2$H\,NMR experiments~\cite{borle85,wohlgemuth80}.
 The $\beta$-carbon peak from POPG overlapped with the g$_2$ peak from glycerol backbone
@@
     \textbf{D)} Headgroup and glycerol backbone C-H bond order parameters for different phospholipids measured in the lamellar liquid disordered phase.
-    The $S_{\rm{CH}}$ magnitudes and signs determined by $^1$H-$^{13}$C NMR spectroscopy (filled squares) for POPE (310~K) and POPG (298~K)
-    measured in this work are shown together with the data previously reported for POPS (298~K) \cite{antila19} and POPC (300~K) \cite{ferreira13,ferreira16}.
-The headgroup and glycerol backbone $S_{\rm{CH}}$ magnitudes measured previously with $^2$H NMR (empty symbols) are also plotted as real values using the signs determined in this work for
+    Filled squares show $S_{\rm{CH}}$ magnitudes and signs determined by $^1$H-$^{13}$C NMR spectroscopy for POPE (310~K) and POPG (298~K), as
+    measured in this work, and from data previously reported for POPS (298~K) \cite{antila19} and POPC (300~K) \cite{ferreira13,ferreira16}.
+Empty symbols show headgroup and glycerol backbone $S_{\rm{CH}}$ magnitudes measured previously with $^2$H NMR and are also plotted as real values using the signs determined in this work for
     POPG with 10nM PIPES (298~K) \cite{borle85},
@@
     DPPE (341~K) \cite{seelig76},
-    E.coli PE and E.coli PG (310~K) \cite{gally81}
+    E. coli PE and E. coli PG (310~K) \cite{gally81}
     .
@@
    \caption{\label{structures}
-     Results from the best simulation model (CHARMM36) simulations demonstrating the differences in conformational ensembles between different lipids. 
-     \textbf{A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurence of different conformations.
-     \textbf{B)} Headgroup and glycerol backbone region order parameters of different different lipids.
+     Results from the best simulation model demonstrating the differences in conformational ensembles between different lipids.
+     The best simulation model (CHARMM36) was selected from the comparison of ten different force fields based on quality evaluation against experimental $S_{\rm{CH}}$ values in Figs.~\ref{HGorderParametersPE} and~\ref{HGorderParametersPOPG}.
+     %For a detailed comparison of all the XX force-fields used with the $S_{\rm{CH}}$ values measured please read the supplementary information.
+     \textbf{A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurrence of different conformations.
+     \textbf{B)} Headgroup and glycerol backbone region order parameters of different lipids.
      \textbf{C)} Relative energies for individual heavy atom dihedral angles estimated from the inverse Boltzmann formula.
-     Angles corresponding energies above 7~$k\mathrm{B}T$ are not shown because they are not observed in simulations.
+     Angles corresponding to energies above 7~$k_\mathrm{B}T$ are not shown because they are not observed in simulations.
      \textbf{D)} Distributions of P--N/C$_\gamma$ vector angle with respect to membrane normal.
@@
 The decrease in the beginning of the S-DROSS curve suggests that the sign of larger g$_2$ order parameter
-is negative and later increase suggests that sign of smaller $\beta$ order parameter is positive (Fig.~\ref{HGorderParameters}B).
-This interpretation is confirmed by SIMPSON calculations in Fig.~\ref{POPGsimpson}.
+is negative and later increase suggests that sign of smaller $\beta$ order parameter is positive as confirmed by the SIMPSON simulations (Fig.~\ref{HGorderParameters}B).
 
-Experimental order parameters of POPC, POPE, POPG and POPS glycerol backbones and headgroups from this and previous studies are collected in Fig.~\ref{HGorderParameters}D, where the signs from $^{13}$C NMR experiments are used also for the $^2$H NMR data from the literature.
-The overall agreement of order parameters determined by different research teams and different techniques for the same lipid headgroup is excellent here and in previous studies \cite{botan15,ollila16,antila19}.
+Experimental order parameters of POPC, POPE, POPG and POPS glycerol backbones and headgroups from this and previous studies are displayed in Fig.~\ref{HGorderParameters}D, where the signs from $^{13}$C NMR experiments are used also for the $^2$H NMR data from the literature.
+The overall agreement of order parameters determined by different research teams and different techniques for the same lipid headgroup is excellent (within $\pm$0.02) here and in previous studies \cite{botan15,ollila16,antila19}.
 Therefore, the differences between lipid types are dictated by the headgroup chemistry rather than inaccuracies in experiments, differences in the acyl chains, or in experimental conditions.
@@
 On the other hand, the $\beta$-carbon order parameter of PG headgroup has a positive sign, in contrast to all the other lipid types.
-This has not been previously observed  in $^2$H NMR experiments detecting only the absolute values~\cite{wohlgemuth80,gally81,borle85}.
+The sign has not been previously known from $^2$H NMR experiments which enable detecting only order parameter absolute values~\cite{wohlgemuth80,gally81,borle85}.
 The glycerol backbone order parameters are similar for all the lipid types, although they move slightly toward positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG.
@@
   \caption{\label{changesWITHsurf}
-    \textbf{A)} Modulation of PC headgroup order parameters and P--N vector angle upon addition of cationic surfactant (dihexadecyldimethylammonium)
+    \textbf{A)} Modulation of PC headgroup order parameters and P--N vector tilt (angle between the P-N vector and the bilayer normal) upon addition of cationic surfactant (dihexadecyldimethylammonium)
     from CHARMM36 simulations compared with the experimental data \cite{scherer89}.
@@
 
-MD simulations could be used to interpret the lipid conformational ensembles from experiments due to their direct connection to the measured C-H bond order parameters.
-To find the best models for this purpose, we took advantage of the power of the NMRlipids project which has gathered a massive number of trajectories using open collaboration.
-We have thereby evaluated all the available MD simulation force fields of PC~\cite{botan15}, PS~\cite{antila19}, PE (Fig.~\ref{HGorderParametersPE}), and PG (Fig.~\ref{HGorderParametersPOPG}) lipids against NMR data in the NMRlipids project.
-The comparison of hundreds of simulation trajectories simulated with approximately 20 different force fields reveal that none of the current parameters can correctly capture the lipid conformational ensemble and reproduce the experimental C-H bond order parameters of all segments.
-However, the best parameters (CHARMM36) correctly capture the essential differences between PC, PE, PG and PS headgroup order parameters (Fig.~\ref{structures} B).
+MD simulations could be used to interpret the lipid conformational ensembles from experiments due to their direct connection to the measured C-H bond order parameters $S_\mathrm{CH}$.
+To find the best MD simulation models (force fields) for this purpose, we took advantage of the power of the NMRlipids Project, an open science collaboration that has gathered a massive number of simulation trajectories, allowing the evaluation of all the available force fields of PC~\cite{botan15}, PS~\cite{antila19}, PE (Fig.~\ref{HGorderParametersPE}), and PG (Fig.~\ref{HGorderParametersPOPG}) lipids against NMR data.
+The comparison of hundreds of simulation trajectories simulated with approximately 20 different force fields revealed that none of them can correctly capture the lipid conformational ensembles and reproduce the experimental $S_\mathrm{CH}$ of all segments within experimental error.
+However, CHARMM36, the force field showing the best agreement with experiments, correctly captures the essential differences between the PC, PE, PG, and PS headgroup order parameters (compare Figs.~\ref{HGorderParameters}D and \ref{structures} B).
 
-To understand the structural origin of distinct order parameters for PG and PS lipids,
+To understand the structural origin of distinct order parameters for the distinct lipids,
 we first calculated the heavy atom dihedral angle distributions from the best performing simulations that reproduced the differences between headgroups~(Fig. \ref{DIHdists}).
@@
 PS and PG lipids than for PC and PE lipids. 
-Rest of the dihedrals are similar between different lipids,
+The rest of the dihedrals are similar between different lipids,
 with the exception of PS lipids for which slightly larger energy for \textit{eclipsed anti} conformation
 in g$_3$-O$_{{\mathrm g}3}$-P-O$_\alpha$ dihedral was observed.
-Therefore we suggest that the main differences between lipid headgroups leading to distinct order parameters occur in the headgroup end, while phosphate region remains similar in different lipids with the exception of PS.
+Therefore we suggest that the main differences between lipid headgroups leading to distinct order parameters occur in the headgroup end, while in the phosphate region they remain similar in different lipids with the exception of PS.
 The increased barriers for dihedral rotations may explain the more rigid headgroup structures in PS~\cite{browning80,buldt81}.
@@
 Therefore, the structures in lipid crystals \cite{buldt81,pascher92} are not representative of the liquid state, and models aiming to explain NMR data using only a few conformations \cite{seelig77c,davis83,Semchyschyn04,akutsu20} are not sufficient to capture the large conformational space of lipids in the liquid lamellar state.
-However, MD simulations with accurate force fields can be expected to capture the correct conformational ensembles of lipids.
+However, conformational ensembles of lipids in MD simulation trajectories with accurate force fields that capture the experimental NMR order parameters and can be considered realistic, and thereby used to interpret the experimental data.
 
@@
 The dihedral angle distributions and relative energies
-in figures \ref{HGorderparametersPCvsSURFdihedrals} and \ref{changesWITHsurf} B) reveal that the
+in figures \ref{HGorderparametersPCvsSURFdihedrals} and \ref{changesWITHsurf} B reveal that the
 addition of positive charge into a membrane 
-decreases the abundance of \textit{trans} states in g$_2$-g$_3$-O$_{g_3}$-P and g$_3$-O$_{g_3}$-P-O$_\alpha$
-dihedrals.
+decreases the abundance of \textit{trans} states in g$_2$-g$_3$-O$_{g_3}$-P and g$_3$-O$_{g_3}$-P-O$_\alpha$, and \textit{cis} states in O$_{g_3}$-P-O$_\alpha$-C$\alpha$ dihedrals.
 The choline region remains essentially unchanged and only minor changes are observed in the other dihedrals even at a surfactant molar fraction of 0.47.
@@
 
-In conclusion, the structural response of lipid headgroups to 
-membrane-bound charges arise
+In conclusion, the MD simulation results suggest that the structural response of lipid headgroups to 
+membrane-bound charges arises
 from mild changes in dihedral angle distributions, rather than restriction of lipids to
@@
 Interpretation of experimental order parameters using MD simulations
-in previous sections suggest that PC, PE, PG and PS lipid headgroups are very flexible,
+in previous sections suggest that PC, PE, PG, and PS lipid headgroups are very flexible,
 allowing them to bind proteins in various different conformations.
@@
 The heavy atom dihedral angle distributions calculated from these conformations 
-(Fig.~\ref{dihedralsFROMpdb} A) reveal that the protein-bound lipids indeed exhibit wide
-range of conformations independently on the headgroup type.
-As for bulk lipid bilayers, only $cis$ conformations
+(Fig.~\ref{dihedralsFROMpdb} A) reveal that the protein-bound lipids indeed exhibit a wide
+range of conformations, independent of the headgroup type.
+As for bulk lipid bilayers (Figs.~\ref{structures}C, \ref{DIHdists}), only {\it cis} conformations
 of P-O$_\alpha$-C$_\alpha$-C$_\beta$ and g$_2$-g$_3$-O$_{g_3}$-P dihedrals are almost
 completely absent in all lipids, and significant differences between
-different headgroups were not observed.
+different headgroups are not observed.
 Structures deviating from lipid crystals 
 have been previously proposed to indicate inaccuracies in lipid structures in PDB \cite{marsh13b,pezeshkian18}. 
-However, we observe large deviations from lipid crystals structures also in conformational ensembles
+However, we observe large deviations from the lipid crystal structures also in conformational ensembles
 that reproduce the NMR data in the liquid lamellar phase, indicating that such deviations
-are realistic also in protein-bound states.
+are realistic also in the protein-bound states.
 
@@
 in a wide range of conformations.
-Therefore, the specific binding of lipids to proteins is not driven by the structural differences between headgroups.
-This is demonstrated in Fig.~\ref{dihedralsFROMpdb} B with two examples where different lipids bound to
+Therefore, the specific binding of lipids to proteins is not driven by differing structural preferences between headgroups.
+This is demonstrated in Fig.~\ref{dihedralsFROMpdb} B with two examples where different lipid types bound to
 different proteins have almost identical headgroup conformations:
@@
 On the other hand, a single lipid headgroup type is capable of accommodating various binding positions and as such would 
-be able to specifically bind to many different kinds of binding sites in different proteins.
+be able to specifically bind to many different kinds of binding sites.
 
@@
 
-C--H bond order parameters from NMR experiments %in this and previous studies \cite{wohlgemuth80,buldt81,seelig87}
-suggest that lipid headgroup conformational ensembles depend on lipid type (PC, PE, PG or PS) 
-and accumulated membrane charge (cationic lipids, surfactants, ions or drugs). 
-Our interpretation of these data, using MD simulations collected
-within the NMRlipids open collaboration,
-revealed that the differences in order parameters can be explained
-by relatively small changes in dihedral angle probability distributions.
-All studied headgroups (PC, PE, PG and PS) are flexible
-and access a similar wide range of conformations with only very few restictions in dihedral orientations
-also when charges are bound to membranes.
-The observed differences in order parameters originate from
-reweighting conformational probabilities rather than changes in accessible structures.
+C--H bond order parameters $S_\mathrm{CH}$ from NMR experiments %in this and previous studies \cite{wohlgemuth80,buldt81,seelig87}
+suggest that lipid headgroup conformational ensembles depend on lipid type (PC, PE, PG, or PS) 
+and membrane charge (cationic lipids, surfactants, accumulated ions or drugs). 
+Our analysis of these data, using MD simulations collected
+within the NMRlipids Project,
+revealed that the differences in $S_\mathrm{CH}$ can be explained
+by small changes in the dihedral angle probability distributions.
+All four studied headgroup types (PC, PE, PG, and PS) are flexible
+and access a similarly wide range of conformations, with very few restrictions in dihedral orientations---also when membranes are charged.
+In conclusion, the observed differences in $S_\mathrm{CH}$ originate from
+slightly reweighed conformational probabilities rather than changes in accessible structures.
 
 The flexibility and wide conformational space of headgroups suggest that protein--bound lipids can
-adapt to various binding sites to optimize the intermolecular lipid--protein interactions.
-We tested this prediction by analyzing the conformations of lipids that are tightly bound to proteins in the PDB.
-Indeed, also protein--bound lipids exhibit wide range of conformations without significant 
-differences between different lipid types. Therefore, the specificify of lipid binding to proteins is not
-regulated by accessible structures of lipids, and a single lipid type can adapt to various binding sites in proteins.
+adapt to various binding sites to optimize the lipid--protein interactions.
+We tested this prediction by analyzing from PDB the conformations of lipids that are tightly bound to proteins.
+We found that also the protein-bound lipids exhibit a wide range of conformations, without significant 
+differences between different lipid types: The specificity of lipid binding to proteins is not
+regulated by accessible lipid structures, but a single lipid type can adapt to various binding pockets.
 
-Our results pave the way toward the understanding of lipid-mediated cell signaling and how lipids regulate membrane protein function in general.
-We suggest that the key to understand selective binding of certain lipid types to proteins
-are intermolecular lipid--protein interactions, rather than conformational restrictions of the lipids.
-On the other hand, broad conformational ensembles in bulk bilayers suggest that lipid crystal structures
-play a minor role while the entropic cost of lipid binding may be significant.
-Finally, we demonstrated the power of open access MD simulation data from the NMRlipids open collaboration
-to complement the data in the PDB for elucidating how complex systems made up of disordered biomolecules behave.
+Our results pave the way toward understanding lipid-mediated cell signaling and how lipids regulate membrane protein function in general.
+We suggest that the key to understanding selective binding of certain lipid types to proteins is in intermolecular lipid--protein interactions, rather than in conformational restrictions inherent to the lipids.
+The broad conformational ensembles we see in bulk bilayers suggest that the structures observed in pure lipid crystals occur seldomly in liquid state.
+Consequently, in protein binding the entropic cost on the lipid may be significant.
+
+Finally, our work has demonstrated the power of open access MD simulation data
+to complement the PDB data in elucidating behavior of complex systems of disordered biomolecules.
 
@@
     %     Put your acknowledgments here.
-AP is grateful to the Centro de
+A.P. is grateful to the Centro de
 Supercomputación de Galicia (CESGA) for use of the Finis
-Terrae computer
-%
-MJ thanks CSC -- IT Center for Science for computational resources and the Emil Aaltonen foundation for financial support.
+Terrae computer.
 
-P.B. was supportedby the Academy of Finland (Grant 311031)
+M.J. thanks CSC -- IT Center for Science for computational resources and the Emil Aaltonen foundation for financial support.
+
+T.M.F. was supported by the Ministry of Economics, Science
+and Digitalisation of the State of Saxony-Anhalt, Germany.
+
+P.B. was supported by the Academy of Finland (Grant 311031).
 
@@
 
-T.J.P acknowledges use of the Iridis high-performance computing resources at the University of Southampton.
+T.J.P. acknowledges use of the Iridis high-performance computing resources at the University of Southampton.
 

### 2021-05-05 CoverLetter.tex

@@
 
-Please find enclosed our manuscript entitled {\it Conformational plasticity of phospholipid headgroups in simulations and experiments}
-by A. Bacle et al. which we would like to submit for consideration of publication in the \textit{ACS Central Science}.
+Please find enclosed our manuscript entitled {\it Conformational plasticity of lipid headgroups in cellular membranes and protein-lipid complexes}
+by A. Bacle et al. which we would like to submit for consideration to be published in \textit{ACS Central Science}.
 
-In the manuscript, we demonstrate that lipids in cellular membranes exhibit more significant plasticity than previously anticipated.
+In the manuscript, we demonstrate that lipids in cellular membranes exhibit a much higher plasticity in terms of allowed conformational states than previously anticipated.
 Our results update the current understanding on how lipids selectively bind to proteins.
@@
 Lipids with different headgroups, the water facing components of cell membranes, are known to regulate cell functions
-by selective binding to different proteins. It is, however, not known if the selectivity is driven by suitable
-lipid conformations that fit to binding pockets in proteins (lock and key model), or by some other interactions.
+by selective binding to different proteins. It is, however, not known if the selectivity is driven by unique
+lipid conformations between the lipid types that fit to binding pockets in proteins (lock and key model), or by some other interactions.
 One of the main obstacles in understanding the selective binding has been the lack of
@@
 In 2013 we set out to solve the conformational ensembles of lipid headgroups by establishing the NMRlipids open collaboration project
-which combines data from molecular dynamics (MD) simulations and NMR experiments.
+which combines data from molecular dynamics (MD) simulations and solid-state NMR experiments.
 Using the extensive data collected within this project, we have now resolved the differences in conformational ensembles
-between most common lipids in eukaryotes and bacteria. 
-Our results reveal significant structural plasticity in lipid headgroup and only minor differences in conformational ensembles
+between the most common lipids in eukaryotes and bacteria. 
+Our results reveal a significant structural plasticity in lipid headgroups and only minor differences of the conformational ensembles
 between lipids with different headgroup chemistry.
-Bioinformatical analysis of protein-bound lipid structures from the protein data bank (PDB) revealed the plasticity in
-lipid headgroups also when bound to proteins.
+Notably, and motivated by our findings on the model systems, by using a bioinformatical analysis of protein-bound lipid structures from the protein data bank (PDB), we found that the plasticity in lipid headgroups is a characteristic feature in lipid-protein complexes.
 
-Novel conclusions from our work are that lipid headgroups can adapt to multiple binding sites in proteins,
+The novel conclusions from our work are that lipid headgroups can adapt to multiple binding sites in proteins,
 and the selectivity of lipid-protein binding is driven by intermolecular interactions rather than by restrictions in lipid conformations.
-Furthermore, we demonstrated the power of conceptually novel open collaboration approach
-to complement the data in the PDB for elucidating how complex systems made up of disordered biomolecules behave.
+Furthermore, we demonstrated the power of a conceptually novel open collaboration approach that extended the realm of Molecular Dynamics simulations to complement the experimental data in the PDB for elucidating how complex systems made up of disordered biomolecules behave.
+
+We believe that our approach and findings are of great interest to understand the connection of lipid chemistry to protein-lipid interactions  and ultimately the role of lipid chemistry in a number of diseases.
+Therefore, we believe that \textit{ACS Central Science} is a highly suitable journal to publish this work.
+
+We now leave it for your own consideration.
 

### 2021-05-05 manuscriptPGPE.tex
Updates from Overleaf

@@
 
-\title{Conformational plasticity of phospholipid headgroups in simulations and experiments} %Title of paper
+\title{Conformational plasticity of lipid headgroups in cellular membranes and protein--lipid complexes} %Title of paper
 
@@
 \author{Pavel Buslaev}
-\affiliation{Nanoscience Center and Department of Chemistry, University of Jyväskylä, P.O. Box 35, 40014 Jyväskylä, Finland}
+\affiliation{Nanoscience Center and Department of Chemistry, University of Jyv{\"a}skyl{\"a}, P.O. Box 35, 40014 Jyv{\"a}skyl{\"a}, Finland}
 \affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, 141701 Dolgoprudny, Russia}
@@
 C--H bond order parameters $S_\mathrm{CH}$ measured using $^2$H\,NMR~\cite{seelig77c,davis83,Semchyschyn04}.
+Because these order parameters are similar as in living cells \cite{gally81,scherer87,seelig90},
+the model membranes can be used to resolve the lipid headgroup conformations in biological conditions.
 According to these experiments, the glycerol backbone conformations are largely similar irrespective of the headgroup \cite{gally81}, and
@@
 we therefore elucidate also the effect of protein binding to headgroup conformations.
-Similarity of glycerol backbone and headgroup order parameters in model membranes and bacteria \cite{gally81,scherer87,seelig90}
-suggest that our results can be used to 
-understand the biological role of lipid headgroups in lipid--protein interactions.
 These interactions are crucial, for example, in lipid-mediated signaling \cite{lemmon08} and for the
@@
      \textbf{C)} Relative energies for individual heavy atom dihedral angles estimated from the inverse Boltzmann formula.
-     Angles corresponding to energies above 7~$k_\mathrm{B}T$ are not shown because they are not observed in simulations.
+     Angles corresponding to energies above 8~$k_\mathrm{B}T$ are not shown because they are not observed in simulations.
      \textbf{D)} Distributions of P--N/C$_\gamma$ vector angle with respect to membrane normal.
@@
 The results in Fig.~\ref{structures}C suggest that all lipid headgroups are very flexible and
-energy costs for rotating individual dihedrals to almost any angle is low (below 7~$k_\mathrm{B}T$).
+energy costs for rotating individual dihedrals to almost any angle is low (below 8~$k_\mathrm{B}T$).
 Only $cis$ states of P-O$_\alpha$-C$_\alpha$-C$_\beta$ and g$_2$-g$_3$-O$_{g_3}$-P have larger relative
@@
 However, the differences between PC and PE in P-O$_\alpha$-C$_\alpha$-C$_\beta$ dihedral
-and P--N vector dipole may be artificial as the $\beta$-carbon order parameter in PC is too negative even in the best available force field, thereby not being equal to the order parameter in PE as observed in experiments \cite{botan15}.
+and P--N vector dipole may be artificial as the $\beta$-carbon order parameter in PC is too negative even in the best available force field~ \cite{botan15}, thereby not being equal to the order parameter in PE as observed in experiments.
 
@@
 addition of positive charge into a membrane 
-decreases the abundance of \textit{trans} states in g$_2$-g$_3$-O$_{g_3}$-P and g$_3$-O$_{g_3}$-P-O$_\alpha$, and \textit{cis} states in O$_{g_3}$-P-O$_\alpha$-C$\alpha$ dihedrals.
+decreases the abundance of \textit{trans} states in g$_2$-g$_3$-O$_{g_3}$-P and g$_3$-O$_{g_3}$-P-O$_\alpha$, and \textit{cis} states in O$_{g_3}$-P-O$_\alpha$-C$_\alpha$ dihedrals.
 The choline region remains essentially unchanged and only minor changes are observed in the other dihedrals even at a surfactant molar fraction of 0.47.

### 2021-05-05 manuscriptPGPEsuppl.tex
Updates from Overleaf

@@
   \centering
-  \includegraphics[width=9.0cm]{./Figs/HGorderparametersPG.pdf}
+  \includegraphics[width=8.0cm]{./Figs/HGorderparametersPG.pdf}
   \caption{\label{HGorderParametersPOPG}
@@
     and {\it Escherichia coli} PG from Ref.~\citenum{gally81}).
-    ///
     Experimental data for DPPG had been measured with 0.1\,M NaCl~\cite{wohlgemuth80}, while all of the other data are with
@@
     $^2$ NMR DPPG and POPG in 10\,mM PIPES buffer.
-    ///
     The light blue areas span 0.04 units around the average of the extremal experimental values,
@@
     from experiments at 298 K \cite{scherer87,macdonald87} and simulations with different force fields
-    (temperatures listed in tables \ref{systemsMIX} and \ref{systemsMIX2} are between 298-310 K).
-    Signs are determined as discussed in Refs. \citenum{botan15,ollila16}.
+    (temperatures listed in tables \ref{systemsMIX} are between 298-310 K).
+    Signs are determined as discussed in Refs. \citenum{botan15} and~\citenum{ollila16}.
   }

### 2021-05-07 manuscriptPGPEsuppl.tex

@@
 
-\title{ Supporting Information:\\Conformational plasticity of phospholipid headgroups in simulations and experiments}
+\title{ Supporting Information:\\Conformational plasticity of lipid headgroups in cellular membranes and protein--lipid complexes}
 
@@
 This approach has been tested against trajectories with explicit hydrogens and the deviations in order parameters are small \cite{buildHvalidation,piggot17}.\\
-%\todo{BuildH program is now cited with a direct link to the GitHub repo. I think that a release to Zenodo would be nice in the final publication.}\\
 
@@
 \noindent
-{\it Am{\'e}lie Bacle} set up, performed and analysed POPC and POPC:POPE (1:1) simulations with the Berger Force Field.
+{\it Am{\'e}lie Bacle} set up, performed and analysed POPC and POPC:POPE (1:1) simulations with the Berger force field.
 %% Please write description of your contribution here
@@
 %% Please write description of your contribution here
-performed the analysis of dihedrals and isomers of lipids. Analysed lipid structures from Protein Data Bank. Prepared panels for figures 2 and 4. Participated in discussions.
+contributed to the analysis of dihedrals and isomers of lipids from MD simulations, and to the analysis of lipid structures from the Protein Data Bank. Prepared panels for figures 2 and 4.
 
@@
 {\it Rebeca Garc{\'i}a Fandi{\~n}o}
-designed and supervised the molecular dynamics simulations carried out by Antonio Pe{\'o}n, and contributed to some discussions.
+designed and supervised the molecular dynamics simulations carried out by Antonio Pe{\'o}n.
 %% Please write description of your contribution here
@@
 {\it Tiago M. Ferreira}
-was responsible for the solid-state NMR experiments/figures and took part in writing the manuscript.
-
-\noindent
-{\it Patrick F.J. Fuchs} supervised Paula Mil{\'a}n Rodr{\'i}guez, Am{\'e}lie Bacle and Chris Papadopoulos, created the buildH software, contributed to many discussions.
-%% Please write description of your contribution here
-
-\noindent
-{\it Ivan Gushchin} supervised Pavel Buslaev and contributed to the analysis of lipid structures in the Protein Data Bank.
-%% Please write description of your contribution here
-
-\noindent
-{\it Matti Javanainen} set up and performed simulations using CHARMM36, Slipids, and MacRog lipid models. He contributed to the organization of the manuscript.
-%% Please write description of your contribution here
-
-\noindent
-{\it Anne M. Kiirikki} set up and performed POPC:POPG(1:1) and POPC:POPG(4:1) simulations with CHARMM36 and POPC:POPG(4:1) with Lipid17ecc. She also contributed to the NMRlipids databank.
+was responsible for the solid-state NMR experiments. Prepared figures~\ref{HGorderParameters},~\ref{POPEspectra}, and~\ref{POPGspectra}, and took part in writing the manuscript.
+
+\noindent
+{\it Patrick F.J.
+Fuchs} supervised Paula Mil{\'a}n Rodr{\'i}guez, Am{\'e}lie Bacle and Chris Papadopoulos, created the buildH software and contributed to the calculation of lipid isomers.
+Provided comments on the manuscript.
+%% Please write description of your contribution here
+
+\noindent
+{\it Ivan Gushchin} supervised Pavel Buslaev and contributed to the analysis of lipid structures in the Protein Data Bank. Provided comments on the manuscript.
+%% Please write description of your contribution here
+
+\noindent
+{\it Matti Javanainen} set up and performed simulations using CHARMM36, Slipids, and MacRog lipid models. Provided comments on the manuscript.
+%% Please write description of your contribution here
+
+\noindent
+{\it Anne M.
+Kiirikki} developed the preliminary version of the NMRlipids databank, which was used in several analyses.
+Set up and performed POPC:POPG(1:1) and POPC:POPG(4:1) simulations with CHARMM36 and POPC:POPG(4:1) with Lipid17ecc.
 %% Please write description of your contribution here
@@
 {\it Josef Melcr}
-created the analysis script \texttt{scripts/calcOrderParameters.py}, 
-prepared the Lipid17ecc topology, 
-and contributed to discussions and manuscript preparation.
-
-\noindent
-{\it Paula Mil{\'a}n Rodr{\'i}guez} set up, performed and analysed POPC and POPC:POPE (1:1) simulations with the MacRog Force Field.
-She also fixed the initial force field (itp) files prior to the simulations.
+%created the analysis script \texttt{scripts/calcOrderParameters.py}, 
+contributed to the preparation of the Lipid17ecc parameters. 
+Provided comments on the manuscript.
+
+\noindent
+{\it Paula Mil{\'a}n Rodr{\'i}guez} corrected errors in available MacRog force field parameters. Set up, performed and analysed POPC and POPC:POPE (1:1) simulations with the MacRog force field.
 %% Please write description of your contribution here
@@
 {\it Markus S. Miettinen}
+made figures~\ref{HGorderParametersPE} and~\ref{HGorderParametersPOPG}. Provided comments on the manuscript.
 %% Please write description of your contribution here
@@
 {\it {\'A}ngel Pi{\~n}eiro}
-Created the opAAUA\_prod.py code for the calculation of order parameters. Contributed to some discussions. 
+created the opAAUA\_prod.py code for the calculation of order parameters from united atom trajectories.
 %% Please write description of your contribution here
@@
 {\it Salla I. Virtanen}
+prepared Lipid17ecc parameters.
 %% Please write description of your contribution here

### 2021-05-07 CoverLetter.tex
Updates from Overleaf

@@
 \pagestyle{empty}
-\noindent O. H. Samuli Ollila \\
+\noindent Dr. O. H. Samuli Ollila \\
 \noindent Institute of Biotechnology \\
@@
 
-Please find enclosed our manuscript entitled {\it Conformational plasticity of lipid headgroups in cellular membranes and protein-lipid complexes}
+Please find enclosed our manuscript entitled {\it Conformational plasticity of lipid headgroups in cellular membranes and protein--lipid complexes}
 by A. Bacle et al. which we would like to submit for consideration to be published in \textit{ACS Central Science}.
@@
 which combines data from molecular dynamics (MD) simulations and solid-state NMR experiments.
-Using the extensive data collected within this project, we have now resolved the differences in conformational ensembles
-between the most common lipids in eukaryotes and bacteria. 
+The extensive data collected within this project has now enabled us to resolve the differences in conformational ensembles between the most common lipid types in eukaryotes and bacteria. 
 Our results reveal a significant structural plasticity in lipid headgroups and only minor differences of the conformational ensembles
 between lipids with different headgroup chemistry.
-Notably, and motivated by our findings on the model systems, by using a bioinformatical analysis of protein-bound lipid structures from the protein data bank (PDB), we found that the plasticity in lipid headgroups is a characteristic feature in lipid-protein complexes.
+Motivated by our findings on the model systems, we used a novel bioinformatical analysis of protein-bound lipid structures from the protein data bank (PDB), and found that the plasticity in lipid headgroups is a characteristic feature also in lipid-protein complexes.
 
@@
 and the selectivity of lipid-protein binding is driven by intermolecular interactions rather than by restrictions in lipid conformations.
-Furthermore, we demonstrated the power of a conceptually novel open collaboration approach that extended the realm of Molecular Dynamics simulations to complement the experimental data in the PDB for elucidating how complex systems made up of disordered biomolecules behave.
+Furthermore, we demonstrated the power of a conceptually novel open collaboration approach that extended the realm of MD simulations to complement the experimental data in the PDB for elucidating how complex systems made up of disordered biomolecules behave.
 
-We believe that our approach and findings are of great interest to understand the connection of lipid chemistry to protein-lipid interactions  and ultimately the role of lipid chemistry in a number of diseases.
+We believe that our approach and findings are of great interest to understand the connection of lipid chemistry to protein-lipid interactions, and ultimately the role of lipid chemistry in a number of diseases.
+Moreover, the open access policy of {\it ACS Central Science} is optimal for the publications from the NMRlipids Project which fosters the open collaboration and open data approaches.
 Therefore, we believe that \textit{ACS Central Science} is a highly suitable journal to publish this work.
@@
 
-O. H. Samuli Ollila
+Dr. O. H. Samuli Ollila
 

### 2021-05-07 manuscriptPGPE.tex
Updates from Overleaf

@@
 
-Structures of protein-bound lipids are available in the Protein Data Bank (PDB) \cite{berman00},
+Structures of protein-bound lipids are available in the Protein Data Bank (PDB, \url{http://www.rcsb.org/}) \cite{berman00},
 and crystal structures of lipids have been determined \cite{buldt81,pascher92},
@@
 \ref{HGorderParameters} C.
-Relative energy costs for turning dihedral angles with respect to the most probable value (lowest energy) were estimated from the inverse Boltzmann formula $\Delta E(\theta) = -kT \left[\ln\left[p(\theta)\right]-\ln\left[p(\theta_0)\right] \right]$, where $p(\theta)$ is the dihedral angle distribution and $\theta_0$ is the most probable angle from MD simulation.
+Relative free energy costs for turning dihedral angles with respect to the most probable value (lowest free energy) were estimated from the inverse Boltzmann formula $\Delta E(\theta) = -kT \left[\ln\left[p(\theta)\right]-\ln\left[p(\theta_0)\right] \right]$, where $p(\theta)$ is the dihedral angle distribution and $\theta_0$ is the most probable angle from MD simulation.
 
@@
      The best simulation model (CHARMM36) was selected from the comparison of ten different force fields based on quality evaluation against experimental $S_{\rm{CH}}$ values in Figs.~\ref{HGorderParametersPE} and~\ref{HGorderParametersPOPG}.
-     %For a detailed comparison of all the XX force-fields used with the $S_{\rm{CH}}$ values measured please read the supplementary information.
      \textbf{A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurrence of different conformations.
      \textbf{B)} Headgroup and glycerol backbone region order parameters of different lipids.
-     \textbf{C)} Relative energies for individual heavy atom dihedral angles estimated from the inverse Boltzmann formula.
-     Angles corresponding to energies above 8~$k_\mathrm{B}T$ are not shown because they are not observed in simulations.
+     \textbf{C)} Relative free energies for individual heavy atom dihedral angles estimated from the inverse Boltzmann formula.
+     Angles corresponding to free energies above 8~$k_\mathrm{B}T$ are not shown because they are not observed in simulations.
      \textbf{D)} Distributions of P--N/C$_\gamma$ vector angle with respect to membrane normal.
@@
     from CHARMM36 simulations compared with the experimental data \cite{scherer89}.
-    \textbf{B)} Relative energies for individual dihedral angles estimated from inverse Boltzmann distributions of heavy atom dihedral angles
+    \textbf{B)} Relative free energies for individual dihedral angles estimated from inverse Boltzmann distributions of heavy atom dihedral angles
     with different amounts of cationic surfactant from CHARMM36 simulations.
@@
 we first calculated the heavy atom dihedral angle distributions from the best performing simulations that reproduced the differences between headgroups~(Fig. \ref{DIHdists}).
-Then, we used the inverse Boltzmann formula to estimate the
+Then, we used the inverse Boltzmann formula to estimate the free
 energy costs for different dihedral angle orientations.
-The results in Fig.~\ref{structures}C suggest that all lipid headgroups are very flexible and
+The results in Fig.~\ref{structures}C suggest that all lipid headgroups are very flexible and free
 energy costs for rotating individual dihedrals to almost any angle is low (below 8~$k_\mathrm{B}T$).
-Only $cis$ states of P-O$_\alpha$-C$_\alpha$-C$_\beta$ and g$_2$-g$_3$-O$_{g_3}$-P have larger relative
+Only $cis$ states of P-O$_\alpha$-C$_\alpha$-C$_\beta$ and g$_2$-g$_3$-O$_{g_3}$-P have larger relative free
 energies and are not observed for any lipids during the simulations.
@@
 whereas PC and PE exhibit symmetric distributions.
-Also, the energy barriers for O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ dihedral
+Also, the free energy barriers for O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ dihedral
 rotations between \textit{gauche} and \textit{trans} states are larger for
@@
 The rest of the dihedrals are similar between different lipids,
-with the exception of PS lipids for which slightly larger energy for \textit{eclipsed anti} conformation
+with the exception of PS lipids for which slightly larger free energy for \textit{eclipsed anti} conformation
 in g$_3$-O$_{{\mathrm g}3}$-P-O$_\alpha$ dihedral was observed.
@@
 In conclusion, all studied lipid headgroups sample very broad conformational ensembles in the liquid lamellar phase and the sampled dihedral angles are within similar ranges for all headgroup types.
-Since the rotation of dihedral angles to almost any position bears relatively low energy cost, the lipid headgroups are able to adopt a wide range of conformations when interacting with proteins, ions, or other biomolecules.
+Since the rotation of dihedral angles to almost any position bears relatively low free energy cost, the lipid headgroups are able to adopt a wide range of conformations when interacting with proteins, ions, or other biomolecules.
 Therefore, the structures in lipid crystals \cite{buldt81,pascher92} are not representative of the liquid state, and models aiming to explain NMR data using only a few conformations \cite{seelig77c,davis83,Semchyschyn04,akutsu20} are not sufficient to capture the large conformational space of lipids in the liquid lamellar state.
@@
 that correctly captured the experimentally measured decrease in PC headgroup order parameters upon addition of cationic surfactants into a bilayer (figure \ref{changesWITHsurf} A).
-The dihedral angle distributions and relative energies
+The dihedral angle distributions and relative free energies
 in figures \ref{HGorderparametersPCvsSURFdihedrals} and \ref{changesWITHsurf} B reveal that the

### 2021-05-10 manuscriptPGPE.tex
Add submitted version

@@
-%% ****** Start of file aiptemplate.tex ****** %
-%%
-%%   This file is part of the files in the distribution of AIP substyles for REVTeX4.
-%%   Version 4.1 of 9 October 2009.
-%%
-%
-% This is a template for producing documents for use with 
-% the REVTEX 4.1 document class and the AIP substyles.
-% 
-% Copy this file to another name and then work on that file.
-% That way, you always have this original template file to use.
+\documentclass[journal=jpcbfk,manuscript=article]{achemso}
 
-%\documentclass[aip,graphicx]{revtex4-1}
-%\documentclass[aip,reprint]{revtex4-1}
-
-%\usepackage{graphicx}
-
-%\draft % marks overfull lines with a black rule on the right
-%\documentclass[pre,aps,floatfix,authordate1-4,twocolumn]{revtex4-1}
-%\documentclass[pre,aps,floatfix,authordate1-4]{revtex4-1}
-
-\documentclass[aps,prl,superscriptaddress,twocolumn]{revtex4}
-
-%\documentclass[aps,prl,preprint,groupedaddress]{revtex4}
-
+\usepackage[version=3]{mhchem}
+\usepackage[T1]{fontenc}
+\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
+%\newcommand{\todo}[1]{\textcolor{red}{#1}}
+\usepackage[obeyFinal]{easy-todo}
 \usepackage{rotating}
-\usepackage{times}
-\usepackage{graphicx}
-\usepackage{setspace}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{epstopdf}
-\usepackage[obeyFinal]{easy-todo}
-\usepackage{csquotes}
+\usepackage{upgreek}				
+\usepackage{xcolor}
+\usepackage{booktabs}
+\usepackage{multirow}
+\usepackage{lmodern}
+\usepackage{microtype}
 \usepackage{xr-hyper}
+\usepackage{soul} % for highlights with \hl{} 
 \usepackage{hyperref} % to make TOC titles clickable
@@
 
-\begin{document}
-
-% Use the \preprint command to place your local institutional report number 
-% on the title page in preprint mode.
-% Multiple \preprint commands are allowed.
-%\preprint{}
-
-\title{Conformational plasticity of lipid headgroups in cellular membranes and protein--lipid complexes} %Title of paper
-
-% repeat the \author .. \affiliation  etc. as needed
-% \email, \thanks, \homepage, \altaffiliation all apply to the current author.
-% Explanatory text should go in the []'s, 
-% actual e-mail address or url should go in the {}'s for \email and \homepage.
-% Please use the appropriate macro for the type of information
-
-% \affiliation command applies to all authors since the last \affiliation command. 
-% The \affiliation command should follow the other information.
-
 \author{Am{\'e}lie Bacle}
-\affiliation{Laboratoire Coop{\'e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit{\'e} de Poitiers, 1 rue Georges Bonnet, 86000 Poitiers, France }
+\affiliation{\tiny Laboratoire Coop{\'e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit{\'e} de Poitiers, 1 rue Georges Bonnet, 86000 Poitiers, France }
 
@@
 \author{O. H. Samuli Ollila}
-\email[]{samuli.ollila@helsinki.fi}
+\email{samuli.ollila@helsinki.fi}
 \affiliation{Institute of Biotechnology, University of Helsinki}
@@
 
-% Collaboration name, if desired (requires use of superscriptaddress option in \documentclass). 
-% \noaffiliation is required (may also be used with the \author command).
-%\collaboration{}
-%\noaffiliation
+\title{Conformational plasticity of lipid headgroups in cellular membranes and protein--lipid complexes} %Title of paper
 
-\date{\today}
+\begin{document}
 
@@
 
-%\pacs{}% insert suggested PACS numbers in braces on next line
-
 \maketitle
 
-% Body of paper goes here. Use proper sectioning commands. 
-% References should be done using the \cite, \ref, and \label commands
-
-%\label{}
 \section{Introduction}
@@
 
-Here, we use natural abundance $^1$H-$^{13}$C solid-state NMR experiments %(enabling the determination of order parameter signs)
-and MD simulations from the NMRlipids Project
+Here, we use natural abundance $^1$H-$^{13}$C solid-state NMR experiments and MD simulations from the NMRlipids Project
 to resolve differences in the conformational ensembles of PC, PE, PG, and PS lipid headgroups.
@@
 
-\section{Methods}
-\subsection{Experimental C--H bond order parameters}
-The headgroup and glycerol backbone C--H bond order parameters of 1-palmitoyl-2-oleoyl-sn-glycero-3-phosphoethanolamine (POPE) and 1-palmitoyl-2-oleoyl-sn-glycero-3-phospho-(1'-rac-glycerol) (POPG), purchased from Avanti polar lipids, were measured using natural abundance $^1$H-$^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
-The samples were prepared by simply mixing the lipids with water until a homogeneous preparation of multi-lamellar vesicles (MLVs) was attained and left to equilibrate for approximately 24 hours before measurements.
-The magnitudes of the order parameters were determined using a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04}, and the order parameter signs from S-DROSS experiments~\cite{gross97} combined with SIMPSON simulations \cite{bak00}.
-%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
-%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
-The set-up of the solid-state NMR experiments were identical as in our previous work~\cite{antila19}.
-The POPE experiments were recorded at 310~K and POPG experiments at 298~K, where the bilayers are in the liquid disordered phase \cite{marsh13}.
+\section{Results and Discussion}
 
-Glycerol backbone peaks from both lipids, and the $\alpha$-carbon peak from POPE in the INEPT spectra 
-were assigned based on previously measured POPC spectra~\cite{ferreira13}.
-The $\beta$-carbon peak from POPE was assigned based on $^{13}$C chemical shift table for amines available
-at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm}.
-The asignment of the $\alpha$, $\beta$, and $\gamma$ peaks of POPG was based on the measured R-PDLF dipolar couplings in comparison to the known order parameters from previous $^2$H\,NMR experiments~\cite{borle85,wohlgemuth80}.
-The $\beta$-carbon peak from POPG overlapped with the g$_2$ peak from glycerol backbone
-because their chemical environments are similar.
+\subsection{Differences between lipid headgroups from $^{13}$C NMR experiments}
 
-\subsection{Molecular dynamics simulations}
-
-Molecular dynamics simulation data were collected and analyzed using
-the methods from the NMRlipids Open Collaboration project (\url{nmrlipids.blogspot.fi}) \cite{botan15,catte16,ollila16,antila19}.
-%\url{github.com/NMRlipids/NMRlipidsIVotherHGs}. 
-Simulation details, accessibility information and quality evaluation of approximately 70 combinations of lipid headgroup and force field simulated for this work are in the supplementary information.
-
-Best models for the interpretation of lipid headgroup conformational ensembles from the experimental data were selected
-using quality evaluation measures defined in the NMRlipids project.
-Conformational ensembles of headgroup and glycerol backbone in PE and PG simulations were evaluated using the C--H bond order parameters \cite{botan15}.
-Interactions between different headgroups were evaluated by monitoring the changes in headgroup order parameters upon mixing the lipids \cite{antila19}.
-The ion binding affinities and response of lipids to bound charge were evaluated by monitoring the changes in lipid headgroup order parameters \cite{catte16,antila19}.
-
-The differences in headgroup conformation ensembles were analyzed calculating the distributions of six dihedral angles, O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$, P-O$_\alpha$-C$_\alpha$-C$_\beta$, O$_{g3}$-P-O$_\alpha$-C$_\alpha$, g$_3$-O$_{g3}$-P-O$_\alpha$, g$_2$-g$_3$-O$_{g3}$-P, and g$_1$-g$_2$-g$_3$-O$_{g3}$, labeled in Fig.
-\ref{HGorderParameters} C.
-Relative free energy costs for turning dihedral angles with respect to the most probable value (lowest free energy) were estimated from the inverse Boltzmann formula $\Delta E(\theta) = -kT \left[\ln\left[p(\theta)\right]-\ln\left[p(\theta_0)\right] \right]$, where $p(\theta)$ is the dihedral angle distribution and $\theta_0$ is the most probable angle from MD simulation.
-
-%\clearpage
 \begin{figure*}[]
   \centering
-%  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
-  %  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCPSPEPG.eps}
-   \includegraphics[width=18.0cm]{./Figs/figure_PG_PE.pdf}
+   \includegraphics[width=\textwidth]{./Figs/figure_PG_PE.pdf}
    \caption{\label{HGorderParameters}
@@
    }
-   %\todo{This is a sketch, Tiago Ferreira will make a new figure.} \\
-%  \todo{D) could be clarified as Fig. 2 in the NMRlipids IVps paper.}
 \end{figure*}
-
-\subsection{Analysis of protein-bound lipid conformations}
-Lipid structures from the Protein Data Bank (PDB, \cite{berman2000})
-were searched using PDBe REST API (\url{www.ebi.ac.uk/pdbe/pdbe-rest-api})
-using the ligand names listed in the supplementary information.
-Dihedral distributions from the lipid structures were calculated
-using the MDAnalysis Python library \cite{agrawal11,gowers16} and
-Jupyter notebook available from {\it scripts/PDBanalysis.ipynb} folder in Ref~\citenum{NMRlipidsIVbgit}.
-Only the structures determined using X-ray crystallography or Cryo-EM with a resolution higher than 3.2~\text{\AA} were used in the analysis.
-Some structures of multimeric proteins contained multiple lipids in the same conformation.
-The conformations were considered identical if all dihedral angles were the same within 3 degrees.
-In these cases, this conformation was counted only once for the subsequent analyses
-to avoid its overweighting.
-
-To demonstrate the existence of similar structures of different lipids bound to different proteins, we searched for pairs having the last five heavy atom dihedrals angles in the headgroup end (excluding g$_1$-g$_2$-g$_3$-O$_{g3}$ from the list in previous section) within
-30 degrees of each other among the structures with resolution higher than 2.5~\text{\AA}. The two most representative examples
-were handpicked from the results.
-
-\section{Results and Discussion}
-
-\begin{figure*}[bt]
-  \centering
-   \includegraphics[width=18.0cm]{./Figs/figure2.eps}
-   \caption{\label{structures}
-     Results from the best simulation model demonstrating the differences in conformational ensembles between different lipids.
-     The best simulation model (CHARMM36) was selected from the comparison of ten different force fields based on quality evaluation against experimental $S_{\rm{CH}}$ values in Figs.~\ref{HGorderParametersPE} and~\ref{HGorderParametersPOPG}.
-     \textbf{A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurrence of different conformations.
-     \textbf{B)} Headgroup and glycerol backbone region order parameters of different lipids.
-     \textbf{C)} Relative free energies for individual heavy atom dihedral angles estimated from the inverse Boltzmann formula.
-     Angles corresponding to free energies above 8~$k_\mathrm{B}T$ are not shown because they are not observed in simulations.
-     \textbf{D)} Distributions of P--N/C$_\gamma$ vector angle with respect to membrane normal.
-  }
-\end{figure*}
-
-\subsection{Differences between lipid headgroups from $^{13}$C NMR experiments}
 
@@
 
-\begin{figure*}[bt]
+\begin{figure*}
   \centering
-  \includegraphics[width=18.0cm]{./Figs/HGorderparametersPCvsSURFchangeDIHEDRALSlog.eps}
-  \caption{\label{changesWITHsurf}
-    \textbf{A)} Modulation of PC headgroup order parameters and P--N vector tilt (angle between the P-N vector and the bilayer normal) upon addition of cationic surfactant (dihexadecyldimethylammonium)
-    from CHARMM36 simulations compared with the experimental data \cite{scherer89}.
-    \textbf{B)} Relative free energies for individual dihedral angles estimated from inverse Boltzmann distributions of heavy atom dihedral angles
-    with different amounts of cationic surfactant from CHARMM36 simulations.
+   \includegraphics[width=\textwidth]{./Figs/figure2.eps}
+   \caption{\label{structures}
+     Results from the best simulation model demonstrating the differences in conformational ensembles between different lipids.
+     The best simulation model (CHARMM36) was selected from the comparison of ten different force fields based on quality evaluation against experimental $S_{\rm{CH}}$ values in Figs.~\ref{HGorderParametersPE} and~\ref{HGorderParametersPOPG}.
+     \textbf{A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurrence of different conformations.
+     \textbf{B)} Headgroup and glycerol backbone region order parameters of different lipids.
+     \textbf{C)} Relative free energies for individual heavy atom dihedral angles estimated from the inverse Boltzmann formula.
+     Angles corresponding to free energies above 8~$k_\mathrm{B}T$ are not shown because they are not observed in simulations.
+     \textbf{D)} Distributions of P--N/C$_\gamma$ vector angle with respect to membrane normal.
   }
@@
 
+\begin{figure*}[bt]
+  \centering
+  \includegraphics[width=\textwidth]{./Figs/HGorderparametersPCvsSURFchangeDIHEDRALSlog.eps}
+  \caption{\label{changesWITHsurf}
+    \textbf{A)} Modulation of PC headgroup order parameters and P--N vector tilt (angle between the P-N vector and the bilayer normal) upon addition of cationic surfactant (dihexadecyldimethylammonium)
+    from CHARMM36 simulations compared with the experimental data \cite{scherer89}.
+    \textbf{B)} Relative free energies for individual dihedral angles estimated from inverse Boltzmann distributions of heavy atom dihedral angles
+    with different amounts of cationic surfactant from CHARMM36 simulations.
+  }
+\end{figure*}
+
 Charged entities, such as lipids, proteins, surfactants, drugs, and ions incorporated in membranes can change the orientation of the headgroup dipole in PC lipids and affect the order parameters of the lipid headgroups~\cite{seelig87}.
@@
 we calculated the heavy atom dihedral angle distributions from 
-the subset of simulations %(CHARMM36)
+the subset of simulations 
 that correctly captured the experimentally measured decrease in PC headgroup order parameters upon addition of cationic surfactants into a bilayer (figure \ref{changesWITHsurf} A).
@@
 
-\begin{figure*}[]
+\begin{figure*}[tb]
   \centering
-  \includegraphics[width=19.0cm]{./Figs/figure4.eps}
+  \includegraphics[width=\textwidth]{./Figs/figure4.eps}
   \caption{\label{dihedralsFROMpdb}
@@
 
-%However, it is important to note that lipids are often not the main target in
-%structures deposited in PDB and therefore their conformations may be less reliable than those of proteins.
-
 \section{Conclusions}
 
-C--H bond order parameters $S_\mathrm{CH}$ from NMR experiments %in this and previous studies \cite{wohlgemuth80,buldt81,seelig87}
+C--H bond order parameters $S_\mathrm{CH}$ from NMR experiments 
 suggest that lipid headgroup conformational ensembles depend on lipid type (PC, PE, PG, or PS) 
@@
 
-% If you have acknowledgments, this puts in the proper section head.
-\begin{acknowledgments}
+\section{Methods}
+\subsection{Experimental C--H bond order parameters}
+The headgroup and glycerol backbone C--H bond order parameters of 1-palmitoyl-2-oleoyl-sn-glycero-3-phosphoethanolamine (POPE) and 1-palmitoyl-2-oleoyl-sn-glycero-3-phospho-(1'-rac-glycerol) (POPG), purchased from Avanti polar lipids, were measured using natural abundance $^1$H-$^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
+The samples were prepared by simply mixing the lipids with water until a homogeneous preparation of multi-lamellar vesicles (MLVs) was attained and left to equilibrate for approximately 24 hours before measurements.
+The magnitudes of the order parameters were determined using a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04}, and the order parameter signs from S-DROSS experiments~\cite{gross97} combined with SIMPSON simulations \cite{bak00}.
+%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
+%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
+The set-up of the solid-state NMR experiments were identical as in our previous work~\cite{antila19}.
+The POPE experiments were recorded at 310~K and POPG experiments at 298~K, where the bilayers are in the liquid disordered phase \cite{marsh13}.
+
+Glycerol backbone peaks from both lipids, and the $\alpha$-carbon peak from POPE in the INEPT spectra 
+were assigned based on previously measured POPC spectra~\cite{ferreira13}.
+The $\beta$-carbon peak from POPE was assigned based on $^{13}$C chemical shift table for amines available
+at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm}.
+The asignment of the $\alpha$, $\beta$, and $\gamma$ peaks of POPG was based on the measured R-PDLF dipolar couplings in comparison to the known order parameters from previous $^2$H\,NMR experiments~\cite{borle85,wohlgemuth80}.
+The $\beta$-carbon peak from POPG overlapped with the g$_2$ peak from glycerol backbone
+because their chemical environments are similar.
+
+\subsection{Molecular dynamics simulations}
+
+Molecular dynamics simulation data were collected and analyzed using
+the methods from the NMRlipids Open Collaboration project (\url{nmrlipids.blogspot.fi}) \cite{botan15,catte16,ollila16,antila19}.
+Simulation details, accessibility information and quality evaluation of approximately 70 combinations of lipid headgroup and force field simulated for this work are in the supplementary information.
+
+Best models for the interpretation of lipid headgroup conformational ensembles from the experimental data were selected
+using quality evaluation measures defined in the NMRlipids project.
+Conformational ensembles of headgroup and glycerol backbone in PE and PG simulations were evaluated using the C--H bond order parameters \cite{botan15}.
+Interactions between different headgroups were evaluated by monitoring the changes in headgroup order parameters upon mixing the lipids \cite{antila19}.
+The ion binding affinities and response of lipids to bound charge were evaluated by monitoring the changes in lipid headgroup order parameters \cite{catte16,antila19}.
+
+The differences in headgroup conformation ensembles were analyzed calculating the distributions of six dihedral angles, O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$, P-O$_\alpha$-C$_\alpha$-C$_\beta$, O$_{g3}$-P-O$_\alpha$-C$_\alpha$, g$_3$-O$_{g3}$-P-O$_\alpha$, g$_2$-g$_3$-O$_{g3}$-P, and g$_1$-g$_2$-g$_3$-O$_{g3}$, labeled in Fig.
+\ref{HGorderParameters} C.
+Relative free energy costs for turning dihedral angles with respect to the most probable value (lowest free energy) were estimated from the inverse Boltzmann formula $\Delta E(\theta) = -kT \left[\ln\left[p(\theta)\right]-\ln\left[p(\theta_0)\right] \right]$, where $p(\theta)$ is the dihedral angle distribution and $\theta_0$ is the most probable angle from MD simulation.
+
+\subsection{Analysis of protein-bound lipid conformations}
+Lipid structures from the Protein Data Bank (PDB, \cite{berman00})
+were searched using PDBe REST API (\url{www.ebi.ac.uk/pdbe/pdbe-rest-api})
+using the ligand names listed in the supplementary information.
+Dihedral distributions from the lipid structures were calculated
+using the MDAnalysis Python library \cite{agrawal11,gowers16} and
+Jupyter notebook available from {\it scripts/PDBanalysis.ipynb} folder in Ref~\citenum{NMRlipidsIVbgit}.
+Only the structures determined using X-ray crystallography or Cryo-EM with a resolution higher than 3.2~\text{\AA} were used in the analysis.
+Some structures of multimeric proteins contained multiple lipids in the same conformation.
+The conformations were considered identical if all dihedral angles were the same within 3 degrees.
+In these cases, this conformation was counted only once for the subsequent analyses
+to avoid its overweighting.
+
+To demonstrate the existence of similar structures of different lipids bound to different proteins, we searched for pairs having the last five heavy atom dihedrals angles in the headgroup end (excluding g$_1$-g$_2$-g$_3$-O$_{g3}$ from the list in previous section) within
+30 degrees of each other among the structures with resolution higher than 2.5~\text{\AA}. The two most representative examples
+were handpicked from the results.
+
+\begin{acknowledgement}
+
   % Put your acknowledgments here.
 A.P. is grateful to the Centro de
-Supercomputación de Galicia (CESGA) for use of the Finis
+Supercomputaci{\'o}n de Galicia (CESGA) for use of the Finis
 Terrae computer.
-
 M.J. thanks CSC -- IT Center for Science for computational resources and the Emil Aaltonen foundation for financial support.
-
 T.M.F. was supported by the Ministry of Economics, Science
 and Digitalisation of the State of Saxony-Anhalt, Germany.
-
 P.B. was supported by the Academy of Finland (Grant 311031).
-
 F.F.-R.
 acknowledges Tecnol\'{o}gico Nacional de M\'{e}xico Proyecto IT16C431, Direcci\'{o}n General de Asuntos del Personal Acad\'{e}mico (DGAPA) Programa de Apoyo a Proyectos de Investigaci\'{o}n e Innovaci\'{o}n Tecnol\'{o}gica (PAPIIT) IG100920, CONACyT Ciencia de Frontera 74884 for financial support and Miztli-Direcci\'{o}n de C\'{o}mputo y de Tecnolog\'{i}as de Informaci\'{o}n y Comunicaci\'{o}n (DGTIC) - Universidad Nacional Aut\'{o}noma de M{\'e}xico (UNAM) (Project LANCAD-UNAM-DGTIC-057) facilities for computing-time allocation.
-
 I.G. was supported by the Ministry of Science and Higher Education of the Russian Federation (agreement \#075-00337-20-03, project FSMG-2020-0003). 
-
 J.J.M. gratefully acknowledges financial support from the Carlsberg 
@@
 Research Computing Center at the University of Chicago.
-
 O.H.S.O, A.M.K, and S.I.V acknowledge CSC -- IT Center for Science for computational resources and Academy of Finland (grants 315596 and 319902) for financial support.
-
 T.J.P. acknowledges use of the Iridis high-performance computing resources at the University of Southampton.
-
 J.M. thanks the Center for Information Technology of the University of Groningen for their support and for providing access to the Peregrine high performance computing cluster.
 
-\end{acknowledgments}
+\end{acknowledgement}
 
@@
 
-%\newpage
-%\section{APPENDIX: The NMR results reported by Tiago Ferreira}
-
-%\listoftodos
-
 \end{document}
-%
-% ****** End of file aiptemplate.tex ******

### 2021-05-10 manuscriptPGPEsuppl.tex
Add submitted version

@@
 \noindent
-{\it Patrick F.J.
-Fuchs} supervised Paula Mil{\'a}n Rodr{\'i}guez, Am{\'e}lie Bacle and Chris Papadopoulos, created the buildH software and contributed to the calculation of lipid isomers.
-Provided comments on the manuscript.
+{\it Patrick F.J. Fuchs} supervised Paula Mil{\'a}n Rodr{\'i}guez, Am{\'e}lie Bacle and Chris Papadopoulos,
+created the buildH software for the calculations of order parameters
+from united atom simulations,
+and contributed to the calculation of lipid isomers. Provided comments on the manuscript.
 %% Please write description of your contribution here

### 2021-05-18 CoverLetter.tex
Update cover letter

@@
 Please find enclosed our manuscript entitled {\it Conformational plasticity of lipid headgroups in cellular membranes and protein--lipid complexes}
-by A. Bacle et al. which we would like to submit for consideration to be published in \textit{ACS Central Science}.
+by A. Bacle et al. which we would like to submit for consideration to be published in \textit{Chemical Science}.
 
 In the manuscript, we demonstrate that lipids in cellular membranes exhibit a much higher plasticity in terms of allowed conformational states than previously anticipated.
-Our results update the current understanding on how lipids selectively bind to proteins.
+Our results update the current understanding on how lipids selectively bind to proteins
+and present a significant contribution to the emerging field studying biomolecular complexes containing disordered molecules.
 
@@
 In 2013 we set out to solve the conformational ensembles of lipid headgroups by establishing the NMRlipids open collaboration project
-which combines data from molecular dynamics (MD) simulations and solid-state NMR experiments.
+which combines data from molecular dynamics (MD) simulations and solid-state NMR experiments (previous publications from the project have gained 209 citations in total).
 The extensive data collected within this project has now enabled us to resolve the differences in conformational ensembles between the most common lipid types in eukaryotes and bacteria. 
@@
 We believe that our approach and findings are of great interest to understand the connection of lipid chemistry to protein-lipid interactions, and ultimately the role of lipid chemistry in a number of diseases.
-Moreover, the open access policy of {\it ACS Central Science} is optimal for the publications from the NMRlipids Project which fosters the open collaboration and open data approaches.
-Therefore, we believe that \textit{ACS Central Science} is a highly suitable journal to publish this work.
+Moreover, the open access policy of {\it Chemical Science} is optimal for the publications from the NMRlipids Project which fosters the open collaboration and open data approaches as all the data and other material are publicly available throughout the project.
+Therefore, we believe that \textit{Chemical Science} is a highly suitable journal to publish this work.
 

### 2021-05-27 manuscriptPGPE.tex
Update title and abstract

@@
 
-\title{Conformational plasticity of lipid headgroups in cellular membranes and protein--lipid complexes} %Title of paper
+\title{Inverse conformational selection in lipid-protein binding} %Title of paper
 
@@
 % insert abstract here
-  The chemistry of lipid headgroups, the water-facing interface of cellular membranes, regulates cell functions via lipid--protein interactions. In the physiologically relevant liquid state,
-  lipid molecules are not fixed to a single structure but sample an ensemble of conformations with Boltzmann-weighted probabilities.
-  However, the distributions of conformations in such ensembles and their lipid-type dependence have not yet been experimentally determined: 
+  Interest in lipid interactions with proteins and other biomolecules is emerging not only in
+  fundamental biochemistry, but also in the field of nanobiotechnology where lipid nanoparticles
+  are commonly used, for example, as carriers in mRNA vaccines.
+  The outward facing components of lipid bilayer, the lipid headgroups, regulate how cellular membranes or lipid nanoparticles
+  interact with approaching substances, such as proteins, drugs, RNA or viruses.
+  %In the physiologically relevant liquid state,
+  %lipid molecules are not fixed to a single structure but sample an ensemble of conformations with Boltzmann-weighted probabilities.
+  Because conformational ensembles of lipids in physiologically relevant liquid state 
+  %However, the distributions of conformations in such ensembles and their lipid-type dependence
+  have not yet been experimentally determined, a crucial piece has been missing from the conception on how they interact with other biomolecules: 
   It is not clear if lipid molecules exchange between a few rigid structures,
@@
   Here, we combine solid-state NMR experiments and molecular dynamics simulations from the NMRlipids Project to resolve the conformational ensembles of headgroups of four key lipid types in various biologically relevant conditions.
-We find all lipid headgroup types to sample a wide range of conformations in both neutral and charged cellular membranes.
-  Differences in the headgroup chemistry do manifest
-  in probability distributions of conformations, but all four lipid types can access almost the whole conformation range.
+We find that all lipid headgroup types sample a wide range of conformations % a wide range of conformations
+  in both neutral and charged cellular membranes, and that differences in the headgroup chemistry manifests only
+  in probability distributions of conformations. %, but all four lipid types can access .
   Together with the analysis of 894 protein-bound lipid structures from the Protein Data Bank (PDB), this suggests that lipids can bind to proteins in a wide range of conformations, which are not limited by the headgroup chemistry.
-Therefore, the selective adsorption of peripheral proteins to membranes and the binding of lipids to transmembrane proteins are likely to be regulated by specific protein--lipid interactions, not by any inherent conformational restrictions of the lipids.
-  Our results pave the way to comprehensive understanding of lipid--protein interactions in biomedical applications, in particular in
-  lipid-mediated signaling.
+  %Therefore, the selective adsorption of peripheral proteins to membranes and the binding of lipids to transmembrane proteins are likely to be regulated by specific protein--lipid interactions, not by any inherent conformational restrictions of the lipids.
+  %Our results pave the way to comprehensive understanding of lipid--protein interactions in biomedical applications, in particular in
+  %lipid-mediated signaling.
+  Our results propose that lipids are able to select a suitable conformation from the wide range of selection to fit the various binding sites in proteins.
+  The proposed {\it inverse conformational selection model} can be also extended to lipid binding to other biomolecules than proteins, such as drug, RNA and viruses.
 

### 2021-05-28 manuscriptPGPE.tex
Update text

@@
 
+\begin{tocentry}
+
+     \includegraphics[height=3.5cm]{../Figs/TOC2.eps}
+%Some journals require a graphical entry for the Table of Contents.
+%This should be laid out ``print ready'' so that the sizing of the
+%text is correct.
+
+%Inside the \texttt{tocentry} environment, the font used is Helvetica
+%8\,pt, as required by \emph{Journal of the American Chemical
+%Society}.
+
+%The surrounding frame is 9\,cm by 3.5\,cm, which is the maximum
+%permitted for  \emph{Journal of the American Chemical Society}
+%graphical table of content entries. The box will not resize if the
+%content is too big: instead it will overflow the edge of the box.
+
+%This box and the associated title will always be printed on a
+%separate page at the end of the document.
+
+\end{tocentry}
+
 \begin{abstract}
@@
 
-Chemical compositions of hydrophilic lipid headgroups vary between different
-organelles and organisms, and different lipid types
-regulate protein functions in many different ways \cite{lee03,vanmeer08}.
-Lipids can directly bind to proteins, or indirectly affect protein
-functions by altering membrane properties such as charge or elasticity \cite{lee03,lemmon08}.
-Specific interactions with certain lipid headgroups
-are known to be essential for the function of several proteins \cite{lee11,lemmon08},
-but it is not clear if the specificity arises from the
+Lipid interactions with other biomolecules is gaining interest
+not only in molecular cell biology \cite{harayama18}, but also in
+nanobiotechnology applications, such as design of antibodies \cite{vigant15}
+and carriers for mRNA vaccines \cite{pardi18,schoenmaker21}.
+Interactions of cellular membranes or lipid nanoparticles with other biomolecules,
+such as proteins, drugs or RNA, are regulated by lipid headgroups,
+the outward facing components of lipid bilayers \cite{harayama18,vanmeer08}.
+%Chemical compositions of hydrophilic  vary between different
+%organelles and organisms, and different lipid types
+%regulate protein functions in many different ways \cite{lee03,vanmeer08}.
+%Lipids can directly bind to proteins, or indirectly affect protein
+%functions by altering membrane properties such as charge or elasticity \cite{lee03,lemmon08}.
+Chemical compositions of lipid headgroups vary between different
+organelles and organisms, and specific interactions with certain lipid headgroups
+are known to be essential for the function of several proteins \cite{harayama18,vanmeer08}.
+However, the driving forces for specific interactions between lipids and membrane binding
+substances are not fully understood because conformational ensembles of lipids in
+physiologically relevant liquid state have not been experimentally determined.
+Therefore, it is not clear if the specificity arises from the
 differences in accessible conformations between lipid types or
@@
 and has been related to various diseases \cite{vance15,calzada16,patel17}.
+PC lipids are also used in mRNA COVID-19 vaccines \cite{schoenmaker21}.
 PS are the most common negatively charged lipids in eukaryotes and PG in bacteria,
@@
 we therefore elucidate also the effect of protein binding to headgroup conformations.
-These interactions are crucial, for example, in lipid-mediated signaling \cite{lemmon08} and for the
-design of phoshopolipid-specific antibodies \cite{vigant15}.
+%These interactions are crucial, for example, in lipid-mediated signaling \cite{lemmon08} and for the
+%design of phoshopolipid-specific antibodies \cite{vigant15}. 
+The resulting {\it inverse conformational selection model} for lipid-protein binding can be also
+applied to understand interactions between membranes and other biomolecules than proteins, such as drugs, RNA, or viruses.
+On the other hand, the detailed understanding of lipid bilayer interfaces can facilitate the design
+of lipid nanoparticle carriers for mRNA vaccines with less side effects.
 
@@
 
-Our results pave the way toward understanding lipid-mediated cell signaling and how lipids regulate membrane protein function in general.
-We suggest that the key to understanding selective binding of certain lipid types to proteins is in intermolecular lipid--protein interactions, rather than in conformational restrictions inherent to the lipids.
+Our results suggest that lipid binding to proteins can be described by the {\it inverse conformational selection model}
+where unbound lipids sample a wide conformational ensemble offering the large selection of conformations
+that can bind to various binding sites in substrates, such as protein, drug, RNA, or virus, that remain conformationally fixed.
+In this model, the binding energy is dominated by
+intermolecular interactions and changes in lipid conformational entropy, while 
+conformational restrictions inherent to the lipids play a minor role.
+%We suggest that the key to understanding selective binding of certain lipid types to proteins is in intermolecular lipid--protein interactions, rather than in conformational restrictions inherent to the lipids.
 The broad conformational ensembles we see in bulk bilayers suggest that the structures observed in pure lipid crystals occur seldomly in liquid state.
 Consequently, in protein binding the entropic cost on the lipid may be significant.
-
-Finally, our work has demonstrated the power of open access MD simulation data
+Applications of the proposed model in molecular cell biology include the understanding of lipid-mediated cell signaling
+and functions of lipid regulated membrane proteins in general. 
+In bionanotechnology, our model facilitates, for example,
+the design of phoshopolipid-specific antibodies and lipid nanoparticle carries with minimal sides effects for mRNA vaccines.
+Furthermore, our work has demonstrated the power of open access MD simulation data
 to complement the PDB data in elucidating behavior of complex systems of disordered biomolecules.

### 2021-05-29 CoverLetter.tex
Update manuscript and cover letter

--- 2021-05-05 CoverLetter.tex
@@
 
-Please find enclosed our manuscript entitled {\it Conformational plasticity of lipid headgroups in cellular membranes and protein-lipid complexes}
-by A. Bacle et al. which we would like to submit for consideration to be published in \textit{ACS Central Science}.
+Please find enclosed our manuscript entitled
+{\it Inverse conformational selection in lipid-protein binding}
+by A. Bacle et al. which we would like to submit for consideration to be published in the \textit{Journal of American Chemical Society}.
 
-In the manuscript, we demonstrate that lipids in cellular membranes exhibit a much higher plasticity in terms of allowed conformational states than previously anticipated.
-Our results update the current understanding on how lipids selectively bind to proteins.
+In the manuscript, we present the {\it Inverse conformational selection model} for lipid-protein binding.
+The model illustrates how lipids in cellular membranes or in lipid nanoparticles can select
+suitable conformations from wide ensemble %sampled in physiologically relevant liquid state
+to bind in various biomolecules, such as proteins, drugs, RNA or viruses.
+Our model is formulated based on solid state NMR experiments, molecular dynamics (MD) simulations,
+and bioinformatical analysis.
+First, the extensive NMR and MD simulation data collected within the NMRlipids project
+(open collaboration project started by us in 2013
+with 210 citations in total) revealed that 
+the most common lipid types in eukaryotes and bacteria exhibit a much higher plasticity in
+terms of allowed conformational states than previously anticipated.
+Motivated by these results, 
+we used a novel bioinformatical analysis of protein-bound lipid structures from the protein data bank (PDB),
+and found that the plasticity in lipid headgroups is a characteristic feature also in lipid-protein complexes.
 
-Lipids with different headgroups, the water facing components of cell membranes, are known to regulate cell functions
-by selective binding to different proteins. It is, however, not known if the selectivity is driven by unique
-lipid conformations between the lipid types that fit to binding pockets in proteins (lock and key model), or by some other interactions.
-One of the main obstacles in understanding the selective binding has been the lack of
-experimental methods to resolve the conformational ensembles of lipids when they are not bound to proteins.
+While lipid-protein interactions, and cell functions regulated by them,  remain
+as a central topic in molecular cell biology, interactions of lipids with proteins and other biomolecules
+is gaining interested also in nanobiotechnology. For example, lipid nanoparticles used as carriers
+in mRNA vaccines against COVID-19 can cause allergic reactions.
+Our {\it Inverse conformational selection model} can be used to understand how
+these nanoparticles interact with RNA cargo and physiological environment,
+thereby facilitating the design of novel vaccine or drug carrier with less side effects.
+Furthermore, the {\it Inverse conformational selection model}
+may apply also to other than lipid molecules, such as intrinsically disordered proteins or sugars.
 
-In 2013 we set out to solve the conformational ensembles of lipid headgroups by establishing the NMRlipids open collaboration project
-which combines data from molecular dynamics (MD) simulations and solid-state NMR experiments.
-Using the extensive data collected within this project, we have now resolved the differences in conformational ensembles
-between the most common lipids in eukaryotes and bacteria. 
-Our results reveal a significant structural plasticity in lipid headgroups and only minor differences of the conformational ensembles
-between lipids with different headgroup chemistry.
-Notably, and motivated by our findings on the model systems, by using a bioinformatical analysis of protein-bound lipid structures from the protein data bank (PDB), we found that the plasticity in lipid headgroups is a characteristic feature in lipid-protein complexes.
-
-The novel conclusions from our work are that lipid headgroups can adapt to multiple binding sites in proteins,
-and the selectivity of lipid-protein binding is driven by intermolecular interactions rather than by restrictions in lipid conformations.
-Furthermore, we demonstrated the power of a conceptually novel open collaboration approach that extended the realm of Molecular Dynamics simulations to complement the experimental data in the PDB for elucidating how complex systems made up of disordered biomolecules behave.
-
-We believe that our approach and findings are of great interest to understand the connection of lipid chemistry to protein-lipid interactions  and ultimately the role of lipid chemistry in a number of diseases.
-Therefore, we believe that \textit{ACS Central Science} is a highly suitable journal to publish this work.
+We believe that the proposed conceptually novel {\it Inverse conformational selection model} 
+contributes to wide scope of research fields, ranging from biophysics and biochemistry to biology and biotechnology.
+Moreover, the combination of the state-of-the-art NMR experiments and MD simulations with original open collaboration
+and bioinformatical approaches further increase the significance of the manuscript.
+%the open access policy of {\it Nature Communications} is optimal for the publications from
+%the NMRlipids Project which fosters the open collaboration and open data approaches as all the data and other material
+%are publicly available throughout the project.
+Therefore, we believe that our manuscript fulfills the high standards for publications in the \textit{Journal of American Chemical Society}.
 

### 2021-05-29 manuscriptPGPE.tex
Update manuscript and cover letter

@@
   %However, the distributions of conformations in such ensembles and their lipid-type dependence
-  have not yet been experimentally determined, a crucial piece has been missing from the conception on how they interact with other biomolecules: 
+  have not yet been experimentally determined, a crucial piece has been missing from the understanding
+  on how they interact with other biomolecules: 
   It is not clear if lipid molecules exchange between a few rigid structures,
@@
   Here, we combine solid-state NMR experiments and molecular dynamics simulations from the NMRlipids Project to resolve the conformational ensembles of headgroups of four key lipid types in various biologically relevant conditions.
-We find that all lipid headgroup types sample a wide range of conformations % a wide range of conformations
+We find that all lipid headgroup types sample a wide range of
+  overlapping conformations % a wide range of conformations
   in both neutral and charged cellular membranes, and that differences in the headgroup chemistry manifests only
   in probability distributions of conformations. %, but all four lipid types can access .
-  Together with the analysis of 894 protein-bound lipid structures from the Protein Data Bank (PDB), this suggests that lipids can bind to proteins in a wide range of conformations, which are not limited by the headgroup chemistry.
+  Together with the analysis of 894 protein-bound lipid structures from the Protein Data Bank (PDB),
+  this suggests that all four key lipid types can bind to proteins in a wide range of conformations,
+  which are not limited by the headgroup chemistry.
   %Therefore, the selective adsorption of peripheral proteins to membranes and the binding of lipids to transmembrane proteins are likely to be regulated by specific protein--lipid interactions, not by any inherent conformational restrictions of the lipids.
@@
   %lipid-mediated signaling.
-  Our results propose that lipids are able to select a suitable conformation from the wide range of selection to fit the various binding sites in proteins.
-  The proposed {\it inverse conformational selection model} can be also extended to lipid binding to other biomolecules than proteins, such as drug, RNA and viruses.
+  Our results propose that lipids are able to select a suitable conformation
+  from the wide range available to them to fit the various binding sites in proteins.
+  The proposed {\it Inverse Conformational Selection Model} will extend also to lipid binding
+  to other targets than proteins, such as drugs, RNA and viruses.
 
@@
 In bionanotechnology, our model facilitates, for example,
-the design of phoshopolipid-specific antibodies and lipid nanoparticle carries with minimal sides effects for mRNA vaccines.
+the design of phoshopolipid-specific antibodies and lipid nanoparticle carriers with minimal sides effects for mRNA vaccines.
 Furthermore, our work has demonstrated the power of open access MD simulation data

### 2021-06-07 manuscriptPGPEsuppl.tex
Fix error in table S3

@@
       POPC:POPG (7:3)        & CHARMM36 \cite{klauda10,venable13}    & 0  & 350     & 25000 & 0   &  310  & 500 & 100 & \citenum{POPC7POPG1charmm36}  \\
-      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}     & 0  & 150:150 & 31500 & 0   &  298  & 500 & 400 & \citenum{CHARMM36POPCPOPG5050} \\ 
+      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}     & 0  & 150:150 & 31500 & 0   &  298  & 400 & 300 & \citenum{CHARMM36POPCPOPG5050} \\ 
       POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}     & 0.1 & 150:150 & 31329 & 57 &  298  & 400 & 300 & \citenum{CHARMM36POPCPOPG5050100mMCaCl} \\
-      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}     & 1.08 & 150:150 & 29766 & 578  &  298  & 500 & 400 & \citenum{CHARMM36POPCPOPG50501000mMCaCl} \\
-      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 0  & 350:88 & 26280 & 0  &  298  & 500 & 400 & \citenum{CHARMM36POPCPOPG4010} \\
-      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 0.1 & 350:88 & 26280 & 47  &  298  & 500 & 400 & \citenum{CHARMM36POPCPOPG4010100mMCaCl} \\
-      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 1.0 & 350:88 & 24927 & 451  &  298  & 500 & 400 & \citenum{CHARMM36POPCPOPG40101000mMCaCl} \\
+      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}     & 1.08 & 150:150 & 29766 & 578  &  298  & 400 & 300 & \citenum{CHARMM36POPCPOPG50501000mMCaCl} \\
+      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 0  & 350:88 & 26280 & 0  &  298  & 400 & 300 & \citenum{CHARMM36POPCPOPG4010} \\
+      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 0.1 & 350:88 & 26280 & 47  &  298  & 400 & 300 & \citenum{CHARMM36POPCPOPG4010100mMCaCl} \\
+      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 1.0 & 350:88 & 24927 & 451  &  298  & 400 & 300 & \citenum{CHARMM36POPCPOPG40101000mMCaCl} \\
       \hline

### 2021-07-01 manuscriptPGPE.tex
Updates based on reviewer comments

@@
 
-     \includegraphics[height=3.5cm]{../Figs/TOC2.eps}
+     \includegraphics[height=3.5cm]{./Figs/TOC2.eps}
 %Some journals require a graphical entry for the Table of Contents.
@@
 To find the best MD simulation models (force fields) for this purpose, we took advantage of the power of the NMRlipids Project, an open science collaboration that has gathered a massive number of simulation trajectories, allowing the evaluation of all the available force fields of PC~\cite{botan15}, PS~\cite{antila19}, PE (Fig.~\ref{HGorderParametersPE}), and PG (Fig.~\ref{HGorderParametersPOPG}) lipids against NMR data.
-The comparison of hundreds of simulation trajectories simulated with approximately 20 different force fields revealed that none of them can correctly capture the lipid conformational ensembles and reproduce the experimental $S_\mathrm{CH}$ of all segments within experimental error.
+The comparison of hundreds of simulation trajectories simulated with approximately 20 different force fields revealed that none of them can correctly capture the lipid conformational ensembles and reproduce the experimental $S_\mathrm{CH}$ of all segments within experimental error (for detailed discussion on the quality of force fields, see section S3 in the supplementary information and Refs.
+\citenum{botan15,ollila16,antila19}).
 However, CHARMM36, the force field showing the best agreement with experiments, correctly captures the essential differences between the PC, PE, PG, and PS headgroup order parameters (compare Figs.~\ref{HGorderParameters}D and \ref{structures} B).

### 2021-07-05 manuscriptPGPE.tex

@@
   %lipid molecules are not fixed to a single structure but sample an ensemble of conformations with Boltzmann-weighted probabilities.
-  Because conformational ensembles of lipids in physiologically relevant liquid state 
+  Because conformational ensembles of lipid headgroups in physiologically relevant liquid state 
   %However, the distributions of conformations in such ensembles and their lipid-type dependence
@@
   %lipid-mediated signaling.
-  Our results propose that lipids are able to select a suitable conformation
+  Our results propose that lipids are able to select a suitable headgroup conformation
   from the wide range available to them to fit the various binding sites in proteins.
@@
 different headgroups are not observed.
-Structures deviating from lipid crystals 
+Protein-bound lipid structures deviating from protein-free lipid crystal structures 
 have been previously proposed to indicate inaccuracies in lipid structures in PDB \cite{marsh13b,pezeshkian18}. 
-However, we observe large deviations from the lipid crystal structures also in conformational ensembles
+However, we observe large deviations from the protein-free lipid crystal structures also in conformational ensembles
 that reproduce the NMR data in the liquid lamellar phase, indicating that such deviations
@@
 
-Our results suggest that lipid binding to proteins can be described by the {\it inverse conformational selection model}
+Our results suggest that lipid headgroup binding to proteins can be described by the {\it inverse conformational selection model}
 where unbound lipids sample a wide conformational ensemble offering the large selection of conformations
@@
 conformational restrictions inherent to the lipids play a minor role.
-%We suggest that the key to understanding selective binding of certain lipid types to proteins is in intermolecular lipid--protein interactions, rather than in conformational restrictions inherent to the lipids.
-The broad conformational ensembles we see in bulk bilayers suggest that the structures observed in pure lipid crystals occur seldomly in liquid state.
-Consequently, in protein binding the entropic cost on the lipid may be significant.
+Here we have focused only on lipid headgroups and potential applications of the proposed model on interactions between hydrophobic acyl chains and proteins are left for future studies. 
 Applications of the proposed model in molecular cell biology include the understanding of lipid-mediated cell signaling

### 2021-07-05 manuscriptPGPE.tex

@@
      \textbf{A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurrence of different conformations.
-     \textbf{B)} Headgroup and glycerol backbone region order parameters of different lipids.
+     \textbf{B)} Headgroup and glycerol backbone region order parameters of different lipids. Colored shaded regions show the range of experimental values for POPC from Fig. \ref{HGorderParameters}.
      \textbf{C)} Relative free energies for individual heavy atom dihedral angles estimated from the inverse Boltzmann formula.
@@
 \citenum{botan15,ollila16,antila19}).
-However, CHARMM36, the force field showing the best agreement with experiments, correctly captures the essential differences between the PC, PE, PG, and PS headgroup order parameters (compare Figs.~\ref{HGorderParameters}D and \ref{structures} B).
+However, CHARMM36, the force field showing the best agreement with experiments, correctly captures the essential differences between the PC, PE, PG, and PS headgroup order parameters, particularly the positive order parameter of $\beta$-carbon in PG and the forking of $\alpha$-carbon in PS, see Fig.~\ref{structures} B.
 

### 2021-07-06 manuscriptPGPE.tex
Updates based on referees' comments.

@@
 the outward facing components of lipid bilayers \cite{harayama18,vanmeer08}.
-%Chemical compositions of hydrophilic  vary between different
-%organelles and organisms, and different lipid types
-%regulate protein functions in many different ways \cite{lee03,vanmeer08}.
-%Lipids can directly bind to proteins, or indirectly affect protein
-%functions by altering membrane properties such as charge or elasticity \cite{lee03,lemmon08}.
 Chemical compositions of lipid headgroups vary between different
@@
 and crystal structures of lipids have been determined \cite{buldt81,pascher92},
-but their relation to the conformational ensembles in bulk membranes in the liquid lamellar phase remains unclear \cite{marsh13b}.
+but their relation to lipids in bulk membranes in the liquid lamellar phase remains unclear \cite{marsh13b}.
+Similarly to disordered proteins~\cite{sormanni17}, lipids sample conformational ensemble which is composed of the set of individual structures whose probability to occur is determined by the Boltzmann distribution.
+In liquid bilayer state, individual phosphatidylcholine lipid samples the conformational ensemble within nanosecond timescale~\cite{ferreira15}.
 The most accurate experimental information on conformational ensembles of lipids
 in this biologically relevant phase are typically derived from NMR experiments, particularly from the
-C--H bond order parameters $S_\mathrm{CH}$ measured using $^2$H\,NMR~\cite{seelig77c,davis83,Semchyschyn04}.
+C--H bond order parameters, $S_\mathrm{CH}=\frac{3}{2} \langle \cos^2 \theta -1 \rangle$, where $\theta$ is the angle between C-H bond and membrane normal, and the average is taken over conformational ensemble of lipids~\cite{seelig77c,davis83,Semchyschyn04}.
 Because these order parameters are similar as in living cells \cite{gally81,scherer87,seelig90},
@@
 
-MD simulations could be used to interpret the lipid conformational ensembles from experiments due to their direct connection to the measured C-H bond order parameters $S_\mathrm{CH}$.
+Ideal
+MD simulations would correctly predict the conformational ensembles of lipids, thereby giving an interpretation for  the experimental C-H bond order parameters, $S_\mathrm{CH}$, due to their direct connection~\cite{ollila16}.
 To find the best MD simulation models (force fields) for this purpose, we took advantage of the power of the NMRlipids Project, an open science collaboration that has gathered a massive number of simulation trajectories, allowing the evaluation of all the available force fields of PC~\cite{botan15}, PS~\cite{antila19}, PE (Fig.~\ref{HGorderParametersPE}), and PG (Fig.~\ref{HGorderParametersPOPG}) lipids against NMR data.
@@
 
-In conclusion, all studied lipid headgroups sample very broad conformational ensembles in the liquid lamellar phase and the sampled dihedral angles are within similar ranges for all headgroup types.
-Since the rotation of dihedral angles to almost any position bears relatively low free energy cost, the lipid headgroups are able to adopt a wide range of conformations when interacting with proteins, ions, or other biomolecules.
-Therefore, the structures in lipid crystals \cite{buldt81,pascher92} are not representative of the liquid state, and models aiming to explain NMR data using only a few conformations \cite{seelig77c,davis83,Semchyschyn04,akutsu20} are not sufficient to capture the large conformational space of lipids in the liquid lamellar state.
+In conclusion, the rotation of dihedral angles to almost any position bears relatively low free energy cost and the sampled dihedral angles are within similar ranges for all studied headgroup types.
+In other words, all studied lipid headgroups sample very broad conformational ensembles in the liquid lamellar phase.
+Therefore, the lipid headgroups are able to adopt a wide range of conformations when interacting with proteins, ions, or other biomolecules.
+Furthermore, the structures in lipid crystals \cite{buldt81,pascher92} are not representative of the liquid state, and models aiming to explain NMR data using only a few conformations \cite{seelig77c,davis83,Semchyschyn04,akutsu20} are not sufficient to capture the large conformational space of lipids in the liquid lamellar state.
 However, conformational ensembles of lipids in MD simulation trajectories with accurate force fields that capture the experimental NMR order parameters and can be considered realistic, and thereby used to interpret the experimental data.
@@
   \caption{\label{changesWITHsurf}
-    \textbf{A)} Modulation of PC headgroup order parameters and P--N vector tilt (angle between the P-N vector and the bilayer normal) upon addition of cationic surfactant (dihexadecyldimethylammonium)
+    \textbf{A)} Modulation of PC headgroup order parameters and P--N vector tilt (average angle between the P-N vector and the bilayer normal) upon addition of cationic surfactant (dihexadecyldimethylammonium)
     from CHARMM36 simulations compared with the experimental data \cite{scherer89}.
@@
 However, the detailed understanding of the structural and energetic responses of lipids to membrane-bound electric charge is still lacking \cite{Semchyschyn04}.
+For example, it is not clear if the membrane-bound charges restrain the lipids into specific conformations or just redistribute the probabilities of states by keeping the accessible conformations similar.
 

### 2021-07-07 manuscriptPGPE.tex
Updates based on referees' comments.

@@
 but their relation to lipids in bulk membranes in the liquid lamellar phase remains unclear \cite{marsh13b}.
-Similarly to disordered proteins~\cite{sormanni17}, lipids sample conformational ensemble which is composed of the set of individual structures whose probability to occur is determined by the Boltzmann distribution.
+Similarly to disordered proteins~\cite{sormanni17}, lipid molecules sample conformational ensembles which are composed of sets of individual structures whose probabilities of occurrences are determined by the Boltzmann distribution.
 In liquid bilayer state, individual phosphatidylcholine lipid samples the conformational ensemble within nanosecond timescale~\cite{ferreira15}.
@@
 in this biologically relevant phase are typically derived from NMR experiments, particularly from the
-C--H bond order parameters, $S_\mathrm{CH}=\frac{3}{2} \langle \cos^2 \theta -1 \rangle$, where $\theta$ is the angle between C-H bond and membrane normal, and the average is taken over conformational ensemble of lipids~\cite{seelig77c,davis83,Semchyschyn04}.
+C--H bond order parameters, $S_\mathrm{CH}=\frac{3}{2} \langle \cos^2 \theta -1 \rangle$, where $\theta$ is the angle between C-H bond and membrane normal, and the average is taken over the conformational ensemble of lipids~\cite{seelig77c,davis83,Semchyschyn04}.
 Because these order parameters are similar as in living cells \cite{gally81,scherer87,seelig90},
@@
 conformational restrictions inherent to the lipids play a minor role.
-Here we have focused only on lipid headgroups and potential applications of the proposed model on interactions between hydrophobic acyl chains and proteins are left for future studies. 
+Here we have focused only on lipid headgroups while potential applications of the proposed model on interactions between hydrophobic acyl chains and proteins are left for future studies. 
 Applications of the proposed model in molecular cell biology include the understanding of lipid-mediated cell signaling

### 2021-07-12 manuscriptPGPE.tex
Updates from Overleaf

@@
   Interest in lipid interactions with proteins and other biomolecules is emerging not only in
-  fundamental biochemistry, but also in the field of nanobiotechnology where lipid nanoparticles
-  are commonly used, for example, as carriers in mRNA vaccines.
-  The outward facing components of lipid bilayer, the lipid headgroups, regulate how cellular membranes or lipid nanoparticles
-  interact with approaching substances, such as proteins, drugs, RNA or viruses.
-  %In the physiologically relevant liquid state,
-  %lipid molecules are not fixed to a single structure but sample an ensemble of conformations with Boltzmann-weighted probabilities.
-  Because conformational ensembles of lipid headgroups in physiologically relevant liquid state 
-  %However, the distributions of conformations in such ensembles and their lipid-type dependence
-  have not yet been experimentally determined, a crucial piece has been missing from the understanding
-  on how they interact with other biomolecules: 
-  It is not clear if lipid molecules exchange between a few rigid structures,
-  or fluctuate freely across a practically continuous spectrum of  conformations.
+  fundamental biochemistry, but also in the field of nanobiotechnology where lipids
+  are commonly used, for example, in carriers of mRNA vaccines.
+  The outward facing components of cellular membranes and lipid nanoparticles, the lipid headgroups, regulate membrane 
+  interactions with approaching substances, such as proteins, drugs, RNA or viruses.
+  Because lipid headgroup conformational ensembles  
+  have not been experimentally determined in physiologically relevant conditions, an essential question on their interactions with other biomolecules remains unanswered: 
+  Do headgroups exchange between a few rigid structures,
+  or fluctuate freely across a practically continuous spectrum of  conformations?
   Here, we combine solid-state NMR experiments and molecular dynamics simulations from the NMRlipids Project to resolve the conformational ensembles of headgroups of four key lipid types in various biologically relevant conditions.
-We find that all lipid headgroup types sample a wide range of
-  overlapping conformations % a wide range of conformations
+We find that lipid headgroups sample a wide range of
+  overlapping conformations
   in both neutral and charged cellular membranes, and that differences in the headgroup chemistry manifests only
-  in probability distributions of conformations. %, but all four lipid types can access .
-  Together with the analysis of 894 protein-bound lipid structures from the Protein Data Bank (PDB),
-  this suggests that all four key lipid types can bind to proteins in a wide range of conformations,
+  in probability distributions of conformations. Furthermore,
+  the analysis of 894 protein-bound lipid structures from the Protein Data Bank (PDB)
+  suggests that lipids can bind to proteins in wide range of conformations,
   which are not limited by the headgroup chemistry.
-  %Therefore, the selective adsorption of peripheral proteins to membranes and the binding of lipids to transmembrane proteins are likely to be regulated by specific protein--lipid interactions, not by any inherent conformational restrictions of the lipids.
-  %Our results pave the way to comprehensive understanding of lipid--protein interactions in biomedical applications, in particular in
-  %lipid-mediated signaling.
-  Our results propose that lipids are able to select a suitable headgroup conformation
+  We propose that lipids can select a suitable headgroup conformation
   from the wide range available to them to fit the various binding sites in proteins.
@@
 
-Lipid interactions with other biomolecules is gaining interest
+Lipid interactions with other biomolecules are gaining interest
 not only in molecular cell biology \cite{harayama18}, but also in
@@
 but their relation to lipids in bulk membranes in the liquid lamellar phase remains unclear \cite{marsh13b}.
-Similarly to disordered proteins~\cite{sormanni17}, lipid molecules sample conformational ensembles which are composed of sets of individual structures whose probabilities of occurrences are determined by the Boltzmann distribution.
-In liquid bilayer state, individual phosphatidylcholine lipid samples the conformational ensemble within nanosecond timescale~\cite{ferreira15}.
+Similarly to disordered proteins~\cite{sormanni17}, lipid molecules sample conformational ensembles that are composed of sets of individual structures whose occurrence probabilities are determined by the Boltzmann distribution~\cite{buslaev16}.
+In liquid bilayer state, individual phosphatidylcholine lipid samples the conformational ensemble on nanosecond timescale~\cite{ferreira15,antila21a}.
 The most accurate experimental information on conformational ensembles of lipids
@@
 C--H bond order parameters, $S_\mathrm{CH}=\frac{3}{2} \langle \cos^2 \theta -1 \rangle$, where $\theta$ is the angle between C-H bond and membrane normal, and the average is taken over the conformational ensemble of lipids~\cite{seelig77c,davis83,Semchyschyn04}.
-Because these order parameters are similar as in living cells \cite{gally81,scherer87,seelig90},
+Because these order parameters are similar to those in living cells \cite{gally81,scherer87,seelig90},
 the model membranes can be used to resolve the lipid headgroup conformations in biological conditions.
@@
    \caption{\label{HGorderParameters}
-     Chemical structure, refocused-INEPT spectrum, 2D R-PDLF spectra, and S-DROSS dipolar slices (from top to bottom) for \textbf{A)} POPE  and \textbf{B)} POPG MLVs measured by $^1$H-$^{13}$C solid-state NMR.
+     Chemical structure, refocused-INEPT spectrum, 2D R-PDLF spectra, and S-DROSS dipolar slices (from top to bottom) for \textbf{A)} POPE  and \textbf{B)} POPG multi-lamellar vesicles measured by $^1$H-$^{13}$C solid-state NMR.
 The S-DROSS dipolar slices show both experimental data (open circles) and the result of NMR numerical simulations (blue solid lines).
@@
      \textbf{C)} Chemical structure of POPC and POPS.
-    \textbf{D)} Headgroup and glycerol backbone C-H bond order parameters for different phospholipids measured in the lamellar liquid disordered phase.
+    \textbf{D)} Headgroup and glycerol backbone C--H bond order parameters $S_mathrm{CH}$ for different phospholipids measured in the lamellar liquid disordered phase.
     Filled squares show $S_{\rm{CH}}$ magnitudes and signs determined by $^1$H-$^{13}$C NMR spectroscopy for POPE (310~K) and POPG (298~K), as
@@
 Empty symbols show headgroup and glycerol backbone $S_{\rm{CH}}$ magnitudes measured previously with $^2$H NMR and are also plotted as real values using the signs determined in this work for
-    POPG with 10nM PIPES (298~K) \cite{borle85},
-    DPPG with 10mM PIPES and 100mM NaCl (314~K) \cite{wohlgemuth80}, 
+    POPG with 10\,mM PIPES (298~K) \cite{borle85},
+    DPPG with 10\,mM PIPES and 100\,mM NaCl (314~K) \cite{wohlgemuth80}, 
     DPPE (341~K) \cite{seelig76},
-    E. coli PE and E. coli PG (310~K) \cite{gally81}
-    .
+    {\it E. coli} PE and {\it E. coli} PG with 10\,mM PIPES and 100\,mM NaCl (310~K) \cite{gally81}.
    }
@@
 To experimentally characterize the differences in headgroup conformational ensembles of lipids that are not
-bound to proteins, we measured the C--H bond order parameters
+bound to proteins, we measured the C--H bond order parameters $S_\mathrm{CH}$
 and their signs of POPG and POPE in the liquid lamellar phase, as we did previously for POPC and POPS \cite{ferreira13,ferreira16,antila19}.
-Determination of headgroup and glycerol backbone order parameters and their signs
+Determination of headgroup and glycerol backbone $S_\mathrm{CH}$ and their signs
 was straightforward from the data in Figs.~\ref{HGorderParameters}, \ref{POPEspectra} and \ref{POPGspectra}
@@
 Based on previous $^2$H NMR measurements \cite{wohlgemuth80,gally81,borle85},
-we assigned the larger order parameter to the g$_2$ carbon
+we assigned the larger $S_\mathrm{CH}$ to the g$_2$ carbon
 and used the literature value for the $\beta$-carbon in SIMPSON simulations to determine the signs.
-The decrease in the beginning of the S-DROSS curve suggests that the sign of larger g$_2$ order parameter
-is negative and later increase suggests that sign of smaller $\beta$ order parameter is positive as confirmed by the SIMPSON simulations (Fig.~\ref{HGorderParameters}B).
+The decrease in the beginning of the S-DROSS curve suggests that the sign of the larger g$_2$ $S_\mathrm{CH}$
+is negative, and the later increase suggests that sign of the smaller $\beta$ $S_\mathrm{CH}$ is positive, as confirmed by the SIMPSON simulations (Fig.~\ref{HGorderParameters}B).
 
-Experimental order parameters of POPC, POPE, POPG and POPS glycerol backbones and headgroups from this and previous studies are displayed in Fig.~\ref{HGorderParameters}D, where the signs from $^{13}$C NMR experiments are used also for the $^2$H NMR data from the literature.
-The overall agreement of order parameters determined by different research teams and different techniques for the same lipid headgroup is excellent (within $\pm$0.02) here and in previous studies \cite{botan15,ollila16,antila19}.
-Therefore, the differences between lipid types are dictated by the headgroup chemistry rather than inaccuracies in experiments, differences in the acyl chains, or in experimental conditions.
+Experimental order parameters of POPC, POPE, POPG, and POPS glycerol backbones and headgroups from this and previous studies are displayed in Fig.~\ref{HGorderParameters}D, where the signs from $^{13}$C NMR experiments are used also for the $^2$H NMR data from the literature.
+The overall agreement of $S_\mathrm{CH}$ determined by different research teams and different techniques for the same lipid headgroup was excellent (within $\pm$0.02) here and in previous studies \cite{botan15,ollila16,antila19}.
+Therefore, the differences between lipid types are dictated by the headgroup chemistry rather than inaccuracies in experiments, or differences in the acyl chains or in the experimental conditions.
 
-The most distinct order parameters are observed for PS headgroups, for which the $\alpha$-carbon order parameter exhibits significant forking (different values for different hydrogens bound to the same carbon) and the $\beta$-carbon has a more negative value than other studied lipid types.
-On the other hand, the $\beta$-carbon order parameter of PG headgroup has a positive sign, in contrast to all the other lipid types.
+The most distinct order parameters are observed for PS headgroups, for which the $\alpha$-carbon $S_\mathrm{CH}$ exhibits significant forking (different values for different hydrogens bound to the same carbon) and the $\beta$-carbon has a more negative value than in other studied lipid types.
+On the other hand, the $\beta$-carbon $S_\mathrm{CH}$ of PG headgroup has a positive sign, in contrast to all the other lipid types.
 The sign has not been previously known from $^2$H NMR experiments which enable detecting only order parameter absolute values~\cite{wohlgemuth80,gally81,borle85}.
@@
    \caption{\label{structures}
-     Results from the best simulation model demonstrating the differences in conformational ensembles between different lipids.
+     Results from the best simulation model demonstrating the differences in conformational ensembles between different lipid types.
      The best simulation model (CHARMM36) was selected from the comparison of ten different force fields based on quality evaluation against experimental $S_{\rm{CH}}$ values in Figs.~\ref{HGorderParametersPE} and~\ref{HGorderParametersPOPG}.
      \textbf{A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurrence of different conformations.
-     \textbf{B)} Headgroup and glycerol backbone region order parameters of different lipids. Colored shaded regions show the range of experimental values for POPC from Fig. \ref{HGorderParameters}.
+     \textbf{B)} Headgroup and glycerol backbone region order parameters of the different lipid types. Colored shaded regions show the range of experimental values from Fig. \ref{HGorderParameters}.
      \textbf{C)} Relative free energies for individual heavy atom dihedral angles estimated from the inverse Boltzmann formula.
      Angles corresponding to free energies above 8~$k_\mathrm{B}T$ are not shown because they are not observed in simulations.
-     \textbf{D)} Distributions of P--N/C$_\gamma$ vector angle with respect to membrane normal.
+     \textbf{D)} Distributions of the P--N/C$_\gamma$ vector angle with respect to the membrane normal.
   }
@@
 The comparison of hundreds of simulation trajectories simulated with approximately 20 different force fields revealed that none of them can correctly capture the lipid conformational ensembles and reproduce the experimental $S_\mathrm{CH}$ of all segments within experimental error (for detailed discussion on the quality of force fields, see section S3 in the supplementary information and Refs.
-\citenum{botan15,ollila16,antila19}).
+\citenum{botan15,ollila16,antila19,antila21a}).
 However, CHARMM36, the force field showing the best agreement with experiments, correctly captures the essential differences between the PC, PE, PG, and PS headgroup order parameters, particularly the positive order parameter of $\beta$-carbon in PG and the forking of $\alpha$-carbon in PS, see Fig.~\ref{structures} B.
@@
 To understand the structural origin of distinct order parameters for the distinct lipids,
-we first calculated the heavy atom dihedral angle distributions from the best performing simulations that reproduced the differences between headgroups~(Fig. \ref{DIHdists}).
+we first calculated and visualized the heavy atom dihedral angle distributions from the best performing simulations that reproduced the differences between headgroups~(Figs.~\ref{structures} B and \ref{DIHdists}).
 Then, we used the inverse Boltzmann formula to estimate the free
@@
 Furthermore, the structures in lipid crystals \cite{buldt81,pascher92} are not representative of the liquid state, and models aiming to explain NMR data using only a few conformations \cite{seelig77c,davis83,Semchyschyn04,akutsu20} are not sufficient to capture the large conformational space of lipids in the liquid lamellar state.
-However, conformational ensembles of lipids in MD simulation trajectories with accurate force fields that capture the experimental NMR order parameters and can be considered realistic, and thereby used to interpret the experimental data.
+However, conformational ensembles of lipids in MD simulation trajectories with accurate force fields that capture the experimental NMR order parameters can be considered realistic, and thereby used to interpret the experimental data.
 
@@
 different headgroups are not observed.
-Protein-bound lipid structures deviating from protein-free lipid crystal structures 
-have been previously proposed to indicate inaccuracies in lipid structures in PDB \cite{marsh13b,pezeshkian18}. 
+Deviation of protein-bound lipid structures from protein-free lipid structures in crystals 
+has been previously proposed to indicate inaccuracies in lipid structures in PDB \cite{marsh13b,pezeshkian18}. 
 However, we observe large deviations from the protein-free lipid crystal structures also in conformational ensembles
@@
 
+\begin{suppinfo}
+
+Additional NMR data, ligand names in PDB used to analyse the protein--bound lipid conformations, evaluation of MD simulations against NMR experiments, dihedral angle distributions from simulations, and description of simulated systes with simulation details.
+
+%This will usually read something like: ``Experimental procedures and
+%characterization data for all new compounds. The class will
+%automatically add a sentence pointing to the information on-line:
+
+\end{suppinfo}
+
 \begin{acknowledgement}

### 2021-07-12 manuscriptPGPEsuppl.tex
Updates from Overleaf

@@
 
-\author{\small Am{\'e}lie Bacle}
-\affiliation{\tiny Laboratoire Coop\'{e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit\'{e} de Poitiers, 1 rue Georges Bonnet, 86000 Poitiers, France }
+\author{Am{\'e}lie Bacle}
+\affiliation{\tiny Laboratoire Coop{\'e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit{\'e} de Poitiers, 1 rue Georges Bonnet, 86000 Poitiers, France }
 
 \author{Pavel Buslaev}
-\affiliation{Nanoscience Center and Department of Chemistry, University of Jyväskylä, P.O. Box 35, 40014 Jyväskylä, Finland}
+\affiliation{Nanoscience Center and Department of Chemistry, University of Jyv{\"a}skyl{\"a}, P.O. Box 35, 40014 Jyv{\"a}skyl{\"a}, Finland}
 \affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, 141701 Dolgoprudny, Russia}
@@
 \affiliation{Center for Research in Biological Chemistry and Molecular Materials (CiQUS), Universidade de Santiago de Compostela, E-15782 Santiago de Compostela, Spain}
-\alsoaffiliation{CIQUP, Centro de Investigação em Qu\'{i}mica, Departamento de Qu\'{i}mica e Bioqu\'{i}mica, Faculdade de Ciências, Universidade do Porto, Porto, Portugal}
+\affiliation{CIQUP, Centro de Investigação em Qu{\'i}mica, Departamento de Qu{\'i}mica e Bioqu{\'i}mica, Faculdade de Ci{\^e}ncias, Universidade do Porto, Porto, Portugal}
 
 \author{Fernando Favela-Rosales}
-\affiliation[Tecnol\'{o}gico Nacional de M\'{e}xico]{Departamento de Ciencias B\'{a}sicas, Tecnol\'{o}gico Nacional de M\'{e}xico, ITS Zacatecas Occidente, M\'{e}xico}
-
-\author{Tiago Ferreira}
-\affiliation{Halle, Germany}
+\affiliation{Departamento de Ciencias B\'{a}sicas, Tecnol\'{o}gico Nacional de M\'{e}xico - ITS Zacatecas Occidente, M\'{e}xico}
+
+\author{Tiago M. Ferreira}
+\affiliation{NMR group - Institute for Physics, Martin Luther University Halle-Wittenberg, 06120 Halle (Saale), Germany}
 
 \author{Patrick F.J. Fuchs}
-\affiliation{Sorbonne Universit\'{e}, Ecole Normale Sup\'{e}rieure, PSL University, CNRS, Laboratoire des Biomol\'{e}cules (LBM), 75005 Paris, France}
-\alsoaffiliation{Universit\'{e} de Paris, UFR Sciences du Vivant, 75013, Paris, France}
+\affiliation{Sorbonne Universit{\'e}, Ecole Normale Sup{\'e}rieure, PSL University, CNRS, Laboratoire des Biomol{\'e}cules (LBM), 75005 Paris, France}
+\affiliation{Universit{\'e} de Paris, UFR Sciences du Vivant, 75013, Paris, France}
 
@@
 \author{Matti Javanainen}
-\affiliation[Czech Academy of Sciences]{Institute of Organic Chemistry and Biochemistry of the 
+\affiliation{Institute of Organic Chemistry and Biochemistry of the 
 Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
@@
 \author{Jesper J. Madsen}
-\affiliation[University of Chicago]{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
-\alsoaffiliation[University of South Florida]{Global and Planetary Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
+\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
+\affiliation{Global and Planetary Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
 
 \author{Josef Melcr}
-\affiliation[Czech Academy of Sciences]{Institute of Organic Chemistry and Biochemistry of the 
-Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
-\alsoaffiliation{Groningen Biomolecular Sciences and Biotechnology Institute 
+\affiliation{Groningen Biomolecular Sciences and Biotechnology Institute 
 and The Zernike Institute for Advanced Materials, 
@@
 
-\author{Paula Milán Rodríguez}
-\affiliation{Sorbonne Université, Ecole Normale Supérieure, PSL University, CNRS, Laboratoire des Biomolécules (LBM), 75005 Paris, France}
+\author{Paula Mil{\'a}n Rodr{\'i}guez}
+\affiliation{Sorbonne Universit{\'e}, Ecole Normale Sup{\'e}rieure, PSL University, CNRS, Laboratoire des Biomol{\'e}cules (LBM), 75005 Paris, France}
 
 \author{Markus S. Miettinen}
+\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
 % \affiliation[Max Planck Institute of Colloids and Interfaces]{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
-\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
 
@@
 \author{Chris G. Papadopoulos}
-\affiliation{Université Paris-Saclay, CEA, CNRS, Institute for Integrative Biology of the Cell (I2BC), 91198, Gif-sur-Yvette, France}
+\affiliation{Universit{\'e} Paris-Saclay, CEA, CNRS, Institute for Integrative Biology of the Cell (I2BC), 91198 Gif-sur-Yvette, France}
 
 \author{Antonio Pe{\'o}n}
-\affiliation{CIQUP, Centro de Investigação em Qu\'{i}mica, Departamento de Qu\'{i}mica e Bioqu\'{i}mica, Faculdade de Ciências, Universidade do Porto, Porto, Portugal}
+\affiliation{CIQUP, Centro de Investigação em Qu{\'i}mica, Departamento de Qu{\'i}mica e Bioqu{\'i}mica, Faculdade de Ci{\^e}ncias, Universidade do Porto, Porto, Portugal}
 
 \author{Thomas J. Piggot}
-\affiliation[University of Southampton]{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, United Kingdom}
+\affiliation{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, United Kingdom}
 
@@
 
-\title{ Supporting Information:\\Conformational plasticity of lipid headgroups in cellular membranes and protein--lipid complexes}
+\title{ Supporting Information:\\Inverse conformational selection in lipid-protein binding}
 

### 2021-07-12 manuscriptPGPE.tex
Small updates

@@
 in this biologically relevant phase are typically derived from NMR experiments, particularly from the
-C--H bond order parameters, $S_\mathrm{CH}=\frac{3}{2} \langle \cos^2 \theta -1 \rangle$, where $\theta$ is the angle between C-H bond and membrane normal, and the average is taken over the conformational ensemble of lipids~\cite{seelig77c,davis83,Semchyschyn04}.
+C--H bond order parameters, $S_\mathrm{CH}=\frac{1}{2} \langle 3\cos^2 \theta -1 \rangle$, where $\theta$ is the angle between C-H bond and membrane normal, and the average is taken over the conformational ensemble of lipids~\cite{seelig77c,davis83,Semchyschyn04}.
 Because these order parameters are similar to those in living cells \cite{gally81,scherer87,seelig90},
@@
      \textbf{C)} Chemical structure of POPC and POPS.
-    \textbf{D)} Headgroup and glycerol backbone C--H bond order parameters $S_mathrm{CH}$ for different phospholipids measured in the lamellar liquid disordered phase.
+    \textbf{D)} Headgroup and glycerol backbone C--H bond order parameters $S_\mathrm{CH}$ for different phospholipids measured in the lamellar liquid disordered phase.
     Filled squares show $S_{\rm{CH}}$ magnitudes and signs determined by $^1$H-$^{13}$C NMR spectroscopy for POPE (310~K) and POPG (298~K), as
@@
 lipids could not be resolved with the currently available force fields and experimental data
-(Figs.~\ref{HGorderparametersPCvsPEPG} and~\ref{HGorderparametersPGvsPCchange}, and Ref.~\cite{melcrova16,antila19,melcr20}).
+(Figs.~\ref{HGorderparametersPCvsPEPG} and~\ref{HGorderparametersPGvsPCchange}, and Ref.~\cite{antila19,melcr20}).
 

### 2021-07-21 manuscriptPGPE.tex
Update to the submitted revised version
Update manuscriptPGPE.tex

--- 2021-07-07 manuscriptPGPE.tex
@@
-\documentclass[journal=jpcbfk,manuscript=article]{achemso}
+%% ****** Start of file aiptemplate.tex ****** %
+%%
+%%   This file is part of the files in the distribution of AIP substyles for REVTeX4.
+%%   Version 4.1 of 9 October 2009.
+%%
+%
+% This is a template for producing documents for use with 
+% the REVTEX 4.1 document class and the AIP substyles.
+% 
+% Copy this file to another name and then work on that file.
+% That way, you always have this original template file to use.
 
-\usepackage[version=3]{mhchem}
-\usepackage[T1]{fontenc}
-\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
-%\newcommand{\todo}[1]{\textcolor{red}{#1}}
+%\documentclass[aip,graphicx]{revtex4-1}
+%\documentclass[aip,reprint]{revtex4-1}
+
+%\usepackage{graphicx}
+
+%\draft % marks overfull lines with a black rule on the right
+%\documentclass[pre,aps,floatfix,authordate1-4,twocolumn]{revtex4-1}
+%\documentclass[pre,aps,floatfix,authordate1-4]{revtex4-1}
+
+\documentclass[aps,prl,superscriptaddress,twocolumn]{revtex4}
+
+%\documentclass[aps,prl,preprint,groupedaddress]{revtex4}
+
+\usepackage{rotating} 
+\usepackage{times}
+\usepackage{graphicx}
+\usepackage{setspace}
+\usepackage{amsmath}
+\usepackage{epstopdf}
 \usepackage[obeyFinal]{easy-todo}
-\usepackage{rotating}
-\usepackage{upgreek}				
-\usepackage{xcolor}
-\usepackage{booktabs}
-\usepackage{multirow}
-\usepackage{lmodern}
-\usepackage{microtype}
+\usepackage{csquotes}
 \usepackage{xr-hyper}
-\usepackage{soul} % for highlights with \hl{} 
 \usepackage{hyperref} % to make TOC titles clickable
@@
 
+\begin{document}
+
+% Use the \preprint command to place your local institutional report number 
+% on the title page in preprint mode.
+% Multiple \preprint commands are allowed.
+%\preprint{}
+
+\title{NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids} %Title of paper
+
+% repeat the \author .. \affiliation  etc. as needed
+% \email, \thanks, \homepage, \altaffiliation all apply to the current author.
+% Explanatory text should go in the []'s, 
+% actual e-mail address or url should go in the {}'s for \email and \homepage.
+% Please use the appropriate macro for the type of information
+
+% \affiliation command applies to all authors since the last \affiliation command. 
+% The \affiliation command should follow the other information.
+
 \author{Am{\'e}lie Bacle}
-\affiliation{\tiny Laboratoire Coop{\'e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit{\'e} de Poitiers, 1 rue Georges Bonnet, 86000 Poitiers, France }
+\affiliation{Paris, France}
 
 \author{Pavel Buslaev}
-\affiliation{Nanoscience Center and Department of Chemistry, University of Jyv{\"a}skyl{\"a}, P.O. Box 35, 40014 Jyv{\"a}skyl{\"a}, Finland}
-\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, 141701 Dolgoprudny, Russia}
+\affiliation{Nanoscience Center and Department of Chemistry, University of Jyv{\"a}skyl{\"a}, Finland}
+\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
 
@@
 \affiliation{Center for Research in Biological Chemistry and Molecular Materials (CiQUS), Universidade de Santiago de Compostela, E-15782 Santiago de Compostela, Spain}
-\affiliation{CIQUP, Centro de Investigação em Qu{\'i}mica, Departamento de Qu{\'i}mica e Bioqu{\'i}mica, Faculdade de Ci{\^e}ncias, Universidade do Porto, Porto, Portugal}
+\affiliation{CIQUP, Centro de Investigação em Química, Departamento de Química e Bioquímica, Faculdade de Ciências, Universidade do Porto, Porto, Portugal}
 
 \author{Fernando Favela-Rosales}
-\affiliation{Departamento de Ciencias B\'{a}sicas, Tecnol\'{o}gico Nacional de M\'{e}xico - ITS Zacatecas Occidente, M\'{e}xico}
+\affiliation{Departamento de Ciencias B\'{a}sicas, Tecnol\'{o}gico Nacional de M\'{e}xico, Campus Zacatecas Occidente, M\'{e}xico}
 
-\author{Tiago M. Ferreira}
-\affiliation{NMR group - Institute for Physics, Martin Luther University Halle-Wittenberg, 06120 Halle (Saale), Germany}
+\author{Tiago Ferreira}
+\affiliation{Halle, Germany}
 
-\author{Patrick F.J. Fuchs}
-\affiliation{Sorbonne Universit{\'e}, Ecole Normale Sup{\'e}rieure, PSL University, CNRS, Laboratoire des Biomol{\'e}cules (LBM), 75005 Paris, France}
-\affiliation{Universit{\'e} de Paris, UFR Sciences du Vivant, 75013, Paris, France}
+\author{Patrick Fuchs}
+\affiliation{Paris, France}
 
 \author{Ivan Gushchin}
-\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, 141701 Dolgoprudny, Russia}
+\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
 
@@
 \author{Jesper J. Madsen}
-\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
-\affiliation{Global and Planetary Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
+\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
+\affiliation{Global and Planetary Health, College of Public Health, University of South Florida, Tampa, Florida 33612, United States of America}
 
 \author{Josef Melcr}
+\affiliation{Institute of Organic Chemistry and Biochemistry of the 
+Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
 \affiliation{Groningen Biomolecular Sciences and Biotechnology Institute 
@@
 
-\author{Paula Mil{\'a}n Rodr{\'i}guez}
-\affiliation{Sorbonne Universit{\'e}, Ecole Normale Sup{\'e}rieure, PSL University, CNRS, Laboratoire des Biomol{\'e}cules (LBM), 75005 Paris, France}
+\author{Paula Milan Rodriguez}
+\affiliation{Paris, France}
 
 \author{Markus S. Miettinen}
+% \affiliation[Max Planck Institute of Colloids and Interfaces]{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
 \affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
-% \affiliation[Max Planck Institute of Colloids and Interfaces]{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
 
 \author{O. H. Samuli Ollila}
-\email{samuli.ollila@helsinki.fi}
+\email[]{samuli.ollila@helsinki.fi}
 \affiliation{Institute of Biotechnology, University of Helsinki}
@@
 \author{Chris G. Papadopoulos}
-\affiliation{Universit{\'e} Paris-Saclay, CEA, CNRS, Institute for Integrative Biology of the Cell (I2BC), 91198 Gif-sur-Yvette, France}
+\affiliation{I2BC - University Paris Sud}
 
 \author{Antonio Pe{\'o}n}
-\affiliation{CIQUP, Centro de Investigação em Qu{\'i}mica, Departamento de Qu{\'i}mica e Bioqu{\'i}mica, Faculdade de Ci{\^e}ncias, Universidade do Porto, Porto, Portugal}
+\affiliation{Spain}
 
@@
 
-\author{Salla I. Virtanen}
-\affiliation{Institute of Biotechnology, University of Helsinki}
+% Collaboration name, if desired (requires use of superscriptaddress option in \documentclass). 
+% \noaffiliation is required (may also be used with the \author command).
+%\collaboration{}
+%\noaffiliation
 
-\title{Inverse conformational selection in lipid-protein binding} %Title of paper
-
-\begin{document}
-
-\begin{tocentry}
-
-     \includegraphics[height=3.5cm]{./Figs/TOC2.eps}
-%Some journals require a graphical entry for the Table of Contents.
-%This should be laid out ``print ready'' so that the sizing of the
-%text is correct.
-
-%Inside the \texttt{tocentry} environment, the font used is Helvetica
-%8\,pt, as required by \emph{Journal of the American Chemical
-%Society}.
-
-%The surrounding frame is 9\,cm by 3.5\,cm, which is the maximum
-%permitted for  \emph{Journal of the American Chemical Society}
-%graphical table of content entries. The box will not resize if the
-%content is too big: instead it will overflow the edge of the box.
-
-%This box and the associated title will always be printed on a
-%separate page at the end of the document.
-
-\end{tocentry}
+\date{\today}
 
@@
 % insert abstract here
-  Interest in lipid interactions with proteins and other biomolecules is emerging not only in
-  fundamental biochemistry, but also in the field of nanobiotechnology where lipid nanoparticles
-  are commonly used, for example, as carriers in mRNA vaccines.
-  The outward facing components of lipid bilayer, the lipid headgroups, regulate how cellular membranes or lipid nanoparticles
-  interact with approaching substances, such as proteins, drugs, RNA or viruses.
-  %In the physiologically relevant liquid state,
-  %lipid molecules are not fixed to a single structure but sample an ensemble of conformations with Boltzmann-weighted probabilities.
-  Because conformational ensembles of lipid headgroups in physiologically relevant liquid state 
-  %However, the distributions of conformations in such ensembles and their lipid-type dependence
-  have not yet been experimentally determined, a crucial piece has been missing from the understanding
-  on how they interact with other biomolecules: 
-  It is not clear if lipid molecules exchange between a few rigid structures,
-  or fluctuate freely across a practically continuous spectrum of  conformations.
-  Here, we combine solid-state NMR experiments and molecular dynamics simulations from the NMRlipids Project to resolve the conformational ensembles of headgroups of four key lipid types in various biologically relevant conditions.
-We find that all lipid headgroup types sample a wide range of
-  overlapping conformations % a wide range of conformations
-  in both neutral and charged cellular membranes, and that differences in the headgroup chemistry manifests only
-  in probability distributions of conformations. %, but all four lipid types can access .
-  Together with the analysis of 894 protein-bound lipid structures from the Protein Data Bank (PDB),
-  this suggests that all four key lipid types can bind to proteins in a wide range of conformations,
-  which are not limited by the headgroup chemistry.
-  %Therefore, the selective adsorption of peripheral proteins to membranes and the binding of lipids to transmembrane proteins are likely to be regulated by specific protein--lipid interactions, not by any inherent conformational restrictions of the lipids.
-  %Our results pave the way to comprehensive understanding of lipid--protein interactions in biomedical applications, in particular in
-  %lipid-mediated signaling.
-  Our results propose that lipids are able to select a suitable headgroup conformation
-  from the wide range available to them to fit the various binding sites in proteins.
-  The proposed {\it Inverse Conformational Selection Model} will extend also to lipid binding
-  to other targets than proteins, such as drugs, RNA and viruses.
+  Chemistry of lipid headgroups, the water facing components of cell membranes that regulate cell functions via lipid--protein interactions, varies between organisms and organelles.
+Because membranes are in liquid state under physiological conditions,
+  individual lipids are not fixed to single conformations but they sample an ensemble of conformations with the Boltzmann weighted probabilities.
+  However, these ensembles and their dependence on lipid types have not yet been experimentally determined. 
+  Therefore, it has not been clear if lipid molecules exchange between few rigid conformations,
+  or if they can freely fluctuate in all possible conformations.
+  Here, we combine solid state NMR experiments and molecular dynamics simulations from the NMRlipids open collaboration to resolve the conformational ensembles of the headgroups of key lipid types in their liquid lamellar phase under various biologically relevant conditions.
+Interpretation of NMR experiments using the plethora of simulation data collected in the NMRlipids project suggests that all lipid headgroups sample a wide range of conformations in neutral and charged cellular membranes.
+  Differences in the headgroup chemistry between different lipid types manifest
+  in probability distributions of conformations, but all lipid types can access almost any of the possible conformations.
+  Together with the analysis of protein-bound lipids from the protein data bank (PDB), this suggests that lipids can bind to proteins in wide range of conformations independently on the headgroup chemistry.
+Therefore, the selective adsorption of proteins to membranes is likely regulated by specific protein--lipid interactions rather than conformational restrictions of the bilayer lipids.
+  Our results pave the way to comprehensive understanding of
+  lipid mediated signaling and lipid--protein interactions in biomedical applications.
 
@@
 
-\maketitle
+%\pacs{}% insert suggested PACS numbers in braces on next line
 
+\maketitle %\maketitle must follow title, authors, abstract and \pacs
+
+% Body of paper goes here. Use proper sectioning commands. 
+% References should be done using the \cite, \ref, and \label commands
+
+%\label{}
 \section{Introduction}
 
-Lipid interactions with other biomolecules is gaining interest
-not only in molecular cell biology \cite{harayama18}, but also in
-nanobiotechnology applications, such as design of antibodies \cite{vigant15}
-and carriers for mRNA vaccines \cite{pardi18,schoenmaker21}.
-Interactions of cellular membranes or lipid nanoparticles with other biomolecules,
-such as proteins, drugs or RNA, are regulated by lipid headgroups,
-the outward facing components of lipid bilayers \cite{harayama18,vanmeer08}.
-Chemical compositions of lipid headgroups vary between different
-organelles and organisms, and specific interactions with certain lipid headgroups
-are known to be essential for the function of several proteins \cite{harayama18,vanmeer08}.
-However, the driving forces for specific interactions between lipids and membrane binding
-substances are not fully understood because conformational ensembles of lipids in
-physiologically relevant liquid state have not been experimentally determined.
-Therefore, it is not clear if the specificity arises from the
+Chemical compositions of hydrophilic lipid headgroups vary between different
+organelles and organisms, and different lipid types
+regulate protein functions in many different ways \cite{lee03,vanmeer08}.
+Lipids can directly bind to proteins or indirectly affect protein
+functions by altering membrane properties such as charge or elasticity \cite{lee03,lemmon08}.
+Specific interactions with certain lipid headgroups
+are known to be essential for the function of several proteins \cite{lee11,lemmon08},
+but it is not clear if the specificity is driven by the
 differences in accessible conformations between lipid types or
-from specific intermolecular lipid--protein interactions.
+by specific lipid--protein interactions.
 
-Structures of protein-bound lipids are available in the Protein Data Bank (PDB, \url{http://www.rcsb.org/}) \cite{berman00},
+Stuctures of protein-bound lipids are available in the protein data bank (PDB) \cite{berman00}
 and crystal structures of lipids have been determined \cite{buldt81,pascher92},
-but their relation to lipids in bulk membranes in the liquid lamellar phase remains unclear \cite{marsh13b}.
-Similarly to disordered proteins~\cite{sormanni17}, lipid molecules sample conformational ensembles which are composed of sets of individual structures whose probabilities of occurrences are determined by the Boltzmann distribution.
-In liquid bilayer state, individual phosphatidylcholine lipid samples the conformational ensemble within nanosecond timescale~\cite{ferreira15}.
-The most accurate experimental information on conformational ensembles of lipids
-in this biologically relevant phase are typically derived from NMR experiments, particularly from the
-C--H bond order parameters, $S_\mathrm{CH}=\frac{3}{2} \langle \cos^2 \theta -1 \rangle$, where $\theta$ is the angle between C-H bond and membrane normal, and the average is taken over the conformational ensemble of lipids~\cite{seelig77c,davis83,Semchyschyn04}.
-Because these order parameters are similar as in living cells \cite{gally81,scherer87,seelig90},
-the model membranes can be used to resolve the lipid headgroup conformations in biological conditions.
-According to these experiments, the glycerol backbone conformations are largely similar irrespective of the headgroup \cite{gally81}, and
-the headgroup conformations are similar in the phosphatidylcholine (PC), phosphatidylethanolamine (PE), and phosphatidylglycerol (PG) lipids,
-whereas the phosphatidylserine (PS) headgroup is more rigid \cite{wohlgemuth80,buldt81}. 
-Notably, however, the signs of $S_\mathrm{CH}$ are not accessible from $^2$H\,NMR experiments \cite{ollila16},
+but their relation to the conformational ensembles in bulk membranes in the liquid lamellar phase remains unclear \cite{marsh13b}.
+%In addition to the changes in lipid conformational ensembles upon binding to proteins,
+%also the experimentally measured response of lipid headgroup to membrane bound charges remains poorly understood 
+%due to the lack of suitable models to interpret the lipid conformational ensembles in liquid lamellar state \cite{Semchyschyn04}.
+Most accurate experimental information on conformational ensembles of lipids
+%in the physiologically relevant
+in liquid lamellar phase are typically derived from NMR experiments, particularly from 
+C--H bond order parameters measured using $^2$H NMR~\cite{seelig77c,davis83,Semchyschyn04}.
+According to these experiments, the glycerol backbone conformations are largely similar irrespectively of the headgroup \cite{gally81}, 
+the headgroup conformations are similar in  phosphatidylcholine (PC), phosphatidylethanolamine (PE) and phosphatidylglycerol (PG) lipids,
+while the headgroup of phosphatidylserine (PS) lipids is more rigid \cite{wohlgemuth80,buldt81}. 
+However, order parameters signs are not accessible in $^2$H NMR experiments \cite{ollila16}
 and universal models to map order parameters to structural ensembles are not available \cite{pezeshkian18,akutsu20}.
-Therefore, it is not clear if the lipid headgroups in a liquid lamellar bilayer exchange between a few restricted conformations, or can fluctuate freely across a wide conformational range.
+Therefore, it is not clear if the conformational ensemble of lipid headgroups in liquid lamellar bilayer
+is composed of exchange between few restricted conformations, or if lipid headgroups can fluctuate freely across a wide range of conformations.
 
-Here, we use natural abundance $^1$H-$^{13}$C solid-state NMR experiments and MD simulations from the NMRlipids Project
-to resolve differences in the conformational ensembles of PC, PE, PG, and PS lipid headgroups.
-Zwitterionic PC are the most common lipids in eukaryotes and PE in bacteria \cite{vanmeer08,sohlenkamp16};
+Here, we use natural abundance $^{13}$C NMR experiments %(enabling the determination of order parameter signs)
+and MD simulations from the NMRlipids open collaboration
+to resolve the differences in the conformational ensembles between PC, PE, PG and PS lipid headgroups.
+We elucidate also the effect of charges and protein binding to lipid headgroup conformations.
+Zwitterionic PC and PE are the most common lipids in eukaryotes and bacteria, respectively \cite{vanmeer08,sohlenkamp16}.
 PE is also the second most abundant glycerophospholipid in eukaryotic cells
 and has been related to various diseases \cite{vance15,calzada16,patel17}.
-PC lipids are also used in mRNA COVID-19 vaccines \cite{schoenmaker21}.
-PS are the most common negatively charged lipids in eukaryotes and PG in bacteria,
-and both are known to affect membrane protein functionality and signaling \cite{lemmon08,leventis10,sohlenkamp16,hariharan18}.
-All the four studied lipid types specifically bind to various proteins \cite{yeagle14};
-we therefore elucidate also the effect of protein binding to headgroup conformations.
-%These interactions are crucial, for example, in lipid-mediated signaling \cite{lemmon08} and for the
-%design of phoshopolipid-specific antibodies \cite{vigant15}. 
-The resulting {\it inverse conformational selection model} for lipid-protein binding can be also
-applied to understand interactions between membranes and other biomolecules than proteins, such as drugs, RNA, or viruses.
-On the other hand, the detailed understanding of lipid bilayer interfaces can facilitate the design
-of lipid nanoparticle carriers for mRNA vaccines with less side effects.
+PS and PG are the most common negatively charged lipids in eukaryotes and bacteria, respectively,
+and their presence affect membrane protein functionality and signaling \cite{lemmon08,leventis10,sohlenkamp16,hariharan18}.
+All the studied lipids specifically bind to various proteins \cite{yeagle14}.
+%We use our results to elucidate also lipid--protein interactions and the effect of charges on lipid conformations.
+
+Similarity of glycerol backbone and headgroup order parameters in model membranes and bacteria \cite{gally81,scherer87,seelig90}
+suggest that our results can be used to 
+%the results from model systems could be used to
+understand the biological role of lipid headgroups in lipid--protein interactions.
+These interactions are crucial, for example, in lipid mediated signaling \cite{lemmon08} and
+design of phoshoplipid-specific antibodies \cite{vigant15}. 
+%The lipid conformational ensembles in liquid lamellar state paves the way toward understanding the
+%specific binding of different lipid types to membrane proteins and how they regulate the protein function.
+
+%Notably, such measurements can be performed also on living cells \cite{gally81,scherer87,seelig90}.
+
+\section{Methods}
+\subsection{Experimental C--H bond order parameters}
+The headgroup and glycerol backbone C--H bond order parameters of 1-palmitoyl-2-oleoyl-sn-glycero-3-phosphoethanolamine (POPE) and 1-palmitoyl-2-oleoyl-sn-glycero-3-phospho-(1'-rac-glycerol) (POPG), purchased from Avanti polar lipids, were measured using natural abundance $^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
+The magnitudes of order parameters were determined from the chemical-shift resolved dipolar splittings using a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04}, and the signs from S-DROSS experiments~\cite{gross97} combined with SIMPSON simulations \cite{bak00}.
+%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
+%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
+The NMR experiments were identical as in our previous work~\cite{antila19}.
+The POPE experiments were recorded at 310~K and POPG experiments at 298~K, where the bilayers are in the liquid disordered phase \cite{marsh13}.
+
+Glycerol backbone peaks from both lipids, and the $\alpha$-carbon peak from POPE in the INEPT spectra 
+were assigned based on previously measured POPC spectra~\cite{ferreira13}.
+The $\beta$-carbon peak from POPE was assigned based on $^{13}$C chemical shift table for amines available
+at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm}.
+\todo{How were $\alpha$ and  $\gamma$-carbon peaks assigned in POPG?}.
+The $\beta$-carbon peak from POPG overlapped with the g$_2$ peak from glycerol backbone
+because their chemical environments are similar.
+\todo{Details to be checked by Tiago}.
+
+\subsection{Molecular dynamics simulations}
+
+Molecular dynamics simulation data were collected and analyzed using
+the methods from the NMRlipids Open Collaboration project (\url{nmrlipids.blogspot.fi}) \cite{botan15,catte16,ollila16,antila19}.
+%\url{github.com/NMRlipids/NMRlipidsIVotherHGs}. 
+Simulation details, accessibility information and quality evaluation of more than 70 systems simulated for this work are in the supplementary information.
+
+Best models for the interpretation of lipid headgroup conformational ensembles from the experimental data were selected
+using quality evaluation measures defined in the NMRlipids project.
+Conformational ensembles of headgroup and glycerol backbone in PE and PG simulations were evaluated using the C-H bond order parameters \cite{botan15}.
+Interactions between different headgroups were evaluated by monitoring the changes in headgroup order parameters upon mixing the lipids \cite{antila19}.
+The ion binding affinities and response of lipids to bound charge were evaluated by monitoring the changes in lipid headgroup order parameters \cite{catte16,antila19}.
+
+%\clearpage
+\begin{figure*}[]
+  \centering
+%  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
+  %  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCPSPEPG.eps}
+   \includegraphics[width=18.0cm]{./Figs/figure1.pdf}
+   \caption{\label{HGorderParameters}
+     Chemical structure, refocused-INEPT spectrum, 2D R-PDLF spectra, and S-DROSS data (from top to bottom) of \textbf{A)} POPE  and \textbf{B)} POPG.
+      Full NMR spectra are shown Figs. \ref{POPEspectra} and \ref{POPGspectra}.
+     \textbf{C)} Chemical structure of POPC and POPS.
+    \textbf{D)} Headgroup and glycerol backbone order parameters from different experiments in lamellar liquid disordered phase.
+    The values and signs for POPE (310~K) and POPG (298~K)
+    measured in this work, and for POPS (298~K) \cite{antila19} and POPC (300~K) \cite{ferreira13,ferreira16}
+    previously using $^{13}$C NMR. The literature values for
+    DOPS with 0.1M of NaCl (303~K) \cite{browning80},
+    POPG with 10nM PIPES (298~K) \cite{borle85},
+    DPPG with 10mM PIPES and 100mM NaCl (314~K) \cite{wohlgemuth80}, 
+    DPPE (341~K) \cite{seelig76},
+    E.coliPE and E.coliPG (310~K) \cite{gally81}
+    are measured using $^2$H NMR. The signs from $^{13}$C NMR are used also for the literature values.
+   }
+   %\todo{This is a sketch, Tiago Ferreira will make a new figure.} \\
+%  \todo{D) could be clarified as Fig. 2 in the NMRlipids IVps paper.}
+\end{figure*}
+
+\subsection{Analysis of protein-bound lipid conformations}
+Lipid structures from Protein Data Bank (PDB, \url{http://www.rcsb.org/})
+were searched using PDBe REST API (\url{www.ebi.ac.uk/pdbe/pdbe-rest-api})
+using the ligand names listed in the supplementary information.
+Heavy atom dihedral distributions from the lipid structures were calculated
+using the MDAnalysis python library \cite{agrawal11,gowers16} and
+Jupyter notebook available from \url{https://github.com/pbuslaev/scr/blob/master/PDB%20analysis.ipynb}.
+Only structures from X-ray and Cryo-EM with higher than 3.2~\AA~resolution were used in the analysis.
+Lipids from same PDBID with all dihedrals angles within 3 degrees were discarded from the analysis
+to avoid overweighting of lipid structures present, e.g., due to rotational symmetries.
+
+To demonstrate the existence of similar structures of different lipids bound to different proteins,
+we searched for pairs having the last five dihedral angles (excluding g$_1$-g$_2$-g$_3$-O$_{g3}$) within
+30 degrees from structures with resolution higher than 2.5~\AA. The two most representative examples
+were handpicked from the results.
 
@@
 
-\subsection{Differences between lipid headgroups from $^{13}$C NMR experiments}
-
-\begin{figure*}[]
+\begin{figure*}[bt]
   \centering
-   \includegraphics[width=\textwidth]{./Figs/figure_PG_PE.pdf}
-   \caption{\label{HGorderParameters}
-     Chemical structure, refocused-INEPT spectrum, 2D R-PDLF spectra, and S-DROSS dipolar slices (from top to bottom) for \textbf{A)} POPE  and \textbf{B)} POPG MLVs measured by $^1$H-$^{13}$C solid-state NMR.
-The S-DROSS dipolar slices show both experimental data (open circles) and the result of NMR numerical simulations (blue solid lines).
-     Full NMR spectra are shown in Figs. \ref{POPEspectra} and \ref{POPGspectra}.
-     \textbf{C)} Chemical structure of POPC and POPS.
-    \textbf{D)} Headgroup and glycerol backbone C-H bond order parameters for different phospholipids measured in the lamellar liquid disordered phase.
-    Filled squares show $S_{\rm{CH}}$ magnitudes and signs determined by $^1$H-$^{13}$C NMR spectroscopy for POPE (310~K) and POPG (298~K), as
-    measured in this work, and from data previously reported for POPS (298~K) \cite{antila19} and POPC (300~K) \cite{ferreira13,ferreira16}.
-Empty symbols show headgroup and glycerol backbone $S_{\rm{CH}}$ magnitudes measured previously with $^2$H NMR and are also plotted as real values using the signs determined in this work for
-    POPG with 10nM PIPES (298~K) \cite{borle85},
-    DPPG with 10mM PIPES and 100mM NaCl (314~K) \cite{wohlgemuth80}, 
-    DPPE (341~K) \cite{seelig76},
-    E. coli PE and E. coli PG (310~K) \cite{gally81}
-    .
+   \includegraphics[width=18.0cm]{./Figs/figure2.eps}
+   \caption{\label{structures}
+     Results from the best simulation model (CHARMM36) simulations demonstrating the differences in conformational ensembles between different lipids. 
+     \textbf{A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurence of different conformations.
+     \textbf{B)} Headgroup and glycerol backbone region order parameters of different different lipids.
+     \textbf{C)} Relative energies for individual heavy atom dihedral angles estimated from the inverse Boltzmann formula.
+     Angles corresponding energies above 7 kT are not shown because they are not observed in simulations.
+     \textbf{D)} Distributions of P-N vector angle with respect to membrane normal.
   }
@@
 
+\subsection{Differences between lipid headgroups from $^{13}$C NMR experiments}
+
 To experimentally characterize the differences in headgroup conformational ensembles of lipids that are not
-bound to proteins, we measured the C--H bond order parameters
+bound to proteins in an electrostatically neutral cell membrane, we measured the C--H bond order parameters
 and their signs of POPG and POPE in the liquid lamellar phase, as we did previously for POPC and POPS \cite{ferreira13,ferreira16,antila19}.
@@
 The decrease in the beginning of the S-DROSS curve suggests that the sign of larger g$_2$ order parameter
-is negative and later increase suggests that sign of smaller $\beta$ order parameter is positive as confirmed by the SIMPSON simulations (Fig.~\ref{HGorderParameters}B).
+is negative and later increase suggests that sign of smaller $\beta$ order parameter is positive (Fig.~\ref{HGorderParameters}B).
+This interpretation is confirmed by SIMPSON calculations in Fig.~\ref{POPGsimpson}.
 
-Experimental order parameters of POPC, POPE, POPG and POPS glycerol backbones and headgroups from this and previous studies are displayed in Fig.~\ref{HGorderParameters}D, where the signs from $^{13}$C NMR experiments are used also for the $^2$H NMR data from the literature.
-The overall agreement of order parameters determined by different research teams and different techniques for the same lipid headgroup is excellent (within $\pm$0.02) here and in previous studies \cite{botan15,ollila16,antila19}.
+Experimental order parameters of POPC, POPE, POPG and POPS glycerol backbones and headgroups from this and previous studies are collected in Fig.~\ref{HGorderParameters}D, where signs determined from $^{13}$C NMR experiments are used also for the $^2$H NMR data from the literature.
+The overall agreement of order parameters determined by different research teams and different techniques for the same lipid headgroup is very good here and in previous studies \cite{botan15,ollila16,antila19}.
 Therefore, the differences between lipid types are dictated by the headgroup chemistry rather than inaccuracies in experiments, differences in the acyl chains, or in experimental conditions.
 
-The most distinct order parameters are observed for PS headgroups, for which the $\alpha$-carbon order parameter exhibits significant forking (different values for different hydrogens bound to the same carbon) and the $\beta$-carbon has a more negative value than other studied lipid types.
+The most distinct order parameters are observed for PS headgroups, for which the $\alpha$-carbon order parameter exhibits significant forking (different values for different hydrogens in the same carbon) and the $\beta$-carbon has more negative value than other studied lipid types.
 On the other hand, the $\beta$-carbon order parameter of PG headgroup has a positive sign, in contrast to all the other lipid types.
-The sign has not been previously known from $^2$H NMR experiments which enable detecting only order parameter absolute values~\cite{wohlgemuth80,gally81,borle85}.
+This has not been previously observed  in $^2$H NMR experiments detecting only the absolute values~\cite{wohlgemuth80,gally81,borle85}.
 The glycerol backbone order parameters are similar for all the lipid types, although they move slightly toward positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG.
@@
 
-\begin{figure*}
+\begin{figure*}[bt]
   \centering
-   \includegraphics[width=\textwidth]{./Figs/figure2.eps}
-   \caption{\label{structures}
-     Results from the best simulation model demonstrating the differences in conformational ensembles between different lipids.
-     The best simulation model (CHARMM36) was selected from the comparison of ten different force fields based on quality evaluation against experimental $S_{\rm{CH}}$ values in Figs.~\ref{HGorderParametersPE} and~\ref{HGorderParametersPOPG}.
-     \textbf{A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurrence of different conformations.
-     \textbf{B)} Headgroup and glycerol backbone region order parameters of different lipids. Colored shaded regions show the range of experimental values for POPC from Fig. \ref{HGorderParameters}.
-     \textbf{C)} Relative free energies for individual heavy atom dihedral angles estimated from the inverse Boltzmann formula.
-     Angles corresponding to free energies above 8~$k_\mathrm{B}T$ are not shown because they are not observed in simulations.
-     \textbf{D)} Distributions of P--N/C$_\gamma$ vector angle with respect to membrane normal.
+  \includegraphics[width=18.0cm]{./Figs/HGorderparametersPCvsSURFchangeDIHEDRALSlog.eps}
+  \caption{\label{changesWITHsurf}
+    \textbf{A)} Modulation of PC headgroup order parameters and P--N vector angle upon addition of cationic surfactant (dihexadecyldimethylammonium)
+    from CHARMM36 simulations compared with the experimental data \cite{scherer89}.
+    \textbf{B)} Relative energies for individual dihedral angles estimated from inverse Boltzmann distributions of heavy atom dihedral angles
+    with different amounts of cationic surfactant from CHARMM36 simulations.
   }
@@
 
-Ideal
-MD simulations would correctly predict the conformational ensembles of lipids, thereby giving an interpretation for  the experimental C-H bond order parameters, $S_\mathrm{CH}$, due to their direct connection~\cite{ollila16}.
-To find the best MD simulation models (force fields) for this purpose, we took advantage of the power of the NMRlipids Project, an open science collaboration that has gathered a massive number of simulation trajectories, allowing the evaluation of all the available force fields of PC~\cite{botan15}, PS~\cite{antila19}, PE (Fig.~\ref{HGorderParametersPE}), and PG (Fig.~\ref{HGorderParametersPOPG}) lipids against NMR data.
-The comparison of hundreds of simulation trajectories simulated with approximately 20 different force fields revealed that none of them can correctly capture the lipid conformational ensembles and reproduce the experimental $S_\mathrm{CH}$ of all segments within experimental error (for detailed discussion on the quality of force fields, see section S3 in the supplementary information and Refs.
-\citenum{botan15,ollila16,antila19}).
-However, CHARMM36, the force field showing the best agreement with experiments, correctly captures the essential differences between the PC, PE, PG, and PS headgroup order parameters, particularly the positive order parameter of $\beta$-carbon in PG and the forking of $\alpha$-carbon in PS, see Fig.~\ref{structures} B.
-
-To understand the structural origin of distinct order parameters for the distinct lipids,
-we first calculated the heavy atom dihedral angle distributions from the best performing simulations that reproduced the differences between headgroups~(Fig. \ref{DIHdists}).
-Then, we used the inverse Boltzmann formula to estimate the free
+To understand the structural origin of distinct order parameters for PG and PS lipids,
+we first calculated the heavy atom dihedral angle distributions from simulations
+that best reproduce the differences between headgroups according to the quality
+evaluation in the SI and Refs. \cite{botan15,antila19} (Fig. \ref{DIHdists}).
+Then, we used the inverse Boltzmann formula to estimate the
 energy costs for different dihedral angle orientations.
-The results in Fig.~\ref{structures}C suggest that all lipid headgroups are very flexible and free
-energy costs for rotating individual dihedrals to almost any angle is low (below 8~$k_\mathrm{B}T$).
-Only $cis$ states of P-O$_\alpha$-C$_\alpha$-C$_\beta$ and g$_2$-g$_3$-O$_{g_3}$-P have larger relative free
+The results in Fig.~\ref{structures}C suggest that all lipid headgroups are very flexible and
+energy costs for rotating individual dihedrals to almost any angle is low (below 7 kT).
+Only $cis$ states of P-O$_\alpha$-C$_\alpha$-C$_\beta$ and g$_2$-g$_3$-O$_{g_3}$-P have larger relative
 energies and are not observed for any lipids during the simulations.
@@
 which prefer \textit{gauche$^-$} conformations for PG and \textit{gauche$^+$} for PS,
-whereas PC and PE exhibit symmetric distributions.
-Also, the free energy barriers for O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ dihedral
+while PC and PE exhibit symmetric distributions.
+Also, the energy barriers for O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ dihedral
 rotations between \textit{gauche} and \textit{trans} states are larger for
 PS and PG lipids than for PC and PE lipids. 
-The rest of the dihedrals are similar between different lipids,
-with the exception of PS lipids for which slightly larger free energy for \textit{eclipsed anti} conformation
+Rest of the dihedrals are similar between different lipids,
+with the exception of PS lipids for which slightly larger energy for ecplised anti conformation
 in g$_3$-O$_{{\mathrm g}3}$-P-O$_\alpha$ dihedral was observed.
-Therefore we suggest that the main differences between lipid headgroups leading to distinct order parameters occur in the headgroup end, while in the phosphate region they remain similar in different lipids with the exception of PS.
+Therefore we suggest that the main differences between lipid headgroups leading to distinct order parameter occur in the choline part, while also changes in phosphate region may contribute in PS lipids.
 The increased barriers for dihedral rotations may explain the more rigid headgroup structures in PS~\cite{browning80,buldt81}.
-Furthermore, the angle between headgroup dipole and membrane normal decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig.~\ref{structures}D).
+Furthermore, the angle between headgroup dipole and membrane normal decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig.~\ref{structures}D)).
 However, the differences between PC and PE in P-O$_\alpha$-C$_\alpha$-C$_\beta$ dihedral
-and P--N vector dipole may be artificial as the $\beta$-carbon order parameter in PC is too negative even in the best available force field~ \cite{botan15}, thereby not being equal to the order parameter in PE as observed in experiments.
+and P--N vector dipole may be artificial as the $\beta$-carbon order parameter in PC is too negative even in the best available force field, thereby not being equal to the order parameter in PE as observed in experiments \cite{botan15}.
 
-In conclusion, the rotation of dihedral angles to almost any position bears relatively low free energy cost and the sampled dihedral angles are within similar ranges for all studied headgroup types.
-In other words, all studied lipid headgroups sample very broad conformational ensembles in the liquid lamellar phase.
-Therefore, the lipid headgroups are able to adopt a wide range of conformations when interacting with proteins, ions, or other biomolecules.
-Furthermore, the structures in lipid crystals \cite{buldt81,pascher92} are not representative of the liquid state, and models aiming to explain NMR data using only a few conformations \cite{seelig77c,davis83,Semchyschyn04,akutsu20} are not sufficient to capture the large conformational space of lipids in the liquid lamellar state.
-However, conformational ensembles of lipids in MD simulation trajectories with accurate force fields that capture the experimental NMR order parameters and can be considered realistic, and thereby used to interpret the experimental data.
+In conclusion, all lipid headgroups sample very broad conformational ensembles in the liquid lamellar phase and the sampled dihedral angles are within approximately same ranges for all headgroup types.
+% (Fig.
+\ref{structures}).
+Because the rotation of dihedral angles to almost any position bears relatively low energy cost,
+%results suggest that all lipid headgroups are very flexible---
+%thereby being
+the lipid headgroups are able to adopt wide range of conformations when interacting with proteins, ions or other biomolecules.
+%Our results support the models proposing free rotation around phosphate group for PC lipids, which decouples the dynamics and conformations between acyl chains and headgroups \cite{klauda08c,antila21b}, and suggests that these models could be applied for all lipids containing the phosphorus group.
+%The wide range of observed conformations suggest that
+Therefore, the structures in lipid crystals \cite{buldt81,pascher92} play only a minor role, and models aiming to explain NMR data using only few conformations \cite{seelig77c,davis83,Semchyschyn04,akutsu20} are not sufficient to capture the large conformational space of lipids in the liquid lamellar state.
 
@@
 
-\begin{figure*}[bt]
-  \centering
-  \includegraphics[width=\textwidth]{./Figs/HGorderparametersPCvsSURFchangeDIHEDRALSlog.eps}
-  \caption{\label{changesWITHsurf}
-    \textbf{A)} Modulation of PC headgroup order parameters and P--N vector tilt (average angle between the P-N vector and the bilayer normal) upon addition of cationic surfactant (dihexadecyldimethylammonium)
-    from CHARMM36 simulations compared with the experimental data \cite{scherer89}.
-    \textbf{B)} Relative free energies for individual dihedral angles estimated from inverse Boltzmann distributions of heavy atom dihedral angles
-    with different amounts of cationic surfactant from CHARMM36 simulations.
-  }
-\end{figure*}
+Charged entities, such as lipids, proteins, surfactants, drugs, and ions incorporated in membranes can reorient the headgroup dipole in PC lipids and affect the order parameters of the lipid headgroups \cite{seelig87}.
+However, the detailed understanding of this structural and energetic response of lipids to membrane-bound electric charge is still lacking \cite{Semchyschyn04}.
 
-Charged entities, such as lipids, proteins, surfactants, drugs, and ions incorporated in membranes can change the orientation of the headgroup dipole in PC lipids and affect the order parameters of the lipid headgroups~\cite{seelig87}.
-However, the detailed understanding of the structural and energetic responses of lipids to membrane-bound electric charge is still lacking \cite{Semchyschyn04}.
-For example, it is not clear if the membrane-bound charges restrain the lipids into specific conformations or just redistribute the probabilities of states by keeping the accessible conformations similar.
-
-To resolve lipid headgroup conformational ensembles in cellular membranes bearing positive charge,
+To resolve lipid headgroup conformational ensembles in cell membranes bearing positive charge,
 we calculated the heavy atom dihedral angle distributions from 
-the subset of simulations 
+the subset of simulations %(CHARMM36)
 that correctly captured the experimentally measured decrease in PC headgroup order parameters upon addition of cationic surfactants into a bilayer (figure \ref{changesWITHsurf} A).
-The dihedral angle distributions and relative free energies
+The dihedral angle distributions and relative energies
 in figures \ref{HGorderparametersPCvsSURFdihedrals} and \ref{changesWITHsurf} B reveal that the
 addition of positive charge into a membrane 
-decreases the abundance of \textit{trans} states in g$_2$-g$_3$-O$_{g_3}$-P and g$_3$-O$_{g_3}$-P-O$_\alpha$, and \textit{cis} states in O$_{g_3}$-P-O$_\alpha$-C$_\alpha$ dihedrals.
-The choline region remains essentially unchanged and only minor changes are observed in the other dihedrals even at a surfactant molar fraction of 0.47.
+decreases the abundance of trans states in g$_2$-g$_3$-O$_{g_3}$-P and g$_3$-O$_{g_3}$-P-O$_\alpha$
+dihedrals.
+The choline region remains essentially unchanged and only minor changes are observed in other dihedrals even though almost half of the molecules in the membrane are cationic surfactants.
 
@@
 The bound Ca$^{2+}$ ion to PC headgroup leads to similar decrease in trans state probability for g$_3$-O$_{g3}$-P-O$_\alpha$ dihedral
-in the most realistic MD simulation models
+in the most realistic MD simulations models
 as observed for cationic surfactants
 (lipid17ecc and CHARMM36 in Figs.~\ref{DIHSwithCAlipid17eccPOPC}, \ref{DIHSwithCAcharmm36POPC} and Fig.~\ref{changesWITHCaClPG}).
-The dihedral distributions of the PG headgroup are more sensitive to the bound ions in the most realistic simulations,
-but upward tilting of the headgroup dipole upon the addition of CaCl$_2$ is weaker than in PC
+The dihedral distributions of PG headgroup are more sensitive to the bound ions in the most realistic simulations,
+but upward tilting of the headgroup dipole upon addition of CaCl$_2$ is weaker than in PC
 (Lipid17 and Slipids in Figs.~\ref{changesWITHCaClPG},~\ref{DIHSwithCAslipidsPOPG} and \ref{DIHSwithCAlipid17POPG}).
-However, the changes in PG lipid dihedrals upon the addition of CaCl$_2$ differ between the best models (Figs.~\ref{DIHSwithCAslipidsPOPG} and~\ref{DIHSwithCAlipid17POPG}), and none of the simulations captures the Ca$^{2+}$ ion binding affinity and conformational ensemble of PG lipids simultaneously.
-Moreover, experimental data to evaluate the response of $\alpha$-carbon order parameters to the added CaCl$_2$ in PG is not available.
-Additionally, the headgroup conformational ensembles in mixtures of PC and charged (PG or PS) or zwitterionic (PE)
-lipids could not be resolved with the currently available force fields and experimental data
-(Figs.~\ref{HGorderparametersPCvsPEPG} and~\ref{HGorderparametersPGvsPCchange}, and Ref.~\cite{melcrova16,antila19,melcr20}).
+However, the changes in PG lipid dihedrals upon addition of CaCl$_2$ differ between the best models (Figs.~\ref{DIHSwithCAslipidsPOPG} and~\ref{DIHSwithCAlipid17POPG}), none of the simulations captures the Ca$^{2+}$ ion binding affinity and conformational ensemble of PG lipids simultaneously, and experimental data to evaluate the response of $\alpha$-carbon order parameters to the added CaCl$_2$ in PG is not available.
+Also, the headgroup conformational ensembles in mixtures of PC and charged (PG or PS) or zwitterionic (PE)
+lipids could not be resolve with the currently available force fields and experimental data
+(Figs.~\ref{HGorderparametersPCvsPEPG} and~\ref{HGorderparametersPGvsPCchange}, and Ref. \cite{antila19,melcr20}).
 
-In conclusion, the MD simulation results suggest that the structural response of lipid headgroups to 
-membrane-bound charges arises
+%Despite the difficulties in simulations to capture the headgroup conformations in mixtures involving charged lipids,
+In conclusion, the structural response of lipid headgroups to 
+%experimentally observed changes in headgroup order parameters upon addition of
+membrane-bound charges arise
+%from relatively small changes in conformational ensembles. These changes eventuate
 from mild changes in dihedral angle distributions, rather than restriction of lipids to
@@
 
-\begin{figure*}[tb]
+\begin{figure*}[]
   \centering
-  \includegraphics[width=\textwidth]{./Figs/figure4.eps}
+  \includegraphics[width=19.0cm]{./Figs/figure4.eps}
   \caption{\label{dihedralsFROMpdb}
@@
     B) The structure of PE headgroup bound to cytochrome {\it bc}$_1$ complex (PDB ID: 1PP9 \cite{huang05})
-    with identical conformation as the PC headgroup bound to yeast Sec14 (PDB ID: 3B7Q \cite{schaaf08}),
+    with identical conformation as the PC headgroup bound to yeast sec14 (PDB ID: 3B7Q \cite{schaaf08}),
     and the structure of PG headgroup bound to bovine cytochrome {\it c} oxidase (PDB ID: 3AG3 \cite{muramoto10})
@@
 Interpretation of experimental order parameters using MD simulations
-in previous sections suggest that PC, PE, PG, and PS lipid headgroups are very flexible,
+in previous sections suggest that PC, PE, PG and PS lipid headgroups are very flexible,
 allowing them to bind proteins in various different conformations.
-To test this prediction, we analyzed the conformations of all protein-bound lipids 
+To test this prediction, we analyzed the protein-bound lipid conformations
 from structures deposited in the PDB \cite{berman00}.
-We found 311 PC, 394 PE, 154 PG, and 35 PS lipid conformations that were bound to different types of proteins 
-(including both integral and peripheral membrane proteins) and were determined as a part of protein structure using crystallography or cryo-EM.
-The full list of structures with bound lipids is available from {\it Data/PDBdistributions/pdbtable.pdf} in Ref.~\cite{NMRlipidsIVbgit}.
+We found 311 PC, 394 PE, 154 PG, and 35 PS protein-bound lipid conformations
+%presenting lipids that are tightly bound to proteins in fixed conformations
+that were determined as a part of protein structure using crystallography or cryo-EM.
 
 The heavy atom dihedral angle distributions calculated from these conformations 
-(Fig.~\ref{dihedralsFROMpdb} A) reveal that the protein-bound lipids indeed exhibit a wide
-range of conformations, independent of the headgroup type.
-As for bulk lipid bilayers (Figs.~\ref{structures}C, \ref{DIHdists}), only {\it cis} conformations
+(Fig.~\ref{dihedralsFROMpdb} A) reveal that the protein-bound lipids indeed exhibit wide
+range of conformations independently on the headgroup type.
+As for bulk lipid bilayers, only $cis$ conformations
 of P-O$_\alpha$-C$_\alpha$-C$_\beta$ and g$_2$-g$_3$-O$_{g_3}$-P dihedrals are almost
 completely absent in all lipids, and significant differences between
-different headgroups are not observed.
-Protein-bound lipid structures deviating from protein-free lipid crystal structures 
+different headgroups were not observed.
+Structures deviating from lipid crystals 
 have been previously proposed to indicate inaccuracies in lipid structures in PDB \cite{marsh13b,pezeshkian18}. 
-However, we observe large deviations from the protein-free lipid crystal structures also in conformational ensembles
+However, we observe large deviations from lipid crystals structures also in conformational ensembles
 that reproduce the NMR data in the liquid lamellar phase, indicating that such deviations
-are realistic also in the protein-bound states.
+are realistic also in protein-bound states.
+%The %range of available angles for
+%O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ and
+%g$_1$-g$_2$-g$_3$-O$_{g_3}$ dihedrals are even less restricted in protein bound state
+%than in liquid lamellar phase where $cis$ and $anti$ eclipsed state were not present (Fig.~\ref{structures}). 
+%Major differences between different lipid headgroups are not observed when bound to proteins.
+%Slight preference for the trans state in O$_\alpha$-C$_\alpha$-C$_\beta$-/C$_\gamma$ dihedral of PC,
+%and for positive angles in P-O$_\alpha$-C$_\alpha$-C$_\gamma$ dihedral of PS lipids with respect
+%to other lipids could be present in the data, but the statistics is not sufficient for conclusions.
+%Therefore, the differences in conformational ensembles between different
+%lipids in liquid lamellar state in Fig.~\ref{structures} are not seen in the protein bound states.
 
 Our results suggest that flexible lipid headgroups can optimize the intermolecular interactions with proteins by binding
-in a wide range of conformations.
-Therefore, the specific binding of lipids to proteins is not driven by differing structural preferences between headgroups.
-This is demonstrated in Fig.~\ref{dihedralsFROMpdb} B with two examples where different lipid types bound to
-different proteins have almost identical headgroup conformations:
-PE in cytochrome {\it bc}$_1$ complex is similar to PC bound to yeast Sec14,
+in wide range of conformations.
+Therefore, the specific binding of lipids to proteins is not driven by the structural differences between headgroups.
+This is demonstrated in Fig.~\ref{dihedralsFROMpdb} B with two examples where different lipids bound to
+different proteins have identical conformations:
+PE in cytochrome {\it bc}$_1$ complex is similar to PC bound to yeast sec14,
 and PG bound to bovine cytochrome {\it c} oxidase is similar to PC bound to pore-forming toxin (FraC).
 On the other hand, a single lipid headgroup type is capable of accommodating various binding positions and as such would 
-be able to specifically bind to many different kinds of binding sites.
+be able to specifically bind to many different kinds of binding sites in different proteins.
+
+%However, it is important to note that lipids are often not the main target in
+%structures deposited in PDB and therefore their conformations may be less reliable than those of proteins.
 
@@
 
-C--H bond order parameters $S_\mathrm{CH}$ from NMR experiments 
-suggest that lipid headgroup conformational ensembles depend on lipid type (PC, PE, PG, or PS) 
-and membrane charge (cationic lipids, surfactants, accumulated ions or drugs). 
-Our analysis of these data, using MD simulations collected
-within the NMRlipids Project,
-revealed that the differences in $S_\mathrm{CH}$ can be explained
-by small changes in the dihedral angle probability distributions.
-All four studied headgroup types (PC, PE, PG, and PS) are flexible
-and access a similarly wide range of conformations, with very few restrictions in dihedral orientations---also when membranes are charged.
-In conclusion, the observed differences in $S_\mathrm{CH}$ originate from
-slightly reweighed conformational probabilities rather than changes in accessible structures.
+C-H bond order parameters from NMR experiments %in this and previous studies \cite{wohlgemuth80,buldt81,seelig87}
+suggest that lipid headgroup conformational ensembles depend on lipid type (PC, PE, PG or PS) 
+and accumulated membrane charge (cationic lipids, surfactants, ions or drugs). 
+Our interpretation of this data, using MD simulations collected
+within the NMRlipids open collaboration,
+revealed that the differences in order parameters can be explained
+by relatively small changes in dihedral angle probability distributions.
+All studied headgroups (PC, PE, PG and PS) are flexible
+and access a similar wide range of conformations with only very few restictions in dihedral orientations
+also when charges are bound to membranes.
+The observed differences in order parameters originate from
+reweighting conformational probabilities rather than changes in accessible structures.
 
 The flexibility and wide conformational space of headgroups suggest that protein--bound lipids can
-adapt to various binding sites to optimize the lipid--protein interactions.
-We tested this prediction by analyzing from PDB the conformations of lipids that are tightly bound to proteins.
-We found that also the protein-bound lipids exhibit a wide range of conformations, without significant 
-differences between different lipid types: The specificity of lipid binding to proteins is not
-regulated by accessible lipid structures, but a single lipid type can adapt to various binding pockets.
+adapt to various binding sites to optimize the intermolecular lipid--protein interactions.
+We tested this prediction by analyzing the conformations of lipids that are tightly bound to proteins in from the PDB.
+Indeed, also protein--bound lipids exhibit wide range of conformations without significant 
+differences between different lipid types. Therefore, the specificify of lipid binding to proteins is not
+regulated by accessible structures of lipids, and a single lipid type can adapt to various binding sites in proteins.
 
-Our results suggest that lipid headgroup binding to proteins can be described by the {\it inverse conformational selection model}
-where unbound lipids sample a wide conformational ensemble offering the large selection of conformations
-that can bind to various binding sites in substrates, such as protein, drug, RNA, or virus, that remain conformationally fixed.
-In this model, the binding energy is dominated by
-intermolecular interactions and changes in lipid conformational entropy, while 
-conformational restrictions inherent to the lipids play a minor role.
-Here we have focused only on lipid headgroups while potential applications of the proposed model on interactions between hydrophobic acyl chains and proteins are left for future studies. 
-Applications of the proposed model in molecular cell biology include the understanding of lipid-mediated cell signaling
-and functions of lipid regulated membrane proteins in general. 
-In bionanotechnology, our model facilitates, for example,
-the design of phoshopolipid-specific antibodies and lipid nanoparticle carriers with minimal sides effects for mRNA vaccines.
-Furthermore, our work has demonstrated the power of open access MD simulation data
-to complement the PDB data in elucidating behavior of complex systems of disordered biomolecules.
+Our results pave the way toward understanding of lipid-mediated cell signaling and how lipids regulate membrane protein function in general.
+We suggest that the key to understand selective binding of certain lipid types to proteins
+are intermolecular lipid-protein interactions, rather than conformational restrictions of the lipids.
+On the other hand, wide conformational ensembles in bulk bilayers suggest that lipid crystal structures
+play a minor role and that entropic cost of lipid binding may be significant.
+Finally, we demonstrated the power of open access MD simulation data from the NMRlipids open collaboration
+to complement the data in the PDB for elucidating how complex systems containing disordered biomolecules behave.
 
-\section{Methods}
-\subsection{Experimental C--H bond order parameters}
-The headgroup and glycerol backbone C--H bond order parameters of 1-palmitoyl-2-oleoyl-sn-glycero-3-phosphoethanolamine (POPE) and 1-palmitoyl-2-oleoyl-sn-glycero-3-phospho-(1'-rac-glycerol) (POPG), purchased from Avanti polar lipids, were measured using natural abundance $^1$H-$^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
-The samples were prepared by simply mixing the lipids with water until a homogeneous preparation of multi-lamellar vesicles (MLVs) was attained and left to equilibrate for approximately 24 hours before measurements.
-The magnitudes of the order parameters were determined using a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04}, and the order parameter signs from S-DROSS experiments~\cite{gross97} combined with SIMPSON simulations \cite{bak00}.
-%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
-%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
-The set-up of the solid-state NMR experiments were identical as in our previous work~\cite{antila19}.
-The POPE experiments were recorded at 310~K and POPG experiments at 298~K, where the bilayers are in the liquid disordered phase \cite{marsh13}.
+% If you have acknowledgments, this puts in the proper section head.
+\begin{acknowledgments}
+AP is grateful to the Centro de
+Supercomputación de Galicia (CESGA) for use of the Finis
+Terrae computer
+    %     Put your acknowledgments here.
 
-Glycerol backbone peaks from both lipids, and the $\alpha$-carbon peak from POPE in the INEPT spectra 
-were assigned based on previously measured POPC spectra~\cite{ferreira13}.
-The $\beta$-carbon peak from POPE was assigned based on $^{13}$C chemical shift table for amines available
-at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm}.
-The asignment of the $\alpha$, $\beta$, and $\gamma$ peaks of POPG was based on the measured R-PDLF dipolar couplings in comparison to the known order parameters from previous $^2$H\,NMR experiments~\cite{borle85,wohlgemuth80}.
-The $\beta$-carbon peak from POPG overlapped with the g$_2$ peak from glycerol backbone
-because their chemical environments are similar.
-
-\subsection{Molecular dynamics simulations}
-
-Molecular dynamics simulation data were collected and analyzed using
-the methods from the NMRlipids Open Collaboration project (\url{nmrlipids.blogspot.fi}) \cite{botan15,catte16,ollila16,antila19}.
-Simulation details, accessibility information and quality evaluation of approximately 70 combinations of lipid headgroup and force field simulated for this work are in the supplementary information.
-
-Best models for the interpretation of lipid headgroup conformational ensembles from the experimental data were selected
-using quality evaluation measures defined in the NMRlipids project.
-Conformational ensembles of headgroup and glycerol backbone in PE and PG simulations were evaluated using the C--H bond order parameters \cite{botan15}.
-Interactions between different headgroups were evaluated by monitoring the changes in headgroup order parameters upon mixing the lipids \cite{antila19}.
-The ion binding affinities and response of lipids to bound charge were evaluated by monitoring the changes in lipid headgroup order parameters \cite{catte16,antila19}.
-
-The differences in headgroup conformation ensembles were analyzed calculating the distributions of six dihedral angles, O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$, P-O$_\alpha$-C$_\alpha$-C$_\beta$, O$_{g3}$-P-O$_\alpha$-C$_\alpha$, g$_3$-O$_{g3}$-P-O$_\alpha$, g$_2$-g$_3$-O$_{g3}$-P, and g$_1$-g$_2$-g$_3$-O$_{g3}$, labeled in Fig.
-\ref{HGorderParameters} C.
-Relative free energy costs for turning dihedral angles with respect to the most probable value (lowest free energy) were estimated from the inverse Boltzmann formula $\Delta E(\theta) = -kT \left[\ln\left[p(\theta)\right]-\ln\left[p(\theta_0)\right] \right]$, where $p(\theta)$ is the dihedral angle distribution and $\theta_0$ is the most probable angle from MD simulation.
-
-\subsection{Analysis of protein-bound lipid conformations}
-Lipid structures from the Protein Data Bank (PDB, \cite{berman00})
-were searched using PDBe REST API (\url{www.ebi.ac.uk/pdbe/pdbe-rest-api})
-using the ligand names listed in the supplementary information.
-Dihedral distributions from the lipid structures were calculated
-using the MDAnalysis Python library \cite{agrawal11,gowers16} and
-Jupyter notebook available from {\it scripts/PDBanalysis.ipynb} folder in Ref~\citenum{NMRlipidsIVbgit}.
-Only the structures determined using X-ray crystallography or Cryo-EM with a resolution higher than 3.2~\text{\AA} were used in the analysis.
-Some structures of multimeric proteins contained multiple lipids in the same conformation.
-The conformations were considered identical if all dihedral angles were the same within 3 degrees.
-In these cases, this conformation was counted only once for the subsequent analyses
-to avoid its overweighting.
-
-To demonstrate the existence of similar structures of different lipids bound to different proteins, we searched for pairs having the last five heavy atom dihedrals angles in the headgroup end (excluding g$_1$-g$_2$-g$_3$-O$_{g3}$ from the list in previous section) within
-30 degrees of each other among the structures with resolution higher than 2.5~\text{\AA}. The two most representative examples
-were handpicked from the results.
-
-\begin{acknowledgement}
-
-  % Put your acknowledgments here.
-A.P. is grateful to the Centro de
-Supercomputaci{\'o}n de Galicia (CESGA) for use of the Finis
-Terrae computer.
-M.J. thanks CSC -- IT Center for Science for computational resources and the Emil Aaltonen foundation for financial support.
-T.M.F. was supported by the Ministry of Economics, Science
-and Digitalisation of the State of Saxony-Anhalt, Germany.
-P.B. was supported by the Academy of Finland (Grant 311031).
-F.F.-R.
-acknowledges Tecnol\'{o}gico Nacional de M\'{e}xico Proyecto IT16C431, Direcci\'{o}n General de Asuntos del Personal Acad\'{e}mico (DGAPA) Programa de Apoyo a Proyectos de Investigaci\'{o}n e Innovaci\'{o}n Tecnol\'{o}gica (PAPIIT) IG100920, CONACyT Ciencia de Frontera 74884 for financial support and Miztli-Direcci\'{o}n de C\'{o}mputo y de Tecnolog\'{i}as de Informaci\'{o}n y Comunicaci\'{o}n (DGTIC) - Universidad Nacional Aut\'{o}noma de M{\'e}xico (UNAM) (Project LANCAD-UNAM-DGTIC-057) facilities for computing-time allocation.
-I.G. was supported by the Ministry of Science and Higher Education of the Russian Federation (agreement \#075-00337-20-03, project FSMG-2020-0003). 
 J.J.M. gratefully acknowledges financial support from the Carlsberg 
@@
 Research Computing Center at the University of Chicago.
-O.H.S.O, A.M.K, and S.I.V acknowledge CSC -- IT Center for Science for computational resources and Academy of Finland (grants 315596 and 319902) for financial support.
-T.J.P. acknowledges use of the Iridis high-performance computing resources at the University of Southampton.
-J.M. thanks the Center for Information Technology of the University of Groningen for their support and for providing access to the Peregrine high performance computing cluster.
 
-\end{acknowledgement}
+\end{acknowledgments}
 
@@
 
+%\newpage
+%\section{APPENDIX: The NMR results reported by Tiago Ferreira}
+
+\listoftodos
+
 \end{document}
+%
+% ****** End of file aiptemplate.tex ******

### 2021-07-21 manuscriptPGPEsuppl.tex
Update manuscriptPGPEsuppl.tex

--- 2021-06-07 manuscriptPGPEsuppl.tex
@@
 \usepackage{microtype}
-\usepackage{xr-hyper}
+\usepackage{xr}
+\externaldocument{manuscriptPGPE}
 \usepackage{soul} % for highlights with \hl{}
+
 \usepackage{hyperref} % to make TOC titles clickable
@@
 
-\makeatletter
-\newcommand*{\addFileDependency}[1]{% argument=file name and extension
-  \typeout{(#1)}
-  \@addtofilelist{#1}
-  \IfFileExists{#1}{}{\typeout{No file #1.}}
-}
-\makeatother
-
-\newcommand*{\myexternaldocument}[1]{%
-    \externaldocument{#1}%
-    \addFileDependency{#1.tex}%
-    \addFileDependency{#1.aux}%
-}
-
-\myexternaldocument{manuscriptPGPE}
-
-\author{\small Am{\'e}lie Bacle}
-\affiliation{\tiny Laboratoire Coop\'{e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit\'{e} de Poitiers, 1 rue Georges Bonnet, 86000 Poitiers, France }
-
-\author{Pavel Buslaev}
-\affiliation{Nanoscience Center and Department of Chemistry, University of Jyväskylä, P.O. Box 35, 40014 Jyväskylä, Finland}
-\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, 141701 Dolgoprudny, Russia}
+\author{\scriptsize Am{\'e}lie Bacle}
+\affiliation{Paris, France}
+
+\author{\scriptsize Pavel Buslaev}
+\affiliation{\tiny Nanoscience Center and Department of Chemistry, University of Jyv{\"a}skyl{\"a}, Finland}
+\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
 
@@
 \affiliation{Center for Research in Biological Chemistry and Molecular Materials (CiQUS), Universidade de Santiago de Compostela, E-15782 Santiago de Compostela, Spain}
-\alsoaffiliation{CIQUP, Centro de Investigação em Qu\'{i}mica, Departamento de Qu\'{i}mica e Bioqu\'{i}mica, Faculdade de Ciências, Universidade do Porto, Porto, Portugal}
+\affiliation{CIQUP, Centro de Investigação em Química, Departamento de Química e Bioquímica, Faculdade de Ciências, Universidade do Porto, Porto, Portugal}
 
 \author{Fernando Favela-Rosales}
-\affiliation[Tecnol\'{o}gico Nacional de M\'{e}xico]{Departamento de Ciencias B\'{a}sicas, Tecnol\'{o}gico Nacional de M\'{e}xico, ITS Zacatecas Occidente, M\'{e}xico}
+\affiliation[Tecnol\'{o}gico Nacional de M\'{e}xico]{Departamento de Ciencias B\'{a}sicas, Tecnol\'{o}gico Nacional de M\'{e}xico, Campus Zacatecas Occidente, M\'{e}xico}
 
@@
 
-\author{Patrick F.J. Fuchs}
-\affiliation{Sorbonne Universit\'{e}, Ecole Normale Sup\'{e}rieure, PSL University, CNRS, Laboratoire des Biomol\'{e}cules (LBM), 75005 Paris, France}
-\alsoaffiliation{Universit\'{e} de Paris, UFR Sciences du Vivant, 75013, Paris, France}
+\author{Patrick Fuchs}
+\affiliation{Paris, France}
 
 \author{Ivan Gushchin}
-\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, 141701 Dolgoprudny, Russia}
+\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
 
@@
 \author{Jesper J. Madsen}
-\affiliation[University of Chicago]{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
-\alsoaffiliation[University of South Florida]{Global and Planetary Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
+\affiliation[University of Chicago]{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
+\alsoaffiliation[University of South Florida]{Global and Planetary Health, College of Public Health, University of South Florida, Tampa, Florida 33612, United States of America}
 
@@
 
-\author{Paula Milán Rodríguez}
-\affiliation{Sorbonne Université, Ecole Normale Supérieure, PSL University, CNRS, Laboratoire des Biomolécules (LBM), 75005 Paris, France}
+\author{Paula Milan Rodriguez}
+\affiliation{Paris, France}
 
@@
 \author{Chris G. Papadopoulos}
-\affiliation{Université Paris-Saclay, CEA, CNRS, Institute for Integrative Biology of the Cell (I2BC), 91198, Gif-sur-Yvette, France}
+\affiliation[]{I2BC - University Paris Sud}
 
 \author{Antonio Pe{\'o}n}
-\affiliation{CIQUP, Centro de Investigação em Qu\'{i}mica, Departamento de Qu\'{i}mica e Bioqu\'{i}mica, Faculdade de Ciências, Universidade do Porto, Porto, Portugal}
+\affiliation[]{Spain}
 
@@
 \affiliation{Departamento de F{\'i}sica Aplicada, Facultade de F{\'i}sica, Universidade de Santiago de Compostela, E-15782 Santiago de Compostela, Spain}
-
-\author{Salla I. Virtanen}
-\affiliation{Institute of Biotechnology, University of Helsinki}
 
@@
 
-\title{ Supporting Information:\\Conformational plasticity of lipid headgroups in cellular membranes and protein--lipid complexes}
+\title{ Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids }
 
@@
 
-\clearpage
 \section{R-PDLF and SDROSS experiments}
@@
 %  \centering
-  \includegraphics[width=0.9\textwidth]{./Figs/POPE_SI.pdf}
+  \includegraphics[width=0.8\textwidth]{./Figs/POPEexperiment.pdf}
   \caption{\label{POPEspectra}
-    Solid-state NMR spectroscopy for determining the C--H bond order parameters, $S_{\rm{CH}}$, in the POPE headgroup and glycerol backbone.
-(A) Chemical structure of POPE showing the labels used to identify different carbons.
-    (B) Refocused INEPT $^{13}$C spectrum of POPE MLVs. 
-    (C) Headgroup and glycerol backbone R-PDLF dipolar slices.
-The arrows indicate the splittings used to determine the C--H bond order parameters by using  $|S_{\rm{CH}}|=\Delta\nu/(0.315 \times d_{\rm{CH}})$.
-The rigid coupling $d_{\rm{CH}}$ used was 22 kHz.
-    (D) Experimental S-DROSS curves giving signs of the order parameters measured.
-Grey lines are a set of slices taken from a region of the spectrum without peaks, i.e.
-grey lines are noise profiles, to highlight the signal-to-noise ratio of the measured modulations.}
+    (A) Chemical structure of POPE with the labeling of headgroup and glycerol backbone carbons.
+    (B) INEPT spectra from POPE sample with the headgroup and glycerol backbone peaks labeled.
+    (C) 2D R-PDLF spectra
+    (D) Dipolar sliced from the  2D R-PDLF spectra with the resulting order parameters on top of figures.
+    (E) Experimetal S-DROSS curves giving signs of the order parameters.
+  }
+  \todo{NMR figures in SI to be updated.}
 \end{figure}
@@
 %  \centering
-  \includegraphics[width=\textwidth]{./Figs/POPG_SI.pdf}
+  \includegraphics[width=0.8\textwidth]{./Figs/PGexpRPDLF.eps}
   \caption{\label{POPGspectra}
-    Solid-state NMR spectroscopy for determining the C--H bond order parameters, $S_{\rm{CH}}$, in the POPG headgroup and glycerol backbone.
-(A) Chemical structure of POPG showing the labels used to identify different carbons.
-    (B) Refocused INEPT $^{13}$C spectrum of POPG MLVs.
-    (C) Headgroup and glycerol backbone R-PDLF dipolar slices.
-The arrows indicate the splittings used to determine the C--H bond order parameters by using  $|S_{\rm{CH}}|=\Delta\nu/(0.315 \times d_{\rm{CH}})$.
-The rigid coupling $d_{\rm{CH}}$ used was 22 kHz.
-    (D) Experimental S-DROSS curves giving signs of the order parameters measured.
-Note that the S-DROSS modulation for $g_3$ is very close to the noise level.
-In this case, we have confirmed that the order parameter sign of $g_3$ was negative by performing a measurement with less points in the indirect dimension, more scans and using a higher MAS frequency of 8 kHz (not shown).
+    (A) Chemical structure of POPG with the labeling of headgroup and glycerol backbone carbons.
+    (B) INEPT spectra from POPG sample with the headgroup and glycerol backbone peaks labeled.
+    (C) 2D R-PDLF spectra
+    (D) Dipolar sliced from the  2D R-PDLF spectra with the resulting order parameters on top of figures.
+    (E) Experimetal S-DROSS curves giving signs of the order parameters.
+  }
+\end{figure}
+
+\begin{figure}[]
+%  \centering
+  \includegraphics[width=0.8\textwidth]{./Figs/SIMPSON.pdf}
+  \caption{\label{POPGsimpson}
+    Simpson simulaton of S-DROSS curve of $\beta$-carbon of POPG.
   }
@@
 \subsection{Conformational ensembles of headgroup and glycerol backbone in PE and PG lipids}
-The quality of PE and PG headgroup conformational ensembles in different simulations are evaluated against NMR experiments in figures \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG} using C-H bond order parameters as in our previous studies for PC and PS lipids \cite{botan15,antila19}.
+
+The quality of PE and PG headgroup conformational ensembles in different simulations against NMR experiments is evaluated in figures \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG} using C-H bond order parameters as in our previous studies for PC and PS lipids \cite{botan15,antila19}.
 Conclusions are the same for all lipids: None of the force fields correctly captures the lipid headgroup conformational ensembles, but CHARMM36 gives results closest to experiments.
@@
   \caption{\label{HGorderParametersPE}
-    C--H bond order parameters, $S_{\rm CH}$, of the PE headgroup ($\beta$ and $\alpha$) and
-    glycerol backbone (g$_3$, g$_2$, g$_1$) carbons from NMR experiments (horizontal lines;
-    POPE and signs this work, DPPE from Ref.~\citenum{seelig76},
-    {\it Escherichia coli} PE from Ref.~\citenum{gally81})
-    and MD simulations with different force fields (symbols).
-    The light blue areas span 0.04 units around the average of the extremal experimental values,
-    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
-    The vertical bars shown for most simulation values are not error bars,
-    but demonstrate that for these systems we had at least two data sets;
-    the ends of the bars mark the extreme values from the sets,
-    the symbol marks the measurement-time-weighted average.
+    The headgroup and glycerol backbone order parameters of PE lipids
+    from experiments (POPE and signs this work, DPPE from Ref.~\citenum{seelig76}
+    and E.coliPE from Ref.~\citenum{gally81}) and simulations with different force fields.
   }
@@
   \centering
-  \includegraphics[width=8.0cm]{./Figs/HGorderparametersPG.pdf}
+  \includegraphics[width=9.0cm]{./Figs/HGorderparametersPG.pdf}
   \caption{\label{HGorderParametersPOPG}
-    C--H bond order parameters, $S_{\rm CH}$, of the PG headgroup ($\beta$ and $\alpha$) and
-    glycerol backbone (g$_3$, g$_2$, g$_1$) carbons
-    from MD simulations with different force fields (symbols) and
-    NMR experiments (horizontal lines;
-    $^{13}$C NMR POPG and signs from this work,
-    $^2$H NMR POPG from Ref.~\citenum{borle85}, % contains 10 mM of PIPES,
-    DPPG from Ref.~\citenum{wohlgemuth80}, % and 10 mM PIPES
-    and {\it Escherichia coli} PG from Ref.~\citenum{gally81}).
-    Experimental data for DPPG had been measured with 0.1\,M NaCl~\cite{wohlgemuth80}, while all of the other data are with
-    (Na$^+$, except for CHARMM36 K$^+$) counterions only.
-    The DPPG is [R],
-    2H POPG beta racemic, alpha 3,3' averaged, which should equal 3,1'~\cite{wohlgemuth80}
-    13C POPG racemic.
-    $^2$ NMR DPPG and POPG in 10\,mM PIPES buffer.
-    The light blue areas span 0.04 units around the average of the extremal experimental values,
-    in accordance with the expected quantitative accuracy of experiments~\cite{ollila16}.
+    The headgroup and glycerol backbone order parameters of PG lipids
+    from experiments (POPG and signs from this work and from Ref.~\citenum{borle85}, %contains 10mM of PIPES,
+    DPPG with 100mM NaCl from Ref.~\citenum{wohlgemuth80},% contains 10mM PIPES and,
+    and E.Coli PG results from Ref.~\citenum{gally81})
+    and simulations with different force fields.
   }
@@
     from experiments at 298 K \cite{scherer87,macdonald87} and simulations with different force fields
-    (temperatures listed in tables \ref{systemsMIX} are between 298-310 K).
-    Signs are determined as discussed in Refs. \citenum{botan15} and~\citenum{ollila16}.
+    (temperatures listed in tables \ref{systemsMIX} and \ref{systemsMIX2} are between 298-310 K).
+    Signs are determined as discussed in Refs. \citenum{botan15,ollila16}.
   }
@@
 headgroup dipole tilts more parallel to the membrane plane after incorporation
-of negative charges into the membrane~\cite{seelig87, scherer87,antila19}.
+of negative charges into the membrane~\cite{seelig87, scherer87,antila18}.
 The response of PC headgroup order parameters to the addition of PE or PG lipids
@@
 The good performance of Berger-OPLS simulations is surprising here because
-headgroup conformational ensemble is not very close to experiments in this model and
+headgroup conformational enemble is not very close to experiments in this model and
 the response of headgroup order parameters
@@
 \clearpage
-\section{Dihedral angle distributions}
-
-\subsection{Dihedral angles of PC, PE, PG and PS headgroups}
-
-%The dihedral angle distributions are shown in Fig. \ref{DIHdists}
-%and relative energies in Fig. \ref{structures} in the main text.
+\section{Dihedral angle distributions and the analysis of relative energies}
+
+\subsection{Dihedral angles and relative energies of PC, PE, PG and PS headgroups}
+
+We estimated the relative energies of each dihedral angle value
+with respect to the most probable value (lowest energy) 
+from the inverse Boltzmann formula $\Delta E(\theta) = -kT \left[\ln\left[p(\theta)\right]-\ln\left[p(\theta_0)\right] \right]$,
+where $p(\theta)$ is the dihedral angle distribution and $\theta_0$ is the most probable angle from MD simulation.
+The dihedral angle distributions are shown in Fig. \ref{DIHdists}
+and relative energies in Fig. \ref{structures} in the main text.
 
@@
   \caption{\label{DIHdists}
-    Heavy atom dihedral angle distributions from CHARMM36 simulations that correctly capture the order parameter differences between lipid headgroups.
+    Heavy atom dihedral angle distributions from CHARMM36 simulations that correctly capture the
+    order parameter differences between the force fields.
   }
@@
 are listed in Tables~\ref{systemsPE} and \ref{systemsPG},
-and lipid mixtures with additional ions in Table~\ref{systemsMIX}.
+and lipid mixtures with additional ions in Tables~\ref{systemsMIX} and \ref{systemsMIX2}.
 The simulations were analyzed using preliminary versions of the NMRlipids databank
@@
 For united atom simulations, we first constructed trajectories including hydrogens with ideal geometry using either \texttt{buildH} program~\cite{buildH} or (\texttt{scratch/opAAUA\_prod.py}) in  Ref.~\citenum{MATCHgit}, and the order parameters were then calculated from these trajectories.
-This approach has been tested against trajectories with explicit hydrogens and the deviations in order parameters are small \cite{buildHvalidation,piggot17}.\\
+This approach has been tested against trajectories with explicit hydrogens and the deviations in order parameters are small \cite{buildH,piggot17}.\\
+%\todo{BuildH program is now cited with a direct link to the GitHub repo. I think that a release to Zenodo would be nice in the final publication.}\\
 
@@
       		& 144	& 5760  %&0    
-		& 310  & 500          & 400          & \citenum{charmm36POPEfiles} \\
+		& 310  & 500          & 400          & \cite{charmm36POPEfiles} \\
       POPE & CHARMM36 \cite{klauda10}           %& 0      
       		& 500       & 25000 %& 0   
-		&  310  & 500 & 100 & \citenum{POPEcharmm} \\
+		&  310  & 500 & 100 & \cite{POPEcharmm} \\
 %   POPE	& CHARMM36 \cite{klauda10}           %& 0.11
 %		& 500       & 25000 & 50  &  310  & 500 & 100 & \cite{POPEcharmm150mMNaCl} \\
-      POPE	& CHARMM36-UA \cite{lee14}         %&0       
+      POPE	& CHARMM36-UA \cite{??}         %&0       
       		& 336	& 15254 %&0    
-		& 310  & 2$\times$200 & 2$\times$100 & \citenum{charmm36uaPOPEfiles}  \\
+		& 310  & 2$\times$200 & 2$\times$100 & \cite{charmm36uaPOPEfiles}  \\
       \hline
@@
       		& 288 	& 9386  %&0    
-		& 336  & 200 & 100 & \citenum{slipidsDPPEfiles}  \\
+		& 336  & 200 & 100 & \cite{slipidsDPPEfiles}  \\
       POPE	& Slipids \cite{jambeck12b} %&0    
-      		& 336	& 15254     %&0    
-		& 310  & 2$\times$200 &  2$\times$100 & \citenum{slipidsPOPEfiles}  \\
+      		& 336	& ?     %&0    
+		& 310  & 2$\times$200 &  2$\times$100 & \cite{slipidsPOPEfiles}  \\
       POPE	& Slipids \cite{jambeck12b}            %& 0
 		& 500 & 25000 %& 0   
-		&  310  & 500 & 100 & \citenum{POPEslipids} \\
+		&  310  & 500 & 100 & \cite{POPEslipids} \\
 %      POPE  & Slipids / {\AA}qvist \cite{jambeck12b,aqvist90}  %& 0.11
@@
       \hline
-      DPPE	& GROMOS-CKP    \cite{piggot11}      %&0
+      DPPE	& GROMOS-CKP    \cite{??}      %&0
           	& 128	& 3655  %&0    
-		& 342  & 2$\times$500 & 2$\times$400 & \citenum{gromosCKPdppe} \\
+		& 342  & 2$\times$500 & 2$\times$400 & \cite{gromosCKPdppe} \\
 %      POPE  & GROMOS-CKP    \cite{??}      %&0
 %    		& 128	& 3552  &0    & 313  & 2$\times$500 & 2$\times$400 & \cite{gromosCKPpope} \\
-      POPE	& GROMOS-CKP    \cite{piggot11}      %&0
+      POPE	& GROMOS-CKP    \cite{??}      %&0
           	& 500	& 25000 %&0    
-		& 310  & 500 & 100 & \citenum{gromosCKPpopeT310} \\
+		& 310  & 500 & 100 & \cite{gromosCKPpopeT310} \\
 %      POPE  & GROMOS-CKP    \cite{??}      &0.11 & 500	& 25000 &50   & 310  & 500 & 100 & \cite{gromosCKPpopeT310150mMNaCl} \\
@@
       \hline
-      POPE	& GROMOS 43A1-S3 \cite{chiu09}     %&0
+      POPE	& GROMOS 43A1-S3 \cite{??}     %&0
           	& 128	& 3552     %&0    
-		& 313  & 2$\times$200 & 2$\times$100 & \citenum{gromos43a1s3POPEfiles}  \\
-      \hline
-      POPE	& OPLS-UA/HG-H \cite{Ulmschneider09}   %&0    
+		& 313  & 2$\times$200 & 2$\times$100 & \cite{gromos43a1s3POPEfiles}  \\
+      \hline
+      POPE	& OPLS-UA/HG-H \cite{??}   %&0    
       		& 128	& 3328     %&0    
-		& 303  & 2$\times$200 & 2$\times$100 & \citenum{OPLSuaWvdWPOPEfiles} \\
-      POPE	& OPLS-UA \cite{Ulmschneider09}            %&0
+		& 303  & 2$\times$200 & 2$\times$100 & \cite{OPLSuaWvdWPOPEfiles} \\
+      POPE	& OPLS-UA \cite{??}            %&0
       		& 128	& 3328     %&0    
-		& 303  & 2$\times$200 & 2$\times$100 & \citenum{OPLSuaPOPEfiles} \\
+		& 303  & 2$\times$200 & 2$\times$100 & \cite{OPLSuaPOPEfiles} \\
       \hline
@@
       		& 144	& 5760     %&0    
-		& 310  & 500 & 350 & \citenum{MacRogPOPEfiles} \\
+		& 310  & 500 & 350 & \cite{MacRogPOPEfiles} \\
 %   POPE	& OPLS-MacRog \cite{rog16}     &0    & 128	& 5120     &0    & 300  & 500 & 300 & \cite{MacRogPOPEfilesT300K} \\
       \hline
-      POPE	& Berger-POPE-2004 \cite{devries04}       %&0    
+      POPE	& Berger-POPE-2004 \cite{??}       %&0    
       		& 128	& 3552  %&0    
-		& 303  & 2$\times$200 & 2$\times$100 & \citenum{bergerPOPEfiles}  \\
-      POPE	& Berger-POPE-2018 \cite{devries04}      %&0    
+		& 303  & 2$\times$200 & 2$\times$100 & \cite{bergerPOPEfiles}  \\
+      POPE	& Berger-POPE-2018 \cite{??}      %&0    
       		& 128	& 3552  %&0    
-		& 303  & 2$\times$200 & 2$\times$100 & \citenum{berger2POPEfiles}  \\
+		& 303  & 2$\times$200 & 2$\times$100 & \cite{berger2POPEfiles}  \\
 %   DOPE  & Berger-POPE-2004 \cite{??}       &0    & 128	& 4789  &0    & 271  & 2$\times$200 & 2$\times$100 & \cite{bergerDOPEfiles}  \\
@@
       		& 500 & 25000 %& 0  
-		&  310  & 500 & 100 & \citenum{POPElipid17} \\
+		&  310  & 500 & 100 & \cite{POPElipid17} \\
 %      POPE             & Lipid17 \cite{gould18} & 0.11   & 500 & 25000 & 50  &  310  & 500 & 100 & \cite{POPElipid17150mMNaCl} \\
@@
       % some footnotes are not visible in typeset-MS (pdf)
-      lipid/counter-ions & force field for lipids / ions &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
-      \hline
-      POPG/K$^+$  & CHARMM36 \cite{venable13}        & 118& 4110    & 298  & 100 & 100 & \citenum{CHARMM36popg}  \\
+      lipid/counter-ions & force field for lipids / ions & NaCl (M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+      \hline
+      POPG/K$^+$  & CHARMM36 \cite{venable13}       &0         & 118& 4110   &0    & 298  & 100 & 100 & \cite{CHARMM36popg}  \\
 %      POPG             & CHARMM36 \cite{venable13}        & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGcharmm150mMNaCl} \\
-%      POPG             & CHARMM36 \cite{venable13}                & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGcharmm} \\
-      \hline
-      POPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}            & 288 	& 10664   & 298  & 250 & 100 & \citenum{slipidsPOPGfiles} \\
-      DPPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}           & 288 	& 11232   & 314  & 200 & 100 & \citenum{slipidsDPPGfiles} \\
-%      DPPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}           & 288 	& 11232   & 298  & 400 & 100 & \citenum{slipidsDPPGfilesT298K} \\
-%      POPG         & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}      & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGslipids} \\
+      POPG             & CHARMM36 \cite{venable13}        & 0              & 500 & 25000 & 0   &  310  & 500 & 100 & \cite{POPGcharmm} \\
+      \hline
+      POPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    &0         & 288 	& 10664   &0     & 298  & 250 & 100 & \cite{slipidsPOPGfiles} \\
+      DPPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    &0         & 288 	& 11232  &0     & 314  & 200 & 100 & \cite{slipidsDPPGfiles} \\
+      DPPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    &0         & 288 	& 11232   &0     & 298  & 400 & 100 & \cite{slipidsDPPGfilesT298K} \\
+      POPG         & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    & 0    & 500 & 25000 & 0  &  310  & 500 & 100 & \cite{POPGslipids} \\
 %      POPG         & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    & 0.11    & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGslipids150mMNaCl} \\
       \hline
-      POPG/Na$^+$         & Lipid17 / Dang \cite{gould18,smith94,dang06}     & 500 & 25000  &  310  & 500 & 100 & \citenum{POPGlipid17} \\
-%      POPG             & Lipid17 \cite{gould18}         & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGlipid17150mMNaCl} \\
-      \hline
-      POPG/Na$^+$             & GROMOS-CKP / GROMOS 54A7 \cite{piggot11,schmid11}         & 500 & 25000 &  310  & 500 & 100 & \citenum{POPGgromosCKP} \\
+      POPG             & LIPID17 / Dang \cite{gould18,smith94,dang06}         & 0              & 500 & 25000 & 0   &  310  & 500 & 100 & \cite{POPGlipid17} \\
+%      POPG             & LIPID17 \cite{gould18}         & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGlipid17150mMNaCl} \\
+      \hline
+      POPG             & GROMOS-CKP \cite{??}         & 0              & 500 & 25000 & 0  &  310  & 500 & 100 & \cite{POPGgromosCKP} \\
 %      POPG             & GROMOS-CKP \cite{??}         & 0.11           & 500 & 25000 & 49 &  310  & 500 & 100 & \cite{POPGgromosCKP150mMNaCl} \\
@@
       % some footnotes are not visible in typeset-MS (pdf)
-      lipid/counter-ions & force field for lipids / ions & CaCl$_2$\,(M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
-      \hline
-%      POPC                   & CHARMM36 \citenum{??}        & 0.11      & 0  & 500     & 25000 & 48  &  310  & 500 & 100 & \cite{POPCcharmm150mMNaCl301K}  \\
+      lipid/counter-ions & force field for lipids / ions & NaCl (M) & CaCl$_2$\,(M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+      \hline
+%      POPC                   & CHARMM36 \cite{??}        & 0.11      & 0  & 500     & 25000 & 48  &  310  & 500 & 100 & \cite{POPCcharmm150mMNaCl301K}  \\
 %      POPC:POPG (7:3)        & CHARMM36 \cite{??}        & 0.11      & 0  & 350     & ?     & ?   &  310  & 500 & 100 & \cite{POPC7POPG1charmm36NaCl}  \\
-      POPC                   & CHARMM36 \cite{klauda10}            & 0  & 500     & 25000 & 0   &  310  & 500 & 100 & \citenum{POPCcharmmT310K}  \\
-      POPC:POPG (7:3)        & CHARMM36 \cite{klauda10,venable13}    & 0  & 350     & 25000 & 0   &  310  & 500 & 100 & \citenum{POPC7POPG1charmm36}  \\
-      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}     & 0  & 150:150 & 31500 & 0   &  298  & 400 & 300 & \citenum{CHARMM36POPCPOPG5050} \\ 
-      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}     & 0.1 & 150:150 & 31329 & 57 &  298  & 400 & 300 & \citenum{CHARMM36POPCPOPG5050100mMCaCl} \\
-      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}     & 1.08 & 150:150 & 29766 & 578  &  298  & 400 & 300 & \citenum{CHARMM36POPCPOPG50501000mMCaCl} \\
-      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 0  & 350:88 & 26280 & 0  &  298  & 400 & 300 & \citenum{CHARMM36POPCPOPG4010} \\
-      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 0.1 & 350:88 & 26280 & 47  &  298  & 400 & 300 & \citenum{CHARMM36POPCPOPG4010100mMCaCl} \\
-      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}    & 1.0 & 350:88 & 24927 & 451  &  298  & 400 & 300 & \citenum{CHARMM36POPCPOPG40101000mMCaCl} \\
-      \hline
-      POPC             & CHARMM36 \cite{klauda10}        &0          & 256 & 8704 & 0  &  300  & 300 & 250 & \citenum{POPCcharmm300K} \\
-      POPC:POPE (1:1)  & CHARMM36 \cite{klauda10,venable13}         & 0  & 128 & 8704 & 0  &  300  & 300 & 250 & \citenum{POPC1POPE1charmm36} \\
-      \hline
-      POPC             & OPLS-MacRog \cite{rog16}     &0           & 128 & 5120 & 0  &  300  & 500 & 300 & \citenum{POPCmacrog300K} \\
-      POPC:POPE (1:1)  & OPLS-MacRog \cite{rog16}     &0           & 128 & 5120 & 0  &  300  & 500 & 300 & \citenum{POPC1POPE1macrogT300K} \\
-      \hline
-      POPC             & Slipid \cite{jambeck12b}     &0          & 512 & 23943 & 0  &  298  & 170 & 100 & \citenum{POPCslipid298K} \\
-      POPC:POPE (1:1)  & Slipid \cite{jambeck12b}     &0          & 128 & 5120  & 0  &  298  & 500 & 300 & \citenum{POPC1POPE1slipidT298K} \\
-     \hline
-      POPC                   & GROMOS-CKP \cite{piggot12} & 0  & 500     & 25000 & 0   &  310  & 500 & 100 & \citenum{POPCgromosCKPT310K}  \\
-      POPC:POPG (7:3)        & GROMOS-CKP / GROMOS54A7 \cite{piggot11,piggot12,schmid11} & 0  & 350:150 & 25000 & 0   &  310  & 500 & 100 & \citenum{POPC7POPG3gromosCKPT310K} \\
-     \hline
-      POPC                   & Slipid \cite{jambeck12b}  & 0      & 500     & 25000 & 0   &  310  & 500 & 100 & \citenum{POPCslipid301K}  \\
-      POPC:POPG (7:3)        & Slipid / {\AA}qvist \cite{jambeck12b,aqvist90} & 0  & 350:150 & 25000 & 0   &  310  & 500 & 100 & \citenum{slipidPOPC70POPG30T310K} \\
-      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0  & 128:128 & 12800 & 0   &  298  & 500 & 400 & \citenum{slipidPOPC50POPG50T298K} \\
-      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0.1  & 128:128 & 12800 & 23  &  298  & 500 & 400 & \citenum{slipidPOPC50POPG50T298K} \\
-      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0.2  & 128:128 & 12800 & 46  &  298  & 1500 & 500 & \citenum{slipidPOPC50POPG50T298K} \\
-      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0.5  & 128:128 & 12800 & 115 &  298  & 1500 & 500 & \citenum{slipidPOPC50POPG50T298K} \\
-      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 1.0  & 128:128 & 12800 & 230 &  298  & 1500 & 500 & \citenum{slipidPOPC50POPG50T298K} \\
-      \hline
-      POPC:POPG (4:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}       & 0  & 350:88 & 26265 & 0  &  298  & 400 & 350 & \citenum{Lipid17POPCPOPG8020} \\
-      POPC:POPG (4:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}    & 0.1& 350:88 & 26124 & 47 &  298  & 400 & 250 & \citenum{Lipid17POPCPOPG8020100mMCaCl} \\
-      POPC:POPG (4:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}    & 1.0& 350:88 & 24840 & 475 &  298  & 1200 & 200 & \citenum{Lipid17POPCPOPG80201000mMCaCl} \\
-      POPC:POPG (1:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}   & 0  & 150:150 & 31572 & 0  &  298  & 320 & 200 & \citenum{Lipid17POPCPOPG5050} \\
-      POPC:POPG (1:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}   & 0.1& 150:150 & 31401 & 57 &  298  & 718 & 198 & \citenum{Lipid17POPCPOPG5050100mMCaCl} \\
-      POPC:POPG (1:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}  & 1.0& 150:150 & 29865 & 569 &  298  & 720 & 200 & \citenum{Lipid17POPCPOPG50501000mMCaCl} \\
-      \hline
-      POPC:POPG (4:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     &0     & 350:88 & 26265 & 0  &  298  & 400 & 300 & \citenum{Lipid17eccPOPCPOPG8020} \\
-      POPC:POPG (4:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 0.1& 350:88 & 26124 & 47 &  298  & 400 & 300 & \citenum{Lipid17eccPOPCPOPG8020100mMCaCl} \\
-      POPC:POPG (4:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 1.0& 350:88 & 24840 & 475 &  298  & 400 & 300 & \citenum{Lipid17eccPOPCPOPG80201000mMCaCl} \\
-      POPC:POPG (1:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 0  & 150:150 & 31572 & 0  &  298  & 347.8 & 333 & \citenum{Lipid17eccPOPCPOPG5050} \\
-      POPC:POPG (1:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 0.1& 150:150 & 29865 & 54 &  298  & 400 & 300 & \citenum{Lipid17eccPOPCPOPG5050100mMCaCl} \\
-      POPC:POPG (1:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     & 1.0& 150:150 & 29865 & 569 &  298  & 600 & 400 & \citenum{Lipid17eccPOPCPOPG50501000mMCaCl} \\
-      \hline
-      POPC             & Berger-POPC-07 \cite{ollila07a} & 0  & 256 & 10240 & 0  &  300  & 300 & 200 & \citenum{POPCberger300K} \\
-      POPC:POPE (1:1)  & Berger-POPE-04~\cite{devries04}  & 0  & 128 & 11008 & 0  &  300  & 300 & 200 & \citenum{POPC1POPE1berger} \\
+      POPC                   & CHARMM36 \cite{klauda10}        & 0         & 0  & 500     & 25000 & 0   &  310  & 500 & 100 & \cite{POPCcharmmT310K}  \\
+      POPC:POPG (7:3)        & CHARMM36 \cite{klauda10,venable13}        & 0         & 0  & 350     & 25000 & 0   &  310  & 500 & 100 & \cite{POPC7POPG1charmm36}  \\
+      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}        &0          & 0  & 150:150 & 31500 & 0   &  298  & 500 & 400 & \cite{CHARMM36POPCPOPG5050} \\ 
+      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}        &0          & 0.1 & 150:150 & 31329 & 57 &  298  & 400 & 300 & \cite{CHARMM36POPCPOPG5050100mMCaCl} \\
+      POPC:POPG (1:1)        & CHARMM36 \cite{klauda10,venable13}        &0          & 1.08 & 150:150 & 29766 & 578  &  298  & 500 & 400 & \cite{CHARMM36POPCPOPG50501000mMCaCl} \\
+      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}        &0          & 0  & 350:88 & 26280 & 0  &  298  & 500 & 400 & \cite{CHARMM36POPCPOPG4010} \\
+      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}        &0          & 0.1 & 350:88 & 26280 & 47  &  298  & 500 & 400 & \cite{CHARMM36POPCPOPG4010100mMCaCl} \\
+      POPC:POPG (4:1)        & CHARMM36 \cite{klauda10,venable13}        &0          & 1.0 & 350:88 & 24927 & 451  &  298  & 500 & 400 & \cite{CHARMM36POPCPOPG40101000mMCaCl} \\
+      \hline
+      POPC             & CHARMM36 \cite{klauda10}        &0          & 0  & 256 & 8704 & 0  &  300  & 300 & 250 & \cite{POPCcharmm300K} \\
+      POPC:POPE (1:1)  & CHARMM36 \cite{klauda10,venable13}        &0          & 0  & 128 & 8704 & 0  &  300  & 300 & 250 & \cite{POPC1POPE1charmm36} \\
+      \hline
+      POPC             & OPLS-MacRog \cite{rog16}     &0          & 0  & 128 & 5120 & 0  &  300  & 500 & 300 & \cite{POPCmacrog300K} \\
+      POPC:POPE (1:1)  & OPLS-MacRog \cite{rog16}     &0          & 0  & 128 & 5120 & 0  &  300  & 500 & 300 & \cite{POPC1POPE1macrogT300K} \\
+      \hline
+      POPC             & Slipid \cite{jambeck12b}     &0          & 0  & 512 & 23943 & 0  &  298  & 170 & 100 & \cite{POPCslipid298K} \\
+      POPC:POPE (1:1)  & Slipid \cite{jambeck12b}     &0          & 0  & 128 & 5120  & 0  &  298  & 500 & 300 & \cite{POPC1POPE1slipidT298K} \\
+     \hline
+      POPC                   & GROMOS-CKP / ?? \cite{??,??}  & 0         & 0  & 500     & 25000 & 0   &  310  & 500 & 100 & \cite{POPCgromosCKPT310K}  \\
+      POPC:POPG (7:3)        & GROMOS-CKP / ?? \cite{??,??}  & 0         & 0  & 350:150 & 25000 & 0   &  310  & 500 & 100 & \cite{POPC7POPG3gromosCKPT310K} \\
+     \hline
+      POPC                   & Slipid \cite{jambeck12b}  & 0         & 0  & 500     & 25000 & 0   &  310  & 500 & 100 & \cite{POPCslipid301K}  \\
+      POPC:POPG (7:3)        & Slipid / {\AA}qvist \cite{jambeck12b,aqvist90}  & 0         & 0  & 350:150 & 25000 & 0   &  310  & 500 & 100 & \cite{slipidPOPC70POPG30T310K} \\
+      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0         & 0  & 128:128 & 12800 & 0   &  298  & 500 & 400 & \cite{slipidPOPC50POPG50T298K} \\
+      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0         & 0.1  & 128:128 & 12800 & 23  &  298  & 500 & 400 & \cite{slipidPOPC50POPG50T298K} \\
+      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0         & 0.2  & 128:128 & 12800 & 46  &  298  & 1500 & 500 & \cite{slipidPOPC50POPG50T298K} \\
+      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0         & 0.5  & 128:128 & 12800 & 115 &  298  & 1500 & 500 & \cite{slipidPOPC50POPG50T298K} \\
+      POPC:POPG (1:1)        & Slipid / Dang \cite{jambeck12b,jambeck2012another,smith94,dang06} & 0         & 1.0  & 128:128 & 12800 & 230 &  298  & 1500 & 500 & \cite{slipidPOPC50POPG50T298K} \\
+    \end{tabular}
+    }
+  \end{minipage}
+    \todo{Citation and ion model for GROMOS-CKP?} \\
+\end{table*}
+%\end{sidewaystable*}
+
+\begin{table*}
+%  \begin{sidewaystable*}[!p]
+  \centering
+  \caption{List of MD simulations with PE and PG lipids mixed with PC.
+    % The salt concentrations calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
+    % these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
+    % The lipid force fields named as in our previous work~\cite{botan15}.
+  }\label{systemsMIX2}
+  \begin{minipage}[t]{\textwidth}
+    \resizebox{\columnwidth}{!}{
+    \begin{tabular}{l c c r r r r r r c c}
+      %\hline
+      % some footnotes are not visible in typeset-MS (pdf)
+      lipid/counter-ions & force field for lipids / ions & NaCl (M) & CaCl$_2$\,(M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+      \hline
+      POPC:POPG (4:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}        &0          & 0  & 350:88 & 26265 & 0  &  298  & 400 & 350 & \cite{Lipid17POPCPOPG8020} \\
+      POPC:POPG (4:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}        &0          & 0.1& 350:88 & 26124 & 47 &  298  & 400 & 250 & \cite{Lipid17POPCPOPG8020100mMCaCl} \\
+      POPC:POPG (4:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}        &0          & 1.0& 350:88 & 24840 & 475 &  298  & 1200 & 200 & \cite{Lipid17POPCPOPG80201000mMCaCl} \\
+      POPC:POPG (1:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}        &0          & 0  & 150:150 & 31572 & 0  &  298  & 320 & 200 & \cite{Lipid17POPCPOPG5050} \\
+      POPC:POPG (1:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}        &0          & 0.1& 150:150 & 31401 & 57 &  298  & 718 & 198 & \cite{Lipid17POPCPOPG5050100mMCaCl} \\
+      POPC:POPG (1:1)        & Lipid17 / Dang \cite{gould18,smith94,dang06}        &0          & 1.0& 150:150 & 29865 & 569 &  298  & 720 & 200 & \cite{Lipid17POPCPOPG50501000mMCaCl} \\
+      \hline
+      POPC:POPG (4:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     &0          & 0  & 350:88 & 26265 & 0  &  298  & 400 & 300 & \cite{Lipid17eccPOPCPOPG8020} \\
+      POPC:POPG (4:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     &0          & 0.1& 350:88 & 26124 & 47 &  298  & 400 & 300 & \cite{Lipid17eccPOPCPOPG8020100mMCaCl} \\
+      POPC:POPG (4:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     &0          & 1.0& 350:88 & 24840 & 475 &  298  & 400 & 300 & \cite{Lipid17eccPOPCPOPG80201000mMCaCl} \\
+      POPC:POPG (1:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     &0          & 0  & 150:150 & 31572 & 0  &  298  & 347.8 & 333 & \cite{Lipid17eccPOPCPOPG5050} \\
+      POPC:POPG (1:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     &0          & 0.1& 150:150 & 29865 & 54 &  298  & 400 & 300 & \cite{Lipid17eccPOPCPOPG5050100mMCaCl} \\
+      POPC:POPG (1:1)        & Lipid17ecc / ECC-ions \cite{pluharova14,kohagen16,martinek18}     &0          & 1.0& 150:150 & 29865 & 569 &  298  & 600 & 400 & \cite{Lipid17eccPOPCPOPG50501000mMCaCl} \\
+      \hline
+      POPC             & Berger \cite{??} \todoi{This is probable not plain berger, correct force filed should be described.}  &0  & 0  & 256 & 10240 & 0  &  300  & 300 & 200 & \cite{POPCberger300K} \\
+      POPC:POPE (1:1)  & Berger \cite{??}  \todoi{This is probable not plain berger, correct force filed should be described.} &0          & 0  & 128 & 11008 & 0  &  300  & 300 & 200 & \cite{POPC1POPE1berger} \\
       %POPC:DOPE (1:1)  & Berger \cite{??}  \todoi{This is probable not plain berger, correct force filed should be described.} &0          & 0  & 128 & 10240 & 0  &  300  & 300 & 200 & \cite{POPC1DOPE1berger} \\
@@
   \end{minipage}
+  \todo{Citation and description for ''Berger'' model?} \\
 \end{table*}
@@
 
-\noindent {\it POPE} A lipid bilayer, consisting of a total of 144 POPE molecules, distributed equally between the two leaflets was set using CHARMM-GUI \cite{lee16}.
-The bilayer was solvated by 5760 water molecules (40 per lipid).
-The random initial configuration and topologies were generated using the CHARMM-GUI web portal, which provides  GROMACS-compatible simulation input files \cite{lee16}.
-CHARMM36 lipid parameters \cite{klauda10} were used for POPE, whereas the CHARMM-specific TIP3P water model \cite{jorgensen83} was used for water.
-The bilayer was simulated for 500~ns using GROMACS 2018.6 at 310~K.
-
-The recommended simulation parameters for CHARMM36 force field in GROMACS were used \cite{lee16}.
-Namely, buffered Verlet lists were used to keep track of neighbouring atoms \cite{Pall13}.
-The Lennard-Jones potential was cut off at 1.2~nm with the forces switched to 0 between 1.0~nm and the cutoff value.
-The smooth PME algorithm \cite{darden93,essman95} was used to account for long-range electrostatics.
-The temperatures of the lipid and the solvent were separately coupled to a Nos\'{e}-Hoover thermostat \cite{nose84,hoover85} with a time constant of 1~ps and a target temperature of 310~K.
-The system was coupled to a semi-isotropic (isotropic on the membrane plane) Parrinello--Rahman barostat \cite{parrinello81} with a time constant of 5~ps, a reference pressure of 1~bar, and compressibility of 4.5$\times$10$^{-5}$~1/bar.
-The bonds with hydrogen atoms were constrained using p-LINCS \cite{hess07,hess97}.
-The simulation files are available at Ref.~\citenum{charmm36POPEfiles}.
-
-%\noindent {\it POPE with additional NaCl} \todo{Simulation details by A. Peon.}.
-%A pure POPE system was built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935].
-A symmetric lipid distribution was used.
-The bilayer contains 500 lipids and 50 CHARMM TIP3P Water model per lipid.
-The simulated system was minimized, equilibrated and production run.
-Production simulation was 500-ns-long, which the last 100 ns were analysed.
-%For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
-For water, the SETTLE method was applied.
-The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps.
-The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1.
-Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
-van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
-Long-range dispersion correction to the pressure and energy was added.
-
-\noindent {\it POPG with 118 lipids} Lipid bilayer containing 118 POPG molecules, 4110 TIP3P water molecules, and 118 potassium ions was build using CHARMM-GUI \cite{lee16}.
+\noindent {\it POPE} \todo{Simulation details by M. Javanainen}.
+
+%\noindent {\it POPE with additional NaCl} \todo{Simulation details by A. Peon.}
+
+\noindent {\it POPG} Lipid bilayer containing 118 POPG molecules, 4110 TIP3P water molecules, and 118 potassium ions was build using CHARMM-GUI \cite{lee16}.
 The system was simulated 100~ns, coupled to 298 K using Nose-Hoover~\cite{nose84,hoover85} thermostat and 1 bar with
@@
 
-\noindent {\it POPG with 500 lipids} A pure POPG system was built with CHARMM-GUI Membrane builder \cite{lee16}.
-A symmetric lipid distribution was used.
-The bilayer contains 500 lipids and 50 CHARMM TIP3P water molecules per lipid.
-The 500-ns-long production simulation was performed after energy minimization and equilibration.
-The last 100 ns was used for analysis.
-For production simulation, a Leap-Frog algorithm was applied for integrating Newton's equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
-For water, the SETTLE method was applied.
-The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat~\cite{bussi07}.
-Solvent and lipids were coupled to separate thermostats with a coupling constant of 0.1 ps.
-The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello-Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 $\times$ 10$^5$ bar$^{-1}$.
-Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
-Van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
-Long-range dispersion correction to the pressure and energy was added.
-
-\noindent {\it POPC:POPE mixtures} 
-A pure POPC system and a 50:50 POPC:POPE mixture were built and equilibrated using CHARMM-GUI~\cite{lee16}.
-They contained 256 lipids (for the mixture 64 lipids per leaflet for each species) and 34 water molecules per lipid.
-No ions were added.
-The production simulations were run for 300 ns with a time step of 2 fs.
-The first 50 ns were discarded for the analysis.
-The simulations were run with the GROMACS 2016.4~\cite{abraham2015gromacs} version.
-The v-rescale thermostat~\cite{bussi07} was used with a temperature of 300 K and a time constant of 1 ps; lipids and water were coupled separately to the heat bath.
-Pressure was kept constant at 1 bar using a semi-isotropic Parrinello--Rahman barostat~\cite{parrinello81} with a time constant of 5.0 ps.
-A real space cut-off of 1.2 nm was employed for electrostatic interactions while the long-range part was evaluated using the PME method~\cite{darden93,essman95}.
-A force-based switching function was used to switch the Lennard-Jones forces to zero over a range of 1--1.2 nm.
-All bonds with hydrogen atoms were constrained with the LINCS algorithm~\cite{hess07,hess97}.
-Water molecules were kept rigid with the SETTLE algorithm~\cite{miyamoto92}.
-The simulation files are available from Ref.~\citenum{POPCcharmm300K} (pure POPC) and~\citenum{POPC1POPE1charmm36} (POPC:POPE mixture).
+\todo{Simulation details for larger simulation by A. Peon.}
+
+%\noindent {\it POPG with additional NaCl} \todo{Simulation details by A. Peon.}
+
+\noindent {\it POPC:POPE mixtures} Data is available at \cite{POPCcharmm300K,POPC1POPE1charmm36}.
+300 K with v-rescale (tau=0.1 ps),
+1 bar with PR semiisotropic (tau=4 ps, compressibility=4.5e-5 bar$^{-1}$),
+PME order 4 and space 0.12,
+rcoulomb and rvdw 1.0,
+128 lipids per leaflet,
+no ion \todo{Full simulation details by Fuchs et al.}
 
@@
 
-\noindent{\it POPC:POPE mixture}
-A POPC/POPE bilayer with its lipids distributed evenly among the leaflets was generated by from a pure POPC bilayer by removing and renaming atoms in the head group region.
-The bilayer contained a total of 100 POPC and 100 POPE lipids, and it was solvated by 45 water molecules per lipid (for a total of 9000 water molecules).
-The Slipids force field \cite{jambeck12,jambeck12b,jambeck13} was used for lipids, and the TIP3P model \cite{jorgensen83} for water.
-A 300~ns-long simulation was performed using GROMACS 2019.4 \cite{abraham2015gromacs}.
-The simulation parameters were equal to those used for the POPC/POPG mixture with additional CaCl$_2$ (see below).
-The simulation data are available at Ref.~\citenum{POPC1POPE1slipidT298K}.
-
 %\noindent {\it POPC:POPG mixture with additional NaCl} \todo{Simulation details by A. Peon.
@@
 
-\noindent {\it POPC:POPG mixture with additional CaCl$_2$} 
-A lipid bilayer consisting of a total of 256 lipids (128 POPC + 128 POPG) spread equally between the two leaflets was generated using CHARMM-GUI \cite{lee16}.
-The membrane was solvated by 50 water molecules per lipid (a total of 12800), and 128 Na$^{+}$ counter ions for the POPG charges.
-The initial random configuration was equilibrated using the CHARMM-GUI protocol and using the CHARMM36 force field \cite{klauda10}.
-Next, additional ions were added to obtain initial CaCl$_2$ concentrations of 0, 100, 200, 500, or 1000~mM (0/0, 23/46, 46/92, 115/230, 230/460 Ca$^{2+}$/Cl$^{-}$ ions) The systems were energy-minimized and equilibrated using the Slipids force field \cite{jambeck12,jambeck12b,jambeck13} before production simulations at 298~K.
-The ion parameters by \citeauthor{dang06} were used \cite{dang06}.
-
-The neighbour lists with a cutoff of 1.0~nm were updated every 10 simulation steps.
-The smooth PME algorithm was used to calculate long-range electrostatics \cite{darden93,essman95}.
-The Lennard-Jones potential was cut off at 1.0~nm, and the dispersion corrections \cite{shirts07} were applied to energy and pressure.
-The stochastic velocity rescaling thermostat \cite{bussi07} with a time constant of 0.5~ps and a target temperature of 298~K was applied separately to lipids and the solvent.
-A constant pressure of 1~bar was maintained by a Parrinello--Rahman barostat \cite{parrinello81} with a time constant of 10~ps and compressibility of 4.5$\times$10$^{-5}$~1/bar.
-The pressure coupling was performed semi-isotropically with the two simulation box vectors aligned along the membrane plane considered isotropic.
-All bonds were constrained using the p-LINCS algorithm \cite{hess97,hess07}.
-
-The systems simulated for 1500~ns (CaCl$_2$-containing systems) or 500~ns (systems without CaCl$_2$).
-The simulations were performed using GROMACS 2019.4 \cite{abraham2015gromacs}, and the simulation files are available at Ref.~\citenum{slipidPOPC50POPG50T298K}.
+\noindent {\it POPC:POPG mixture with additional CaCl} \todo{Simulation details by M. Javanainen.}
 
@@
 
-\noindent {\it POPC:POPE mixtures} 
-Two systems were simulated using the Berger force field~\cite{Berger97}, pure POPC and a mixture 50:50 POPC:POPE.
-For POPC, parameters with the Bachar~\cite{bachar04} correction for the dihedrals next to double bond were used (Berger-POPC-07).
-For POPE, we additionnally used the de Vries modification implementing a repulsion of the hydrogen atoms located on the amino group of ethanolamine~\cite{devries04}.
-Starting from a PDB file of a pure POPC system with Berger atom names, a 50:50 POPC:POPE mixture was built by mutating randomly methyl groups to hydrogens (POPC -> POPE).
-Each system contained 256 lipids (for the mixture 64 lipids per leaflet for each species) and about 40 water molecules per lipid.
-No ions were added.
-Both systems were minimized and equilibrated.
-The production simulations were run for 300 ns with a time step of 2 fs.
-The first 100 ns were discarded for the analysis.
-The simulations were run with GROMACS 4.5.3~\cite{pronk13} version.
-The v-rescale thermostat~\cite{bussi07} was used with a temperature of 300 K and a time constant of 0.1 ps; lipids and water were coupled separately to the heat bath.
-Pressure was kept constant at 1 bar using a semi–isotropic Parrinello–Rahman barostat~\cite{parrinello81} with a time constant of 4.0 ps.
-A real space cut-off of 1.0 nm was employed for van der Waals and electrostatic interactions.
-The long-range part of electrostatic interactions was evaluated using the PME method~\cite{darden93,essman95} with a grid spacing of 0.12 nm and an interpolation order of 4.
-All bonds with hydrogen atoms were constrained with the LINCS algorithm~\cite{hess07,hess97}.
-Water molecules were kept rigid with the SETTLE algorithm~\cite{miyamoto92}.
-The simulation files are available from
-Ref.~\citenum{POPCberger300K} (pure POPC) and~\citenum{POPC1POPE1berger} (POPC:POPE mixture).
+\noindent {\it POPC:POPE mixtures} Data is available at \cite{POPCberger300K,POPC1POPE1berger}. 
+300 K with v-rescale (tau=0.1 ps),
+1 bar with PR semiisotropic (tau=4 ps, compressibility=4.5e-5 bar$^{-1}$),
+PME order 4 and space 0.12,
+rcoulomb and rvdw 1.0,
+128 lipids per leaflet,
+no ion \todo{Simulation details by Fuchs et al.}
 
@@
 
-\noindent {\it POPE} POPE simulations with the OPLS-UA force field were performed for 200 ns using GROMACS 4.5.7.
-POPE parameters were constructed by modifying the OPLS-UA POPC of Ulmschneider and Ulmschneider \cite{Ulmschneider09} with standard OPLS lysine parameters.
-The starting membrane structure, containing 64 POPE lipids per leaflet, was created by modifying an OPLS-UA POPC membrane.
-Simulations employed a 1.0 nm cut-off with PME and no dispersion correction.
-Further simulation details and data can be found at \citenum{OPLSuaPOPEfiles}.
-
-\noindent {\it POPE with vdW interaction in H (OPLS-UA/HG-H)} In addition to the OPLS simulations mentioned above, further simulations employing the same starting structure and simulation settings were employed using slightly modified parameters.
-These modified parameters, termed OPLS-UA/HG-H, were designed to increase the area per lipid through employing a small repulsive potential on the ethanolamine hydrogen atoms, as per the approach of de Vries with the Berger POPE parameters \cite{devries04}.
-Further details and data of these simulations can be found at Ref.\citenum{OPLSuaWvdWPOPEfiles}.
-
-\subsection{GROMOS-CKP}
-
-\noindent {\it POPE} 
-Pure POPE system was built with CHARMM-GUI Membrane builder \cite{lee16}.
-Symmetric lipid distribution was used.
-The bilayer contains 500 lipids and 50 SPC Water model per lipid.
-GROMOS-CKP force field was used for lipids \cite{piggot11}.
-%and GROMOS54A7 for the rest.
-The 500-ns-long production simulation was performed after energy minimization and equilibration.
-The last 100 ns was used for analysis.
-For production simulation, a Leap-Frog algorithm was applied for integrating Newton's equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
-For water, the SETTLE method was applied.
-The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat \cite{bussi07}.
-Solvent and lipids were coupled to separate thermostats with a coupling constant of 0.1 ps.
-The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello-Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 $\times$ 10$^5$ bar$^{-1}$.
-Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
-van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
-Long-range dispersion correction to the pressure and energy was added.
-The simulation data and files are available at Ref. \citenum{gromosCKPpopeT310}.
-
-%\noindent {\it DOPE} Data is available at \cite{gromosCKPdope}.
-
-\noindent {\it DPPE} GROMOS-CKP DPPE simulations were performed for 500 ns using GROMACS 5.0.6.
-These simulations used the original GROMOS-CKP parameters \cite{piggot11} but with the charges in the ethanolamine head group taken from a GROMOS 53A6 lysine side-chain rather than a PC lipid head group, as done in the original parameters.
-The starting structure contained 64 lipids per leaflet and was taken from a previous GROMOS-CKP simulation \cite{piggot11}.
-Simulations employed standard GROMOS-CKP settings with PME employed for the long-range electrostatic interactions.
-Further details and simulation data can be found at Ref.\citenum{gromosCKPdppe}.
-
-\noindent {\it POPG} 
-A pure L-POPG system was built with CHARMM-GUI Membrane builder \cite{lee16}.
-A symmetric lipid distribution was used.
-The bilayer contains 500 lipids and 50 SPC Water model per lipid with 500 sodium ions.
-GROMOS-CKP force field was used for lipid \cite{piggot11} and GROMOS54A7 \cite{schmid11} for the rest.
-The 500-ns-long production simulation was performed after energy minimization and equilibration.
-The last 100 ns was used for analysis.
-The simulation details are the same as in POPE system.
-%For production simulation, a Leap-Frog algorithm was applied for integrating Newton's equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
-For water, the SETTLE method was applied.
-The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat \cite{bussi07}.
-%Solvent and lipids were coupled to separate thermostats with a coupling constant of 0.1 ps.
-The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello-Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 $\times$ 10$^5$ bar$^{-1}$.
-Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
-Van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
-Long-range dispersion correction to the pressure and energy was added.
-The data and simulation files are available at Ref.~\citenum{POPGgromosCKP}.
-
-\noindent {\it POPC:POPG mixture} 
-Symmetric membranes of pure POPC and POPC:POPG 7:3 mixtures containing 500 lipids and 50 SPC water molecules per lipid were composed with CHARMM-GUI \cite{lee16}.
-The POPC:POPG 7:3 mixture contained also 150 sodium counter ions to balance the charge of PG lipids.
-GROMOS-CKP force field~\cite{piggot11,piggot12} was used for lipids and GROMOS54A7 \cite{schmid11} for the rest.
-Other simulation details are as in POPE and POPG systems.
+\noindent {\it POPE} Data is available at \cite{OPLSuaPOPEfiles}. \todo{Simulation details by  T. Piggot.}
+%From Zenodo: The PE parameters were constructed by modifying the OPLS-UA POPC of Ulmschneider and Ulmschneider with the standard OPLS lysine parameters.
+
+\noindent {\it POPE with vdW interaction in H (OPLS-UA/HG-H)} Data is available at \cite{OPLSuaWvdWPOPEfiles}. \todo{Simulation details by  T. Piggot.}
+%From Zenodo: The PE parameters were constructed by modifying the OPLS-UA POPC of Ulmschneider and Ulmschneider with the standard OPLS lysine parameters and the addition of LJ parameters on the hydrogen atoms in the head group (designed to increase the area per lipid).
+
+\subsection{GROMOS-CKP and GROMOS-CKPM}
+
+\noindent {\it POPE} Data is available at \cite{gromosCKPpopeT310}. \todo{Simulation details by  A. Peon.}
+
+%\noindent {\it DOPE} Data is available at \cite{gromosCKPdope}. \todo{Simulation details by  T. Piggot.}
+
+\noindent {\it DPPE} Data is available at \cite{gromosCKPdppe}. \todo{Simulation details by  T. Piggot.}
+
+\noindent {\it POPG} \todo{Simulation details by A. Peon.}
+
+\noindent {\it POPC:POPG mixture} \todo{Simulation details by A. Peon.}
 
@@
 
-\noindent {\it POPE} A bilayer patch with a total of 144 POPE molecules, distributed evenly between two leaflets, was created by reordering atoms in a final structure of a POPE simulation performed using the CHARMM36 force field.
-The bilayer was hydrated by 40 water molecules per lipid for a total of 5760 water molecules.
-The OPLS-based MacRog force field was used \cite{Kulig15b} for lipids and the TIP3P model \cite{jorgensen83} for water.
-
-Buffered Verlet lists \cite{Pall13} were used to keep track of neighbours.
-The Lennard-Jones potential was cut off at 1.0~nm, and dispersion corrections were applied to energy and pressure \cite{shirts07}.
-Smooth PME algorithm was used to calculate long-range electrostatics \cite{darden93,essman95}.
-The Nos\'{e}--Hoover thermostat was used to keep the temperatures of the lipids and water at 310~K.
-These groups were coupled separately, and a time constant of 0.4~ps was used.
-The Parrinello--Rahman barostat~\cite{parrinello81} was used to keep the pressures in the membrane plane as well as normal to it constant at 1~bar.
-For the barostat, a time constant of 10~ps was used, and the membrane compressibility was set to 4.5$\times$10$^{-5}$~1/bar.
-All bonds were constrained using p-LINCS \cite{hess97,hess07}.
-
-The system was simulated for 500~ns using GROMACS 2019.2 \cite{abraham2015gromacs}, and the simulation data are available at Ref.~\citenum{MacRogPOPEfiles}.
-
-\noindent {\it POPC:POPE mixtures} The initial force field parameter files in GROMACS format were taken from Ref.~\citenum{rog16}. 
-However, a number of errors were detected in the published files so we fixed them in the following way. 
-For POPE, the two aliphatic chains sn-1 and sn-2 were switched (in the file the lipid was in fact OPPE); 
-one atom (named C27) was not connected to the previous atom in the aliphatic chain so bonds, angles, dihedrals, 
-pairs were included to create the connection; two atoms were called C27 thus one of them was renamed C28. 
-For POPC, some impropers were missing leading to non planar systems for double bonds 
-(in particular the carbonyls of sn-1 and sn-2 as well as the double bond of the oleoyl chain), 
-we thus added back those impropers. The resulting fixed files can be found from Refs.~\citenum{POPCmacrog300K} (popc\_fixed.itp) and~\citenum{POPC1POPE1macrogT300K} (pope\_fixed.itp). 
-To check our fix, one simulation of pure POPC was also run and the order parameter compared to previous published results~\cite{botan15}. We found very similar values.
-All the details of the procedure described here are described 
-in file {\it scratch/report\_results\_comparison.pdf} in Ref.~\citenum{NMRlipidsIVbgit}.
-
-Two systems of pure POPC and 50:50 POPC:POPE mixture were built using CHARMM-GUI~\cite{lee16}.
-They contained 128 lipids and 5120 TIP3 water molecules.
-The initial PDB file was modified to match OPLS-MacRog nomenclature and atom order.
-No ions were added.
-The production simulations were run for 500 ns with a time step of 2 fs.
-The first 200 ns were discarded for the analysis.
-The simulations were run with the GROMACS 2018.5~\cite{abraham2015gromacs} version.
-The v-rescale thermostat~\cite{bussi07} was used with a temperature of 300 K and a time constant of 0.1 ps; lipids and water were coupled separately to the heat bath.
-Pressure was kept constant at 1 bar using a semi-isotropic Parrinello-Rahman barostat~\cite{parrinello81} with a time constant of 4.0 ps.
-Long-range electrostatic interactions were calculated using the PME method~\cite{darden93,essman95}.
-A real space cut-off of 1.0 nm was employed with a grid spacing of 0.1 nm in the reciprocal space.
-Lennard-Jones potentials were cut off at 1.0 nm.
-All covalent bonds in lipids were constrained using the LINCS algorithm~\cite{hess07,hess97}.
-Water molecules were kept rigid with the SETTLE algorithm~\cite{miyamoto92}.
-
-The data for pure POPC are available from Refs.~\citenum{POPCmacrog300K}, and for the POPC:POPE mixture from Ref.~\citenum{POPC1POPE1macrogT300K}.
+\noindent {\it POPE} \todo{Simulation details by M. Javanainen} % and P. Fuchs.}
+
+\noindent {\it POPC:POPE mixtures} \todo{Simulation details by P. Fuchs.}
 
@@
 
-\noindent {\it POPE} 
-A pure POPE system built with CHARMM-GUI was used to create the Amber input files.
-Amber prmtop and inpcrd files were obtained with LEaP from AmberTools 16 distribution (\url{http://ambermd.org/}).
-ParmEd version 3.0.0~\cite{shirts17} was used to convert the Amber prmtop and inpcrd files into Gromacs top and gro files. 
-After conversion, the incorrect dihedrals with type 1 were changed to type 9
-(for details, see discussion in \url{https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/12}).
-
-The bilayer contains 500 lipids and 50 TIP3P waters per lipid.
-The 500-ns-long production simulation was performed after energy minimization and equilibration.
-The last 100 ns was used for analysis.
-For production simulation, a Leap-Frog algorithm was applied for integrating Newton's equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
-For water, the SETTLE method was applied.
-The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat~\cite{bussi07}.
-Solvent and lipids were coupled to separate thermostats with a coupling constant of 0.1 ps.
-The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello-Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 $\times$ 10$^5$ bar$^{-1}$.
-Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
-van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
-Long-range dispersion correction to the pressure and energy was added.
-
-\noindent {\it POPG} 
-A pure POPG system built with CHARMM-GUI was used to create the Amber input files.
-The system was further processed and simulated similarly to the POPE system.
-
-%Amber prmtop and inpcrd files were obtained with LEaP from AmberTools 16 distribution (http://ambermd.org/).
-Then, in order to create the Gromacs input files, ParmEd (v 3.0.0, Swails J.
-ParmEd [Internet].
-https://github.com/ParmEd/ParmEd; https://dx.doi.org/10.1007/s10822-016-9977-1) was used to convert the Amber prmtop and inpcrd files into Gromacs top and gro files.
-%The bilayer contains 500 lipids and 50 TIP3P waters per lipid and 500 sodium ions.
-The simulated system was minimized, equilibrated and production run.
-Production simulation was 500-ns-long, which the last 100 ns were analysed.
-%For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs.
-For water, the SETTLE method was applied.
-The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps.
-The pressure was kept constant with a semi-isotropic scheme.
-The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1.
-Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh.
-van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm.
-Long-range dispersion correction to the pressure and energy was added.
+\noindent {\it POPE} \todo{Simulation details by A. Peon.}
+
+\noindent {\it POPG} \todo{Simulation details by A. Peon.}
 
 \noindent {\it POPC:POPG 4:1 and POPC:POPG 1:1 mixtures with different CaCl$_2$ concentrations}
-Initial structures were build by removing appropriate amount of lipids from POPC:POPG 7:3 mixture available from Ref.~\citenum{POPCPOPG73lipid17},
-which was build similarly to pure POPG and POPE systems.
-Force field parameters from the same reference 
-(Amber prmtop and inpcrd files were obtained with LEaP from AmberTools 16 distribution (\url{http://ambermd.org/})
-and converted to Gromacs top and gro files with ParmEd version 3.0.0~\cite{shirts17})
-were used, except that incorrect dihedrals with type 1 were changed to type 9
+Initial structures were build by removing appropriate amount of lipids from POPC:POPG 7:3 mixture available from Ref.~\citenum{POPCPOPG73lipid17}.
+Force field parameters from the same reference were used \todo{We still need description from A. Peon how these were obtained},
+except that incorrect dihedrals with type 1 were changed to type 9
 (for details, see discussion in \url{https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/12}).
@@
 \noindent
-{\it Am{\'e}lie Bacle} set up, performed and analysed POPC and POPC:POPE (1:1) simulations with the Berger force field.
+{\it Am{\'e}lie Bacle}
 %% Please write description of your contribution here
@@
 %% Please write description of your contribution here
-contributed to the analysis of dihedrals and isomers of lipids from MD simulations, and to the analysis of lipid structures from the Protein Data Bank. Prepared panels for figures 2 and 4.
 
@@
 {\it Rebeca Garc{\'i}a Fandi{\~n}o}
-designed and supervised the molecular dynamics simulations carried out by Antonio Pe{\'o}n.
 %% Please write description of your contribution here
@@
 {\it Fernando Favela-Rosales}
-set up and performed DPPE, POPG and DPPG simulations with the Slipids Force Field.
 %% Please write description of your contribution here
@@
 {\it Tiago M. Ferreira}
-was responsible for the solid-state NMR experiments. Prepared figures~\ref{HGorderParameters},~\ref{POPEspectra}, and~\ref{POPGspectra}, and took part in writing the manuscript.
-
-\noindent
-{\it Patrick F.J. Fuchs} supervised Paula Mil{\'a}n Rodr{\'i}guez, Am{\'e}lie Bacle and Chris Papadopoulos,
-created the buildH software for the calculations of order parameters
-from united atom simulations,
-and contributed to the calculation of lipid isomers. Provided comments on the manuscript.
-%% Please write description of your contribution here
-
-\noindent
-{\it Ivan Gushchin} supervised Pavel Buslaev and contributed to the analysis of lipid structures in the Protein Data Bank. Provided comments on the manuscript.
-%% Please write description of your contribution here
-
-\noindent
-{\it Matti Javanainen} set up and performed simulations using CHARMM36, Slipids, and MacRog lipid models. Provided comments on the manuscript.
-%% Please write description of your contribution here
-
-\noindent
-{\it Anne M.
-Kiirikki} developed the preliminary version of the NMRlipids databank, which was used in several analyses.
-Set up and performed POPC:POPG(1:1) and POPC:POPG(4:1) simulations with CHARMM36 and POPC:POPG(4:1) with Lipid17ecc.
+%% Please write description of your contribution here
+
+\noindent
+{\it Patrick Fuchs}
+%% Please write description of your contribution here
+
+\noindent
+{\it Ivan Gushchin}
+%% Please write description of your contribution here
+
+\noindent
+{\it Matti Javanainen}
+%% Please write description of your contribution here
+
+\noindent
+{\it Anne M. Kiirikki}
 %% Please write description of your contribution here
@@
 {\it Josef Melcr}
-%created the analysis script \texttt{scripts/calcOrderParameters.py}, 
-contributed to the preparation of the Lipid17ecc parameters. 
-Provided comments on the manuscript.
-
-\noindent
-{\it Paula Mil{\'a}n Rodr{\'i}guez} corrected errors in available MacRog force field parameters. Set up, performed and analysed POPC and POPC:POPE (1:1) simulations with the MacRog force field.
+%% Please write description of your contribution here
+
+\noindent
+{\it Paula Milan Rodriguez}
 %% Please write description of your contribution here
@@
 {\it Markus S. Miettinen}
-made figures~\ref{HGorderParametersPE} and~\ref{HGorderParametersPOPG}. Provided comments on the manuscript.
 %% Please write description of your contribution here
@@
 \noindent
-{\it Chris G. Papadopoulos} set up, performed and analysed POPC and POPC:POPE (1:1) simulations with the CHARMM36 Force Field.
-%% Please write description of your contribution here
-
-\noindent
-{\it Antonio Pe{\'o}n} set up, performed and analysed POPC, POPE, POPG and POPC:POPG (7:3) simulations with the CHARMM36, SLIPIDS, GROMOS-CKP and Lipid17 Force Fields.
+{\it Chris G. Papadopoulos}
+%% Please write description of your contribution here
+
+\noindent
+{\it Antonio Pe{\'o}n}
 %% Please write description of your contribution here
@@
 {\it {\'A}ngel Pi{\~n}eiro}
-created the opAAUA\_prod.py code for the calculation of order parameters from united atom trajectories.
-%% Please write description of your contribution here
-
-\noindent
-{\it Salla I. Virtanen}
-prepared Lipid17ecc parameters.
 %% Please write description of your contribution here

### 2021-07-28 manuscriptPGPE.tex
Update author information and minor fixes for the last submission.

--- 2021-07-12 manuscriptPGPE.tex
@@
 \author{Am{\'e}lie Bacle}
-\affiliation{\tiny Laboratoire Coop{\'e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit{\'e} de Poitiers, 1 rue Georges Bonnet, 86000 Poitiers, France }
+\affiliation{\tiny Laboratoire Coop{\'e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit{\'e} de Poitiers, 1 rue Georges Bonnet,  Poitiers, 86000, France }
 
 \author{Pavel Buslaev}
-\affiliation{Nanoscience Center and Department of Chemistry, University of Jyv{\"a}skyl{\"a}, P.O. Box 35, 40014 Jyv{\"a}skyl{\"a}, Finland}
-\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, 141701 Dolgoprudny, Russia}
+\affiliation{Nanoscience Center and Department of Chemistry, University of Jyv{\"a}skyl{\"a}, P.O. Box 35, Jyv{\"a}skyl{\"a}, 40014 , Finland}
+\alsoaffiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology,  Dolgoprudny, 141701, Russia}
 
-\author{Rebeca Garc{\'i}a Fandi{\~n}o}
-\affiliation{Center for Research in Biological Chemistry and Molecular Materials (CiQUS), Universidade de Santiago de Compostela, E-15782 Santiago de Compostela, Spain}
-\affiliation{CIQUP, Centro de Investigação em Qu{\'i}mica, Departamento de Qu{\'i}mica e Bioqu{\'i}mica, Faculdade de Ci{\^e}ncias, Universidade do Porto, Porto, Portugal}
+\author{Rebeca Garc{i}a-Fandi{n}o}
+\affiliation{Center for Research in Biological Chemistry and Molecular Materials (CiQUS), Universidade de Santiago de Compostela,  Santiago de Compostela, E-15782, Spain}
+\alsoaffiliation{CIQUP, Centro de Investigação em Qu{\'i}mica, Departamento de Qu{\'i}mica e Bioqu{\'i}mica, Faculdade de Ci{\^e}ncias, Universidade do Porto, Porto, 4169-007, Portugal}
 
 \author{Fernando Favela-Rosales}
-\affiliation{Departamento de Ciencias B\'{a}sicas, Tecnol\'{o}gico Nacional de M\'{e}xico - ITS Zacatecas Occidente, M\'{e}xico}
+\affiliation{Departamento de Ciencias B\'{a}sicas, Tecnol\'{o}gico Nacional de M\'{e}xico - ITS Zacatecas Occidente, Sombrerete, Zacatecas, 99102, M\'{e}xico}
 
-\author{Tiago M. Ferreira}
-\affiliation{NMR group - Institute for Physics, Martin Luther University Halle-Wittenberg, 06120 Halle (Saale), Germany}
+\author{Tiago Mendes Ferreira}
+\affiliation{NMR group - Institute for Physics, Martin Luther University Halle-Wittenberg,  Halle (Saale), 06120, Germany}
 
 \author{Patrick F.J. Fuchs}
-\affiliation{Sorbonne Universit{\'e}, Ecole Normale Sup{\'e}rieure, PSL University, CNRS, Laboratoire des Biomol{\'e}cules (LBM), 75005 Paris, France}
-\affiliation{Universit{\'e} de Paris, UFR Sciences du Vivant, 75013, Paris, France}
+\affiliation{Sorbonne Universit{\'e}, Ecole Normale Sup{\'e}rieure, PSL University, CNRS, Laboratoire des Biomol{\'e}cules (LBM), Paris, 75005, France}
+\alsoaffiliation{Universit{\'e} de Paris, UFR Sciences du Vivant, Paris, 75013, France}
 
 \author{Ivan Gushchin}
-\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, 141701 Dolgoprudny, Russia}
+\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, Dolgoprudny, 141701, Russia}
 
@@
 \affiliation{Institute of Organic Chemistry and Biochemistry of the 
-Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
+Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2,  Prague, CZ-16610, Czech Republic}
 
 \author{Anne M. Kiirikki}
-\affiliation{Institute of Biotechnology, University of Helsinki}
+\affiliation{Institute of Biotechnology, University of Helsinki, Helsinki, 00014, Finland}
 
 \author{Jesper J. Madsen}
-\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
-\affiliation{Global and Planetary Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
+\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, 60637, United States of America}
+\alsoaffiliation{Global and Planetary Health, College of Public Health, University of South Florida, Tampa, Florida, 33612, United States of America}
 
@@
 and The Zernike Institute for Advanced Materials, 
-University of Groningen, 9747 AG Groningen, The Netherlands}
+University of Groningen, Groningen, 9747 AG, The Netherlands}
 
 \author{Paula Mil{\'a}n Rodr{\'i}guez}
-\affiliation{Sorbonne Universit{\'e}, Ecole Normale Sup{\'e}rieure, PSL University, CNRS, Laboratoire des Biomol{\'e}cules (LBM), 75005 Paris, France}
+\affiliation{Sorbonne Universit{\'e}, Ecole Normale Sup{\'e}rieure, PSL University, CNRS, Laboratoire des Biomol{\'e}cules (LBM), Paris, 75005, France}
 
 \author{Markus S. Miettinen}
-\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
-% \affiliation[Max Planck Institute of Colloids and Interfaces]{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
+\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, Potsdam, 14424, Germany}
 
@@
 \email{samuli.ollila@helsinki.fi}
-\affiliation{Institute of Biotechnology, University of Helsinki}
+\affiliation{Institute of Biotechnology, University of Helsinki, Helsinki, 00014, Finland}
 
 \author{Chris G. Papadopoulos}
-\affiliation{Universit{\'e} Paris-Saclay, CEA, CNRS, Institute for Integrative Biology of the Cell (I2BC), 91198 Gif-sur-Yvette, France}
+\affiliation{Universit{\'e} Paris-Saclay, CEA, CNRS, Institute for Integrative Biology of the Cell (I2BC), Gif-sur-Yvette, 91198, France}
 
 \author{Antonio Pe{\'o}n}
-\affiliation{CIQUP, Centro de Investigação em Qu{\'i}mica, Departamento de Qu{\'i}mica e Bioqu{\'i}mica, Faculdade de Ci{\^e}ncias, Universidade do Porto, Porto, Portugal}
+\affiliation{CIQUP, Centro de Investigação em Qu{\'i}mica, Departamento de Qu{\'i}mica e Bioqu{\'i}mica, Faculdade de Ci{\^e}ncias, Universidade do Porto, Porto, 4169-007, Portugal}
 
 \author{Thomas J. Piggot}
-\affiliation{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, United Kingdom}
+\affiliation{Chemistry, University of Southampton, Highfield, Southampton, SO17 1BJ, United Kingdom}
 
 \author{{\'A}ngel Pi{\~n}eiro}
-\affiliation{Departamento de F{\'i}sica Aplicada, Facultade de F{\'i}sica, Universidade de Santiago de Compostela, E-15782 Santiago de Compostela, Spain}
+\affiliation{Departamento de F{\'i}sica Aplicada, Facultade de F{\'i}sica, Universidade de Santiago de Compostela, Santiago de Compostela, E-15782, Spain}
 
 \author{Salla I. Virtanen}
-\affiliation{Institute of Biotechnology, University of Helsinki}
+\affiliation{Institute of Biotechnology, University of Helsinki, Helsinki, 00014, Finland}
 
@@
 
-     \includegraphics[height=3.5cm]{./Figs/TOC2.eps}
+     \includegraphics[height=3.5cm]{./Figs/TOC3.png}
 %Some journals require a graphical entry for the Table of Contents.
@@
 To understand the structural origin of distinct order parameters for the distinct lipids,
-we first calculated and visualized the heavy atom dihedral angle distributions from the best performing simulations that reproduced the differences between headgroups~(Figs.~\ref{structures} B and \ref{DIHdists}).
+we first calculated and visualized the heavy atom dihedral angle distributions from the best performing simulations that reproduced the differences between headgroups~(Figs.~\ref{structures} A and \ref{DIHdists}).
 Then, we used the inverse Boltzmann formula to estimate the free
@@
   % Put your acknowledgments here.
-A.P. is grateful to the Centro de
-Supercomputaci{\'o}n de Galicia (CESGA) for use of the Finis
-Terrae computer.
 M.J. thanks CSC -- IT Center for Science for computational resources and the Emil Aaltonen foundation for financial support.
@@
 T.J.P. acknowledges use of the Iridis high-performance computing resources at the University of Southampton.
-J.M. thanks the Center for Information Technology of the University of Groningen for their support and for providing access to the Peregrine high performance computing cluster.
+J.M.
+thanks the Center for Information Technology of the University of Groningen for their support and for providing access to the Peregrine high performance computing cluster.
+R.G.-F.
+and A.
+P.
+acknowledge the financial support from Fundação para a Ciência e Tecnologia (FCT) and FEDER European funds, through the project PTDC/BIA-BFS/30579/2017 (POCI-01-0145-30579) and UIDB/00081/2020.
+R.G.-F.
+thanks Ministerio de Ciencia, Innovación y Universidades for a “Ramón y Cajal” contract (RYC-2016-20335), and also to the Spanish Agencia Estatal de Investigación (AEI) and the ERDF (RTI2018-098795-A-I00) and the Xunta de Galicia and the ERDF (ED431F 2020/05), and Centro singular de investigación de Galicia accreditation 2016-2019, ED431G/09).
+Á.
+P.
+acknowledges to the Spanish Agencia Estatal de Investigación (AEI) and the ERDF (PID2019-111327GB-I00).
+R.G.-F., A.
+P.
+and Á.
+P.
+acknowledge to the Centro de Supercomputación de Galicia (CESGA) for technical support and computing time.
 

### 2021-07-28 manuscriptPGPEsuppl.tex
Update author information and minor fixes for the last submission.

--- 2021-07-12 manuscriptPGPEsuppl.tex
@@
 \author{Am{\'e}lie Bacle}
-\affiliation{\tiny Laboratoire Coop{\'e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit{\'e} de Poitiers, 1 rue Georges Bonnet, 86000 Poitiers, France }
+\affiliation{\tiny Laboratoire Coop{\'e}ratif "Lipotoxicity and Channelopathies - ConicMeds", Universit{\'e} de Poitiers, 1 rue Georges Bonnet,  Poitiers, 86000, France }
 
 \author{Pavel Buslaev}
-\affiliation{Nanoscience Center and Department of Chemistry, University of Jyv{\"a}skyl{\"a}, P.O. Box 35, 40014 Jyv{\"a}skyl{\"a}, Finland}
-\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, 141701 Dolgoprudny, Russia}
-
-\author{Rebeca Garc{\'i}a Fandi{\~n}o}
-\affiliation{Center for Research in Biological Chemistry and Molecular Materials (CiQUS), Universidade de Santiago de Compostela, E-15782 Santiago de Compostela, Spain}
-\affiliation{CIQUP, Centro de Investigação em Qu{\'i}mica, Departamento de Qu{\'i}mica e Bioqu{\'i}mica, Faculdade de Ci{\^e}ncias, Universidade do Porto, Porto, Portugal}
+\affiliation{Nanoscience Center and Department of Chemistry, University of Jyv{\"a}skyl{\"a}, P.O. Box 35, Jyv{\"a}skyl{\"a}, 40014 , Finland}
+\alsoaffiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology,  Dolgoprudny, 141701, Russia}
+
+\author{Rebeca Garc{i}a-Fandi{n}o}
+\affiliation{Center for Research in Biological Chemistry and Molecular Materials (CiQUS), Universidade de Santiago de Compostela,  Santiago de Compostela, E-15782, Spain}
+\alsoaffiliation{CIQUP, Centro de Investigação em Qu{\'i}mica, Departamento de Qu{\'i}mica e Bioqu{\'i}mica, Faculdade de Ci{\^e}ncias, Universidade do Porto, Porto, 4169-007, Portugal}
 
 \author{Fernando Favela-Rosales}
-\affiliation{Departamento de Ciencias B\'{a}sicas, Tecnol\'{o}gico Nacional de M\'{e}xico - ITS Zacatecas Occidente, M\'{e}xico}
-
-\author{Tiago M. Ferreira}
-\affiliation{NMR group - Institute for Physics, Martin Luther University Halle-Wittenberg, 06120 Halle (Saale), Germany}
+\affiliation{Departamento de Ciencias B\'{a}sicas, Tecnol\'{o}gico Nacional de M\'{e}xico - ITS Zacatecas Occidente, Sombrerete, Zacatecas, 99102, M\'{e}xico}
+
+\author{Tiago Mendes Ferreira}
+\affiliation{NMR group - Institute for Physics, Martin Luther University Halle-Wittenberg,  Halle (Saale), 06120, Germany}
 
 \author{Patrick F.J. Fuchs}
-\affiliation{Sorbonne Universit{\'e}, Ecole Normale Sup{\'e}rieure, PSL University, CNRS, Laboratoire des Biomol{\'e}cules (LBM), 75005 Paris, France}
-\affiliation{Universit{\'e} de Paris, UFR Sciences du Vivant, 75013, Paris, France}
+\affiliation{Sorbonne Universit{\'e}, Ecole Normale Sup{\'e}rieure, PSL University, CNRS, Laboratoire des Biomol{\'e}cules (LBM), Paris, 75005, France}
+\alsoaffiliation{Universit{\'e} de Paris, UFR Sciences du Vivant, Paris, 75013, France}
 
 \author{Ivan Gushchin}
-\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, 141701 Dolgoprudny, Russia}
+\affiliation{Research Center for Molecular Mechanisms of Aging and Age-related Diseases, Moscow Institute of Physics and Technology, Dolgoprudny, 141701, Russia}
 
@@
 \email{samuli.ollila@helsinki.fi}
-\affiliation{Institute of Biotechnology, University of Helsinki}
+\affiliation{Institute of Biotechnology, University of Helsinki, Helsinki, 00014, Finland}
 
 \author{Chris G. Papadopoulos}
-\affiliation{Universit{\'e} Paris-Saclay, CEA, CNRS, Institute for Integrative Biology of the Cell (I2BC), 91198 Gif-sur-Yvette, France}
+\affiliation{Universit{\'e} Paris-Saclay, CEA, CNRS, Institute for Integrative Biology of the Cell (I2BC), Gif-sur-Yvette, 91198, France}
 
 \author{Antonio Pe{\'o}n}
-\affiliation{CIQUP, Centro de Investigação em Qu{\'i}mica, Departamento de Qu{\'i}mica e Bioqu{\'i}mica, Faculdade de Ci{\^e}ncias, Universidade do Porto, Porto, Portugal}
+\affiliation{CIQUP, Centro de Investigação em Qu{\'i}mica, Departamento de Qu{\'i}mica e Bioqu{\'i}mica, Faculdade de Ci{\^e}ncias, Universidade do Porto, Porto, 4169-007, Portugal}
 
 \author{Thomas J. Piggot}
-\affiliation{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, United Kingdom}
+\affiliation{Chemistry, University of Southampton, Highfield, Southampton, SO17 1BJ, United Kingdom}
 
 \author{{\'A}ngel Pi{\~n}eiro}
-\affiliation{Departamento de F{\'i}sica Aplicada, Facultade de F{\'i}sica, Universidade de Santiago de Compostela, E-15782 Santiago de Compostela, Spain}
+\affiliation{Departamento de F{\'i}sica Aplicada, Facultade de F{\'i}sica, Universidade de Santiago de Compostela, Santiago de Compostela, E-15782, Spain}
 
 \author{Salla I. Virtanen}
-\affiliation{Institute of Biotechnology, University of Helsinki}
-
-%\author{O. H. Samuli Ollila}
-%\email{samuli.ollila@helsinki.fi}
-%%\homepage[]{Your web page}
-%\affiliation{Institute of Organic Chemistry and Biochemistry,
-%Academy of Sciences of the Czech Republic, 
-%Prague 6, Czech Republic}
-%\affiliation{Institute of Biotechnology, University of Helsinki}
+\affiliation{Institute of Biotechnology, University of Helsinki, Helsinki, 00014, Finland}
 
@@
 \noindent
-{\it Rebeca Garc{\'i}a Fandi{\~n}o}
+{\it Rebeca Garc{i}a-Fandi{n}o}
 designed and supervised the molecular dynamics simulations carried out by Antonio Pe{\'o}n.

### 2021-07-28 manuscriptPGPE.tex
update pdfs

@@
 P.
-acknowledge the financial support from Fundação para a Ciência e Tecnologia (FCT) and FEDER European funds, through the project PTDC/BIA-BFS/30579/2017 (POCI-01-0145-30579) and UIDB/00081/2020.
+acknowledge the financial support from Funda{\c c}{\=a}o para a Ci{\^ e}ncia e Tecnologia (FCT) and FEDER European funds, through the project PTDC/BIA-BFS/30579/2017 (POCI-01-0145-30579) and UIDB/00081/2020.
 R.G.-F.
-thanks Ministerio de Ciencia, Innovación y Universidades for a “Ramón y Cajal” contract (RYC-2016-20335), and also to the Spanish Agencia Estatal de Investigación (AEI) and the ERDF (RTI2018-098795-A-I00) and the Xunta de Galicia and the ERDF (ED431F 2020/05), and Centro singular de investigación de Galicia accreditation 2016-2019, ED431G/09).
-Á.
+thanks Ministerio de Ciencia, Innovaci{\'o}n y Universidades for a ''Ram{\'o}n y Cajal'' contract (RYC-2016-20335), and also to the Spanish Agencia Estatal de Investigaci{\'o}n (AEI) and the ERDF (RTI2018-098795-A-I00) and the Xunta de Galicia and the ERDF (ED431F 2020/05), and Centro singular de investigaci{\'o}n de Galicia accreditation 2016-2019, ED431G/09).
+{\'A}.
 P.
-acknowledges to the Spanish Agencia Estatal de Investigación (AEI) and the ERDF (PID2019-111327GB-I00).
+acknowledges to the Spanish Agencia Estatal de Investigaci{\'o}n (AEI) and the ERDF (PID2019-111327GB-I00).
 R.G.-F., A.
 P.
-and Á.
+and {\'A}.
 P.
-acknowledge to the Centro de Supercomputación de Galicia (CESGA) for technical support and computing time.
+acknowledge to the Centro de Supercomputaci{\'o}n de Galicia (CESGA) for technical support and computing time.
 
